### 2024-03-21 paper.tex
Commit early draft of paper
This is still a rough draft of the paper, but does contain the central ideas for GPU stroke expansion. There's also a crate with Rust code for generating the figures and doing experiments (including numerical robustness).

\documentclass[format=acmsmall]{acmart}
\begin{document}
\title{GPU-friendly Stroke Expansion}
\author{Raph Levien}
\author{Arman Uguray}
\affiliation{
    \institution{Google}
    \city{San Francisco}
    \state{CA}
    \country{USA}
}

\begin{abstract}
    Vector graphics includes both filled and stroked paths as the main primitives. While there have been many proposed techniques for rendering filled paths on GPU, stroked paths have proved more elusive. This paper presents a technique for performing stroke expansion, namely the generation of an outline representing the stroke of the given input path, an operation which is considered global, but which we nonetheless implement using a fully parallel algorithm. Our method relies on several novel techniques, including an encoding of the input suitable for parallel decoding and processing, an Euler spiral based method for computing flattened parallel curves. While we focus on flattened output consisting of polylines, the methods can be adapted to preserving curves in the output as well.
\end{abstract}

\maketitle
\thispagestyle{empty}
\pagestyle{plain}

\section{Introduction}

Rendering of vector graphics documents requires handling both filled and stroked primitives. There is substantial literature on GPU rendering of filled paths, but many fewer published techniques for strokes. It is a more challenging problem, especially for parallel computation, because path segments cannot be processed independently of each other; the \emph{joins} between adjacent path segments depend on context. In addition, the path topology affects the rendered result, in particular whether subpaths are open or closed. In the former case, the endpoints are rendered with \emph{caps}. Joins and caps can have multiple styles, including different styles within the same document.

The \cite{Nehab2020} paper gives a comprehensive survey of techniques for stroke expansion, and an algorithm which, for a number of reasons, is only suitable for implementation on CPU (though adapting it to GPU is listed as promising future work). It classifies techniques into \emph{local,} where each path segment generates closed geometry (which may be triangles or other primitives), and \emph{global,} where the overall result is a closed outline of the stroked path, but the partial result from each segment is in general open. Our technique is considered global in this scheme, yet allows independent processing of each path segment.

A number of factors contribute to an algorithm being ``GPU-friendly.'' In addition to simply being able to process the input segments in parallel, such an algorithm also avoids divergent control flow (including subdivision for special events such as inflection points and cusps), uses straightforward numerical techniques not subject to particular problems when evaluated using 32-bit floating point numbers.

We propose that the correctness of stroke outlines be divided into \emph{weak correctness} and \emph{strong correctness.} The latter is defined as the computing the outline of a line swept along the segment, maintaining normal orientation, combined with stroke caps and joins. Weak correctness, by contrast, only requires the parallel curves of stroke segments, combined with caps and the outer contours of joins. For sufficiently well-behaved input (particularly when the curvature measured at endpoints of path segements does not exceed the reciprol of the half linewidth), the two notions are equivalent. As described in detail in \cite{Nehab2020}, very few existing implementations actually implement the strong version, so document authors have become accustomed to not depending on behavior at endpoints. Standards for graphics formats also provide an ``out,'' enabling the weaker behavior. [TODO: find language in SVG standard]. While our method emphasizes weak correctness, it likely can be extended to the stronger sense, by implementing evolutes and inner join contours.

\section{Input encoding}

One contribution of this paper is a highly efficient encoding for paths to be rendered. Goals include conciseness, low cost to encode on the CPU side, and fully parallel processing by the GPU. There are two major aspects to the encoding: a variable size encoding of paths, suitable for both filling and stroking, and an extension to stroking. The input is separated into a \emph{tag stream}, with one byte per path element, and another stream for the coordinate data. The index into coordinate data is computed by a prefix sum from a size value derived from the tag byte.

\subsection{Path encoding}

The tag byte is a number of bit fields, packed into a byte.

TODO: make this into a diagram.

\begin{itemize}
    \item 0-1: coordinate count
    \item 2: subpath end bit
    \item 3: 16/32 bit coordinates
    \item 4: path end bit
    \item 5: transform bit
    \item 6: line style bit
\end{itemize}

For most path segments, the coordinate count indicates the type of path segment: 1 for lines, 2 for quadratic Beziers, and 3 for cubic Beziers. The subpath bit is set on the last segment of a subpath, and the path bit is additionally set on the last bit of a path. The latter is not directly interpreted by the path flattening code, but is forwarded to later rendering stages, for example, the path id can be used to select a color or other paint style.

The size is readily inferred from the tag byte: the number of coordinate pairs is the coordinate count plus one if the subpath end bit is set. The nuber of bytes per coordinate pair is 4 for 16 bit coordinates or 8 for 32 bit coordinates.

It is relatively straightforward to convert the standard moveto/lineto/curveto path representation into this encoding. Subpath boundaries are moved from the beginning of a subpath (moveto) to the end. A separate segment must be generated for closepath if the ending point does not coincide with the start point.

The handling of the subpath bit allows for overlap in coordinates, so that, except at subpath boundaries, the first coordinate pair of each path segment overlaps with the last coordinate pair of the previous one. Subpath handling is all computed in the prefix sum; no special handling is needed on the GPU side.

\section{Euler spirals and their parallel curves}

It is common to approximate cubic Béziers to some intermediate curve format more conducive to offsetting and flattening. A number of published solutions (\cite{Yzerman2020}, \cite{Nehab2020}) use quadratic Béziers, as it is well suited for computation of parallel curves. Even so, this curve has some disadvantages. In particular, it cannot model an inflection point, so the source curve must be subdivided at inflection points.

Like these other approaches, we also use an intermediate curve, but our choice is an Euler spiral. In some ways it is similar to quadratic Béziers -- it also has $O(n^4)$ scaling and is best computed using geometric Hermite interpoliation -- but differs in others. It has no difficulty modeling an inflection point. Further, its parallel curve has a particularly simple mathematical definition and clean behavior regarding cusps.

An Euler spiral segment is defined as having curvature linear in arc length.

The parallel curve of the Euler spiral (also known as ``clothoid'') was characterized by Wieleitner well over a hundred years ago\cite{Wieleitner1907}, and has a straightforward closed-form Cesàro representation, curvature as a function of arc length. TODO: present closed-form solution.

\section{Flattened parallel curves}

The geometry of a stroke outline consists of joins, caps, and the two parallel curves on either side of the input path segemnts, offset by the half linewidth. The former are not particularly difficult to calculate, but parallel curves of cubic Béziers are notoriously tricky. Analytically, is is a tenth order algebraic curve, which is not particularly feasible to compute directly.

Conceptually, generating a flattened stroke outline consists of computing the parallel curve of the input curve segment followed by \emph{flattening,} the generation of a polyline that approximates the parallel curve with sufficient accuracy (which can be measured as Fréchet distance). However, these two stages can be fused for additional performance, obviating the need to store a representation of the intermediate curve.

Using a subpixel Fréchet distance bound guarantees that the rendered image does not deviate visibly from the exact rendering. Another choice would be uniform steps in tangent angle, as chosen by \cite{Kilgard2020}. However, at small curvature, the stroked path can be off by several pixels, and at large curvature there may be considerably more subdivision than needed for faithful rendering.

Proposed image: show circular arcs with polar stroking vs near-optimal flattening.

\subsection{Optimal flattening}

The optimal flattening of a curve, with a maximum error constraint, is as follows. An optimal flattening contains the minimum number of subdivisions required to meet that constraint. Further, the error is the minimum possible with that number of subdivisions.

For a convex curve, the optimal flattening is tractable to compute, though not necessarily very fast. When a curve is monotonic, error is also monotonic with respect to both subdivision points. Thus, any standard root finding technique can find a subdivision with a given error (bisection is the best known, but the ITP method \cite{Oliveira2020} is better). An similar outer loop with a root finding technique can find the error which minimizes the global error for the entire flattening, which, due to monotonicity, is that for which the error of the final segment equals that of the previous segments. Subdividing at the inflection point adds at most one additional point; it is possible to optimize further, but we do not consider that (though the near-optimal flattening we will later present is not subject to this constraint).

\subsection{Density-based subdivision}

It is possible to analytically determine the optimal subdivision in the limit as the error approaches zero. Considering the local region of a curve as having constant curvature $\kappa$, when the distance between two subdivision points is $\Delta x$, the error is $\frac{1}{2}\Delta x^2 \kappa$. Thus, in the continuous limit, the number of subdivision points per unit length is $\sqrt{\frac{\kappa}{2\Delta y}}$, where $\Delta y$ is the target error tolerance. We call $\sqrt{\kappa}$ the \emph{subdivision density}, and scaling it by $\frac{1}{\sqrt{2\Delta y}}$ (which is constant for the curve) yields the number of subdivision points per unit length.

Sampling the curvature at a point is problematic. At an inflection point, the curvature is zero, indicating that no subdivision is needed. Conversely, at a cusp, curvature is infinite, so a technique based on sampling the curvature and advancing by the amount predicted by a circular arc at the sample point would fail to advance (the sampling density would be infinite at that point). In both cases, the sampling density varies along the curve.

Our solution is, instead of \emph{sampling} the subdivision density to determine a subdivision interval, we instead compute its \emph{integral.} Since the sampling density is in subdivisions per unit length (after proper scaling to take into account the error tolerance), the integral over the arc length between two subdivision points should be one. Computing the integral over the length of the curve segments gives the number of subdivisions needed. Rounding up, then dividing that quantity evenly by the number of subdivisions, then computing the inverse integral, gives the location of subdivision points in a nearly optimal flattening.

The integral is expressed over arc length, but for convenience and efficient computation can be re-parametrized as desired. For computing the parallel curve of an Euler spiral, the natural parametrization is the arc length of the source curve. In this case, the conversion between the two is a simple (and readily invertible) equation, as given by \cite{Wieleitner1907}, but even that is not necessary in general; by computing the integral, and its inverse, in terms of the convenient parametrization, no explicit conversion to or from arc length is needed.

\subsection{The subdivision density integral}

The subdivision density for an Euler spiral, normalized so that its inflection point is at -1 and the cusp of the parallel curve is at 1, is simply $1 - \sqrt{|1-s^2|}$. This function is plotted in Figure\ref{fig:subdiv_density}.

\begin{figure}
    \includegraphics{subdiv_density}
    \caption{Subdivision density for the parallel curve of an Euler spiral}
    \Description{A graph of subdivision density. The range from -1 to 1 is a semicircle, and the value from 1 rises sharply from zero to an asymptote of unity slope.}
    \label{fig:subdiv_density}
\end{figure}

The subdivision density integral for the parallel curve of an Euler spiral is given as follows:

\[
    f(x) = \int_0^x\sqrt{|u^2 - 1|} du
\]

This integral has a closed-form analytic solution:

\[
    f(x) = \left\{
        \begin{array}{rl}
            \frac{1}{2}(x\sqrt{|x^2-1|} + \sin^{-1}x) & \text{if } |x| \leq 1 \\
            \frac{1}{2}(x\sqrt{|x^2-1|} - \cosh^{-1}x + \frac{\pi}{4} & \text{if } x \geq 1
        \end{array}
    \right.
\]

Values for $x<-1$ follow from the odd symmetry of the function.

\subsection{Approximation of the subdivision density integral}

The subdivision density integral [TODO: probably should use section or equation number reference] is fairly straightforward to compute in the forward direction, but not invertible using a straightforward closed-form equation. Numerical techniques are possible, but require multiple iterations to achieve sufficient accuracy, so are slower. In this subsection, we present a straightforward and accurate approximation, based on piecewise easily invertible functions. If higher flattening quality is desired at the expense of slower computation, this approximation can be used to determine a good initial value for numeric techniques; two iterations of Newton solving are enough to refine this guess to within 32-bit floating point accuracy.

The approximation is given as follows:

\[ 
    f_\mathit{approx}(x) = \left\{
        \begin{array}{rl}
            \frac{\sin c_1 x}{c_1} & \text{if } x < 0.8 \\
            \frac{\sqrt{8}}{3}(x-1)^{1.5} + \frac{\pi}{4} & \text{if } 0.8 \leq x < 1.25 \\
            0.6406x^2 - 0.81x + c_2 & \text{if } 1.25 \leq x < 2.1 \\
            0.5x^2 - 0.156x + c_3 & \text{if } x \geq 2.1
        \end{array}
        \right.
\]
\[
    \begin{array}{ll}
        c_1 = & 1.0976991822760038 \\
        c_2 = & 0.9148117935952064 \\
        c_3 = & 0.16145779359520596
    \end{array}
\]

The primary rationale for the constants is for the approximation to be continuous. The other parameters were determined empirically; further automated optimization is possible but is unlikely to result in dramatic improvement. Further, this approximation is given for positive values. Negative values follow by symmetry, as the function is odd.

The exact integral and the approximation given above are shown in Figure~\ref{fig:espc}. Visually, it is clear that the agreement is close, and in numerical testing the worst case discrepancy between approximate and exact results is approximately 6\%.

\begin{figure}
    \includegraphics{espc}
    \caption{Integral of subdivision density for Euler spiral parallel curve, and its approximation}
    \Description{A graph of the integral of subdivision density. It is a wriggly diagonal line, with flat parts at -1 and 1. The approximation is nearly indistinguishable from the exact value.}
    \label{fig:espc}
\end{figure}

\section{Evolutes}

In the principled, correct specification for stroking\cite{Nehab2020}, parallel curves are sufficient only for segments in which the curvature does not exceed the reciprocal half width. When it does, additional segments must be drawn, including evolutes of the original curve. In general, the evolute of a cubic Bézier is a very complex curve, requiring approximation techniques. By contrast, the evolute of an Euler spiral ($\kappa = as$) is another spiral with a simple Cesàro equation, namely $\kappa = -a^{-1}s^{-3}$, an instance of the general result that the evolute of a log-aesthetic curve is another log-aesthetic curve\cite{Yoshida2012}.

Flattening this evolute is also straightforward; the subdivision density is proportional to $s^{-0.5}$ where $s$ is the arc length parameter of the underlying Euler spiral (and translated so $s = 0$ is the inflection point). Thus, the integral is $2\sqrt{s}$, and the inverse integral is just squaring. Thus, flattening the evolute of an Euler spiral is simpler than flattening its parallel curve.

TODO: figure showing ES, its evolute, and subdivision points for flattening.

\section{Conversion from cubic Béziers to Euler spirals}

The Euler spiral segment representation of a curve is useful for computing near-optimal flattened parallel curves, but standard APIs and document formats overwhelmingly prefer cubic Béziers as the path representation.

Many techniques for stroke expansion described in the literature apply some lowering of cubic Bézier curves to a simpler curve type, more tractable for evaluating parallel curve. Computing parallel curves directly on cubic Bézier curve segments is not very tractable. In particular, the widely cited Tiller-Hanson algorithm\cite{Tiller1984} performs well for quadratic Béziers but significantly worse for cubics.

A typical pattern for converting from one curve type to another is \emph{adaptive subdivision.} (TODO: is there a good general reference for this?) An approximate curve is found in the parameter space of the target curve family. The error of the approximation is measured. If the error exceeds the specified tolerance, the curve is subdivided (typically at $t = 0.5$), otherwise the approximation is accepted. Subdivisions are also indicated at special points; for example, since quadratic Béziers cannot represent inflection points, and geometric Hermite interpolation is numerically unstable if the input curve is not convex, lowering to quadratic Béziers also requires calculation of inflection points and subdividing there. A good example of this pattern is Nehab\cite{Nehab2020}. One advantage of Euler spirals over quadratic Béziers is that they can represent inflection points just fine, so it is not necessary to compute those for additional subdivision.

The approach in this paper is another variant of adaptive subdivision, with two twists. First, it's not necessary to actually generate the approximate curve to measure the error. Rather, a straightforward closed-form formula accurately predicts it. The second twist is that, since compute shader languages on GPUs typically don't support recursion, the stack is represented explicitly and the conceptual recursion is represented as iterative control flow. This is an entirely standard technique, but with a clever encoding the entire state of the stack can be represented in two words, each level of the stack requiring a mere single bit.

\subsection{Error prediction}

TODO: provide formula: area, halo, center

\subsection{Geometric Hermite interpolation}

Given tangent angles relative to the chord, finding the Euler spiral segment that minimizes total curvature variation is a form of geometric Hermite interpolation. There are a number of published solutions to this problem, all involving nontrivial numerical solving techniques: gradient descent\cite{Kimia2003}, bisection\cite{Walton2009}, or Newton iteration\cite{Connor2014}. Our approach is much more direct. We observe that the Euler spiral parameters are a smooth function of the two input parameters, and determine a 2D Taylor's series for this function. Many of the terms are zero because of symmetry. Preserving terms up to 7th order provides excellent accuracy (TODO: quantify more carefully) with modest computational expense.

TODO: the technique used is a refinement of what was sketched in the Python notebook attached to the "cleaner parallel curves with Euler spirals" blog post. It's basically a 2D Taylor's series. Previously we had the secant method from section 8.2 of my thesis; not sure whether we want to cite that in the submission for anonymity reasons. It's functionally pretty similar to the \cite{Connor2014} reference.

\subsection{Unrolled recursion}

The entire recursion stack for adaptive subdivision can be represented in two words: dt and scaled\_t0. The range is dt * scaled\_t0 .. dt * (scaled\_t0 + 1). Initial values of 1 and 0 represent the range (0, 1). Pushing the stack, subdividing the range in half, is representing by halving dt and doubling scaled\_t0, which leaves the start of the range invariant but its size halved. After accepting an approximation, the next range is determined by incrementing scaled\_t0, then repeatedly popping the stack by doubling dt and halving scaled\_t0, as long as the latter is odd. This can in fact be achieved without iteration by using the "count trailing zeros" intrinsic.

\section{Discussion}

\bibliographystyle{ACM-Reference-Format}
\bibliography{paper}

\end{document}

### 2024-03-21 flatten/src/cubic32.rs

@@ -1,0 +1,86 @@
+// Copyright 2024 the raphlinus.github.io Authors
+// SPDX-License-Identifier: Apache-2.0
+
+//! f32 versions of cubics, tweaked for subdivision
+
+#[derive(Clone, Copy, Debug)]
+pub struct Point {
+    pub x: f32,
+    pub y: f32,
+}
+
+impl Point {
+    pub fn new(x: f32, y: f32) -> Self {
+        Point { x, y }
+    }
+
+    pub fn from_kurbo(p: kurbo::Point) -> Self {
+        Point::new(p.x as f32, p.y as f32)
+    }
+
+    pub fn hypot(self) -> f32 {
+        self.x.hypot(self.y)
+    }
+
+    pub fn hypot2(self) -> f32 {
+        self.x * self.x + self.y * self.y
+    }
+
+    pub fn atan2(self) -> f32 {
+        self.y.atan2(self.x)
+    }
+}
+
+// Note: we don't have separate Vec2 like kurbo
+impl core::ops::Add for Point {
+    type Output = Point;
+
+    fn add(self, rhs: Point) -> Point {
+        Point::new(self.x + rhs.x, self.y + rhs.y)
+    }
+}
+
+impl core::ops::Sub for Point {
+    type Output = Point;
+
+    fn sub(self, rhs: Point) -> Point {
+        Point::new(self.x - rhs.x, self.y - rhs.y)
+    }
+}
+
+impl core::ops::Mul<f32> for Point {
+    type Output = Point;
+
+    fn mul(self, rhs: f32) -> Point {
+        Point::new(self.x * rhs, self.y * rhs)
+    }
+}
+
+pub struct Cubic {
+    pub p0: Point,
+    pub p1: Point,
+    pub p2: Point,
+    pub p3: Point,
+}
+
+impl Cubic {
+    pub fn new(p0: Point, p1: Point, p2: Point, p3: Point) -> Self {
+        Cubic { p0, p1, p2, p3 }
+    }
+
+    /// Evaluate point and derivative.
+    ///
+    /// Note that the second value returned is 1/3 the actual derivative,
+    /// to reduce multiplication.
+    pub fn eval_and_deriv(&self, t: f32) -> (Point, Point) {
+        let m = 1.0 - t;
+        let mm = m * m;
+        let mt = m * t;
+        let tt = t * t;
+        let p =
[…]

### 2024-03-21 flatten/src/evolute.rs

@@ -1,0 +1,58 @@
+// Copyright 2024 the Vello Authors
+// SPDX-License-Identifier: Apache-2.0
+
+//! Math for flattening of Euler evolute
+
+use kurbo::{BezPath, Point};
+
+use crate::euler::{EulerParams, EulerSeg};
+
+fn eval_es_params_evolute(params: &EulerParams, t: f64) -> Point {
+    let offset = -1.0 / params.curvature(t);
+    params.eval_with_offset(t, offset)
+}
+
+fn eval_es_evolute(es: &EulerSeg, t: f64) -> Point {
+    let chord = es.p1 - es.p0;
+    let Point { x, y } = eval_es_params_evolute(&es.params, t);
+    Point::new(
+        es.p0.x + chord.x * x - chord.y * y,
+        es.p0.y + chord.x * y + chord.y * x,
+    )
+}
+
+/// Flatten the evolute of an Euler spiral segment.
+pub fn flatten_es_evolute(es: &EulerSeg, tol: f64) -> BezPath {
+    let mut path = BezPath::new();
+    // TODO: smart flatten logic
+    const N: usize = 10;
+    for i in 0..=N {
+        let t = i as f64 / N as f64;
+        let p = eval_es_evolute(es, t);
+        if i == 0 {
+            path.move_to(p);
+        } else {
+            path.line_to(p)
+        }
+    }
+    path
+}
+
+/// Flatten an Euler spiral segment.
+///
+/// This is duplicative of other work but maybe useful for exploration.
+pub fn flatten_es(es: &EulerSeg, tol: f64) -> BezPath {
+    let mut path = BezPath::new();
+    // TODO: smart flatten logic
+    const N: usize = 10;
+    for i in 0..=N {
+        let t = i as f64 / N as f64;
+        let p = es.eval(t);
+        if i == 0 {
+            path.move_to(p);
+        } else {
+            path.line_to(p)
+        }
+    }
+    path
+}

### 2024-03-21 flatten/src/flatten.rs

@@ -1,0 +1,155 @@
+// Copyright 2024 the Vello Authors
+// SPDX-License-Identifier: Apache-2.0
+
+//! Math for flattening of Euler spiral parallel curve
+
+use std::f64::consts::{FRAC_1_SQRT_2, FRAC_PI_2, FRAC_PI_4};
+
+use kurbo::BezPath;
+
+use crate::euler::EulerSeg;
+
+pub fn flatten_offset(iter: impl Iterator<Item = EulerSeg>, offset: f64) -> BezPath {
+    let mut result = BezPath::new();
+    let mut first = true;
+    for es in iter {
+        if core::mem::take(&mut first) {
+            result.move_to(es.eval_with_offset(0.0, offset));
+        }
+        let scale = es.p0.distance(es.p1);
+        let tol = 1.0;
+        let (k0, k1) = (es.params.k0 - 0.5 * es.params.k1, es.params.k1);
+        // compute forward integral to determine number of subdivisions
+        let dist_scaled = offset / scale;
+        let a = -2.0 * dist_scaled * k1;
+        let b = -1.0 - 2.0 * dist_scaled * k0;
+        let int0 = espc_int_approx(b);
+        let int1 = espc_int_approx(a + b);
+        let integral = int1 - int0;
+        let k_peak = k0 - k1 * b / a;
+        let integrand_peak = (k_peak * (k_peak * dist_scaled + 1.0)).abs().sqrt();
+        let scaled_int = integral * integrand_peak / a;
+        let n_frac = 0.5 * (scale / tol).sqrt() * scaled_int;
+        let n = n_frac.ceil();
+        for i in 0..n as usize {
+            let t = (i + 1) as f64 / n;
+            let inv = espc_int_inv_approx(integral * t + int0);
+            let s = (inv - b) / a;
+            result.line_to(es.eval_with_offset(s, offset));
+        }
+    }
+    result
+}
+
+/// Integral of sqrt(|x^2 - 1|), analytical.
+fn espc_integral(x: f64) -> f64 {
+    let d = 1.0 - x * x;
+    let ds = d.abs().sqrt();
+    let a = if d >= 0.0 {
+        x.asin()
+    } else {
+        FRAC_PI_2.copysign(x) - (ds + x.abs()).ln().copysign(x)
+    };
[…]

### 2024-03-21 flatten/src/flatten32.rs

@@ -1,0 +1,270 @@
+// Copyright 2024 the Vello Authors
+// SPDX-License-Identifier: Apache-2.0
+
+// 32 bit versions of flattening integral, intended to match GPU-side
+// computation.
+
+use std::f32::consts::{FRAC_1_SQRT_2, FRAC_PI_4};
+
+use kurbo::BezPath;
+
+use crate::cubic32::Point;
+use crate::euler32::EulerSeg;
+
+/// A robustness strategy for the ESPC integral
+enum EspcRobust {
+    /// Both k1 and dist are large enough to divide by robustly.
+    Normal,
+    /// k1 is low, so model curve as a circular arc.
+    LowK1,
+    /// dist is low, so model curve as just an Euler spiral.
+    LowDist,
+}
+
+fn to_kurbo(p: Point) -> kurbo::Point {
+    kurbo::Point::new(p.x as f64, p.y as f64)
+}
+
+// Note: generates a kurbo-compatible BezPath for debugging reasons, but
+// actual output can be f32 LineSoup without much trouble.
+pub fn flatten_offset(iter: impl Iterator<Item = EulerSeg>, offset: f32, tol: f32) -> BezPath {
+    let mut result = BezPath::new();
+    let mut first = true;
+    for es in iter {
+        if core::mem::take(&mut first) {
+            result.move_to(to_kurbo(es.eval_with_offset(0.0, offset)));
+        }
+        let scale = (es.p0 - es.p1).hypot();
+        let (k0, k1) = (es.params.k0 - 0.5 * es.params.k1, es.params.k1);
+        // Note: scaling by ch is missing from earlier implementations. The math
+        // should be validated carefully.
+        let dist_scaled = offset * es.params.ch / scale;
+        // The number of subdivisions for curvature = 1
+        let scale_multiplier = 0.5 * FRAC_1_SQRT_2 * (scale / (es.params.ch * tol)).sqrt();
+        // TODO: tune these thresholds
+        const K1_THRESH: f32 = 1e-3;
+        const DIST_THRESH: f32 = 1e-3;
+        let mut a = 0.0;
+        let mut b = 0.0;
+        let mut integral = 0.0;
+        let mut int0 = 0.0;
+        let (n_frac, robust) = if k1.abs() < K1_THRESH {
+            let k = k0 + 0.5 * k1;
+            let n_frac = (k * (k * dist_scaled + 1.0)).abs().sqrt();
[…]

### 2024-03-21 flatten/src/main.rs

@@ -1,0 +1,125 @@
+// Copyright 2024 the Vello Authors
+// SPDX-License-Identifier: Apache-2.0
+
+use cubic32::{Cubic, Point};
+use euler::EulerSeg;
+use euler32::CubicToEulerIter;
+use flatten::{espc_integral_inv, n_subdiv_analytic};
+use flatten32::flatten_offset;
+
+use crate::euler::EulerParams;
+
+mod cubic32;
+mod euler;
+mod euler32;
+mod evolute;
+mod flatten;
+mod flatten32;
+
+fn main_est_flatten_err() {
+    let th0 = 0.101;
+    let th1 = 0.1;
+    let offset = 0.5;
+    let ep = EulerParams::from_angles(th0, th1);
+    println!("{ep:?}");
+    for i in 0..=10 {
+        let t = i as f64 / 10.0;
+        println!(
+            "{t}: {} {:?}",
+            ep.eval_th(t),
+            ep.eval_with_offset(t, offset)
+        );
+    }
+    let th = 0.05;
+    println!("{:?}", ep.inv_th(th));
+    for i in -10..=10 {
+        let dist = i as f64 / 10.0;
+        let exact_err = ep.exact_flatten_err_seg(dist, 0.0, 1.0);
+        println!(
+            "dist={dist:.2}: est {:.6} numeric {:.6} exact {:.6}",
+            ep.est_flatten_err(dist),
+            ep.numeric_flatten_err(dist),
+            exact_err
+        );
+    }
+}
+
+#[allow(unused)]
+fn main_invert_espc_int() {
+    for i in 0..100 {
+        let x = i as f64 / 50.0;
+        let _ = espc_integral_inv(x);
+    }
+}
+
+#[allow(unused)]
+/// Main entry point for ESPC experiments.
+fn main_espc() {
+    let k0 = 0.01;
+    let dist = 1e-4;
+    let scale = 1.0;
+    let tol = 1.0;
+    for i in 0..10 {
+        let k1 = if i == 0 { 0.0 } else { 0.1f64.powi(i) };
+        // Note: this isn't necessarily accurate when k1 and dist are tiny
+        let analytic = n_subdiv_analytic(k0, k1, scale, dist, tol);
+        let approx_f32 =
+            flatten32::n_subdiv_robust(k0 as f32, k1 as f32, scale as f32, dist as f32, tol as f32);
[…]

### 2024-03-23 paper.tex
Write section on cubic to euler error metric
It might be clunky, and lacks visuals, but the core ideas are there.

There's potentially more to be said about robustness work in that area, but my current thinking is to keep all that at a pretty high level, and concise.

@@
 
-TODO: provide formula: area, halo, center
+A key step in approximating one curve with another is evaluating the error of the approximation.
+A common approach (used in \cite{Nehab2020} among others) is to generate the approximate curve, the measure the distance, often by sampling at multiple points on the curve.
+All this is potentially slow, with the additional risk of underestimating the error due to undersampling.
+
+Our approach is different.
+In short, we perform a straightforward analytical computation to accurately estimate the error.
+Our approach to the error metric has two major facets.
+First, we obtain a cubic Bézier which is a very good fit to the Euler spiral, then we estimate the distance between that and the source cubic.
+Due to the triangle inequality, the sum of these is a conservative estimate of the true Fréchet distance between the cubic and the Euler spiral.
+
+For mathematical convenience, the error estimation is done with the chord normalized to unit distance; the actual error is scaled linearly by the actual chord length.
+
+The cubic Bézier approximating the Euler spiral is that with the distance of each control point from the endpoint is $d = \frac{2}{3(1 + \cos \theta)}$, where $\theta$ is the angle of the endpoint tangent relative to the chord.
+This is a generalization of the standard approximation of an arc.
+It should be noted that this is not in general the closest possible fit, but it is computationally tractable and has near-uniform parametrization.
+In general it has quintic scaling; if an Euler spiral segment is divided in half, the error of this cubic fit decreases by a factor of 32.
+A good estimate of this error is:
+
+\[
+    4.6255\times10^{-6}|\theta_0+\theta_1|^5 + 7.5\times10^{-3}|\theta_0+\theta_1|^2|\theta_0-\theta_1|
+\]
+
+Possible image for this: the heatmap from the cleaner parallel curves blog post.
+
+The distance between two cubic Bézier segments can be further broken down into two terms; in most cases, the difference in area accurately predicts the Fréchet distance between the curves.
+Two cubic Béziers with the same area and same tangents tend to be relatively close to each other, but there is some error stemming from the difference in parametrization.
+Area of a cubic Bézier segment is straightforward to compute using Green's theorem:
+
+\[
+    a = \tfrac{3}{20}(2d_0\sin \theta_0 + 2d_1\sin \theta_1 - d_0 d_1\sin(\theta_0+\theta_1))
+\]
+
+The conservative Fréchet distance estimate is 1.55 the absolute difference in area between the source cubic and the Euler spiral approximation.
+The final term for imbalance is as follows, with $\bar{d}_0$ and $\bar{d}_1$ representing the distance from the endpoints of the Euler approximation and $d_0$ and $d_1$ the corresponding distance in the source cubic segment, as in the area calculation above:
+
+\[
+    (0.005|\theta_0+\theta_1| + 0.07|\theta_0 - \theta_1|)\sqrt{(\bar{d}_0 - d_0)^2 + (\bar{d}_1 - d_1)^2}
+\]
+
+The total estimated error is the sum of these three terms.
 

### 2024-03-27 paper.tex

@@
 This paper presents a technique for performing stroke expansion, namely the generation of an outline representing the stroke of the given input path, an operation which is considered global, but which we nonetheless implement using a fully parallel algorithm.
-Our method relies on several novel techniques, including an encoding of the input suitable for parallel decoding and processing, an Euler spiral based method for computing flattened parallel curves.
-While we focus on flattened output consisting of polylines, the methods can be adapted to preserving curves in the output as well.
+We introduce several novel techniques, including an encoding of stroked primitives suitable for parallel processing and an Euler spiral based method for computing flattened parallel curves.
+Our method produces a near-optimal number of line segments for the filled outline of a stroked cubic Bézier and can be fully evaluated on a GPU with minimal preprocessing.
 \end{abstract}
@@
 
+TODO: brief description of stroke to fill conversion.
+
 The \cite{Nehab2020} paper gives a comprehensive survey of techniques for stroke expansion, and an algorithm which, for a number of reasons, is only suitable for implementation on CPU (though adapting it to GPU is listed as promising future work).
@@
 
-A number of factors contribute to an algorithm being ``GPU-friendly.'' In addition to simply being able to process the input segments in parallel, such an algorithm also avoids divergent control flow (including subdivision for special events such as inflection points and cusps), uses straightforward numerical techniques not subject to particular problems when evaluated using 32-bit floating point numbers.
+A number of factors contribute to an algorithm being ``GPU-friendly.'' In addition to simply being able to process the input segments in parallel, such an algorithm also avoids divergent control flow (avoiding the need for explicit subdivision at special events such as inflection points and cusps) and uses robust numerical techniques not subject to particular problems when evaluated using 32-bit floating point numbers.
 
-We propose that the correctness of stroke outlines be divided into \emph{weak correctness} and \emph{strong correctness.} The latter is defined as the computing the outline of a line swept along the segment, maintaining normal orientation, combined with stroke caps and joins.
+We propose that the correctness of stroke outlines be divided into \emph{weak correctness} and \emph{strong correctness.} We define strong correctness as the computation of the outline of a line swept along the segment, maintaining normal orientation, combined with stroke caps and joins.
 Weak correctness, by contrast, only requires the parallel curves of stroke segments, combined with caps and the outer contours of joins.
-For sufficiently well-behaved input (particularly when the curvature measured at endpoints of path segements does not exceed the reciprol of the half linewidth), the two notions are equivalent.
+Weak correctness, by contrast, only requires the parallel curves of stroke segments, combined with caps and the outer contours of joins.
+The two notions are equivalent for sufficiently well-behaved input, in particular when the curvature measured at endpoints of path segments does not exceed the reciprocal of the half linewidth.
 As described in detail in \cite{Nehab2020}, very few existing implementations actually implement the strong version, so document authors have become accustomed to not depending on behavior at endpoints.
@@
 [TODO: find language in SVG standard].
-While our method emphasizes weak correctness, it likely can be extended to the stronger sense, by implementing evolutes and inner join contours.
+While our current implementation emphasizes weak correctness, we believe it can be extended to the stronger sense, by implementing evolutes and inner join contours.
+
+TODO: discussion of prior art.
+
+\begin{itemize}
+    \item Nehab 2020
+    \item Kilgard 2020: Polar stroking (discussion of error metric can move here)
+    \item Yzerman 2020: combined offset + flatten using quadratics
+    \item Blog post of Dragoș?
+    \item Curve-to-curve offsetting?
+Tiller and Hanson (which is good for quadratics, terrible for cubics), Shape control (pretty good but has robustness issues).
+Hain (misses error tolerance in cusp case, as described in Yzerman).
+The "Comparing Offset Curve Approximation Methods" paper is flawed.
+\end{itemize}
 
@@
 The total estimated error is the sum of these three terms.
+We have validated this error metric in randomized testing.
+For values of $\theta$ between $0$ and $0.5$, and values of $d$ between $0$ and $0.6$, this estimate is always conservative, and it is also tight: over Bézier curves generated randomly with parameters drawn from a uniform distribution this range, the mean ratio of estimated to true error is 1.55 (TODO: verify).
 

### 2024-03-30 flatten/src/euler_arc.rs
Some experiments with arc fitting
There's more to be done, but checking in what I have. That includes very careful calculation of the area of an arc segment, and inverting that.

Also some explorations of area of an Euler spiral parallel curve. The result of that exploration is that fine-tuning area is not profitable, as it scales the same as the error of the underlying Euler spiral to arc approximation. A far better approach is to apply a small modifier to the error metric. The current `arc_main` contains a by-hand tester to fit that metric.

@@ -1,0 +1,145 @@
+// Copyright 2024 the Vello Authors
+// SPDX-License-Identifier: Apache-2.0
+
+//! Logic for lowering Euler spirals to arcs
+
+use std::f64::consts::FRAC_PI_2;
+
+use kurbo::{common::GAUSS_LEGENDRE_COEFFS_32, Vec2};
+
+use crate::euler::{EulerParams, EulerSeg};
+
+/// Area of a circle segment.
+///
+/// The `th` argument is the total angle for the arc.
+pub fn arc_area(th: f64) -> f64 {
+    // The trig has numerical robustness problems even for reasonable
+    // small inputs; use Taylor series instead
+    if th.abs() < 0.022 {
+        let t2 = th * th;
+        (2. / 3. + (4. / 45.) * t2 + (4. / 315.) * t2 * t2) * th
+    } else {
+        let sth = th.sin();
+        (th - sth * th.cos()) / (sth * sth)
+    }
+}
+
+pub fn inv_arc_area(a: f64) -> f64 {
+    const SCALE: f64 = 0.949;
+    let mut th = (1.5 / SCALE) * (SCALE * a).atan();
+    let thabs = th.abs();
+    let n_iter = if thabs < 1e-5 {
+        0
+    } else if thabs < 0.15 {
+        1
+    } else if thabs < 0.1 {
+        2
+    } else if thabs < 0.5 {
+        3
+    } else if thabs < 1.4 {
+        4
+    } else if thabs < 3.2 {
+        5
+    } else {
+        // TODO: more - and possibly fit the curve
+        6
+    };
+    for _ in 0..n_iter {
+        let err = arc_area(th) - a;
+        let sth = th.sin();
+        let deriv = (2. - 2. * th * th.cos() / sth) / (sth * sth);
+        th -= err / deriv;
+    }
+    th
+}
+
+// Possibly should be method of EulerParams
+fn euler_area(params: &EulerParams) -> f64 {
+    let mut sum = 0.0;
+    for (wi, xi) in GAUSS_LEGENDRE_COEFFS_32 {
+        let t = 0.5 + 0.5 * xi;
+        let p = params.eval(t);
+        sum += wi * p.y * params.eval_th(t).cos();
+    }
+    0.5 * sum / params.ch
+}
+
+fn fit_euler_to_arc(params: &EulerParams) -> f64 {
+    let a = euler_area(params);
+    println!("area = {a}");
[…]

### 2024-03-30 flatten/src/main.rs

@@ -12,6 +12,7 @@ use crate::euler::EulerParams;
 mod cubic32;
 mod euler;
 mod euler32;
+mod euler_arc;
 mod evolute;
 mod flatten;
 mod flatten32;
@@ -99,6 +100,7 @@ fn main() {
         }
         "espc" => main_espc(),
         "est_flatten_err" => main_est_flatten_err(),
+        "arc" => euler_arc::arc_main(),
         _ => println!("unknown figure type"),
     }
 }

### 2024-04-02 paper.tex
Writeup of error metrics and and arc approximation
This gets a bit more of my thinking down on paper about error metrics, and also presents the most important results for approximation by arcs. There are quite a few citations, the references are expanded.

It also starts a future work section.

@@
 This paper presents a technique for performing stroke expansion, namely the generation of an outline representing the stroke of the given input path, an operation which is considered global, but which we nonetheless implement using a fully parallel algorithm.
-We introduce several novel techniques, including an encoding of stroked primitives suitable for parallel processing and an Euler spiral based method for computing flattened parallel curves.
+We introduce several novel techniques, including an encoding of stroked primitives suitable for parallel processing and an Euler spiral based method for computing both flattened parallel curves and approximations to arc segments.
 Our method produces a near-optimal number of line segments for the filled outline of a stroked cubic Bézier and can be fully evaluated on a GPU with minimal preprocessing.
@@
 We propose that the correctness of stroke outlines be divided into \emph{weak correctness} and \emph{strong correctness.} We define strong correctness as the computation of the outline of a line swept along the segment, maintaining normal orientation, combined with stroke caps and joins.
-Weak correctness, by contrast, only requires the parallel curves of stroke segments, combined with caps and the outer contours of joins.
 Weak correctness, by contrast, only requires the parallel curves of stroke segments, combined with caps and the outer contours of joins.
@@
 
+\section{Flattening and arc approximation of curves}
+
+% Note: this is being written as a fairly stand-alone section. No doubt some reorganization should happen
+
+The core problem in stroke expansion is approximating the desired curve by segments of some other curve, usually a simpler one.
+These segments must be within an error tolerance of the source curve, and ideally there are not many more of them than optimal.
+We consider a number of source to target pairs, most importantly cubic Béziers to Euler spirals, and Euler spiral parallel curves to either lines or arcs.
+
+There are generally three approaches to such curve approximation.
+The most straightforward but also least efficient is ``cut then measure,'' usually combined with adaptive subdivision.
+In this technique, a candidate approximate curve is produced, then the error is measured against the source curve, usually by sampling a number of points along both curves and determining a maximum error (or perhaps some other error norm).
+If the error is within tolerance, the approximation is accepted.
+Otherwise, the curve is subdivided (usually at t = 0.5) and each half is recursively approximated.
+A substantial fraction of all curve approximation methods in the literature are of this form, including \cite{Nehab2020}.
+The main disadvantage is the cost of computing the error metric.
+Another risk is underestimating the error due to inadequate sampling; this is a particular problem when the source curve contains a cusp.
+
+% Note: would be cool to run profiler on Nehab code to determine fraction of time computing error metric.
+
+The next approach is similar, but uses an \emph{error metric} to estimate the error.
+Ideally such a metric is a closed-form computation rather than requiring iteration.
+A good error metric is conservative, yet tight, in that it never underestimates the error (which would allow results exceeding the error bound to slip through), and does not significantly overestimate the error, which would result in more subdivision than optimal.
+
+By far the most efficient approach is an \emph{invertible} error metric.
+In this approach, the error metric has an analytic inverse, or at least a good numerical approximation.
+Because the metric is invertible, it can predict the number of subdivisions needed, as well as the parameter value for each subdivision.
+If the error metric is accurate, then approximation is near-optimal.
+One example of an invertible error metric is angle step, used in polar stroking\cite{Kilgard2020}; the number of subdivisions is the total angle subsumed by the curve divided by the angle step size, and the parameter value for each subdivision is the result of solving for a tangent direction.
+Another widely used invertible error metric is Wang's formula (\cite{Goldman2003}, Section 5.6.3), which gives a bound on the flattening error based on the second derivative of the curve.
+This metric is conservative but works well in practice; among other applications, it is used in Skia for path flattening.
+One limitation of Wang's method is that it only applies to the flattening of filled outlines, not parallel curves as is required for strokes.
+
+% Note: an alternate cite for the central result of Wang's formula is Sederberg's lecture notes.
+It doesn't use the name "Wang's formula." In one way, it is a cleaner result, as it doesn't have the rounding up to a power of 2.
+But it is also less of an archival cite.
+% https://scholarsarchive.byu.edu/cgi/viewcontent.cgi?article=1000&context=facpub#section.10.6
+
+\subsection{Error metrics for flattening}
+
+The distance between a circular arc segment of length $s$, with angle between arc and chord of $\theta$, is exactly $(1-\cos \theta)s/(2\theta)$.
+The curvature is $\kappa = \frac{2\theta}{s}$ (equivalently, $\theta = \frac{\kappa s}{2s}$), and this remains constant even as the arc is subdivided.
+Rewriting, $d = (1-\cos\frac{\kappa}{2s})\frac{s}{\kappa}$.
+From this, we can derive a precise, invertible error metric.
+Subdividing the arc in to $n$ segments, the distance error for each segment is $\frac{1}{\kappa}(1-\cos \frac{s\kappa}{2n})$.
+Solving for n, we get:
+
+\[
+n = \frac{s\kappa}{2\cos^{-1}\left( 1- d\kappa \right)}
+\]
+
+To flatten a finite arc, round up $n$ to the nearest integer. This will cause the error to decrease, so will still be within the error bounds.
+
+Note that the number of subdivisions is proportional to the arc length.
+Another way of stating this relationship is that the \emph{subdivision density}, the number of subdivisions per unit of arc length, is constant.
+
+The error metric for flattening an arc is exact.
+It always yields the minimum number of subdivisions needed to flatten the curve, and the flattening error is the least possible with given that number of subdivisions.
+For general curves, an exact error bound is not feasible, and we resort to an approximation.
+Again the circular arc provides a good example.
+Applying the small angle approximation $\cos \theta \approx 1 - \theta^2/2$, the approximate distance error is $d = \frac{\kappa s^2}{8n^2}$, and solving for $n$ we get $n = s\sqrt{\frac{\kappa}{8d}}$.
+Note that this estimate is \emph{conservative,} in that it will always request more subdivision and thus produce a lower error than the exact metric.
+
+We are of course concerned with the flattening of more general curves (ultimately the parallel curve of a cubic Bézier), not simply circular arcs.
+It is tempting to sample the curvature and plug it into the formula above, but this can both dramatically undershoot the error (predicting that no subdivision is needed at an inflection point) and overshoot it (requiring infinite subdivision at a cusp; and cusps).
+Apparently, some kind of average curvature is needed.
+
+We propose the following error metric to estimate the maximum distance between an arbitrary curve of length $\hat{s}$ and its chord.
+
+\[
+    d \approx \frac{1}{8}\left(\int_0^{\hat{s}} \sqrt{|\kappa(s)|}ds \right)^2
+\]
+
+% Should s-hat be the integration variable instead? Is there a better choice of symbol?
+
+This formula is the same as the approximate error metric for circular arcs, except that instead of a constant curvature value, a norm-like average with an exponent of 1/2 is used (it is not considered a true norm because the triangle inequality does not hold).
+We chose this particular formulation because it tends to produce invertible error metrics, and because it works well in practice.
+
+This formula also has a meaningful interpretation: the quantity under the integral sign is the subdivision density, and represents the number of subdivisions per unit length of an optimal flattening as the error tolerance approaches zero.
+In particular, the number of subdivisions is:
+
+\[
+    n = \left(\int_0^{\hat{s}} \sqrt{|\kappa(s)|}ds \right)\sqrt{\frac{1}{8d}}
+\]
+
+In addition, if the function represented by the integral is invertible, then the corresponding error metric is invertible.
+Evaluate the function to determine the number of subdivision points, then evenly divide the result, using the inverse of the function to map these values back into parameter values for the source curve being approximated.
+
+\subsection{Error metrics for flattening Euler spirals}
+
+We choose Euler spiral segments for our intermediate curve representation precisely because their simple formulation in terms of curvature (Cesàro equation) results in similarly simple subdivision density integrals.
+
+An Euler spiral segment is defined by $\kappa(s) = as+b$, or alternatively $\kappa(s) = a(s-s_0)$, where $s_0 = -b/a$ is the location of the inflection point.
+Applying the above error metric, the subdivision density is simply $\sqrt{|a(s-s_0)|}$.
+The integral is $\frac{2}{3}\sqrt{a}(s-s_0)^1.5$, which is readily invertible.
+
+An immediate consequence is that flattening an Euler spiral by choosing subdivision points $s_i = a\cdot i^\frac{2}{3}$ produces a near-optimal flattening, as can visually be seen in Figure\ref{fig:es_flatten}.
+
+\begin{figure}
+    \includegraphics[scale=0.3]{es_flatten}
+    \caption{Flattening of an Euler spiral using points spaced by power-law}
+    \Description{An Euler spiral along with its flattened representation. The flattened curve has an error}
+    \label{fig:es_flatten}
+\end{figure}
+
 \section{Euler spirals and their parallel curves}
@@
 
-\subsection{Density-based subdivision}
-
-It is possible to analytically determine the optimal subdivision in the limit as the error approaches zero.
-Considering the local region of a curve as having constant curvature $\kappa$, when the distance between two subdivision points is $\Delta x$, the error is $\frac{1}{2}\Delta x^2 \kappa$.
-Thus, in the continuous limit, the number of subdivision points per unit length is $\sqrt{\frac{\kappa}{2\Delta y}}$, where $\Delta y$ is the target error tolerance.
-We call $\sqrt{\kappa}$ the \emph{subdivision density}, and scaling it by $\frac{1}{\sqrt{2\Delta y}}$ (which is constant for the curve) yields the number of subdivision points per unit length.
-
-Sampling the curvature at a point is problematic.
-At an inflection point, the curvature is zero, indicating that no subdivision is needed.
-Conversely, at a cusp, curvature is infinite, so a technique based on sampling the curvature and advancing by the amount predicted by a circular arc at the sample point would fail to advance (the sampling density would be infinite at that point).
-In both cases, the sampling density varies along the curve.
-
-Our solution is, instead of \emph{sampling} the subdivision density to determine a subdivision interval, we instead compute its \emph{integral.} Since the sampling density is in subdivisions per unit length (after proper scaling to take into account the error tolerance), the integral over the arc length between two subdivision points should be one.
-Computing the integral over the length of the curve segments gives the number of subdivisions needed.
-Rounding up, then dividing that quantity evenly by the number of subdivisions, then computing the inverse integral, gives the location of subdivision points in a nearly optimal flattening.
-
-The integral is expressed over arc length, but for convenience and efficient computation can be re-parametrized as desired.
-For computing the parallel curve of an Euler spiral, the natural parametrization is the arc length of the source curve.
-In this case, the conversion between the two is a simple (and readily invertible) equation, as given by \cite{Wieleitner1907}, but even that is not necessary in general; by computing the integral, and its inverse, in terms of the convenient parametrization, no explicit conversion to or from arc length is needed.
-
 \subsection{The subdivision density integral}
 
-The subdivision density for an Euler spiral, normalized so that its inflection point is at -1 and the cusp of the parallel curve is at 1, is simply $1 - \sqrt{|1-s^2|}$.
+The subdivision density for the parallel curve of an Euler spiral, normalized so that its inflection point is at -1 and the cusp of the parallel curve is at 1, is simply $1 - \sqrt{|1-s^2|}$.
 This function is plotted in Figure\ref{fig:subdiv_density}.
@@
 \begin{figure}
-    \includegraphics{subdiv_density}
+    \includegraphics[scale=0.6]{subdiv_density}
     \caption{Subdivision density for the parallel curve of an Euler spiral}
@@
             \frac{1}{2}(x\sqrt{|x^2-1|} + \sin^{-1}x) & \text{if } |x| \leq 1 \\
-            \frac{1}{2}(x\sqrt{|x^2-1|} - \cosh^{-1}x + \frac{\pi}{4} & \text{if } x \geq 1
+            \frac{1}{2}(x\sqrt{|x^2-1|} - \cosh^{-1}x + \frac{\pi}{4}) & \text{if } x \geq 1
         \end{array}
@@
 
+\section{Error metrics for approximation by arcs}
+
+The problem of approximating a curve by a sequence of arc segments has extensive literature, but none of the published solutions are quite suitable for our application.
+The specific problem of approximating an Euler spiral by arcs is considered in \cite{Meek2004} using a ``cut then measure'' adaptive subdivision scheme, but their solution has poor quality; it scales as $O(1/n^2)$, while $O(1/n^3)$ is attainable.
+The result was improved ``slightly'' by \cite{Narayan2014}.
+The literature also contains optimal results, namely \cite{Maier2014} and \cite{Nuntawisuttiwong2021}, but at considerable cost; both approaches claim $O(n^2)$ time complexity.
+The through-line for all these results is that they are solving a harder problem, adopting the constraint that the generated arc sequence is $G^1$ continuous.
+While desirable for many applications, for simply rendering a stroked outline, it is not needed.
+Even with this constraint relaxed, the angle discontinuities of an arc approximation are tiny compared to the flattening to lines.
+
+% I also expect that the Fréchet distance error of a solution not constrained by G1 continuity will be better by some constant factor (somewhere around 2-4, just as a gut feeling) than the G1-continuous optimum solution.
+This is a major reason why we have such favorable segment counts, even in comparison to Euler spirals.
+However, providing evidence for that claim feels more in scope for another paper.
+
+Our approach is based on a simple error metric, similar in flavor to the one for flattening to line segments.
+The details of the metric (in particular, tuning of constants) were obtained empirically, though we suspect that more rigorous analytic bounds could be obtained.
+In practice it works very well indeed; the best way to observe that is an interactive testing tool, which is provided in the supplemental materials.
+
+The proposed error metric is as follows. The estimated distance error for a curve of length $\hat{s}$ is:
+
+\[
+    d \approx \frac{1}{120}\left(\int_0^{\hat{s}} \sqrt[3]{|\kappa'(s)|}ds \right)^3
+\]
+
+For an Euler spiral segment, $\kappa'(s)$ is constant and thus this error metric becomes nearly trivial.
+With n subdivisions, the estimated distance is simply $\frac{s^3\kappa'}{120n^3}$.
+Solving for $n$, we get $n = s\sqrt[3]{\frac{|\kappa'|}{120d}}$ subdivisions, and those are divided evenly by arc length, as the subdivision density is constant across the curve, just as is the case for flattening arcs to lines.
+
+TODO: image. This will likely strongly resemble Fig. 8a from Narayan2014.
+
+Remarkably, the approximation of an Euler spiral parallel curve by arc segments is almost as simple as that for Euler spirals to arcs.
+As in flattening to lines, the parameter for the curve is the arc length of the originating Euler spiral.
+The subdivision density is then constant, and only a small tweak is needed to the formula for computing the number of subdivisions, taking into account the additional curvature variation from the offset by $h$ (the half line-width).
+The revised formula is:
+
+\[
+    n = s\sqrt[3]{\frac{|\kappa'|(1+0.4|hs\kappa'|)}{120d}}
+\]
+
+This formula was determined empirically by curve-fitting measured error values from fitting Euler spiral parallel curves to arcs, but was also inspired by applying the general error metric formula to the analytical equations for Euler spiral parallel curve, and dropping higher order terms.
+A more rigorous derivation, ideally with firm error bounds, remains as future work.
+
+One consequence of this formula is that, since $h$ appears under absolute value, the same arc approximation can be used for both sides of a stroke.
+
+% Make improved quality claim here? I think it may be too subtle.
+
 \section{Evolutes}
@@
 In particular, the widely cited Tiller-Hanson algorithm\cite{Tiller1984} performs well for quadratic Béziers but significantly worse for cubics.
+
+% TODO: the discussion of adaptive subdivision maybe duplicates the earlier section on "measure then cut."
 
@@
 We have validated this error metric in randomized testing.
-For values of $\theta$ between $0$ and $0.5$, and values of $d$ between $0$ and $0.6$, this estimate is always conservative, and it is also tight: over Bézier curves generated randomly with parameters drawn from a uniform distribution this range, the mean ratio of estimated to true error is 1.55 (TODO: verify).
+For values of $\theta$ between $0$ and $0.5$, and values of $d$ between $0$ and $0.6$, this estimate is always conservative, and it is also tight: over Bézier curves generated randomly with parameters drawn from a uniform distribution this range, the mean ratio of estimated to true error is 1.656.
+
+% Note about supplemental materials.
+This ratio is produced by examples/euler3 in Raph's local branch of euler_err_experiment of kurbo (git 99b00).
+This should be cleaned up and moved into the code repo.
 
@@
 Preserving terms up to 7th order provides excellent accuracy (TODO: quantify more carefully) with modest computational expense.
+
+TODO: A good cite for this is \cite{Reif2021}. Our solution is similar in flavor, but is more precise at modest cost as it contains more polynomial terms.
 
@@
 After accepting an approximation, the next range is determined by incrementing scaled\_t0, then repeatedly popping the stack by doubling dt and halving scaled\_t0, as long as the latter is odd.
-This can in fact be achieved without iteration by using the "count trailing zeros" intrinsic.
+This can in fact be achieved without iteration by using the ``count trailing zeros'' intrinsic.
+
+\section{Future work}
+
+\begin{itemize}
+    \item The GPU implementation sequentializes some work that could be parallel, and is thus not as well load-balanced as it might be.
+An appealing future direction is to split the pipeline into separate stages, executed by the GPU as nodes in a work graph\cite{Patel2024}.
+This structure is appealing because load balancing is done by hardware.
+
+    \item Many of the error metrics were empirically determined.
+The mathematical theory behind them should be developed more rigorously, and that process will likely uncover opportunities to fine-tune the technique.
+\end{itemize}
 

### 2024-04-03 paper.tex
Add arc_chord figure

@@
 
-The distance between a circular arc segment of length $s$, with angle between arc and chord of $\theta$, is exactly $(1-\cos \theta)s/(2\theta)$.
+The distance between a circular arc segment of length $s$, with angle between arc and chord of $\theta$ (see Figure~\ref{fig:arc_chord}), is exactly $(1-\cos \theta)s/(2\theta)$.
 The curvature is $\kappa = \frac{2\theta}{s}$ (equivalently, $\theta = \frac{\kappa s}{2s}$), and this remains constant even as the arc is subdivided.
@@
 Solving for n, we get:
+
+\begin{figure}
+    \includegraphics[scale=0.5]{arc_chord}
+    \caption{A circular arc segment with notations for angles ($\theta$), arc length ($s$), and distance to chord ($d$)}
+    \Description{A circular arc above its chord, with labels at the endpoints for angle from chord ($\theta$), the length of the arc ($s$) and the maximum distance from the arc to its chord ($d$).}
+    \label{fig:arc_chord}
+\end{figure}
+
+% TODO: I might supplement this figure with one subdivided with n=3
 
@@
 Applying the above error metric, the subdivision density is simply $\sqrt{|a(s-s_0)|}$.
-The integral is $\frac{2}{3}\sqrt{a}(s-s_0)^1.5$, which is readily invertible.
+The integral is $\frac{2}{3}\sqrt{a}(s-s_0)^{1.5}$, which is readily invertible.
 

### 2024-04-03 flatten/src/euler.rs
Start testbed stroke implementation
This copies much of the kurbo stroke implementation but adapts it for lowering. There are some improvements to the f64 flatten implementation, but it doesn't yet reflect all the robustness in the f32 version.

Both lowering to lines and arcs is supported.

@@ -272,6 +272,7 @@ impl EulerSeg {
 impl Iterator for CubicToEulerIter {
     type Item = EulerSeg;
 
+    // TODO: adapt robustness logic form euler32
     fn next(&mut self) -> Option<EulerSeg> {
         let t0 = (self.t0 as f64) * self.dt;
         if t0 == 1.0 {

### 2024-04-03 flatten/src/euler_arc.rs

@@ -3,9 +3,7 @@
 
 //! Logic for lowering Euler spirals to arcs
 
-use std::f64::consts::FRAC_PI_2;
-
-use kurbo::{common::GAUSS_LEGENDRE_COEFFS_32, Vec2};
+use kurbo::common::GAUSS_LEGENDRE_COEFFS_32;
 
 use crate::euler::{EulerParams, EulerSeg};
 
@@ -106,6 +104,7 @@ fn espc_arc_error(params: &EulerParams, d: f64, k: f64) -> f64 {
 //
 // We could wire up the precise area/k0 mapping, but it's probably more productive
 // to measure the error and tweak error metrics accordingly.
+#[allow(unused)]
 fn espc_to_arc_k(params: &EulerParams, d: f64) -> f64 {
     let chord = 1.0;
     let s = chord / params.ch;
@@ -132,7 +131,10 @@ fn arc_main_err() {
     let err = euler_arc_error(&ep, th);
     let est_err = 0.05 * dk;
     let est_err_k1 = -ep.k1 * 1. / 120.;
-    println!("err = {err:.3e}, scaled = {:.3e}, est = {est_err:.3e}, est k1 = {est_err_k1:.e}", err * ep.ch);
+    println!(
+        "err = {err:.3e}, scaled = {:.3e}, est = {est_err:.3e}, est k1 = {est_err_k1:.e}",
+        err * ep.ch
+    );
 }
 
 pub fn arc_main() {

### 2024-04-03 flatten/src/flatten.rs

@@ -5,38 +5,96 @@
 
 use std::f64::consts::{FRAC_1_SQRT_2, FRAC_PI_2, FRAC_PI_4};
 
-use kurbo::BezPath;
+use kurbo::{BezPath, Point};
 
 use crate::euler::EulerSeg;
 
-pub fn flatten_offset(iter: impl Iterator<Item = EulerSeg>, offset: f64) -> BezPath {
-    let mut result = BezPath::new();
-    let mut first = true;
-    for es in iter {
-        if core::mem::take(&mut first) {
-            result.move_to(es.eval_with_offset(0.0, offset));
-        }
-        let scale = es.p0.distance(es.p1);
-        let tol = 1.0;
-        let (k0, k1) = (es.params.k0 - 0.5 * es.params.k1, es.params.k1);
-        // compute forward integral to determine number of subdivisions
-        let dist_scaled = offset / scale;
-        let a = -2.0 * dist_scaled * k1;
-        let b = -1.0 - 2.0 * dist_scaled * k0;
-        let int0 = espc_int_approx(b);
+/// A robustness strategy for the ESPC integral
+enum EspcRobust {
+    /// Both k1 and dist are large enough to divide by robustly.
+    Normal,
+    /// k1 is low, so model curve as a circular arc.
+    LowK1,
+    /// dist is low, so model curve as just an Euler spiral.
+    LowDist,
+}
+
+/// Flatten an Euler spiral parallel curve to lines.
+///
+/// Invoke the callback for each point in the flattening. This includes the
+/// end point but not the start point.
+pub fn flatten_offset(es: &EulerSeg, offset: f64, tol: f64, mut f: impl FnMut(Point)) {
+    let scale = (es.p0 - es.p1).hypot();
+    let (k0, k1) = (es.params.k0 - 0.5 * es.params.k1, es.params.k1);
+    // Note: scaling by ch is missing from earlier implementations. The math
+    // should be validated carefully.
+    let dist_scaled = offset * es.params.ch / scale;
+    // The number of subdivisions for curvature = 1
+    let scale_multiplier = 0.5 * FRAC_1_SQRT_2 * (scale / (es.params.ch * tol)).sqrt();
+    // TODO: tune these thresholds
+    const K1_THRESH: f64 = 1e-3;
+    const DIST_THRESH: f64 = 1e-3;
+    let mut a = 0.0;
+    let mut b = 0.0;
+    let mut integral = 0.0;
+    let mut int0 = 0.0;
[…]

### 2024-04-03 flatten/src/main.rs

@@ -9,6 +9,7 @@ use flatten32::flatten_offset;
 
 use crate::euler::EulerParams;
 
+mod arc_segment;
 mod cubic32;
 mod euler;
 mod euler32;
@@ -16,6 +17,7 @@ mod euler_arc;
 mod evolute;
 mod flatten;
 mod flatten32;
+mod stroke;
 
 fn main_est_flatten_err() {
     let th0 = 0.101;
@@ -101,6 +103,7 @@ fn main() {
         "espc" => main_espc(),
         "est_flatten_err" => main_est_flatten_err(),
         "arc" => euler_arc::arc_main(),
+        "stroke" => stroke::stroke_main(),
         _ => println!("unknown figure type"),
     }
 }

### 2024-04-03 flatten/src/stroke.rs

@@ -1,0 +1,605 @@
+// Copyright 2024 the Vello Authors
+// SPDX-License-Identifier: Apache-2.0
+
+//! CPU-side stroke to fill expansion.
+//!
+//! Much of the contents of this file is copy-pasted from kurbo, but adapted for lowering
+//! to lines or arcs.
+
+use std::f64::consts::PI;
+use std::fmt::Write;
+
+use kurbo::{
+    common::solve_quadratic, Affine, Arc, BezPath, Cap, CubicBez, Join, Line, PathEl, PathSeg,
+    Point, QuadBez, Stroke, Vec2,
+};
+
+use crate::{
+    arc_segment::ArcSegment,
+    euler::{CubicToEulerIter, EulerSeg},
+    flatten::flatten_offset,
+};
+
+trait Lowering: Sized + Copy + Clone {
+    // Possibly useful for debugging but not currently used.
+    fn to_bez(&self, path: &mut BezPath);
+
+    fn to_svg_string(&self, svg: &mut String, last_pt: Option<Point>);
+
+    fn lower_espc(es: &EulerSeg, path: &mut LoweredPath<Self>, dist: f64, tol: f64);
+
+    fn lower_arc(arc: &ArcSegment, segs: &mut Vec<Self>, tol: f64);
+
+    fn line(line: Line) -> Self;
+
+    fn reverse(self) -> Self;
+
+    fn end_point(self) -> Point;
+}
+
+#[derive(Debug)]
+struct LoweredPath<L: Lowering> {
+    path: Vec<L>,
+    last_pt: Point,
+}
+
+impl<L: Lowering> Default for LoweredPath<L> {
+    fn default() -> Self {
+        LoweredPath {
+            path: vec![],
+            last_pt: Point::default(),
+        }
+    }
+}
+
+impl<L: Lowering> LoweredPath<L> {
+    fn move_to(&mut self, p: Point) {
+        self.last_pt = p;
+    }
+
+    fn line_to(&mut self, p: Point) {
+        self.path.push(L::line(Line::new(self.last_pt, p)));
+        self.last_pt = p;
+    }
+
+    #[allow(unused)]
+    fn to_bez(&self) -> BezPath {
+        let mut result = BezPath::new();
+        for seg in &self.path {
+            seg.to_bez(&mut result);
+        }
+        result
+    }
+
+    fn to_svg(&self) -> String {
+        let mut svg = String::new();
+        let mut last_pt = None;
[…]

### 2024-04-04 flatten/src/main.rs
Wire up command line arguments to tests
This lets us stroke-expand the paths in SVG files. Current version is very minimal.

@@ -1,6 +1,7 @@
 // Copyright 2024 the Vello Authors
 // SPDX-License-Identifier: Apache-2.0
 
+use clap::Parser;
 use cubic32::{Cubic, Point};
 use euler::EulerSeg;
 use euler32::CubicToEulerIter;
@@ -18,6 +19,18 @@ mod evolute;
 mod flatten;
 mod flatten32;
 mod stroke;
+mod svg;
+
+#[derive(Parser)]
+enum Args {
+    Evolute,
+    Cubic,
+    Espc,
+    EstFlattenErr,
+    Arc,
+    Stroke,
+    Svg(svg::SvgArgs),
+}
 
 fn main_est_flatten_err() {
     let th0 = 0.101;
@@ -86,10 +99,10 @@ fn euler_evolute_main() {
 }
 
 fn main() {
-    let arg = std::env::args().nth(1).expect("provide figure type as arg");
-    match &*arg {
-        "evolute" => euler_evolute_main(),
-        "cubic" => {
+    let args = Args::parse();
+    match args {
+        Args::Evolute => euler_evolute_main(),
+        Args::Cubic => {
             let c = Cubic::new(
                 Point::new(0.0, 0.0),
                 Point::new(0.0, 0.0),
@@ -100,11 +113,11 @@ fn main() {
             let path = flatten_offset(iter, 0.0, 0.1);
             println!("{}", path.to_svg());
         }
-        "espc" => main_espc(),
-        "est_flatten_err" => main_est_flatten_err(),
-        "arc" => euler_arc::arc_main(),
-        "stroke" => stroke::stroke_main(),
-        _ => println!("unknown figure type"),
+        Args::Espc => main_espc(),
+        Args::EstFlattenErr => main_est_flatten_err(),
+        Args::Arc => euler_arc::arc_main(),
+        Args::Stroke => stroke::stroke_main(),
+        Args::Svg(args) => svg::svg_main(args),
     }
 }
 

### 2024-04-04 flatten/src/stroke.rs

@@ -20,7 +20,7 @@ use crate::{
     flatten::flatten_offset,
 };
 
-trait Lowering: Sized + Copy + Clone {
+pub trait Lowering: Sized + Copy + Clone {
     // Possibly useful for debugging but not currently used.
     fn to_bez(&self, path: &mut BezPath);
 
@@ -38,7 +38,7 @@ trait Lowering: Sized + Copy + Clone {
 }
 
 #[derive(Debug)]
-struct LoweredPath<L: Lowering> {
+pub struct LoweredPath<L: Lowering> {
     path: Vec<L>,
     last_pt: Point,
 }
@@ -71,7 +71,7 @@ impl<L: Lowering> LoweredPath<L> {
         result
     }
 
-    fn to_svg(&self) -> String {
+    pub fn to_svg(&self) -> String {
         let mut svg = String::new();
         let mut last_pt = None;
         for seg in &self.path {
@@ -290,7 +290,7 @@ struct StrokeCtx<L: Lowering> {
 }
 
 /// Version of stroke expansion for styles with no dashes.
-fn stroke_undashed<L: Lowering>(
+pub fn stroke_undashed<L: Lowering>(
     path: impl IntoIterator<Item = PathEl>,
     style: &Stroke,
     tolerance: f64,
@@ -599,7 +599,8 @@ impl<L: Lowering> StrokeCtx<L> {
 pub fn stroke_main() {
     let path = BezPath::from_svg("M 10 10 Q 50 10 100 20 Q 100 100 50 100").unwrap();
     //let path = BezPath::from_svg("M 10 10 L 100 20 L 50 100").unwrap();
-    let style = Stroke::new(10.0).with_caps(Cap::Round);
-    let stroked: LoweredPath<ArcSegment> = stroke_undashed(path, &style, 0.1);
+    let style = Stroke::new(50.0).with_caps(Cap::Round);
+    let stroked: LoweredPath<ArcSegment> = stroke_undashed(path, &style, 1.0);
     println!("{}", stroked.to_svg());
+    //println!("{} segments", stroked.path.len());
 }

### 2024-04-04 flatten/src/svg.rs

@@ -1,0 +1,139 @@
+// Copyright 2024 the Vello Authors
+// SPDX-License-Identifier: Apache-2.0
+
+//! A very basic utility to extract SVG paths.
+
+use std::time::Instant;
+use std::{error::Error, io::Write};
+use std::str::FromStr;
+
+use kurbo::{BezPath, Line, Stroke};
+use roxmltree::{Document, Node};
+
+use crate::arc_segment::ArcSegment;
+use crate::stroke::{stroke_undashed, LoweredPath, Lowering};
+
+pub struct StyledPath {
+    pub path: BezPath,
+    pub style: Stroke,
+    // can add color etc, but it's irrelevant for stroke expansion
+}
+
+pub struct SvgScene {
+    pub paths: Vec<StyledPath>,
+    // maybe viewport and extra stuff
+}
+
+impl SvgScene {
+    pub fn load(xml_string: &str) -> Result<SvgScene, Box<dyn Error>> {
+        let doc = Document::parse(xml_string)?;
+        let root = doc.root_element();
+        let mut scene = SvgScene { paths: vec![] };
+        for node in root.children() {
+            parse_rec(node, &mut scene)?;
+        }
+        Ok(scene)
+    }
+}
+
+fn parse_rec(node: Node, scene: &mut SvgScene) -> Result<(), Box<dyn Error>> {
+    match node.tag_name().name() {
+        "g" => {
+            for child in node.children() {
+                parse_rec(child, scene)?;
+            }
+        }
+        "path" => {
+            let d = node.attribute("d").ok_or("missing 'd'")?;
+            let path = BezPath::from_svg(d)?;
+            let width = node
+                .attribute("stroke-width")
+                .map(|a| f64::from_str(a).unwrap_or(1.0))
+                .unwrap_or(4.0);
+            let style = Stroke::new(width);
+            // TODO: cap and join styles
+            scene.paths.push(StyledPath { path, style });
+        }
+        _ => (),
+    }
+    Ok(())
+}
+
+impl SvgScene {
+    fn expand<L: Lowering>(&self, tolerance: f64) -> Vec<LoweredPath<L>> {
+        self.paths
+            .iter()
+            .map(|path| stroke_undashed(&path.path, &path.style, tolerance))
+            .collect()
+    }
+
[…]

### 2024-04-04 flatten/src/main.rs
Wire up skia's stroke expander
Use the skia-safe crate to be able to call into Skia. There's a bunch of rough edges, but it's enough to start getting comparisons.

@@ -18,6 +18,8 @@ mod euler_arc;
 mod evolute;
 mod flatten;
 mod flatten32;
+#[cfg(feature = "skia-safe")]
+mod skia;
 mod stroke;
 mod svg;
 

### 2024-04-04 flatten/src/stroke.rs

@@ -39,7 +39,7 @@ pub trait Lowering: Sized + Copy + Clone {
 
 #[derive(Debug)]
 pub struct LoweredPath<L: Lowering> {
-    path: Vec<L>,
+    pub path: Vec<L>,
     last_pt: Point,
 }
 

### 2024-04-04 flatten/src/svg.rs

@@ -90,6 +90,8 @@ pub struct SvgArgs {
     tolerance: Option<f64>,
     #[arg(short, long)]
     primitive: Option<String>,
+    #[arg(short, long)]
+    output_file: Option<String>,
 }
 
 #[derive(Clone, Parser)]
@@ -114,26 +116,38 @@ pub fn svg_main(args: SvgArgs) {
         Some("a") => PrimType::Arc,
         _ => PrimType::Arc,
     };
-    for _ in 0..1 {
+    for _ in 0..10 {
         let start = Instant::now();
         match prim_type {
             PrimType::Line => {
-                let _paths: Vec<LoweredPath<Line>> = scene.expand(tolerance);
+                let paths: Vec<LoweredPath<Line>> = scene.expand(tolerance);
+                let total_segs: usize = paths.iter().map(|p| p.path.len()).sum();
+                println!("{total_segs} lines, {:?}", start.elapsed());
             }
             PrimType::Arc => {
-                let _paths: Vec<LoweredPath<ArcSegment>> = scene.expand(tolerance);
+                let paths: Vec<LoweredPath<ArcSegment>> = scene.expand(tolerance);
+                let total_segs: usize = paths.iter().map(|p| p.path.len()).sum();
+                println!("{total_segs} arcs, {:?}", start.elapsed());
             }
         }
-        eprintln!("{:?}", start.elapsed());
     }
-    match prim_type {
-        PrimType::Line => {
-            let paths: Vec<LoweredPath<Line>> = scene.expand(tolerance);
-            scene.to_svg(&mut std::io::stdout().lock(), &paths).unwrap();
+    if let Some(out_file_name) = args.output_file {
+        let mut f = std::fs::File::create(out_file_name).unwrap();
+        match prim_type {
+            PrimType::Line => {
+                let paths: Vec<LoweredPath<Line>> = scene.expand(tolerance);
+                scene.to_svg(&mut f, &paths).unwrap();
+            }
+            PrimType::Arc => {
+                let paths: Vec<LoweredPath<ArcSegment>> = scene.expand(tolerance);
+                scene.to_svg(&mut f, &paths).unwrap();
+            }
         }
-        PrimType::Arc => {
-            let paths: Vec<LoweredPath<ArcSegment>> = scene.expand(tolerance);
-            scene.to_svg(&mut std::io::stdout().lock(), &paths).unwrap();
-        }
+    }
+    #[cfg(feature = "skia-safe")]
[…]

### 2024-04-05 flatten/src/euler.rs
Evolutes
This is a working implementation of evolutes. There are some limitations. The lowering of evolutes is crude. Also, there isn't an option to enable or disable them.

@@ -229,6 +229,11 @@ impl EulerParams {
         let n = sum * (1.0 / N as f64) * 0.125f64.sqrt() / self.ch;
         n * n
     }
+
+    fn eval_evolute(&self, t: f64) -> Point {
+        let offset = -1.0 / self.curvature(t);
+        self.eval_with_offset(t, offset)
+    }
 }
 
 impl EulerSeg {
@@ -262,6 +267,15 @@ impl EulerSeg {
         let chord = self.p1 - self.p0;
         let scaled = offset / chord.hypot();
         let Point { x, y } = self.params.eval_with_offset(t, scaled);
+        Point::new(
+            self.p0.x + chord.x * x - chord.y * y,
+            self.p0.y + chord.x * y + chord.y * x,
+        )
+    }
+
+    pub fn eval_evolute(&self, t: f64) -> Point {
+        let chord = self.p1 - self.p0;
+        let Point { x, y } = self.params.eval_evolute(t);
         Point::new(
             self.p0.x + chord.x * x - chord.y * y,
             self.p0.y + chord.x * y + chord.y * x,

### 2024-04-05 flatten/src/evolute.rs

@@ -3,32 +3,18 @@
 
 //! Math for flattening of Euler evolute
 
-use kurbo::{BezPath, Point};
+use kurbo::BezPath;
 
-use crate::euler::{EulerParams, EulerSeg};
-
-fn eval_es_params_evolute(params: &EulerParams, t: f64) -> Point {
-    let offset = -1.0 / params.curvature(t);
-    params.eval_with_offset(t, offset)
-}
-
-fn eval_es_evolute(es: &EulerSeg, t: f64) -> Point {
-    let chord = es.p1 - es.p0;
-    let Point { x, y } = eval_es_params_evolute(&es.params, t);
-    Point::new(
-        es.p0.x + chord.x * x - chord.y * y,
-        es.p0.y + chord.x * y + chord.y * x,
-    )
-}
+use crate::euler::EulerSeg;
 
 /// Flatten the evolute of an Euler spiral segment.
-pub fn flatten_es_evolute(es: &EulerSeg, tol: f64) -> BezPath {
+pub fn flatten_es_evolute(es: &EulerSeg, _tol: f64) -> BezPath {
     let mut path = BezPath::new();
     // TODO: smart flatten logic
     const N: usize = 10;
     for i in 0..=N {
         let t = i as f64 / N as f64;
-        let p = eval_es_evolute(es, t);
+        let p = es.eval_evolute(t);
         if i == 0 {
             path.move_to(p);
         } else {
@@ -41,7 +27,7 @@ pub fn flatten_es_evolute(es: &EulerSeg, tol: f64) -> BezPath {
 /// Flatten an Euler spiral segment.
 ///
 /// This is duplicative of other work but maybe useful for exploration.
-pub fn flatten_es(es: &EulerSeg, tol: f64) -> BezPath {
+pub fn flatten_es(es: &EulerSeg, _tol: f64) -> BezPath {
     let mut path = BezPath::new();
     // TODO: smart flatten logic
     const N: usize = 10;

### 2024-04-05 flatten/src/flatten.rs

@@ -3,7 +3,10 @@
 
 //! Math for flattening of Euler spiral parallel curve
 
-use std::f64::consts::{FRAC_1_SQRT_2, FRAC_PI_2, FRAC_PI_4};
+use std::{
+    f64::consts::{FRAC_1_SQRT_2, FRAC_PI_2, FRAC_PI_4},
+    ops::Range,
+};
 
 use kurbo::{BezPath, Point};
 
@@ -23,9 +26,17 @@ enum EspcRobust {
 ///
 /// Invoke the callback for each point in the flattening. This includes the
 /// end point but not the start point.
-pub fn flatten_offset(es: &EulerSeg, offset: f64, tol: f64, mut f: impl FnMut(Point)) {
+pub fn flatten_offset(
+    es: &EulerSeg,
+    range: Range<f64>,
+    offset: f64,
+    tol: f64,
+    mut f: impl FnMut(Point),
+) {
     let scale = (es.p0 - es.p1).hypot();
-    let (k0, k1) = (es.params.k0 - 0.5 * es.params.k1, es.params.k1);
+    let range_size = range.end - range.start;
+    let k0 = es.params.k0 + (range.start - 0.5) * es.params.k1;
+    let k1 = es.params.k1 * range_size;
     // Note: scaling by ch is missing from earlier implementations. The math
     // should be validated carefully.
     let dist_scaled = offset * es.params.ch / scale;
@@ -64,7 +75,7 @@ pub fn flatten_offset(es: &EulerSeg, offset: f64, tol: f64, mut f: impl FnMut(Po
         let n_frac = scaled_int;
         (n_frac, EspcRobust::Normal)
     };
-    let n = (n_frac * scale_multiplier).ceil().max(1.0);
+    let n = (n_frac * scale_multiplier * range_size).ceil().max(1.0);
     for i in 0..n as usize - 1 {
         let t = (i + 1) as f64 / n;
         let s = match robust {
@@ -80,9 +91,9 @@ pub fn flatten_offset(es: &EulerSeg, offset: f64, tol: f64, mut f: impl FnMut(Po
                 (inv - b) / a
             }
         };
-        f(es.eval_with_offset(s, offset));
+        f(es.eval_with_offset(range.start + range_size * s, offset));
     }
-    f(es.eval_with_offset(1.0, offset));
+    f(es.eval_with_offset(range.end, offset));
 }
 
[…]

### 2024-04-05 flatten/src/stroke.rs

@@ -6,8 +6,8 @@
 //! Much of the contents of this file is copy-pasted from kurbo, but adapted for lowering
 //! to lines or arcs.
 
-use std::f64::consts::PI;
 use std::fmt::Write;
+use std::{f64::consts::PI, ops::Range};
 
 use kurbo::{
     common::solve_quadratic, Affine, Arc, BezPath, Cap, CubicBez, Join, Line, PathEl, PathSeg,
@@ -20,17 +20,35 @@ use crate::{
     flatten::flatten_offset,
 };
 
-pub trait Lowering: Sized + Copy + Clone {
+pub trait Lowering: Sized + Copy + Clone + core::fmt::Debug {
     // Possibly useful for debugging but not currently used.
     fn to_bez(&self, path: &mut BezPath);
 
     fn to_svg_string(&self, svg: &mut String, last_pt: Option<Point>);
 
-    fn lower_espc(es: &EulerSeg, path: &mut LoweredPath<Self>, dist: f64, tol: f64);
+    fn line(line: Line) -> Self;
 
     fn lower_arc(arc: &ArcSegment, segs: &mut Vec<Self>, tol: f64);
 
-    fn line(line: Line) -> Self;
+    // Note: assumes last_pt in the path is set, and is continuous with it.
+    fn lower_espc(
+        es: &EulerSeg,
+        path: &mut LoweredPath<Self>,
+        range: Range<f64>,
+        dist: f64,
+        tol: f64,
+    );
+
+    fn lower_es_evolute(es: &EulerSeg, path: &mut LoweredPath<Self>, range: Range<f64>, _tol: f64) {
+        // placeholder implementation
+        // TODO: pull error metric rabbit from hat
+        const N: usize = 5;
+        let dt = (range.end - range.start) / N as f64;
+        for i in 0..N {
+            let t = range.start + (i + 1) as f64 * dt;
+            path.line_to(es.eval_evolute(t));
+        }
+    }
 
     fn reverse(self) -> Self;
 
@@ -43,11 +61,28 @@ pub struct LoweredPath<L: Lowering> {
     last_pt: Point,
 }
 
+struct StrokeContour<L: Lowering> {
+    main_path: LoweredPath<L>,
+    // the next two items are in play past the cusp
+    rev_parallel: Option<LoweredPath<L>>,
+    evolute: Option<LoweredPath<L>>,
+}
+
 impl<L: Lowering> Default for LoweredPath<L> {
     fn default() -> Self {
[…]

### 2024-04-05 flatten/src/svg.rs

@@ -3,11 +3,11 @@
 
 //! A very basic utility to extract SVG paths.
 
+use std::str::FromStr;
 use std::time::Instant;
 use std::{error::Error, io::Write};
-use std::str::FromStr;
 
-use kurbo::{BezPath, Line, Stroke};
+use kurbo::{BezPath, Cap, Join, Line, Stroke};
 use roxmltree::{Document, Node};
 
 use crate::arc_segment::ArcSegment;
@@ -50,8 +50,24 @@ fn parse_rec(node: Node, scene: &mut SvgScene) -> Result<(), Box<dyn Error>> {
                 .attribute("stroke-width")
                 .map(|a| f64::from_str(a).unwrap_or(1.0))
                 .unwrap_or(4.0);
-            let style = Stroke::new(width);
-            // TODO: cap and join styles
+            let mut cap = Cap::Butt;
+            if let Some(linecap) = node.attribute("stroke-linecap") {
+                match linecap {
+                    "round" => cap = Cap::Round,
+                    "square" => cap = Cap::Square,
+                    _ => (),
+                }
+            }
+            let mut join = Join::Miter;
+            if let Some(linejoin) = node.attribute("stroke-linejoin") {
+                match linejoin {
+                    "round" => join = Join::Round,
+                    "bevel" => join = Join::Bevel,
+                    _ => (),
+                }
+            }
+            // TODO: miter limit
+            let style = Stroke::new(width).with_caps(cap).with_join(join);
             scene.paths.push(StyledPath { path, style });
         }
         _ => (),
@@ -67,7 +83,11 @@ impl SvgScene {
             .collect()
     }
 
-    fn to_svg(&self, out: &mut impl Write, paths: &[LoweredPath<impl Lowering>]) -> Result<(), Box<dyn Error>> {
+    fn to_svg(
+        &self,
+        out: &mut impl Write,
+        paths: &[LoweredPath<impl Lowering>],
+    ) -> Result<(), Box<dyn Error>> {
         // these should probably be fields in the scene
         let width = 1024;
         let height = 1024;
@@ -116,7 +136,7 @@ pub fn svg_main(args: SvgArgs) {
         Some("a") => PrimType::Arc,
         _ => PrimType::Arc,
     };
-    for _ in 0..10 {
+    for _ in 0..1 {
         let start = Instant::now();
         match prim_type {
             PrimType::Line => {

### 2024-04-06 flatten/src/evolute.rs
Smart flattening of evolutes
Use fancy error metric to flatten evolutes to lines. Smart lowering to arcs still remains.

@@ -5,16 +5,24 @@
 
 use kurbo::BezPath;
 
-use crate::euler::EulerSeg;
+use crate::euler::{EulerParams, EulerSeg};
 
 /// Flatten the evolute of an Euler spiral segment.
-pub fn flatten_es_evolute(es: &EulerSeg, _tol: f64) -> BezPath {
+pub fn flatten_es_evolute(es: &EulerSeg, tol: f64) -> BezPath {
     let mut path = BezPath::new();
-    // TODO: smart flatten logic
-    const N: usize = 10;
-    for i in 0..=N {
-        let t = i as f64 / N as f64;
-        let p = es.eval_evolute(t);
+    let arc_len = (es.p1 - es.p0).length() / es.params.ch;
+    let ratio = es.params.k0 / es.params.k1;
+    let rho_int_0 = (0.5 * (ratio - 0.5)).abs().sqrt();
+    let rho_int_1 = (0.5 * (ratio + 0.5)).abs().sqrt();
+    let rho_int = rho_int_1 - rho_int_0;
+    let n_subdiv = (rho_int.abs() * (arc_len / tol).sqrt()).ceil().max(1.0);
+    let n = n_subdiv as usize;
+    let sign2 = 2.0f64.copysign(ratio);
+    for i in 0..=n {
+        let t = i as f64 / n_subdiv;
+        let u = rho_int_0 + t * rho_int;
+        let s = sign2 * u * u + 0.5 - ratio;
+        let p = es.eval_evolute(s);
         if i == 0 {
             path.move_to(p);
         } else {
@@ -41,4 +49,65 @@ pub fn flatten_es(es: &EulerSeg, _tol: f64) -> BezPath {
         }
     }
     path
+}
+
+#[allow(unused)]
+pub fn euler_evolute_scratch() {
+    // This is a sketchpad for numerically verifying the subdivision
+    // density of the evolute of an Euler spiral.
+    let es_params = EulerParams::from_angles(0.7, 1.0);
+    let es = EulerSeg::from_params(
+        kurbo::Point::new(100., 100.),
+        kurbo::Point::new(300., 100.),
+        es_params,
+    );
+    let arc_len = (es.p1 - es.p0).length() / es.params.ch;
+    println!("analytic arc length: {arc_len}");
+    const N: usize = 10;
+    let rho_int_0 = (0.5 * arc_len * (es.params.k0 / es.params.k1 - 0.5)).sqrt();
+    for i in 0..=N {
+        let t = i as f64 / N as f64;
[…]

### 2024-04-06 flatten/src/main.rs

@@ -3,7 +3,6 @@
 
 use clap::Parser;
 use cubic32::{Cubic, Point};
-use euler::EulerSeg;
 use euler32::CubicToEulerIter;
 use flatten::{espc_integral_inv, n_subdiv_analytic};
 use flatten32::flatten_offset;
@@ -87,23 +86,10 @@ fn main_espc() {
     }
 }
 
-fn euler_evolute_main() {
-    let es_params = EulerParams::from_angles(0.7, 1.0);
-    let es = EulerSeg::from_params(
-        kurbo::Point::new(100., 100.),
-        kurbo::Point::new(300., 100.),
-        es_params,
-    );
-    let path = evolute::flatten_es_evolute(&es, 1.0);
-    println!("{}", path.to_svg());
-    let path = evolute::flatten_es(&es, 1.0);
-    println!("{}", path.to_svg());
-}
-
 fn main() {
     let args = Args::parse();
     match args {
-        Args::Evolute => euler_evolute_main(),
+        Args::Evolute => evolute::euler_evolute_main(),
         Args::Cubic => {
             let c = Cubic::new(
                 Point::new(0.0, 0.0),

### 2024-04-06 flatten/src/stroke.rs

@@ -39,14 +39,23 @@ pub trait Lowering: Sized + Copy + Clone + core::fmt::Debug {
         tol: f64,
     );
 
-    fn lower_es_evolute(es: &EulerSeg, path: &mut LoweredPath<Self>, range: Range<f64>, _tol: f64) {
-        // placeholder implementation
-        // TODO: pull error metric rabbit from hat
-        const N: usize = 5;
-        let dt = (range.end - range.start) / N as f64;
-        for i in 0..N {
-            let t = range.start + (i + 1) as f64 * dt;
-            path.line_to(es.eval_evolute(t));
+    fn lower_es_evolute(es: &EulerSeg, path: &mut LoweredPath<Self>, range: Range<f64>, tol: f64) {
+        let range_size = range.end - range.start;
+        let arc_len = (es.p1 - es.p0).length() / es.params.ch;
+        let ratio = es.params.k0 / es.params.k1;
+        let rho_int_0 = (0.5 * (ratio + range.start - 0.5)).abs().sqrt();
+        let rho_int_1 = (0.5 * (ratio + range.end - 0.5)).abs().sqrt();
+        let rho_int = rho_int_1 - rho_int_0;
+        let n_subdiv = (range_size * rho_int.abs() * (arc_len / tol).sqrt())
+            .ceil()
+            .max(1.0);
+        let n = n_subdiv as usize;
+        let sign2 = 2.0f64.copysign(ratio);
+        for i in 1..=n {
+            let t = i as f64 / n_subdiv;
+            let u = rho_int_0 + t * rho_int;
+            let s = range.start + range_size * (sign2 * u * u + 0.5 - ratio);
+            path.line_to(es.eval_evolute(s));
         }
     }
 
@@ -686,7 +695,6 @@ impl<L: Lowering + core::fmt::Debug> StrokeContour<L> {
             self.finalize();
             L::lower_espc(es, &mut self.main_path, 0.0..t, h, tolerance);
             if t < 1.0 {
-                println!("start cusp {:?}", self.main_path.last_pt);
                 let (evolute, rev_parallel) = self.start();
                 L::lower_es_evolute(es, evolute, t..1.0, tolerance);
                 L::lower_espc(es, rev_parallel, t..1.0, h, tolerance);

### 2024-04-06 flatten/src/evolute.rs
Checkpoint ES evolute lowering to arcs
Fine-tuning the math

@@ -3,9 +3,9 @@
 
 //! Math for flattening of Euler evolute
 
-use kurbo::BezPath;
+use kurbo::{Arc, BezPath};
 
-use crate::euler::{EulerParams, EulerSeg};
+use crate::{arc_segment::ArcSegment, euler::{EulerParams, EulerSeg}};
 
 /// Flatten the evolute of an Euler spiral segment.
 pub fn flatten_es_evolute(es: &EulerSeg, tol: f64) -> BezPath {
@@ -26,11 +26,52 @@ pub fn flatten_es_evolute(es: &EulerSeg, tol: f64) -> BezPath {
         if i == 0 {
             path.move_to(p);
         } else {
-            path.line_to(p)
+            path.line_to(p);
         }
     }
     path
 }
+
+pub fn lower_es_evolute_arc(es: &EulerSeg, tol: f64) -> BezPath {
+    let mut path = BezPath::new();
+    let arc_len = (es.p1 - es.p0).length() / es.params.ch;
+    let k0 = es.params.k0;
+    let k1 = es.params.k1;
+    println!("k0 {k0} k1 {k1}");
+    let rho_int_0 = ((27. / 40.) / (k0 - 0.5 * k1)).cbrt();
+    let rho_int_1 = ((27. / 40.) / (k0 + 0.5 * k1)).cbrt();
+    let rho_int = rho_int_1 - rho_int_0;
+    let n_subdiv = (rho_int.abs() * (arc_len / tol).cbrt()).ceil().max(1.0);
+    let n = n_subdiv as usize;
+    let mut last_s = 0.0;
+    let mut last_s_bar = k0 / k1 - 0.5;
+    let mut last_r = last_s_bar.abs().sqrt();
+    let mut p0 = es.eval_evolute(0.0);
+    path.move_to(p0);
+    for i in 1..=n {
+        let t = i as f64 / n_subdiv;
+        let u = rho_int_0 + t * rho_int;
+        let s = ((27. / 40.) / u.powi(3) - k0) / k1 + 0.5;
+        let p1 = es.eval_evolute(s);
+        let avg_k = k0 + (0.5 * (s + last_s) - 0.5) * k1;
+        let s_bar = k0 / k1 + s - 0.5;
+        let r = s_bar.abs().sqrt();
+        let new_k = 2. * (r - last_r).powi(2) / (1. / s_bar - 1. / last_s_bar);
+        let new_k = -2. * new_k * k1.abs();
[…]

### 2024-04-07 flatten/src/euler.rs
Lower evolutes to arcs
The code now lowers Euler spiral segments to arcs, and it works well. The lowering logic is both in evolute.rs (lower_es_evolute_arc) and the Lowering implementation for arcs in stroke.rs

The PR also fixes a range bug in ESPC to arc lowering.

@@ -123,6 +123,10 @@ impl EulerParams {
         (self.k0 + 0.5 * self.k1 * (t - 1.0)) * t - self.th0
     }
 
+    /// Curvature scaled to a unit chord.
+    ///
+    /// Curvature for an arbitrary scaling can be determined by dividing by
+    /// the chord length.
     pub fn curvature(&self, t: f64) -> f64 {
         let raw_k = self.k0 + self.k1 * (t - 0.5);
         raw_k * self.ch

### 2024-04-07 flatten/src/evolute.rs

@@ -43,9 +43,8 @@ pub fn lower_es_evolute_arc(es: &EulerSeg, tol: f64) -> BezPath {
     let rho_int = rho_int_1 - rho_int_0;
     let n_subdiv = (rho_int.abs() * (arc_len / tol).cbrt()).ceil().max(1.0);
     let n = n_subdiv as usize;
-    let mut last_s = 0.0;
-    let mut last_s_bar = k0 / k1 - 0.5;
-    let mut last_r = last_s_bar.abs().sqrt();
+    let mut last_k = k0 - 0.5 * k1;
+    let mut last_r2 = last_k.abs().sqrt();
     let mut p0 = es.eval_evolute(0.0);
     path.move_to(p0);
     for i in 1..=n {
@@ -53,20 +52,15 @@ pub fn lower_es_evolute_arc(es: &EulerSeg, tol: f64) -> BezPath {
         let u = rho_int_0 + t * rho_int;
         let s = ((27. / 40.) / u.powi(3) - k0) / k1 + 0.5;
         let p1 = es.eval_evolute(s);
-        let avg_k = k0 + (0.5 * (s + last_s) - 0.5) * k1;
-        let s_bar = k0 / k1 + s - 0.5;
-        let r = s_bar.abs().sqrt();
-        let new_k = 2. * (r - last_r).powi(2) / (1. / s_bar - 1. / last_s_bar);
-        let new_k = -2. * new_k * k1.abs();
-        println!("{t} {s} old {} new {new_k}", avg_k * (s - last_s));
-        let old_k = avg_k * (s - last_s);
-        let arc = ArcSegment::new(p0, p1, new_k);
+        let k = k0 + (s - 0.5) * k1;
+        let r2 = k.abs().sqrt();
+        let es_k = -4. * (r2 - last_r2).powi(2) / (k1 / k - k1 / last_k);
+        let arc = ArcSegment::new(p0, p1, es_k);
         if let Some(arc) = Arc::from_svg_arc(&arc.to_svg_arc()) {
             path.extend(arc.append_iter(0.1));
         }
-        last_s = s;
-        last_s_bar = s_bar;
-        last_r = r;
+        last_k = k;
+        last_r2 = r2;
         p0 = p1;
     }
     path

### 2024-04-07 flatten/src/stroke.rs

@@ -196,7 +196,6 @@ impl Lowering for Line {
         dist: f64,
         tol: f64,
     ) {
-        // TODO: take range into account
         flatten_offset(es, range, dist, tol, |p| path.line_to(p));
     }
 
@@ -279,6 +278,40 @@ impl Lowering for ArcSegment {
             let t = t0 + 0.5 * dt - 0.5;
             let k = es.params.k0 + t * es.params.k1;
             path.path.push(ArcSegment::new(p0, p1, k * dt));
+            p0 = p1;
+        }
+        path.last_pt = p0;
+    }
+
+    fn lower_es_evolute(es: &EulerSeg, path: &mut LoweredPath<Self>, range: Range<f64>, tol: f64) {
+        let range_size = range.end - range.start;
+        let arc_len = (es.p1 - es.p0).length() / es.params.ch;
+        let k0 = es.params.k0;
+        let k1 = es.params.k1;
+        let rho_int_0 = (1. / (k0 + (range.start - 0.5) * k1)).cbrt();
+        let rho_int_1 = (1. / (k0 + (range.end - 0.5) * k1)).cbrt();
+        let rho_int = rho_int_1 - rho_int_0;
+        let est_err = rho_int.abs() * arc_len / tol;
+        // TODO: apply scaling factor to est_err to tune actual error.
+        // Math suggests 27. / 40. but eyeball error seems too large.
+        let n_subdiv = (est_err.cbrt() * range_size).ceil().max(1.0);
+        let n = n_subdiv as usize;
+        let mut p0 = path.last_pt;
+        let mut last_k = k0 + (range.start - 0.5) * k1;
+        let mut last_r = last_k.abs().sqrt();
+        let sign4 = 4.0f64.copysign(-k0);
+        for i in 1..=n {
+            let t = i as f64 / n_subdiv;
+            let u = rho_int_0 + t * rho_int;
+            let s_0_1 = (1. / u.powi(3) - k0) / k1 + 0.5;
+            let s = range.start + range_size * s_0_1;
+            let p1 = es.eval_evolute(s);
+            let k = k0 + (s - 0.5) * k1;
+            let r = k.abs().sqrt();
+            let es_k = sign4 * (r - last_r).powi(2) / (k1 / k - k1 / last_k);
+            path.path.push(ArcSegment::new(p0, p1, es_k));
+            last_k = k;
+            last_r = r;
             p0 = p1;
         }
         path.last_pt = p0;

### 2024-04-08 paper.tex
Address review feedback in paper
This is a fairly small set of changes in response to review, including new insight into the hybrid nature of flattening in the Skia and Rive renderers.

@@
 As described in detail in \cite{Nehab2020}, very few existing implementations actually implement the strong version, so document authors have become accustomed to not depending on behavior at endpoints.
-Standards for graphics formats also provide an ``out,'' enabling the weaker behavior.
-[TODO: find language in SVG standard].
+Standards for graphics formats, such as SVG\cite{Svg2}, also provide an ``out,'' enabling the weaker behavior.
 While our current implementation emphasizes weak correctness, we believe it can be extended to the stronger sense, by implementing evolutes and inner join contours.
@@
 The core problem in stroke expansion is approximating the desired curve by segments of some other curve, usually a simpler one.
-These segments must be within an error tolerance of the source curve, and ideally there are not many more of them than optimal.
+These segments must be within an error tolerance of the source curve, and ideally close to a minimal number of them.
 We consider a number of source to target pairs, most importantly cubic Béziers to Euler spirals, and Euler spiral parallel curves to either lines or arcs.
@@
 If the error is within tolerance, the approximation is accepted.
-Otherwise, the curve is subdivided (usually at t = 0.5) and each half is recursively approximated.
+Otherwise, the curve is subdivided (usually at $t = 0.5$) and each half is recursively approximated.
 A substantial fraction of all curve approximation methods in the literature are of this form, including \cite{Nehab2020}.
@@
 This metric is conservative but works well in practice; among other applications, it is used in Skia for path flattening.
-One limitation of Wang's method is that it only applies to the flattening of filled outlines, not parallel curves as is required for strokes.
+One limitation of Wang's method is that it only applies to the flattening of filled outlines; when applied naively to generation of parallel curves, it can undershoot substantially, especially near cusps.
 
@@
 Using a subpixel Fréchet distance bound guarantees that the rendered image does not deviate visibly from the exact rendering.
-Another choice would be uniform steps in tangent angle, as chosen by \cite{Kilgard2020}.
+Another choice would be uniform steps in tangent angle, as chosen by polar stroking\cite{Kilgard2020}.
 However, at small curvature, the stroked path can be off by several pixels, and at large curvature there may be considerably more subdivision than needed for faithful rendering.
+It should be noted, to avoid the undershoot at low curvature, both the Skia\cite{Skia} and Rive\cite{Rive} renderers use a hybrid of the Wang and polar stroking error metrics.
 

### 2024-04-10 paper.tex

@@
 
-The distance between a circular arc segment of length $s$, with angle between arc and chord of $\theta$ (see Figure~\ref{fig:arc_chord}), is exactly $(1-\cos \theta)s/(2\theta)$.
-The curvature is $\kappa = \frac{2\theta}{s}$ (equivalently, $\theta = \frac{\kappa s}{2s}$), and this remains constant even as the arc is subdivided.
-Rewriting, $d = (1-\cos\frac{\kappa}{2s})\frac{s}{\kappa}$.
+The distance between a circular arc segment of length $s$ and its chord, with angle between arc and chord of $\theta$ (see Figure~\ref{fig:arc_chord}), is exactly $(1-\cos \theta)\frac{s}{2\theta}$.
+The curvature is $\kappa = \frac{2\theta}{s}$ (equivalently, $\theta = \frac{\kappa s}{2}$), and this remains constant even as the arc is subdivided.
+Rewriting, $d = (1-\cos\frac{\kappa s}{2})\frac{1}{\kappa}$.
 From this, we can derive a precise, invertible error metric.
-Subdividing the arc in to $n$ segments, the distance error for each segment is $\frac{1}{\kappa}(1-\cos \frac{s\kappa}{2n})$.
+Subdividing the arc in to $n$ segments, the distance error for each segment is $\frac{1}{\kappa}(1-\cos \frac{\kappa s}{2n})$.
 Solving for n, we get:
@@
 The former are not particularly difficult to calculate, but parallel curves of cubic Béziers are notoriously tricky.
-Analytically, is is a tenth order algebraic curve, which is not particularly feasible to compute directly.
+Analytically, it is a tenth order algebraic curve, which is not particularly feasible to compute directly.
 
@@
 The literature also contains optimal results, namely \cite{Maier2014} and \cite{Nuntawisuttiwong2021}, but at considerable cost; both approaches claim $O(n^2)$ time complexity.
-The through-line for all these results is that they are solving a harder problem, adopting the constraint that the generated arc sequence is $G^1$ continuous.
-While desirable for many applications, for simply rendering a stroked outline, it is not needed.
-Even with this constraint relaxed, the angle discontinuities of an arc approximation are tiny compared to the flattening to lines.
+The through-line for all these results is that they are solving a harder problem: adopting the constraint that the generated arc sequence is $G^1$ continuous.
+While desirable for many applications, this constraint is not needed for rendering a stroke outline.
+Even with this constraint relaxed, the angle discontinuities of an arc approximation are tiny compared to flattening to lines.
 

### 2024-04-12 paper.tex
First half of the GPU implementation section
- Started Section 8. Please look at sections 8 and 8.1.
- I moved the existing input and path encoding drafts into section 8.
  Some changes to this will follow.
- I put down a lot of TODOs in the text that can use your feedback. Some
  of them are for tasks for next week, others are genuine questions.
- Stubs for incomplete Section 8 subsections.

@@
 \documentclass[format=acmsmall]{acmart}
+\usepackage[linesnumbered]{algorithm2e}
+
 \begin{document}
@@
 
+TODO: the introduction should summarize the paper's contributions and the outline of the rest of the paper.
 TODO: discussion of prior art.
@@
 \end{itemize}
-
-\section{Input encoding}
-
-One contribution of this paper is a highly efficient encoding for paths to be rendered.
-Goals include conciseness, low cost to encode on the CPU side, and fully parallel processing by the GPU.
-There are two major aspects to the encoding: a variable size encoding of paths, suitable for both filling and stroking, and an extension to stroking.
-The input is separated into a \emph{tag stream}, with one byte per path element, and another stream for the coordinate data.
-The index into coordinate data is computed by a prefix sum from a size value derived from the tag byte.
-
-\subsection{Path encoding}
-
-The tag byte is a number of bit fields, packed into a byte.
-
-TODO: make this into a diagram.
-
-\begin{itemize}
-    \item 0-1: coordinate count
-    \item 2: subpath end bit
-    \item 3: 16/32 bit coordinates
-    \item 4: path end bit
-    \item 5: transform bit
-    \item 6: line style bit
-\end{itemize}
-
-For most path segments, the coordinate count indicates the type of path segment: 1 for lines, 2 for quadratic Beziers, and 3 for cubic Beziers.
-The subpath bit is set on the last segment of a subpath, and the path bit is additionally set on the last bit of a path.
-The latter is not directly interpreted by the path flattening code, but is forwarded to later rendering stages, for example, the path id can be used to select a color or other paint style.
-
-The size is readily inferred from the tag byte: the number of coordinate pairs is the coordinate count plus one if the subpath end bit is set.
-The nuber of bytes per coordinate pair is 4 for 16 bit coordinates or 8 for 32 bit coordinates.
-
-It is relatively straightforward to convert the standard moveto/lineto/curveto path representation into this encoding.
-Subpath boundaries are moved from the beginning of a subpath (moveto) to the end.
-A separate segment must be generated for closepath if the ending point does not coincide with the start point.
-
-The handling of the subpath bit allows for overlap in coordinates, so that, except at subpath boundaries, the first coordinate pair of each path segment overlaps with the last coordinate pair of the previous one.
-Subpath handling is all computed in the prefix sum; no special handling is needed on the GPU side.
 
@@
 
+\section{GPU Implementation}
+
+We applied the techniques outlined in Section 7 to implement a data-parallel algorithm that can convert stroked 2D Bézier paths into flat geometry suitable for GPU rendering.
+We focused primarily on the global stroke-to-fill conversion problem in order to generate polygonal outlines for rasterization.
+Since the Euler spiral approximation can fit any source cubic Bézier and its parallel offset curves, our implementation can be used to render both filled and stroked primitives within a satisfactory error tolerance.
+
+We aimed to satisfy the following the criteria:
+\begin{enumerate}
+  \item The implementation must be able to handle a large number of inputs at real-time rates. 
+  \item The stroke outlines must meet the weak correctness definition.
+  \item The implementation must support the standard SVG stroke end cap (\emph{butt}, \emph{square}, \emph{round}) and join styles (\emph{bevel}, \emph{miter}, \emph{round}).
+  \item The CPU must do no work beyond the basic preperation of the input path data for GPU consumption.
+\end{enumerate}
+
+Our implementation consists of a single GPU compute shader that satisfies these criteria.
+We coupled this with a simple and efficient data layout for parallel processing.
+Notably, our implementation is fully data-parallel on input path segments and does not require any computationally expensive curve evaluation on the CPU.
+We implemented our shader on general-purpose GPU compute primitives supported by all modern graphics APIs, particularly Metal, Vulkan, D3D12, and WebGPU[?
+citation?].
+As such, the ideas presented in this section are easily portable.
+
+The rest of this section describes the design of our GPU pipeline and input encoding scheme.
+
+% TODO: WGSL, minimal compute requirements
+% TODO: workgroup layout/dimensions, parallelism
+% TODO: - impl details unordered across shapes. This must also be highlighted.
+%       - Watertightness
+%       - error tolerance used.
+% TODO: copy other TODOs
+% TODO: dashing as future work
+
+\subsection{Pipeline Design}
+
+An SVG path consists of a sequence of instructions (or "verbs").
+The \emph{lineto} and \emph{curveto} instructions denote an individual \emph{path segment} consisting of either a straight line or a cubic Bézier.
+A given subsequence of path segments can be bounded by a \emph{moveto} and/or \emph{closepath} instruction which mark the beginning and end of a \emph{subpath}.
+A subpath must form a closed contour if painted as a fill.
+When painted as a stroke, each subpath must begin and end with a cap according to the desired cap style.
+In addition, each path segment in a stroked subpath must be connected to adjacent path segments of the same subpath with a join.
+For any given path, the cap and join styles do not vary across subpaths or individual path segments.
+
+The standard organizational primitive for a compute shader is the \emph{workgroup}.
+A workgroup can be organized as a 1, 2, or 3-dimensional grid of individual threads that can cooperate using shared memory.
+Multiple workgroups can be \emph{dispatched} as part of a larger global grid, also of 1, 2, or 3 dimensions.
+
+Our compute shader is arranged as a 1-dimensional grid, parallelized over input path segments.
+We chose a workgroup size of 256 as it works well over a wide range of GPUs.
+Each thread in the global grid is assigned to process a single path segment.
+We encode the individual path segments contiguously in a GPU storage buffer such that each element has 1:1 correspondance to a global thread ID in the dispatch.
+This layout easily scales to tens of thousands of input path segments, which can be distributed across any number of paths and subpaths.
+
+% TODO: caps/joins figure highlighting mapping of segments, joins, and caps to threads, encoding.
+% TODO: visualize per-segment flattened output for a fill and for a stroke from the perspective of one thread.
+% TODO: finish this \begin{algorithm} snippet
+\begin{algorithm}
+  $tid \gets$ global invocation ID\;
+  $(style, segment) \gets readScene(tid)$\;
+  \eIf{$style$ is $fill$}{
+      flatten source curve
+  }{\If{you're sad}{
+      just keep going
+  }}
+\caption{The control flow of our compute shader}
+\end{algorithm}
+
+Each thread outputs a polyline that fits a subset of the rendered subpath's outer contour, represented by the path segment.
+For a filled primitive, a thread only outputs the flattened approximation of the path segment.
+For a stroke, a thread outputs the polylines approximating multiple curves: a) the two parallel curves on both sides of the source segment, offset by the desired stroke \emph{halfwidth}, and b) the inner and outer joins that connect the path segment to the next adjacent segment.
+We also store metadata in our input encoding that marks the position of a path segment within the containing subpath.
+We use this information to determine whether to output an end cap instead of a join.
+We encode one additional segment designated as the \emph{stroke cap marker} for every subpath.
+The thread assigned to the marker segment only outputs a start cap, which completes the stroke outline.
+
+The joins between adjacent segments need to form a continuous outline when combined with the segments.
+This requires that a thread have access to the tangent vector at the end of its assigned segment \emph{and} the start tangent of the next adjacent segment.
+We require that the input segments within a subpath are stored in order which simplifies the access to the adjacent segment down to a simple read at the next array offset.
+Co-operation across threads is not necessary.
+
+A property of this data-parallel structure is that the output is unordered.
+The cubic-Bézier-to-Euler-spiral conversion is sequential within a single thread, so it is possible to preserve ordering at the scope of an individual path segment, if desired.
+However, ordering cannot be guaranteed across path segments.
+We refer to this output data structure as \emph{line soup}.
+We believe this structure can be adapted to other primitives, such as circular arc segments or triangles.
+We incorporated our compute shader into our own massively-parallel GPU rasterizer that spatially sorts the line soup in subsequent binning and tiling stages.
+As such, maintaining any sort of ordering with respect to the input path data structure was not necessary.
+
+% TODO: Talk about the inner join here. We detect inner vs outer outline based on join tangent cross product. We just draw a straight line for the inner join.
+% TODO: line soup is bump allocated. Mention this here? I'm leaning towards discussing that in the Limitations section below, which will focus on memory allocation and estimation.
+% TODO: while parallel on input, sequential per curve. However iterations are bounded (? needs measurement). Should this discussion happen here or down in the Performance section?
+
+\subsection{Input encoding}
+
+% TODO: not sure how much to emphasize the path tag encoding.
+The input processing monoids benefit the vello scene API (fragments) the TagMonoid could generally be computed on the CPU (with minimal offset tracking) It's not that relevant to the stroke processing.
+The main bit to mention is that it's a struct-of-arrays.
+% TODO: talk about sanitization of input: closepaths are handled ahead of time, stroke cap marker.
+% TODO: discuss style encoding in greater detail
+
+One contribution of this paper is a highly efficient encoding for paths to be rendered.
+Goals include conciseness, low cost to encode on the CPU side, and fully parallel processing by the GPU.
+There are two major aspects to the encoding: a variable size encoding of paths, suitable for both filling and stroking, and an extension to stroking.
+The input is separated into a \emph{tag stream}, with one byte per path element, and another stream for the coordinate data.
+The index into coordinate data is computed by a prefix sum from a size value derived from the tag byte.
+
+\subsection{Path encoding}
+
+The tag byte is a number of bit fields, packed into a byte.
+
+TODO: make this into a diagram.
+
+\begin{itemize}
+    \item 0-1: coordinate count
+    \item 2: subpath end bit
+    \item 3: 16/32 bit coordinates
+    \item 4: path end bit
+    \item 5: transform bit
+    \item 6: line style bit
+\end{itemize}
+
+For most path segments, the coordinate count indicates the type of path segment: 1 for lines, 2 for quadratic Beziers, and 3 for cubic Beziers.
+The subpath bit is set on the last segment of a subpath, and the path bit is additionally set on the last bit of a path.
+The latter is not directly interpreted by the path flattening code, but is forwarded to later rendering stages, for example, the path id can be used to select a color or other paint style.
+
+The size is readily inferred from the tag byte: the number of coordinate pairs is the coordinate count plus one if the subpath end bit is set.
+The nuber of bytes per coordinate pair is 4 for 16 bit coordinates or 8 for 32 bit coordinates.
+
+It is relatively straightforward to convert the standard moveto/lineto/curveto path representation into this encoding.
+Subpath boundaries are moved from the beginning of a subpath (moveto) to the end.
+A separate segment must be generated for closepath if the ending point does not coincide with the start point.
+
+The handling of the subpath bit allows for overlap in coordinates, so that, except at subpath boundaries, the first coordinate pair of each path segment overlaps with the last coordinate pair of the previous one.
+Subpath handling is all computed in the prefix sum; no special handling is needed on the GPU side.
+
+\subsection{Algorithm}
+
+% TODO: explain flatten_euler
+
+\subsection{Limitations}
+
+% TODO: memory usage discussion: the impact of lines and how arcs might help.
+or work graphs, or maybe mesh shaders.
+Limitation: memory must be allocated up-front.
+bump estimation, violates the no pre-curve-evaluation criterium but simpler techniques (like Wang's) exist and we use it but way over-estimates.
+This is a limitation of the GPU execution model.
+Explain this trade-off.
+  % -> TODO: Memory usage graphs for input scene buffers and LineSoup.
+
+\section{Performance}
+
+% TODO: We integrated our algorithm into a fully GPU-driven rendering pipeline, Vello WGSL
+% TODO: 3 alternative implementations: cubic->ES (for load balancing eval), cubic->ES->LineSoup (current), cubic->ES->ArcSoup
+% TODO: GPUs tested, data sets. Test strategy
+% TODO: divergence (reference prim count)
+
+\section{Discussion}
+
+TODO
+
 \section{Future work}
@@
 
-\section{Discussion}
+% TODO: mention dashing as future work?
 

### 2024-04-13 bench/src/main.rs
bench: Plumbing for a single sample

@@ -1,0 +1,148 @@
+// Copyright 2024 the Vello Authors
+// SPDX-License-Identifier: Apache-2.0
+
+use {
+    anyhow::{anyhow, bail, Context, Result},
+    vello::{
+        kurbo::{Affine, Vec2},
+        util::RenderContext,
+    },
+};
+
+const WIDTH: u32 = 1000;
+const HEIGHT: u32 = 1000;
+
+struct Bench {
+    context: RenderContext,
+    dev: usize,
+    renderer: vello::Renderer,
+    render_target: wgpu::Texture,
+}
+
+struct Scenes {
+    tests: scenes::SceneSet,
+}
+
+#[derive(Debug)]
+struct SceneSamples {
+    samples: Vec<Vec<wgpu_profiler::GpuTimerQueryResult>>,
+}
+
+impl Bench {
+    async fn new() -> Result<Self> {
+        let mut context = RenderContext::new()
+            .or_else(|_| bail!("failed to initialize render context"))?;
+        let dev = context
+            .device(None)
+            .await
+            .ok_or_else(|| anyhow!("failed to initialize device"))?;
+        let device = &context.devices[dev].device;
+        let mut renderer = vello::Renderer::new(
+            device,
+            vello::RendererOptions {
+                surface_format: None,
+                use_cpu: false,
+                num_init_threads: std::num::NonZeroUsize::new(1),
+                antialiasing_support: vello::AaSupport::area_only(),
+            },
+        )
+        .or_else(|_| bail!("failed to initialize renderer"))?;
+        renderer
+            .profiler
+            .change_settings(wgpu_profiler::GpuProfilerSettings {
+                enable_timer_queries: true,
+                enable_debug_groups: true,
+                ..Default::default()
+            })
+            .context("failed to enable timer queries")?;
+        let render_target = device.create_texture(&wgpu::TextureDescriptor {
+            label: Some("Target texture"),
+            size: wgpu::Extent3d {
+                width: WIDTH,
+                height: HEIGHT,
+                depth_or_array_layers: 1,
+            },
+            mip_level_count: 1,
+            sample_count: 1,
+            dimension: wgpu::TextureDimension::D2,
+            format: wgpu::TextureFormat::Rgba8Unorm,
+            usage: wgpu::TextureUsages::STORAGE_BINDING,
+            view_formats: &[],
+        });
+        Ok(Bench { context, dev, renderer, render_target })
+    }
+
[…]

### 2024-04-14 bench/src/main.rs
bench: Sample and analyze "flatten" stage

@@ -3,12 +3,14 @@
 
 use {
     anyhow::{anyhow, bail, Context, Result},
+    std::{fmt, time::Duration},
     vello::{
         kurbo::{Affine, Vec2},
         util::RenderContext,
     },
 };
 
+const SAMPLE_COUNT: usize = 300;
 const WIDTH: u32 = 1000;
 const HEIGHT: u32 = 1000;
 
@@ -24,14 +26,23 @@ struct Scenes {
 }
 
 #[derive(Debug)]
-struct SceneSamples {
+struct SceneQueryResults {
     samples: Vec<Vec<wgpu_profiler::GpuTimerQueryResult>>,
+}
+
+#[derive(Debug)]
+struct Stats {
+    flatten_deltas: Vec<f64>,
+    min: f64,
+    max: f64,
+    median: f64,
+    mean: f64,
 }
 
 impl Bench {
     async fn new() -> Result<Self> {
-        let mut context = RenderContext::new()
-            .or_else(|_| bail!("failed to initialize render context"))?;
+        let mut context =
+            RenderContext::new().or_else(|_| bail!("failed to initialize render context"))?;
         let dev = context
             .device(None)
             .await
@@ -69,10 +80,19 @@ impl Bench {
             usage: wgpu::TextureUsages::STORAGE_BINDING,
             view_formats: &[],
         });
-        Ok(Bench { context, dev, renderer, render_target })
+        Ok(Bench {
+            context,
+            dev,
+            renderer,
+            render_target,
+        })
     }
 
-    fn sample(&mut self, scene_to_sample: &mut scenes::ExampleScene, count: usize) -> Result<SceneSamples> {
+    fn sample(
+        &mut self,
+        scene_to_sample: &mut scenes::ExampleScene,
+        count: usize,
+    ) -> Result<SceneQueryResults> {
         // TODO: sample CPU encoding and end-to-end render times too.
         let mut text = scenes::SimpleText::new();
         let mut images = scenes::ImageCache::new();
@@ -95,7 +115,7 @@ impl Bench {
                 let factor = Vec2::new(WIDTH as f64, HEIGHT as f64);
                 let scale_factor = (factor.x / res.x).min(factor.y / res.y);
                 Affine::scale(scale_factor)
-            },
+            }
             None => Affine::IDENTITY,
         };
 
@@ -103,7 +123,9 @@ impl Bench {
         scene.append(&fragment, Some(transform));
 
         let render_params = vello::RenderParams {
[…]

### 2024-04-14 bench/src/main.rs
RTX 4090 measurements
Handle nested queries when in TIMESTAMP_QUERY_INSIDE_PASSES mode.

@@ -178,10 +178,24 @@ impl SceneQueryResults {
         let mut mean = 0.;
         for sample in &self.samples {
             for query in sample {
-                // TODO: this should process other stages too.
-                if query.label != "flatten" {
+                // When TIMESTAMP_QUERY_INSIDE_PASSES is supported:
+                // TODO: this could process stages other than "flatten"
+                let query = if !query.nested_queries.is_empty() {
+                    let mut flatten = None;
+                    for nq in &query.nested_queries {
+                        if nq.label == "flatten" {
+                            flatten = Some(nq);
+                        }
+                    }
+                    flatten
+                } else if query.label == "flatten" {
+                    Some(query)
+                } else {
+                    None
+                };
+                let Some(query) = query else {
                     continue;
-                }
+                };
                 let delta = query.time.end - query.time.start;
                 if delta < min {
                     min = delta;

### 2024-04-14 bench/src/main.rs
bench: support SVGs

@@ -21,10 +21,6 @@ struct Bench {
     render_target: wgpu::Texture,
 }
 
-struct Scenes {
-    tests: scenes::SceneSet,
-}
-
 #[derive(Debug)]
 struct SceneQueryResults {
     samples: Vec<Vec<wgpu_profiler::GpuTimerQueryResult>>,
@@ -90,7 +86,7 @@ impl Bench {
 
     fn sample(
         &mut self,
-        scene_to_sample: &mut scenes::ExampleScene,
+        scene: &mut scenes::ExampleScene,
         count: usize,
     ) -> Result<SceneQueryResults> {
         // TODO: sample CPU encoding and end-to-end render times too.
@@ -106,9 +102,7 @@ impl Bench {
             complexity: 15,
         };
         let mut fragment = vello::Scene::new();
-        scene_to_sample
-            .function
-            .render(&mut fragment, &mut scene_params);
+        scene.function.render(&mut fragment, &mut scene_params);
 
         let transform = match scene_params.resolution {
             Some(res) => {
@@ -159,14 +153,6 @@ impl Bench {
             samples.push(result);
         }
         Ok(SceneQueryResults { samples })
-    }
-}
-
-impl Scenes {
-    fn new() -> Self {
-        Scenes {
-            tests: scenes::test_scenes(),
-        }
     }
 }
 
@@ -265,16 +251,40 @@ impl fmt::Display for Stats {
     }
 }
 
+fn svg_scenes() -> Result<scenes::SceneSet> {
+    let mut svg_paths = vec![];
+    for file in std::fs::read_dir("./svgs")? {
+        let entry = file?;
+        if let Some(extension) = std::path::Path::new(&entry.file_name()).extension() {
+            if extension == "svg" {
+                svg_paths.push(entry.path());
+            }
+        }
+    }
+    scenes::scene_from_files(&svg_paths)
+}
+
+fn benchmark_scenes(bench: &mut Bench, scenes: &mut scenes::SceneSet, suffix: &str) -> Result<()> {
+    for scene in &mut scenes.scenes {
+        let samples = bench.sample(scene, SAMPLE_COUNT)?;
+        let stats = samples.analyze();
+        println!("{}{},{}", scene.config.name, suffix, stats);
+    }
+    Ok(())
+}
+
 #[pollster::main]
 async fn main() -> Result<()> {
     let mut bench = Bench::new().await?;
[…]

### 2024-04-16 paper.tex
Apply sigconf, authordraft template
Address feedback

@@
-\documentclass[format=acmsmall]{acmart}
+\documentclass[sigconf, authordraft]{acmart}
 \usepackage[linesnumbered]{algorithm2e}
-
 \begin{document}
@@
   \item The implementation must support the standard SVG stroke end cap (\emph{butt}, \emph{square}, \emph{round}) and join styles (\emph{bevel}, \emph{miter}, \emph{round}).
-  \item The CPU must do no work beyond the basic preperation of the input path data for GPU consumption.
+  \item The CPU must do no work beyond the basic preparation of the input path data for GPU consumption.
 \end{enumerate}
@@
 We chose a workgroup size of 256 as it works well over a wide range of GPUs.
+However, this choice of workgroup size is not particularly important for our algorithm as it does not require any cross-thread communication.
 Each thread in the global grid is assigned to process a single path segment.
 We encode the individual path segments contiguously in a GPU storage buffer such that each element has 1:1 correspondance to a global thread ID in the dispatch.
-This layout easily scales to tens of thousands of input path segments, which can be distributed across any number of paths and subpaths.
+This layout easily scales to hundreds of thousands of input path segments, which can be distributed across any number of paths and subpaths.
 
-% TODO: caps/joins figure highlighting mapping of segments, joins, and caps to threads, encoding.
+% TODO: caps/joins figure highlighting mapping of segments, joins, and caps to threads, encoding. Make sure to show the straight-line inner join.
 % TODO: visualize per-segment flattened output for a fill and for a stroke from the perspective of one thread.
@@
 For a filled primitive, a thread only outputs the flattened approximation of the path segment.
-For a stroke, a thread outputs the polylines approximating multiple curves: a) the two parallel curves on both sides of the source segment, offset by the desired stroke \emph{halfwidth}, and b) the inner and outer joins that connect the path segment to the next adjacent segment.
+For a stroke, a thread outputs the polylines approximating multiple curves: a) the two parallel curves on both sides of the source segment, offset by the desired stroke \emph{halfwidth}, and b) the joins that connect the path segment to the next adjacent segment.
 We also store metadata in our input encoding that marks the position of a path segment within the containing subpath.
@@
 This requires that a thread have access to the tangent vector at the end of its assigned segment \emph{and} the start tangent of the next adjacent segment.
-We require that the input segments within a subpath are stored in order which simplifies the access to the adjacent segment down to a simple read at the next array offset.
-Co-operation across threads is not necessary.
+We require that the input segments within a subpath are stored in order which simplifies the access to the adjacent segment down to a buffer read at the next array offset.
+Notably, this scheme does not cooperation across threads.
+While we take care to output the \emph{outer} join segment (where the offset curves meet at a reflex angle) we output the inner join as a line segment.
 
@@
 
-% TODO: Talk about the inner join here. We detect inner vs outer outline based on join tangent cross product. We just draw a straight line for the inner join.
 % TODO: line soup is bump allocated. Mention this here? I'm leaning towards discussing that in the Limitations section below, which will focus on memory allocation and estimation.
 % TODO: while parallel on input, sequential per curve. However iterations are bounded (? needs measurement). Should this discussion happen here or down in the Performance section?
+%    raph@: I don't have strong feelings on where the discussion happens. However, we want to make the following points:
+%         * The concept is currently sequential only for cubic->ES, all else is parallelizable
+%         * The current impl is sequential per curve, but that's just an impl choice.
+Personally I'm a fan of discussing this in future work related to work graphs, as I consider load balancing to be a detail
+%         * Parallel lines per ES makes sense but parallel arcs per ES much less so, just because of the prim count. Thus, such detailed discussion may make sense downstream of the prim count results.
 
@@
 % TODO: We integrated our algorithm into a fully GPU-driven rendering pipeline, Vello WGSL
-% TODO: 3 alternative implementations: cubic->ES (for load balancing eval), cubic->ES->LineSoup (current), cubic->ES->ArcSoup
-% TODO: GPUs tested, data sets. Test strategy
-% TODO: divergence (reference prim count)
+% TODO: 3 alternative implementations: cubic->ES->LineSoup (current), cubic->ES->ArcSoup, less importantly: cubic->ES (for load balancing eval).
+% TODO: Take a measurement with ablated bbox atomic
+% TODO: discuss: GPUs tested, data sets. Test strategy
+% TODO: discuss: divergence (reference prim count)
 
@@
 % TODO: mention dashing as future work?
+raph@: I think so.
+When we describe dashing, it would be good to say that the ES representation is already optimal for dashing, as it's parametrized by arc length.
 

### 2024-04-18 paper.tex
First cut at CPU performance measurement
This just compares against Skia, which is not especially informative.

Skia has a bug that causes over-subdivision at fine tolerances. Our code could also be more optimized; particularly low-hanging fruit includes not evaluating the Euler spiral twice for both sides of a stroke.

The SVG parser now takes into account the scale, which is especially important for the waves example.

@@
 
+We measured the timing of our CPU implementation, generating both line and arc primitives, against Skia's stroker, which outputs quadratic Bézier primitives.
+The results are shown in Figure~\ref{fig:timings}.
+The time shown is the total for test files from the \emph{timings} dataset of \cite{Nehab2020}, and the tolerance in all cases was 0.25.
+TODO: measure against other implementations.
+
+\begin{figure}
+    \includegraphics[scale=0.6]{timings}
+    \caption{Timing for stroke expansion}
+    \Description{A stacked bar chart of the timings for stroke expansion}
+    \label{fig:timings}
+\end{figure}
+
+We also compared the number of primitives generated, for varying tolerance values from 1.0 to 0.001.
+The number of arcs generated is significantly less than the number of line primitives.
+The number of arcs and quadratic Bézier primitives is comparable at practical tolerances, but the count of quadratic Béziers becomes more favorable at finer tolerances.
+These vertical axis shows the sum of output segment counts for test files from the \emph{timings} dataset.
+
+\begin{figure}
+    \includegraphics[scale=0.6]{prim_count}
+    \caption{Primitive count for stroke expansion}
+    \Description{A line chart showing the primitive count}
+    \label{fig:prim_count}
+\end{figure}
+
+NOTE: the increase in Skia numbers at tolerances below 0.01 or so is the result of a precision bug in Skia, uncovered during preparation of this chart.
+We expect that to be fixed, and will revise the chart.
+
 \section{Discussion}

### 2024-04-18 flatten/src/evolute.rs

@@ -5,7 +5,10 @@
 
 use kurbo::{Arc, BezPath};
 
-use crate::{arc_segment::ArcSegment, euler::{EulerParams, EulerSeg}};
+use crate::{
+    arc_segment::ArcSegment,
+    euler::{EulerParams, EulerSeg},
+};
 
 /// Flatten the evolute of an Euler spiral segment.
 pub fn flatten_es_evolute(es: &EulerSeg, tol: f64) -> BezPath {
@@ -65,7 +68,6 @@ pub fn lower_es_evolute_arc(es: &EulerSeg, tol: f64) -> BezPath {
     }
     path
 }
-
 
 /// Flatten an Euler spiral segment.
 ///

### 2024-04-18 flatten/src/main.rs

@@ -1,7 +1,7 @@
 // Copyright 2024 the Vello Authors
 // SPDX-License-Identifier: Apache-2.0
 
-use clap::Parser;
+use clap::{Arg, Parser};
 use cubic32::{Cubic, Point};
 use euler32::CubicToEulerIter;
 use flatten::{espc_integral_inv, n_subdiv_analytic};
@@ -17,6 +17,7 @@ mod euler_arc;
 mod evolute;
 mod flatten;
 mod flatten32;
+mod perf_graph;
 #[cfg(feature = "skia-safe")]
 mod skia;
 mod stroke;
@@ -24,11 +25,12 @@ mod svg;
 
 #[derive(Parser)]
 enum Args {
-    Evolute,
+    Arc,
     Cubic,
+    Evolute,
     Espc,
     EstFlattenErr,
-    Arc,
+    PrimCountGraph(perf_graph::PrimCountArgs),
     Stroke,
     Svg(svg::SvgArgs),
 }
@@ -89,7 +91,7 @@ fn main_espc() {
 fn main() {
     let args = Args::parse();
     match args {
-        Args::Evolute => evolute::euler_evolute_main(),
+        Args::Arc => euler_arc::arc_main(),
         Args::Cubic => {
             let c = Cubic::new(
                 Point::new(0.0, 0.0),
@@ -103,7 +105,8 @@ fn main() {
         }
         Args::Espc => main_espc(),
         Args::EstFlattenErr => main_est_flatten_err(),
-        Args::Arc => euler_arc::arc_main(),
+        Args::Evolute => evolute::euler_evolute_main(),
+        Args::PrimCountGraph(args) => perf_graph::perf_graph(args),
         Args::Stroke => stroke::stroke_main(),
         Args::Svg(args) => svg::svg_main(args),
     }

### 2024-04-18 flatten/src/perf_graph.rs

@@ -1,0 +1,136 @@
+// Copyright 2024 the Vello Authors
+// SPDX-License-Identifier: Apache-2.0
+
+//! A module to generate data for the primitive count graph.
+
+use std::time::{Duration, Instant};
+
+use clap::Parser;
+use kurbo::Line;
+
+use crate::{arc_segment::ArcSegment, stroke::LoweredPath, svg::SvgScene};
+
+#[derive(Parser)]
+pub struct PrimCountArgs {
+    dir: String,
+    chart_type: String,
+}
+
+#[derive(Debug)]
+enum PrimType {
+    Line,
+    Arc,
+    // Not really a primitive type, but useful for comparison
+    #[cfg(feature = "skia-safe")]
+    Skia,
+}
+
+#[derive(PartialEq)]
+enum GraphType {
+    Bar,
+    Sum,
+}
+
+pub fn perf_graph(args: PrimCountArgs) {
+    // TODO: set as arg?
+    let graph_type = match &*args.chart_type {
+        "count" => GraphType::Sum,
+        "timing" => GraphType::Bar,
+        _ => panic!("unknown chart type"),
+    };
+    let mut scenes = vec![];
+    let mut names = vec![];
+    for entry in std::fs::read_dir(&args.dir).unwrap() {
+        let entry = entry.unwrap();
+        let path = entry.path();
+        let xml_str = std::fs::read_to_string(&path).unwrap();
+        names.push(path);
+        let scene = SvgScene::load(&xml_str).unwrap();
+        scenes.push(scene);
+    }
+    let tolerances = if args.chart_type == "timing" {
+        &[0.25]
+    } else {
+        &[1.0, 0.5, 0.2, 0.1, 0.05, 0.02, 0.01, 0.005, 0.002, 0.001][..]
+    };
+    let n_iter = if args.chart_type == "timing" { 10 } else { 1 };
+    if args.chart_type == "timing" {
+        print!("prim");
+        for name in &names {
+            let stem = name
+                .to_str()
+                .unwrap()
+                .rsplit('/')
+                .nth(0)
+                .unwrap()
+                .strip_suffix(".svg")
+                .unwrap();
+            print!(" {stem}");
+        }
+        println!();
+    }
+    for prim_type in [
+        PrimType::Line,
+        PrimType::Arc,
+        #[cfg(feature = "skia-safe")]
+        PrimType::Skia,
+    ] {
+        match graph_type {
+            GraphType::Sum => println!("\"{:?}\"", prim_type),
+            _ => (),
+        }
[…]

### 2024-04-18 flatten/src/svg.rs

@@ -16,6 +16,7 @@ use crate::stroke::{stroke_undashed, LoweredPath, Lowering};
 pub struct StyledPath {
     pub path: BezPath,
     pub style: Stroke,
+    pub scale: f64,
     // can add color etc, but it's irrelevant for stroke expansion
 }
 
@@ -30,17 +31,157 @@ impl SvgScene {
         let root = doc.root_element();
         let mut scene = SvgScene { paths: vec![] };
         for node in root.children() {
-            parse_rec(node, &mut scene)?;
+            parse_rec(node, &mut scene, 1.0)?;
         }
         Ok(scene)
     }
 }
 
-fn parse_rec(node: Node, scene: &mut SvgScene) -> Result<(), Box<dyn Error>> {
+#[derive(Debug, PartialEq)]
+enum Token<'a> {
+    Ident(&'a str),
+    Number(f64),
+    Symbol(char),
+}
+
+struct Lexer<'a>(&'a str);
+
+fn scan_num(s: &str) -> usize {
+    let mut i = 0;
+    let b = s.as_bytes()[i];
+    if b == b'-' {
+        i += 1;
+    }
+    while i < s.len() {
+        let b = s.as_bytes()[i];
+        // a bit sloppy, we allow multiple decimal points
+        if (b'0'..=b'9').contains(&b) || b == b'.' {
+            i += 1;
+        } else {
+            break;
+        }
+    }
+    if i < s.len() {
+        let b = s.as_bytes()[i];
+        if b == b'e' || b == b'E' {
+            i += 1;
+            if i < s.len() {
+                let b = s.as_bytes()[i];
+                if b == b'-' || b == b'+' {
+                    i += 1;
+                }
+            }
+            while i < s.len() {
+                let b = s.as_bytes()[i];
+                if !(b'0'..=b'9').contains(&b) {
+                    break;
+                }
+                i += 1;
+            }
+        }
+    }
+    i
+}
+
+fn scan_ident(s: &str) -> usize {
+    let mut i = 1;
+    while i < s.len() {
+        let b = s.as_bytes()[i];
+        if b == b'_'
+            || (b'0'..=b'9').contains(&b)
+            || (b'a'..=b'z').contains(&b)
+            || (b'A'..=b'Z').contains(&b)
+        {
+            i += 1;
+        } else {
+            break;
+        }
+    }
+    i
+}
+
+impl<'a> Iterator for Lexer<'a> {
[…]

### 2024-04-22 bench/src/main.rs
[bench] Command-line args and sampling tweaks

@@ -3,6 +3,7 @@
 
 use {
     anyhow::{anyhow, bail, Context, Result},
+    clap::Parser,
     std::{fmt, time::Duration},
     vello::{
         kurbo::{Affine, Vec2},
@@ -10,9 +11,9 @@ use {
     },
 };
 
-const SAMPLE_COUNT: usize = 300;
-const WIDTH: u32 = 1000;
-const HEIGHT: u32 = 1000;
+const SAMPLE_COUNT: usize = 1000;
+const WIDTH: u32 = 2048;
+const HEIGHT: u32 = 2048;
 
 struct Bench {
     context: RenderContext,
@@ -37,8 +38,7 @@ struct Stats {
 
 impl Bench {
     async fn new() -> Result<Self> {
-        let mut context =
-            RenderContext::new().or_else(|_| bail!("failed to initialize render context"))?;
+        let mut context = RenderContext::new();
         let dev = context
             .device(None)
             .await
@@ -251,9 +251,9 @@ impl fmt::Display for Stats {
     }
 }
 
-fn svg_scenes() -> Result<scenes::SceneSet> {
+fn svg_scenes(path: &str) -> Result<scenes::SceneSet> {
     let mut svg_paths = vec![];
-    for file in std::fs::read_dir("./svgs")? {
+    for file in std::fs::read_dir(path)? {
         let entry = file?;
         if let Some(extension) = std::path::Path::new(&entry.file_name()).extension() {
             if extension == "svg" {
@@ -273,18 +273,26 @@ fn benchmark_scenes(bench: &mut Bench, scenes: &mut scenes::SceneSet, suffix: &s
     Ok(())
 }
 
+#[derive(Parser)]
+enum Args {
+    VelloTestScenes,
+    Svg(SvgArgs),
+}
+
+#[derive(Parser)]
+struct SvgArgs {
+    directory: String,
+}
+
 #[pollster::main]
 async fn main() -> Result<()> {
+    let args = Args::parse();
     let mut bench = Bench::new().await?;
-    let mut test_scenes = scenes::test_scenes();
-    let svg_scenes = svg_scenes();
-
+    let (mut scenes, suffix) = match args {
+        Args::VelloTestScenes => (scenes::test_scenes(), ""),
+        Args::Svg(svg_args) => (svg_scenes(&svg_args.directory)?, ".svg"),
+    };
     println!("samples: {}", SAMPLE_COUNT);
[…]

### 2024-04-22 paper.tex
A bit more writing on cusp rendering
Also adds a bit to the discussion of primitive count.

This commit also checks in the prim_count and timings files (used to generate the plots). The primitive counts are deterministic, but the timings can vary. In any case, I'm leaning toward checking in intermediate artifacts.

@@
 
+\subsection{Cusp handling}
+
+A significant concern with rendering stroked outlines of cubic Béziers is dealing with cusps and near-cusps in the input curve.
+A Bézier curve is expressed in parametric form, and the derivative with respect to the parameter can be zero or nearly so, causing serious numerical problems for many stroke expansion algorithms.
+In the limit, as the Bézier curve describes a semi-cubical parabola, the curvature can become infinite.
+
+Our approach is to leverage the conversion to Euler spiral segments, which have finite curvature.
+The geometric Hermite interpolation depends on accurate tangents at the endpoints, which in turn are derived from derivatives.
+The tangents are not well defined when the derivative is zero, and are not numerically stable when near-zero.
+Thus, the algorithm has one major mechanism to deal with numerical robustness in these cases: when sampling the cubic Bézier, if the derivative is near zero (as determined by testing against an epsilon threshold), then the derivative is sampled at a slightly perturbed parameter value.
+
+In practice, when rendering a cubic Bézier with a cusp, the region near the cusp is rendered with one Euler spiral segment approximately in shape of the top of a question mark, as shown in Figure~\ref{fig:cusp_rendering}.
+Its parallel curve is well defined, and has the correct shape for the outline of the stroke, given of course that the distance is within tolerance (as enforced by the error metric).
+
+\begin{figure}
+    \includegraphics[scale=0.24]{cusp_rendering}
+    \caption{Rendering of a cubic Bézier cusp using Euler spiral segments}
+    \Description{A cubic Bézier with a cusp is approximated by Euler spiral segments with finite curvature, including one with a question mark shape.
+The resulting stroke outlines have the correct shape.}
+    \label{fig:cusp_rendering}
+\end{figure}
+
+As recommended in both \cite{Nehab2020} and \cite{Kilgard2020}, the rendering of a cusp with infinite curvature matches that of a near-cusp.
+The code to detect near-zero derivatives and re-evaluate is a small amount of logic that does not cause divergence issues, as opposed to extensive a separate ``regularization'' filter in \cite{Nehab2020} (Section 3.1 of the paper) or similar special-case handling.
+
 \section{GPU Implementation}
@@
 
+As observed in \cite{Nehab2020}, Skia is the fastest stroke expansion algorithm by a considerable margin.
+The speed is attained with some compromises, in particular an imprecise error estimation that can sometimes generate outlines exceeding the given tolerance.
+We confirmed that this behavior is still present.
+The Nehab paper proposes a more careful error metric, but a significant fraction of the total computation time is dedicated to evaluating it.
+
 \begin{figure}

### 2024-04-22 paper.tex
Input encoding section text

--- 2024-04-16 paper.tex
@@
 \usepackage[linesnumbered]{algorithm2e}
+\usepackage[format=plain, labelfont={bf}, textfont=normal]{caption}
+\usepackage{booktabs,xcolor}
 \begin{document}
@@
 
-\subsection{Pipeline Design}
+\subsection{Pipeline Design} \label{subsec:pipeline-design}
 
@@
   \eIf{$style$ is $fill$}{
-      flatten source curve
-  }{\If{you're sad}{
-      just keep going
-  }}
-\caption{The control flow of our compute shader}
+      output source curve\;
+  }{
+      \tcp{Handle stroked segment}
+      \eIf{$segment$ is \emph{stroke cap marker}}{
+          \eIf{path is open}{output start cap\;}{output nothing\;}
+      }{
+          output both offset curves\;
+          $neighbor \gets readScene(tid + 1)$\;
+          \eIf{$neighbor$ is not \emph{stroke cap marker} OR path is closed}{
+              output join\;
+          }{
+              output end cap\;
+          }
+      }
+  }
+\caption{The control flow the compute shader}
+\label{alg:shader-flow}
 \end{algorithm}
@@
 We require that the input segments within a subpath are stored in order which simplifies the access to the adjacent segment down to a buffer read at the next array offset.
-Notably, this scheme does not cooperation across threads.
-While we take care to output the \emph{outer} join segment (where the offset curves meet at a reflex angle) we output the inner join as a line segment.
+While we take care to output the \emph{outer} join segment (where the offset curves meet at a reflex angle) according to the join style, we output the inner join simply as a line segment.
 
@@
 
-% TODO: not sure how much to emphasize the path tag encoding.
-The input processing monoids benefit the vello scene API (fragments) the TagMonoid could generally be computed on the CPU (with minimal offset tracking) It's not that relevant to the stroke processing.
-The main bit to mention is that it's a struct-of-arrays.
-% TODO: talk about sanitization of input: closepaths are handled ahead of time, stroke cap marker.
-% TODO: discuss style encoding in greater detail
+The input to our pipeline consists of a sequence of paths, each with an associated 2D affine transformation matrix and a \emph{style} data structure containing information about the stroke style or fill rule.
+We encode paths, transforms, and styles as parallel input streams.
+Unlike a conventional structural-of-arrays arrangement, we permit a one-to-many mapping across elements between certain streams: each transform and style entry can apply to one or more path entries with an array index greater than or equal to transform or style index.
 
-One contribution of this paper is a highly efficient encoding for paths to be rendered.
-Goals include conciseness, low cost to encode on the CPU side, and fully parallel processing by the GPU.
-There are two major aspects to the encoding: a variable size encoding of paths, suitable for both filling and stroking, and an extension to stroking.
-The input is separated into a \emph{tag stream}, with one byte per path element, and another stream for the coordinate data.
-The index into coordinate data is computed by a prefix sum from a size value derived from the tag byte.
+Paths get encoded as two parallel streams: a \emph{path tag} stream containing an element for every segment in every subpath, and a \emph{path data} stream of point coordinates.
+Each segment/verb contributes a variable number of points: 1 for \emph{moveto}/\emph{lineto}, 2 for quadratic Béziers, and 3 for cubic Béziers.
+This variable length stream encoding allows for a highly compact representation of SVG-type content in memory.
+A GPU thread assigned to a path segment must be able to reference the corresponding element in each stream in order to obtain the right curve coordinates, transform, and style information.
+This is done by computing a set of stream offsets for every element in the path tag stream.
 
-\subsection{Path encoding}
+% TODO: Figure: streams and tagmonoid relationship, plus sizes of the elements, plus monoid bits
 
-The tag byte is a number of bit fields, packed into a byte.
+We compute the offsets on the GPU in a separate compute shader that runs prior to the flattening stage.
+A path tag consists of 8 bits that store \emph{stream offset increments}, such that the inclusive prefix sums of all increments for every element in the tag stream yield the stream offsets.
+We call this structure the \emph{tag monoid} and have implemented the computation as a general parallel prefix scan.[TODO: citation?]
 
-TODO: make this into a diagram.
+\begin{table}
+    \begin{tabular}{|c|l|}
+    \hline
+        \textbf{Path Tag Bits} & \multicolumn{1}{|c|}{\textbf{Monoid Fields}} \\
+    \hline
+        0-1 & coordinate count \\
+        2 & subpath end bit \\
+        3 & 16/32 bit coordinates \\
+        4 & path bit \\
+        5 & transform bit \\
+        6 & style bit \\
+    \hline
+    \end{tabular}
+    \caption{
+        \emph{Coordinate count} indicates the type of path segment: 1 for lines, 2 for quadratic Béziers, 3 for cubic Béziers.
+The \emph{subpath end bit} is set on the last segment of a subpath and the \emph{path bit} is additionally set on the last tag of a path.
+The number of curve control points is given by \emph{coordinate count}, plus $1$ if the subpath end bit is set.
+The number of bytes per coordinate pair is 4 for 16 bit or 8 for 32 bit coordinates.
+    }
+    \label{table:pathtag}
+\end{table}
 
-\begin{itemize}
-    \item 0-1: coordinate count
-    \item 2: subpath end bit
-    \item 3: 16/32 bit coordinates
-    \item 4: path end bit
-    \item 5: transform bit
-    \item 6: line style bit
-\end{itemize}
+Table~\ref{table:pathtag} shows the individual fields of the 8-bit path tag.
+The offset increments for the transform and style streams can be stored in a single bit that is set to $1$ only if a path tag marks the beginning of a new transform and/or style entry.
+We also encode an additional \emph{path bit} in the final segment of all paths, which allows us to access additional streams (such as a fill color or gradient parameters) using a per-path offset in later rendering stages.
 
-For most path segments, the coordinate count indicates the type of path segment: 1 for lines, 2 for quadratic Beziers, and 3 for cubic Beziers.
-The subpath bit is set on the last segment of a subpath, and the path bit is additionally set on the last bit of a path.
-The latter is not directly interpreted by the path flattening code, but is forwarded to later rendering stages, for example, the path id can be used to select a color or other paint style.
+The handling of the subpath bit allows for overlap in coordinates, so that except at subpath boundaries, the first coordinate pair of each segment overlaps with the last coordinate pair of the previous one.
+During encoding, subpath boundaries are moved from the beginning of a subpath (moveto) to the end.
+An additional line segment is inserted into the tag stream for fills (closepath) if the final point does not coincide with the start point.
+The final segment of every subpath has the \emph{subpath end bit} set to $1$.
+For strokes, the subpath end segment represents the \emph{stroke cap marker} defined in Section \ref{subsec:pipeline-design}.
+This special segment is assigned a coordinate count of $2$, and the corresponding two points in the path data stream encode the tangent vector of the subpath's initial segment, used to correctly draw the start cap or the connecting join in a closed contour.
 
-The size is readily inferred from the tag byte: the number of coordinate pairs is the coordinate count plus one if the subpath end bit is set.
-The nuber of bytes per coordinate pair is 4 for 16 bit coordinates or 8 for 32 bit coordinates.
-
-It is relatively straightforward to convert the standard moveto/lineto/curveto path representation into this encoding.
-Subpath boundaries are moved from the beginning of a subpath (moveto) to the end.
-A separate segment must be generated for closepath if the ending point does not coincide with the start point.
-
-The handling of the subpath bit allows for overlap in coordinates, so that, except at subpath boundaries, the first coordinate pair of each path segment overlaps with the last coordinate pair of the previous one.
-Subpath handling is all computed in the prefix sum; no special handling is needed on the GPU side.
-
-\subsection{Algorithm}
+\subsection{Flattening Algorithm}
 
 % TODO: explain flatten_euler
-
-\subsection{Limitations}
 
@@
   % -> TODO: Memory usage graphs for input scene buffers and LineSoup.
+% TODO: pathological cases: our implementation doesn't handle culling.
 
@@
 \section{Discussion}
+\subsection{Limitations}
 

### 2024-04-23 paper.tex
Fix typo

@@
   }
-\caption{The control flow the compute shader}
+\caption{The control flow of the compute shader}
 \label{alg:shader-flow}
@@
 We encode paths, transforms, and styles as parallel input streams.
-Unlike a conventional structural-of-arrays arrangement, we permit a one-to-many mapping across elements between certain streams: each transform and style entry can apply to one or more path entries with an array index greater than or equal to transform or style index.
+Unlike a conventional structure-of-arrays arrangement, we permit a one-to-many mapping across elements between certain streams: each transform and style entry can apply to one or more path entries with an array index greater than or equal to transform or style index.
 
@@
 The handling of the subpath bit allows for overlap in coordinates, so that except at subpath boundaries, the first coordinate pair of each segment overlaps with the last coordinate pair of the previous one.
-During encoding, subpath boundaries are moved from the beginning of a subpath (moveto) to the end.
-An additional line segment is inserted into the tag stream for fills (closepath) if the final point does not coincide with the start point.
-The final segment of every subpath has the \emph{subpath end bit} set to $1$.
+During the CPU-side encoding, subpath boundaries are moved from the beginning of a subpath (moveto) to the end.
+The encoding process inserts an additional line segment into the tag and coordinate streams to close the subpath if the final point does not coincide with the start point.
+The \emph{subpath end bit} is set to $1$ for the final segment of every subpath.
 For strokes, the subpath end segment represents the \emph{stroke cap marker} defined in Section \ref{subsec:pipeline-design}.

### 2024-04-23 paper.tex
Address review feedback
A few small nits.

--- 2024-04-22 paper.tex
@@
 As recommended in both \cite{Nehab2020} and \cite{Kilgard2020}, the rendering of a cusp with infinite curvature matches that of a near-cusp.
-The code to detect near-zero derivatives and re-evaluate is a small amount of logic that does not cause divergence issues, as opposed to extensive a separate ``regularization'' filter in \cite{Nehab2020} (Section 3.1 of the paper) or similar special-case handling.
+The code to detect near-zero derivatives and re-evaluate is a small amount of non-divergent logic, unlike the additional ``regularization'' pass proposed in Section 3.1 of \cite{Nehab2020} to handle special cases.
 
@@
 
-We measured the timing of our CPU implementation, generating both line and arc primitives, against Skia's stroker, which outputs quadratic Bézier primitives.
+We measured the timing of our CPU implementation, generating both line and arc primitives, against Skia's stroker, which outputs a mix of straight lines and quadratic Bézier primitives.
 The results are shown in Figure~\ref{fig:timings}.
@@
 
-As observed in \cite{Nehab2020}, Skia is the fastest stroke expansion algorithm by a considerable margin.
+Similarly to \cite{Nehab2020}'s observations, Skia is the fastest stroke expansion algorithm we measured.
 The speed is attained with some compromises, in particular an imprecise error estimation that can sometimes generate outlines exceeding the given tolerance.

### 2024-04-23 paper.tex

--- 2024-04-23 paper.tex
@@
 
+\subsection{Cusp handling}
+
+A significant concern with rendering stroked outlines of cubic Béziers is dealing with cusps and near-cusps in the input curve.
+A Bézier curve is expressed in parametric form, and the derivative with respect to the parameter can be zero or nearly so, causing serious numerical problems for many stroke expansion algorithms.
+In the limit, as the Bézier curve describes a semi-cubical parabola, the curvature can become infinite.
+
+Our approach is to leverage the conversion to Euler spiral segments, which have finite curvature.
+The geometric Hermite interpolation depends on accurate tangents at the endpoints, which in turn are derived from derivatives.
+The tangents are not well defined when the derivative is zero, and are not numerically stable when near-zero.
+Thus, the algorithm has one major mechanism to deal with numerical robustness in these cases: when sampling the cubic Bézier, if the derivative is near zero (as determined by testing against an epsilon threshold), then the derivative is sampled at a slightly perturbed parameter value.
+
+In practice, when rendering a cubic Bézier with a cusp, the region near the cusp is rendered with one Euler spiral segment approximately in shape of the top of a question mark, as shown in Figure~\ref{fig:cusp_rendering}.
+Its parallel curve is well defined, and has the correct shape for the outline of the stroke, given of course that the distance is within tolerance (as enforced by the error metric).
+
+\begin{figure}
+    \includegraphics[scale=0.24]{cusp_rendering}
+    \caption{Rendering of a cubic Bézier cusp using Euler spiral segments}
+    \Description{A cubic Bézier with a cusp is approximated by Euler spiral segments with finite curvature, including one with a question mark shape.
+The resulting stroke outlines have the correct shape.}
+    \label{fig:cusp_rendering}
+\end{figure}
+
+As recommended in both \cite{Nehab2020} and \cite{Kilgard2020}, the rendering of a cusp with infinite curvature matches that of a near-cusp.
+The code to detect near-zero derivatives and re-evaluate is a small amount of non-divergent logic, unlike the additional ``regularization'' pass proposed in Section 3.1 of \cite{Nehab2020} to handle special cases.
+
 \section{GPU Implementation}
@@
 
+We measured the timing of our CPU implementation, generating both line and arc primitives, against Skia's stroker, which outputs a mix of straight lines and quadratic Bézier primitives.
+The results are shown in Figure~\ref{fig:timings}.
+The time shown is the total for test files from the \emph{timings} dataset of \cite{Nehab2020}, and the tolerance in all cases was 0.25.
+TODO: measure against other implementations.
+
+Similarly to \cite{Nehab2020}'s observations, Skia is the fastest stroke expansion algorithm we measured.
+The speed is attained with some compromises, in particular an imprecise error estimation that can sometimes generate outlines exceeding the given tolerance.
+We confirmed that this behavior is still present.
+The Nehab paper proposes a more careful error metric, but a significant fraction of the total computation time is dedicated to evaluating it.
+
+\begin{figure}
+    \includegraphics[scale=0.6]{timings}
+    \caption{Timing for stroke expansion}
+    \Description{A stacked bar chart of the timings for stroke expansion}
+    \label{fig:timings}
+\end{figure}
+
+We also compared the number of primitives generated, for varying tolerance values from 1.0 to 0.001.
+The number of arcs generated is significantly less than the number of line primitives.
+The number of arcs and quadratic Bézier primitives is comparable at practical tolerances, but the count of quadratic Béziers becomes more favorable at finer tolerances.
+These vertical axis shows the sum of output segment counts for test files from the \emph{timings} dataset.
+
+\begin{figure}
+    \includegraphics[scale=0.6]{prim_count}
+    \caption{Primitive count for stroke expansion}
+    \Description{A line chart showing the primitive count}
+    \label{fig:prim_count}
+\end{figure}
+
+NOTE: the increase in Skia numbers at tolerances below 0.01 or so is the result of a precision bug in Skia, uncovered during preparation of this chart.
+We expect that to be fixed, and will revise the chart.
+
 \section{Discussion}

### 2024-04-24 paper.tex

@@
 
-% TODO: caps/joins figure highlighting mapping of segments, joins, and caps to threads, encoding. Make sure to show the straight-line inner join.
-% TODO: visualize per-segment flattened output for a fill and for a stroke from the perspective of one thread.
-% TODO: finish this \begin{algorithm} snippet
+\begin{center}
+\begin{figure}
+    \includegraphics[scale=0.18]{stroke_threads}
+    \caption{Visualization of contour outlines and the IDs of GPU threads that flatten them.
+The subpath contains two adjacent cubic Bézier curves rendered as both a stroke and a fill.
+In the stroked version, thread $n$ flattens the offset curves of the first curve (blue) and the join segment with the miter style (dark blue).
+Thread $n + 1$ flattens the offset curves of the second curve (orange) and the end cap with the round style (dark orange).
+Thread $n + 2$ is assigned the \emph{subpath end segment} and flattens the start cap (red).
+When the same path is rendered as a fill, threads $n$ and $n + 1$ each flatten the source curves, while thread $n + 2$ outputs the \emph{close} segment that joins the end of the subpath to its starting point.
+The black arrows represent the winding direction of the output segments.}
+    \Description{TODO}
+    \label{fig:stroke_threads}
+\end{figure}
+\end{center}
+
 \begin{algorithm}
@@
 
-Each thread outputs a polyline that fits a subset of the rendered subpath's outer contour, represented by the path segment.
+Each thread outputs a polyline that fits a subset of the rendered subpath's outer contour, represented by the path segment (see Figure \ref{fig:stroke_threads}).
 For a filled primitive, a thread only outputs the flattened approximation of the path segment.
@@
 % TODO: explain flatten_euler
+% TODO: visualize per-segment flattened output for a fill and for a stroke from the perspective of one thread.
 

### 2024-04-24 bench/src/main.rs
Add option to filter tests by name

@@ -251,12 +251,37 @@ impl fmt::Display for Stats {
     }
 }
 
-fn svg_scenes(path: &str) -> Result<scenes::SceneSet> {
+fn test_scenes(matches: &Option<String>) -> scenes::SceneSet {
+    let filters: Vec<&str> = matches
+        .as_ref()
+        .map(|v| v.split(",").collect())
+        .unwrap_or(vec![]);
+    let scenes = scenes::test_scenes();
+    return scenes::SceneSet {
+        scenes: scenes
+            .scenes
+            .into_iter()
+            .filter(|s| filters.is_empty() || filters.iter().any(|f| s.config.name.contains(f)))
+            .collect(),
+    };
+}
+
+fn svg_scenes(args: &SvgArgs) -> Result<scenes::SceneSet> {
+    let filters: Vec<&str> = args
+        .matches
+        .as_ref()
+        .map(|v| v.split(",").collect())
+        .unwrap_or(vec![]);
     let mut svg_paths = vec![];
-    for file in std::fs::read_dir(path)? {
+    for file in std::fs::read_dir(&args.directory)? {
         let entry = file?;
         if let Some(extension) = std::path::Path::new(&entry.file_name()).extension() {
-            if extension == "svg" {
+            if extension == "svg"
+                && (filters.is_empty()
+                    || filters
+                        .iter()
+                        .any(|f| entry.file_name().into_string().unwrap().contains(f)))
+            {
                 svg_paths.push(entry.path());
             }
         }
@@ -275,13 +300,24 @@ fn benchmark_scenes(bench: &mut Bench, scenes: &mut scenes::SceneSet, suffix: &s
 
 #[derive(Parser)]
 enum Args {
-    VelloTestScenes,
+    VelloTestScenes(VelloTestScenesArgs),
     Svg(SvgArgs),
+}
+
+#[derive(Parser)]
+struct VelloTestScenesArgs {
+    /// Comma separated list of names to filter on
+    #[arg(short, long)]
+    matches: Option<String>,
 }
 
 #[derive(Parser)]
 struct SvgArgs {
     directory: String,
+
+    /// Comma separated list of names to filter on
+    #[arg(short, long)]
+    matches: Option<String>,
 }
 
 #[pollster::main]
@@ -289,8 +325,8 @@ async fn main() -> Result<()> {
     let args = Args::parse();
     let mut bench = Bench::new().await?;
[…]

### 2024-04-24 bench/src/main.rs
[bench]: Support e2e timings

@@ -4,7 +4,10 @@
 use {
     anyhow::{anyhow, bail, Context, Result},
     clap::Parser,
-    std::{fmt, time::Duration},
+    std::{
+        fmt,
+        time::{Duration, Instant},
+    },
     vello::{
         kurbo::{Affine, Vec2},
         util::RenderContext,
@@ -22,14 +25,19 @@ struct Bench {
     render_target: wgpu::Texture,
 }
 
+type GpuTimerQuerySamples = Vec<Vec<wgpu_profiler::GpuTimerQueryResult>>;
+
 #[derive(Debug)]
 struct SceneQueryResults {
-    samples: Vec<Vec<wgpu_profiler::GpuTimerQueryResult>>,
+    prep_time: Duration,
+    e2e_samples: Vec<Duration>,
+    gpu_samples: GpuTimerQuerySamples,
 }
 
 #[derive(Debug)]
 struct Stats {
-    flatten_deltas: Vec<f64>,
+    prep_time: f64,
+    deltas: Vec<f64>,
     min: f64,
     max: f64,
     median: f64,
@@ -101,6 +109,8 @@ impl Bench {
             interactive: false,
             complexity: 15,
         };
+
+        let prep_start_time = Instant::now();
         let mut fragment = vello::Scene::new();
         scene.function.render(&mut fragment, &mut scene_params);
 
@@ -124,7 +134,14 @@ impl Bench {
             height: HEIGHT,
             antialiasing_method: vello::AaConfig::Area,
         };
-        self.sample_scene(&scene, &render_params, count)
+
+        let prep_end_time = Instant::now();
+        let (e2e_samples, gpu_samples) = self.sample_scene(&scene, &render_params, count)?;
+        Ok(SceneQueryResults {
+            prep_time: prep_end_time - prep_start_time,
+            e2e_samples,
+            gpu_samples,
+        })
     }
 
     fn sample_scene(
@@ -132,66 +149,82 @@ impl Bench {
         scene: &vello::Scene,
         params: &vello::RenderParams,
         count: usize,
-    ) -> Result<SceneQueryResults> {
+    ) -> Result<(Vec<Duration>, GpuTimerQuerySamples)> {
         let view = self
             .render_target
             .create_view(&wgpu::TextureViewDescriptor::default());
         let device = &self.context.devices[self.dev].device;
         let queue = &self.context.devices[self.dev].queue;
-        let mut samples = vec![];
+        let mut timer_query_samples = vec![];
[…]

### 2024-04-24 paper.tex
Abstract
WIP - note, haven't generated the PDF yet

@@
     Vector graphics includes both filled and stroked paths as the main primitives.
-While there have been many proposed techniques for rendering filled paths on GPU, stroked paths have proved more elusive.
-This paper presents a technique for performing stroke expansion, namely the generation of an outline representing the stroke of the given input path, an operation which is considered global, but which we nonetheless implement using a fully parallel algorithm.
-We introduce several novel techniques, including an encoding of stroked primitives suitable for parallel processing and an Euler spiral based method for computing both flattened parallel curves and approximations to arc segments.
-Our method produces a near-optimal number of line segments for the filled outline of a stroked cubic Bézier and can be fully evaluated on a GPU with minimal preprocessing.
+While there are many techniques for rendering filled paths on GPU, stroked paths have proved more elusive.
+This paper presents a technique for performing stroke expansion, namely the generation of the outline representing the stroke of the given input path.
+Stroke expansion is a global problem, with challenging constraints on continuity and correctness.
+Nonetheless, we implement it using a fully parallel algorithm suitable for execution in a GPU compute shader, with minimal preprocessing.
+The output of our method can be either line or circular arc segments, both of which are well suited to GPU rendering, and the number of segments is minimal.
+We introduce several novel techniques, including an encoding of vector graphics primitives suitable for parallel processing, and an Euler spiral based method for computing approximations to parallel curves and evolutes.
 \end{abstract}

### 2024-04-24 bench/src/main.rs
[bench] Uncurse GPU timer queries
Encode the pipeline dispatches 3 times per submit to keep the GPU warm.
The timestamp queries are inaccurate otherwise.

@@ -15,8 +15,8 @@ use {
 };
 
 const SAMPLE_COUNT: usize = 1000;
-const WIDTH: u32 = 2048;
-const HEIGHT: u32 = 2048;
+const WIDTH: u32 = 2088;
+const HEIGHT: u32 = 1600;
 
 struct Bench {
     context: RenderContext,
@@ -97,7 +97,6 @@ impl Bench {
         scene: &mut scenes::ExampleScene,
         count: usize,
     ) -> Result<SceneQueryResults> {
-        // TODO: sample CPU encoding and end-to-end render times too.
         let mut text = scenes::SimpleText::new();
         let mut images = scenes::ImageCache::new();
         let mut scene_params = scenes::SceneParams {
@@ -157,14 +156,18 @@ impl Bench {
         let queue = &self.context.devices[self.dev].queue;
         let mut timer_query_samples = vec![];
         let mut end_to_end_samples = vec![];
+
+        // Because GPUs are cursed, we must submit at least 3 renders per frame to get accurate
+        // results.
+        let min_passes = 3;
         for _ in 0..count {
             let start_time = Instant::now();
-            self.renderer
-                .render_to_texture(device, queue, scene, &view, params)
-                .or_else(|e| bail!("failed to render scene {:?}", e))?;
+            for _ in 0..min_passes {
+                self.renderer
+                    .render_to_texture(device, queue, scene, &view, params)
+                    .or_else(|e| bail!("failed to render scene {:?}", e))?;
+            }
             device.poll(wgpu::Maintain::Wait);
-
-            //std::thread::sleep(Duration::from_millis(16));
 
             let end_time = Instant::now();
             let timer_query_result = self
@@ -173,7 +176,7 @@ impl Bench {
                 .process_finished_frame(queue.get_timestamp_period());
             let result =
                 timer_query_result.ok_or_else(|| anyhow!("no timer query was recorded"))?;
-            end_to_end_samples.push(end_time - start_time);
+            end_to_end_samples.push((end_time - start_time) / min_passes);
             timer_query_samples.push(result);
         }
         Ok((end_to_end_samples, timer_query_samples))

### 2024-04-25 bench/src/lib.rs
[bench] Support Android

@@ -1,0 +1,427 @@
+// Copyright 2024 the Vello Authors
+// SPDX-License-Identifier: Apache-2.0
+
+use {
+    anyhow::{anyhow, bail, Context, Result},
+    clap::Parser,
+    std::{
+        fmt,
+        time::{Duration, Instant},
+    },
+    vello::{
+        kurbo::{Affine, Vec2},
+        util::RenderContext,
+    },
+};
+
+const SAMPLE_COUNT: usize = 1000;
+const WIDTH: u32 = 2088;
+const HEIGHT: u32 = 1600;
+
+struct Bench {
+    context: RenderContext,
+    dev: usize,
+    renderer: vello::Renderer,
+    render_target: wgpu::Texture,
+}
+
+type GpuTimerQuerySamples = Vec<Vec<wgpu_profiler::GpuTimerQueryResult>>;
+
+#[derive(Debug)]
+struct SceneQueryResults {
+    prep_time: Duration,
+    e2e_samples: Vec<Duration>,
+    gpu_samples: GpuTimerQuerySamples,
+}
+
+#[derive(Debug)]
+struct Stats {
+    deltas: Vec<f64>,
+    min: f64,
+    max: f64,
+    median: f64,
+    mean: f64,
+}
+
+#[derive(Debug)]
+struct BenchResult {
+    prep_time: f64,
+    end_to_end: Stats,
+    pipeline_stage: Option<Stats>,
+}
+
+impl Bench {
+    async fn new() -> Result<Self> {
+        let mut context = RenderContext::new();
+        let dev = context
+            .device(None)
+            .await
+            .ok_or_else(|| anyhow!("failed to initialize device"))?;
+        let device = &context.devices[dev].device;
+        let mut renderer = vello::Renderer::new(
+            device,
+            vello::RendererOptions {
+                surface_format: None,
+                use_cpu: false,
+                num_init_threads: std::num::NonZeroUsize::new(1),
+                antialiasing_support: vello::AaSupport::area_only(),
+            },
+        )
+        .or_else(|_| bail!("failed to initialize renderer"))?;
+        renderer
+            .profiler
+            .change_settings(wgpu_profiler::GpuProfilerSettings {
+                enable_timer_queries: true,
+                enable_debug_groups: true,
+                ..Default::default()
+            })
+            .context("failed to enable timer queries")?;
+        let render_target = device.create_texture(&wgpu::TextureDescriptor {
+            label: Some("Target texture"),
[…]

### 2024-04-25 bench/src/main.rs

@@ -1,388 +1,10 @@
 // Copyright 2024 the Vello Authors
 // SPDX-License-Identifier: Apache-2.0
 
-use {
-    anyhow::{anyhow, bail, Context, Result},
-    clap::Parser,
-    std::{
-        fmt,
-        time::{Duration, Instant},
-    },
-    vello::{
-        kurbo::{Affine, Vec2},
-        util::RenderContext,
-    },
-};
-
-const SAMPLE_COUNT: usize = 1000;
-const WIDTH: u32 = 2088;
-const HEIGHT: u32 = 1600;
-
-struct Bench {
-    context: RenderContext,
-    dev: usize,
-    renderer: vello::Renderer,
-    render_target: wgpu::Texture,
-}
-
-type GpuTimerQuerySamples = Vec<Vec<wgpu_profiler::GpuTimerQueryResult>>;
-
-#[derive(Debug)]
-struct SceneQueryResults {
-    prep_time: Duration,
-    e2e_samples: Vec<Duration>,
-    gpu_samples: GpuTimerQuerySamples,
-}
-
-#[derive(Debug)]
-struct Stats {
-    prep_time: f64,
-    deltas: Vec<f64>,
-    min: f64,
-    max: f64,
-    median: f64,
-    mean: f64,
-}
-
-impl Bench {
-    async fn new() -> Result<Self> {
-        let mut context = RenderContext::new();
-        let dev = context
-            .device(None)
-            .await
-            .ok_or_else(|| anyhow!("failed to initialize device"))?;
-        let device = &context.devices[dev].device;
-        let mut renderer = vello::Renderer::new(
-            device,
-            vello::RendererOptions {
-                surface_format: None,
-                use_cpu: false,
-                num_init_threads: std::num::NonZeroUsize::new(1),
-                antialiasing_support: vello::AaSupport::area_only(),
-            },
-        )
-        .or_else(|_| bail!("failed to initialize renderer"))?;
-        renderer
-            .profiler
-            .change_settings(wgpu_profiler::GpuProfilerSettings {
-                enable_timer_queries: true,
-                enable_debug_groups: true,
-                ..Default::default()
-            })
-            .context("failed to enable timer queries")?;
-        let render_target = device.create_texture(&wgpu::TextureDescriptor {
-            label: Some("Target texture"),
-            size: wgpu::Extent3d {
-                width: WIDTH,
-                height: HEIGHT,
-                depth_or_array_layers: 1,
[…]

### 2024-04-25 bench/src/lib.rs
[bench] Android (Mali-G78) SVG timings

@@ -424,4 +424,15 @@ fn android_main() {
         };
         pollster::block_on(run(&cli)).expect("failed");
     }
+    println!("vellobench: measuring timings SVGs");
+    {
+        let cli = Cli {
+            stage: Some("flatten".to_owned()),
+            command: Commands::Svg(SvgArgs {
+                directory: "/data/local/tmp/svgs/timings".to_owned(),
+                matches: None,
+            }),
+        };
+        pollster::block_on(run(&cli)).expect("failed");
+    }
 }

### 2024-04-25 paper.tex
Checkpoint writing
Cusp, introduction, prior work, add appendix for Geometric Hermite interpolation.

@@
 
-TODO: brief description of stroke to fill conversion.
+Stroke expansion is the process of generating an outline that, when filled, produces the \emph{stroke} of a given input path.
+Informally, a line of the same width as the stroke is swept along the path, normal to it, and also decorated with joins and caps.
+Though stroke rendering is referenced in many 2D graphics standards, it has lacked a precise formal definition until recent work, primarily \cite{Nehab2020} and \cite{Kilgard2020}.
 
@@
 
-TODO: the introduction should summarize the paper's contributions and the outline of the rest of the paper.
-TODO: discussion of prior art.
+A great number of rendering techniques for strokes have been proposed.
+Some are \emph{local,} in that they break the stroke up into smaller closed pieces, the union of which forms an approximation to the stroke.
+Included in this category is polar stroking \cite{Kilgard2020}.
+Others are based on stroke expansion, generating an explicit outline, which is then filled to produce the stroke rendering.
+Such techniques are further divided into \emph{flattening} approaches, where the outline is approximated by a polyline, and \emph{curve-based} approaches, exemplified by the approach in \cite{Nehab2020}.
+In both case, the outline contains approximations of parallel curves of the input curve segments, and, when targeting strong correctness, evolutes as well.
+For a more detailed survey of existing stroke rendering techniques, the reader is directed to Section 4 of \cite{Nehab2020}.
 
-\begin{itemize}
-    \item Nehab 2020
-    \item Kilgard 2020: Polar stroking (discussion of error metric can move here)
-    \item Yzerman 2020: combined offset + flatten using quadratics
-    \item Blog post of Dragoș?
-    \item Curve-to-curve offsetting?
-Tiller and Hanson (which is good for quadratics, terrible for cubics), Shape control (pretty good but has robustness issues).
-Hain (misses error tolerance in cusp case, as described in Yzerman).
-The "Comparing Offset Curve Approximation Methods" paper is flawed.
-\end{itemize}
+The subproblem of approximating parallel curves has also spawned extensive literature, none of which is entirely satisfying, especially when it comes to algorithms that can be efficiently evaluated on GPU.
+An example of a reasonably good algorithm that computes flattened parallel curves is \cite{Yzerman2020}, as used in the Blend2D rendering library.
+For producing curved outlines, the Tiller and Hanson algorithm\cite{Tiller1984} is commonly implemented and cited, but its performance is quite poor when applied to cubic Béziers.
+The quadratic Bézier version is adequate, and forms the basis of the algorithm in \cite{Nehab2020}.
+A variant is also used in Skia\cite{Skia}; instead of lowering the cubic Bézier to a quadratic approximation and then computing the parallel curve, Skia approximates the parallel curve directly with a quadratic Bézier, then measures the error to determine whether further subdivision is necessary.
+
+This paper presents a comprehensive solution to stroke expansion, well suited to GPU implementation.
+It can produce both flattened polylines or an approximation consisting of circular arcs.
+The best choice depends on the capabilities of the path rendering mechanism following stroke expansion, but as we show, outlines consisting of circular arcs have many fewer segments and are correspondingly faster to produce.
+The algorithm is based on Euler spiral segments as an intermediate representation, with an iterative algorithm based on a straightforward error metric for conversion from cubic Béziers.
+We have also devised a compact binary encoding of paths, suitable for fully parallel computation of stroke outlines, while requiring minimal CPU-side processing.
+Our algorithm has been implemented in GPU compute shaders, integrated in a full rendering engine for vector graphics, and shows a dramatic speedup over CPU methods.
 
@@
 Given tangent angles relative to the chord, finding the Euler spiral segment that minimizes total curvature variation is a form of geometric Hermite interpolation.
-There are a number of published solutions to this problem, all involving nontrivial numerical solving techniques: gradient descent\cite{Kimia2003}, bisection\cite{Walton2009}, or Newton iteration\cite{Connor2014}.
-Our approach is much more direct.
-We observe that the Euler spiral parameters are a smooth function of the two input parameters, and determine a 2D Taylor's series for this function.
-Many of the terms are zero because of symmetry.
-Preserving terms up to 7th order provides excellent accuracy (TODO: quantify more carefully) with modest computational expense.
-
-TODO: A good cite for this is \cite{Reif2021}. Our solution is similar in flavor, but is more precise at modest cost as it contains more polynomial terms.
-
-TODO: the technique used is a refinement of what was sketched in the Python notebook attached to the "cleaner parallel curves with Euler spirals" blog post.
-It's basically a 2D Taylor's series.
-Previously we had the secant method from section 8.2 of my thesis; not sure whether we want to cite that in the submission for anonymity reasons.
-It's functionally pretty similar to the \cite{Connor2014} reference.
+There are a number of published solutions to this problem, involving nontrivial numerical solving techniques: gradient descent\cite{Kimia2003}, bisection\cite{Walton2009}, or Newton iteration\cite{Connor2014}.
+A more direct approach is to approximate the function as a reasonably low-order polynomial in terms of the endpoint angles.
+Our approach is to use a 7th order polynomial, which is more precise than the 3rd order polynomial proposed in \cite{Reif2021}, at only modest increased cost.
+The exact coefficients used are presented in Appendix \ref{appendix:gh}.
 
@@
 
-A significant concern with rendering stroked outlines of cubic Béziers is dealing with cusps and near-cusps in the input curve.
+There are two types of cusps that must be handled in the stroke expansion problem.
+One is when the input cubic Bézier contains a cusp (or near-cusp, as there are numerical robustness issues), and one is when the parallel curve contains cusps.
+This section will describe both in order.
+
 A Bézier curve is expressed in parametric form, and the derivative with respect to the parameter can be zero or nearly so, causing serious numerical problems for many stroke expansion algorithms.
@@
 The code to detect near-zero derivatives and re-evaluate is a small amount of non-divergent logic, unlike the additional ``regularization'' pass proposed in Section 3.1 of \cite{Nehab2020} to handle special cases.
+
+\subsubsection{Cusps in parallel curves}
+
+Even when the input curve is smooth, its parallel curve contains a cusp when the (signed) curvature equals the offset distance (the half-width of the stroke).
+In published techniques, dealing with these cusps is a nontrivial effort, and involves numerical methods that are not GPU friendly.
+In particular, detecting locations in the cubic Bézier where the curvature crosses a given quantity is a medium-degree polynomial, and in general requires numerical techniques for root finding.
+In the approach of \cite{Nehab2020}, the root finder not only requires a hybrid Newton/bisection method, which requires iteration, but is also recursive in that it uses the roots of a polynomial of one degree lower as a subroutine.
+In general, polynomials up to cubic can be considered GPU-friendly, while finding cusps in cubic Béziers requires a higher degree.
+
+In the simplest case, weak stroke correctness with line segments as the output primitive, no additional work is needed -- the flattening algorithm for Euler spiral parallel curves will naturally generate a point within the given error tolerance of the true cusp.
+However, generation of arc segments and drawing the evolutes needed for strong correctness both require subdivision at the parallel curve cusps.
+
+Fortunately, finding the cusp in the parallel curve of an Euler spiral parallel curve is a simple linear equation, and there is at most one cusp in any such segment.
+Arguably, Euler spirals are a reasonably good solution to problem of determining cusps, as a piecewise linear approximation to curvature, even aside from their usefulness in flattening.
 
@@
 
+\appendix
+
+\section{Geometric Hermite interpolation for Euler spiral}
+\label{appendix:gh}
+
+This appendix gives the detailed algorithm for Geometric Hermite interpolation, determining Euler spiral segment parameters given the tangent angles at the endpoints relative to the chord.
+
+The Euler spiral is represented by $\kappa(s) = k_0 + k_1s$, where $s$ ranges from 0 to 1, ie the arc length is unit normalized.
+This segment is scaled, rotated, and translated into place so that its endpoints match the desired locations.
+Immediately, $k_0$ can be determined as $\theta_0 + \theta_1$.
+Therefore, the remaining parameters to compute are $k_1$ and the ratio of chord length to arc length.
+The latter \emph{could} be computed from $k_0$ and $k_1$ by numerical integratoiun, but it is more efficient to obtain it at the same time as $k_1$.
+Setting $k = \theta_0 + \theta_1$ and $\Delta = \theta_1 - \theta_0$, we have:
+
+\[
+    \setlength\arraycolsep{2pt}
+    \begin{array}{rl}
+    k_1 = & 6\Delta \\
+     - & \Delta^3 / 70 \\
+     - & \Delta^5 / 10780 \\
+     + & \Delta^7 \cdot 2.769178184818219\times 10^{-7} \\
+     - & k^2\Delta / 10 \\
+     + & k^2\Delta^3 / 4200 \\
+     + & k^2\Delta^5 \cdot 1.6959677820260655\times 10^{-5} \\
+     - & k^4\Delta / 1400 \\
+     + & k^4\Delta^3 \cdot 6.84915970574303\times 10^{-5} \\
+     - & k^6\Delta \cdot 7.936475029053326\times 10^{-6}
+    \end{array}
+\]
+
+Similarly, the formula for the chord to arc length ratio:
+
+\[
+    \setlength\arraycolsep{2pt}
+    \begin{array}{rl}
+    c = & 1 \\
+    - & \Delta^2 / 40 \\
+    + & \Delta^4 \cdot 0.00034226190482569864 \\
+    - & \Delta^6 \cdot 1.9349474568904524\times 10^{-6} \\
+    - & k^2 / 24 \\
+    + & k^2\Delta^2 \cdot 0.0024702380951963226 \\
+    - & k^2\Delta^4 \cdot 3.7297408997537985\times 10^{-5} \\
+    + & k^4 / 1920 \\
+    - & k^4\Delta^2 \cdot 4.87350869747975\times 10^{-5} \\
+    - & k^6 \cdot 3.1001936068463107\times 10^{-6}
+    \end{array}
+\]
+
+Note that the latter quantity is equal to $\mbox{sinc}(k/2)$ when $\Delta = 0$.
+
+The coefficients were determined by numerical differentiation, using a Newton-style solver as the source of truth. The resulting formulas are extremely accurate over a wide range of inputs.
+
+% TODO: we can definitely quantify that more precisely.
+
 \end{document}

### 2024-04-26 paper.tex
Add figures for error metric
This adds figures for the cubic to Euler spiral error metric, as well as code and gnuplot scripts to draw them. It's a bit minimal still, but does at least add a visual to support the math and the claim.

@@
 
-Possible image for this: the heatmap from the cleaner parallel curves blog post.
-
 The distance between two cubic Bézier segments can be further broken down into two terms; in most cases, the difference in area accurately predicts the Fréchet distance between the curves.
@@
 For values of $\theta$ between $0$ and $0.5$, and values of $d$ between $0$ and $0.6$, this estimate is always conservative, and it is also tight: over Bézier curves generated randomly with parameters drawn from a uniform distribution this range, the mean ratio of estimated to true error is 1.656.
+
+A visualization of combined error metric is shown in Figure~\ref{fig:cubic_err}, comparing measured and approximate error for a slice of the parameter space, fixing the endpoint angles to 0.1 and 0.2, and varying the distance from both endpoints to the control point.
+
+\begin{figure}
+    \includegraphics[scale=0.75]{cubic_err_measured} \includegraphics[scale=0.75]{cubic_err_approx}
+    \caption{Comparision of measured (top) and approximated (bottom) error for cubic Bézier to Euler spiral conversion. Scale is log (base 10) of the error.}
+    \Description{TODO}
+    \label{fig:cubic_err}
+\end{figure}
 

### 2024-04-26 flatten/src/main.rs

@@ -11,6 +11,7 @@ use crate::euler::EulerParams;
 
 mod arc_segment;
 mod cubic32;
+mod cubic_err_plot;
 mod euler;
 mod euler32;
 mod euler_arc;
@@ -27,6 +28,7 @@ mod svg;
 enum Args {
     Arc,
     Cubic,
+    CubicErr(cubic_err_plot::CubicErrPlot),
     Evolute,
     Espc,
     EstFlattenErr,
@@ -103,6 +105,7 @@ fn main() {
             let path = flatten_offset(iter, 0.0, 0.1);
             println!("{}", path.to_svg());
         }
+        Args::CubicErr(ce) => cubic_err_plot::cubic_err_plot(ce),
         Args::Espc => main_espc(),
         Args::EstFlattenErr => main_est_flatten_err(),
         Args::Evolute => evolute::euler_evolute_main(),

### 2024-04-27 flatten/src/main.rs
Add mmark generation utility
Creates an svg file, which can then be run through the prim-count-graph utility. Also starts a skeletal README.

@@ -1,7 +1,7 @@
 // Copyright 2024 the Vello Authors
 // SPDX-License-Identifier: Apache-2.0
 
-use clap::{Arg, Parser};
+use clap::Parser;
 use cubic32::{Cubic, Point};
 use euler32::CubicToEulerIter;
 use flatten::{espc_integral_inv, n_subdiv_analytic};
@@ -17,6 +17,7 @@ mod euler_arc;
 mod evolute;
 mod flatten;
 mod flatten32;
+mod mmark;
 mod perf_graph;
 #[cfg(feature = "skia-safe")]
 mod skia;
@@ -30,6 +31,7 @@ enum Args {
     Evolute,
     Espc,
     EstFlattenErr,
+    Mmark(mmark::MmarkArgs),
     PrimCountGraph(perf_graph::PrimCountArgs),
     Stroke,
     Svg(svg::SvgArgs),
@@ -106,6 +108,7 @@ fn main() {
         Args::Espc => main_espc(),
         Args::EstFlattenErr => main_est_flatten_err(),
         Args::Evolute => evolute::euler_evolute_main(),
+        Args::Mmark(args) => mmark::mmark_main(args),
         Args::PrimCountGraph(args) => perf_graph::perf_graph(args),
         Args::Stroke => stroke::stroke_main(),
         Args::Svg(args) => svg::svg_main(args),

### 2024-04-27 paper.tex
Results section, GPU performance plots
Still needs a little bit of discussion. Also needs a section on encoding
and bump estimation.

@@
 Some are \emph{local,} in that they break the stroke up into smaller closed pieces, the union of which forms an approximation to the stroke.
-Included in this category is polar stroking \cite{Kilgard2020}.
+Included in this category is polar stroking \cite{Kilgard2020}, which proposes a GPU-friendly subdivision scheme based on an angle step error.
 Others are based on stroke expansion, generating an explicit outline, which is then filled to produce the stroke rendering.
 Such techniques are further divided into \emph{flattening} approaches, where the outline is approximated by a polyline, and \emph{curve-based} approaches, exemplified by the approach in \cite{Nehab2020}.
-In both case, the outline contains approximations of parallel curves of the input curve segments, and, when targeting strong correctness, evolutes as well.
+In both cases, the outline contains approximations of parallel curves of the input curve segments, and, as proposed in \cite{Nehab2020}, their evolutes when targeting strong correctness.
 For a more detailed survey of existing stroke rendering techniques, the reader is directed to Section 4 of \cite{Nehab2020}.
@@
 There are two types of cusps that must be handled in the stroke expansion problem.
-One is when the input cubic Bézier contains a cusp (or near-cusp, as there are numerical robustness issues), and one is when the parallel curve contains cusps.
+One is when the input cubic Bézier contains a cusp (or near-cusp, which is prone to numerical robustness issues), and one is when the parallel curve contains cusps.
 This section will describe both in order.
@@
 Fortunately, finding the cusp in the parallel curve of an Euler spiral parallel curve is a simple linear equation, and there is at most one cusp in any such segment.
-Arguably, Euler spirals are a reasonably good solution to problem of determining cusps, as a piecewise linear approximation to curvature, even aside from their usefulness in flattening.
+Euler spirals are thus a good solution for determining cusps as a piecewise linear approximation to curvature.
 
-\section{GPU Implementation}
+\section{GPU Implementation} \label{section:gpu-impl}
 
@@
 
-Our implementation consists of a single GPU compute shader that satisfies these criteria.
+One of our artifacts is a GPU compute shader that satisfies these criteria.
 We coupled this with a simple and efficient data layout for parallel processing.
@@
 citation?].
-As such, the ideas presented in this section are easily portable.
+As such, the ideas presented in this section are portable to various GPU platforms.
 
@@
 
-% TODO: WGSL, minimal compute requirements
-% TODO: workgroup layout/dimensions, parallelism
-% TODO: - impl details unordered across shapes. This must also be highlighted.
-%       - Watertightness
-%       - error tolerance used.
-% TODO: copy other TODOs
-% TODO: dashing as future work
+% TODO: - Watertightness
 
@@
 Thread $n + 1$ flattens the offset curves of the second curve (orange) and the end cap with the round style (dark orange).
-Thread $n + 2$ is assigned the \emph{subpath end segment} and flattens the start cap (red).
+Thread $n + 2$ handles the \emph{subpath end segment} and flattens the start cap (red).
 When the same path is rendered as a fill, threads $n$ and $n + 1$ each flatten the source curves, while thread $n + 2$ outputs the \emph{close} segment that joins the end of the subpath to its starting point.
@@
 We refer to this output data structure as \emph{line soup}.
-We believe this structure can be adapted to other primitives, such as circular arc segments or triangles.
-We incorporated our compute shader into our own massively-parallel GPU rasterizer that spatially sorts the line soup in subsequent binning and tiling stages.
-As such, maintaining any sort of ordering with respect to the input path data structure was not necessary.
+We integrated the compute shader as a stage in a larger compute shader pipeline (discussed in Section \ref{subsection:gpu-results}) that spatially sorts the line soup in subsequent binning and tiling stages prior to rasterization.
 
-% TODO: line soup is bump allocated. Mention this here? I'm leaning towards discussing that in the Limitations section below, which will focus on memory allocation and estimation.
-% TODO: while parallel on input, sequential per curve. However iterations are bounded (? needs measurement). Should this discussion happen here or down in the Performance section?
-%    raph@: I don't have strong feelings on where the discussion happens. However, we want to make the following points:
-%         * The concept is currently sequential only for cubic->ES, all else is parallelizable
-%         * The current impl is sequential per curve, but that's just an impl choice.
-Personally I'm a fan of discussing this in future work related to work graphs, as I consider load balancing to be a detail
-%         * Parallel lines per ES makes sense but parallel arcs per ES much less so, just because of the prim count. Thus, such detailed discussion may make sense downstream of the prim count results.
+The compute shader writes the line soup output to a GPU storage buffer.
+Each line soup entry contains the viewport coordinates of the two endpoints of a singe line segment.
+We also implemented a variant that outputs circular arc segments instead of lines, in which the output entries contain the arc endpoints in addition to its curvature.
 
@@
 \begin{table}
+    \caption{
+        \emph{Coordinate count} indicates the type of path segment: 1 for lines, 2 for quadratic Béziers, 3 for cubic Béziers.
+The \emph{subpath end bit} is set on the last segment of a subpath and the \emph{path bit} is additionally set on the last tag of a path.
+The number of curve control points is given by \emph{coordinate count}, plus $1$ if the subpath end bit is set.
+The number of bytes per coordinate pair is 4 for 16 bit or 8 for 32 bit coordinates.
+    }
     \begin{tabular}{|c|l|}
@@
     \end{tabular}
-    \caption{
-        \emph{Coordinate count} indicates the type of path segment: 1 for lines, 2 for quadratic Béziers, 3 for cubic Béziers.
-The \emph{subpath end bit} is set on the last segment of a subpath and the \emph{path bit} is additionally set on the last tag of a path.
-The number of curve control points is given by \emph{coordinate count}, plus $1$ if the subpath end bit is set.
-The number of bytes per coordinate pair is 4 for 16 bit or 8 for 32 bit coordinates.
-    }
     \label{table:pathtag}
@@
 
-\section{Performance}
+\section{Results}
 
@@
 % TODO: discuss: divergence (reference prim count)
+
+We implemented two versions of our stroking method: a sequential CPU stroker and the GPU compute shader outlined in Section \ref{section:gpu-impl}.
+Both versions can generate stroked outlines with line and arc primitives.
+We focused our evaluation on the number of output primitives and execution time, using the \emph{timings} dataset provided by Nehab2020\cite{Nehab2020}, in addition to our own curve-intensive scenes to stress the GPU implementation.
+We presents our findings in the remainder of this section.
+
+\subsection{CPU: Primitive Count and Timing}
 
@@
     \includegraphics[scale=0.6]{timings}
-    \caption{Timing for stroke expansion}
+    \caption{CPU timings for stroke expansion, comparing two versions of our stroker (outputting line and arc primitives) against Skia's curve-to-curve stroker. Measured on an Apple M1 Max}
     \Description{A stacked bar chart of the timings for stroke expansion}
@@
 
+\subsection{GPU: Execution Time} \label{subsection:gpu-results}
+
+We evaluated the runtime performance of our compute shader on 4 GPU models and multiple form factors: Google Pixel 6 equipped with the Arm \emph{Mali-G78 MP20} mobile GPU, the Apple \emph{M1 Max} integrated laptop GPU, as well as two discrete desktop GPUs: the mid-2015 NVIDIA {GeForce GTX 980Ti} and the late-2022 NVIDIA {GeForce RTX 4090}.
+We authored our entire GPU pipeline in the \emph{WebGPU Shading Language} (WGSL) and used the \emph{wgpu}\cite{wgpu} framework to natively interoperate with the native graphics APIs (we tested the M1 Max on macOS via the Metal API; the mobile and desktop GPUs were tested via the Vulkan API on Android and Ubuntu systems).
+
+We instrumented our pipeline to obtain fine-grained pipeline stage execution times using the GPU \emph{timestamp query} feature supported by both Metal and Vulkan.
+This allowed us to collect measurements for the curve flattening stage in isolation.
+We ran the compute pipeline on the \emph{timings} SVG files provided by Nehab2020.
+The largest of these SVGs is \emph{waves} (see rendering \emph{(a)} in Figure \ref{fig:renderings}) which contains 42 stroked paths and a total of 13,308 path segments.
+
+In order to stress the GPU with a higher workload, we made two additional test scenes.
+The first is a very large stroked path with 503,304 individually styled dashed segments which we adapted from Skia's open-source test corpus.
+We used two variants of this scene with \emph{butt} and \emph{round} cap styles.
+For the second scene we adapted the \emph{Canvas Paths} test case from MotionMark\cite{todo:mmark}, which renders a large number of randomly generated stroked line, quadratic, and cubic Bézier paths.
+We made two versions with different sizes: mmark-35k and mmark-60k with 30,000 and 60,000 paths respectively.
+Table \ref{table:bump-counts} shows the payload sizes of our largest test scenes.
+
+\begin{table}
+    \caption{Comparison of input and output segment counts in the more intensive GPU test cases. \emph{Segments} are input path segments. \emph{Lines} and \emph{Arcs} are the output primitives.}
+    \begin{tabular}{|l|c|c|c|}
+    \hline
+        \multicolumn{1}{|c|}{\textbf{Test}} & \textbf{Input Segments} & \textbf{Lines} & \textbf{Arcs} \\
+    \hline
+        waves.svg & 13,308 & 406,059 & 111,965 \\
+        mmark-35k & 105,022 & 1,039,899 & 500,351 \\
+        mmark-60k & 179,929 & 1,784,896 & 858,162 \\
+        long dash (butt) & 503,304 & 1,060,476 & 1,060,476 \\
+        long dash (round) & 503,304 & 2,013,216 & 1,060,476 \\
+    \hline
+    \end{tabular}
+    \label{table:bump-counts}
+\end{table}
+
+The test scenes were rendered to a GPU texture with dimensions 2088x1600.
+The contents of each scene were uniformly scaled up (by encoding a transform) to fit the size of the texture, in order to increase the required number of output primitives.
+As with the CPU timings, we used an error tolerance of 0.25 in all cases.
+We recorded the execution times across 3,000 GPU submissions for each scene.
+
+\begin{figure}
+    \includegraphics[scale=0.65]{nehab_gpu_timings}
+    \caption{Mean GPU execution times of the compute shader that generates stroked and filled outlines, for the Nehab2020 \emph{timings} data set, in microseconds.
+The hatched bars show the timings for the version that outputs line primitives while the solid bars show that for arcs.
+Mali-G78 (the lowest end GPU we tested) converts \emph{waves} (the largest scene in the set) to lines in ~3ms.
+The compute shader executes at least an order of magnitude faster on the other GPUs.}
+    \Description{A stacked bar chart of the GPU timings for stroke expansion from Nehab2020 timings SVGs}
+    \label{fig:nehab-gpu-timings}
+\end{figure}
+
+\begin{figure}
+    \includegraphics[scale=0.66]{test_scenes_gpu_timings}
+    \caption{Mean GPU execution times of the compute shader for the stress test scenes.}
+    \Description{A stacked bar chart of the GPU timings for stroke expansion from our test scenes}
+    \label{fig:test-scenes-gpu-timings}
+\end{figure}
+
+Figures \ref{fig:nehab-gpu-timings} and \ref{fig:test-scenes-gpu-timings} show the execution times of our compute shader.
+TODO: brief assessment.
+
+TODO
+% TODO: line soup is bump allocated. talk about allocation and estimation.
+
 \section{Discussion}
 \subsection{Limitations}
-
-TODO
 
@@
 When we describe dashing, it would be good to say that the ES representation is already optimal for dashing, as it's parametrized by arc length.
+% TODO: triangles
+
+% TODO: while parallel on input, sequential per curve. However iterations are bounded (? needs measurement). Should this discussion happen here or down in the Performance section?
+%    raph@: I don't have strong feelings on where the discussion happens. However, we want to make the following points:
+%         * The concept is currently sequential only for cubic->ES, all else is parallelizable
+%         * The current impl is sequential per curve, but that's just an impl choice.
+Personally I'm a fan of discussing this in future work related to work graphs, as I consider load balancing to be a detail
+%         * Parallel lines per ES makes sense but parallel arcs per ES much less so, just because of the prim count. Thus, such detailed discussion may make sense downstream of the prim count results.
+
+%TODO: \ref{fig:renderings}
 
@@
 
-The Euler spiral is represented by $\kappa(s) = k_0 + k_1s$, where $s$ ranges from 0 to 1, ie the arc length is unit normalized.
+The Euler spiral is represented by $\kappa(s) = k_0 + k_1s$, where $s$ ranges from 0 to 1, i.e.
+the arc length is unit normalized.
 This segment is scaled, rotated, and translated into place so that its endpoints match the desired locations.
@@
 Therefore, the remaining parameters to compute are $k_1$ and the ratio of chord length to arc length.
-The latter \emph{could} be computed from $k_0$ and $k_1$ by numerical integratoiun, but it is more efficient to obtain it at the same time as $k_1$.
+The latter \emph{could} be computed from $k_0$ and $k_1$ by numerical integration, but it is more efficient to obtain it at the same time as $k_1$.
 Setting $k = \theta_0 + \theta_1$ and $\Delta = \theta_1 - \theta_0$, we have:

### 2024-04-27 paper.tex
Address review comments
Minor changes suggested in review.

--- 2024-04-26 paper.tex
@@
 Fortunately, finding the cusp in the parallel curve of an Euler spiral parallel curve is a simple linear equation, and there is at most one cusp in any such segment.
-Arguably, Euler spirals are a reasonably good solution to problem of determining cusps, as a piecewise linear approximation to curvature, even aside from their usefulness in flattening.
+Euler spirals are thus a good solution for determining cusps as a piecewise linear approximation to curvature.
 

### 2024-04-27 paper.tex
Test renderings figure

--- 2024-04-27 paper.tex
@@
 This layout easily scales to hundreds of thousands of input path segments, which can be distributed across any number of paths and subpaths.
+We submit the kernel to the GPU as a single dispatch with as many workgroups as needed to handle the entire input.
 
@@
 We also implemented a variant that outputs circular arc segments instead of lines, in which the output entries contain the arc endpoints in addition to its curvature.
+For both primitive types, our renderer also outputs a 32-bit \emph{path ID} used by downstream pipeline stages to identify the path that a segment belongs to, in order to retrieve the correct paint (solid or gradient color) for rendering.
 
@@
 The offset increments for the transform and style streams can be stored in a single bit that is set to $1$ only if a path tag marks the beginning of a new transform and/or style entry.
-We also encode an additional \emph{path bit} in the final segment of all paths, which allows us to access additional streams (such as a fill color or gradient parameters) using a per-path offset in later rendering stages.
+We also encode an additional \emph{path bit} in the final segment of all paths, which allows us to access additional streams (such as fill color or gradient parameters) using a per-path offset in later rendering stages.
 
@@
 
-% TODO: We integrated our algorithm into a fully GPU-driven rendering pipeline, Vello WGSL
-% TODO: 3 alternative implementations: cubic->ES->LineSoup (current), cubic->ES->ArcSoup, less importantly: cubic->ES (for load balancing eval).
-% TODO: Take a measurement with ablated bbox atomic
-% TODO: discuss: GPUs tested, data sets. Test strategy
 % TODO: discuss: divergence (reference prim count)
 
-We implemented two versions of our stroking method: a sequential CPU stroker and the GPU compute shader outlined in Section \ref{section:gpu-impl}.
+We present two versions of our stroking method: a sequential CPU stroker and the GPU compute shader outlined in Section \ref{section:gpu-impl}.
 Both versions can generate stroked outlines with line and arc primitives.
@@
 This allowed us to collect measurements for the curve flattening stage in isolation.
-We ran the compute pipeline on the \emph{timings} SVG files provided by Nehab2020.
+We ran the compute pipeline on the \emph{timings} SVG files provided by Nehab\cite{Nehab2020}.
 The largest of these SVGs is \emph{waves} (see rendering \emph{(a)} in Figure \ref{fig:renderings}) which contains 42 stroked paths and a total of 13,308 path segments.
@@
     \includegraphics[scale=0.65]{nehab_gpu_timings}
-    \caption{Mean GPU execution times of the compute shader that generates stroked and filled outlines, for the Nehab2020 \emph{timings} data set, in microseconds.
+    \caption{Mean GPU execution times of the compute shader that performs stroke expansion and flattening when run on the Nehab\cite{Nehab2020} \emph{timings} data set, in microseconds.
 The hatched bars show the timings for the version that outputs line primitives while the solid bars show that for arcs.
@@
 The compute shader executes at least an order of magnitude faster on the other GPUs.}
-    \Description{A stacked bar chart of the GPU timings for stroke expansion from Nehab2020 timings SVGs}
+    \Description{A stacked bar chart of the GPU timings for stroke expansion from Nehab\cite{Nehab2020} timings SVGs}
     \label{fig:nehab-gpu-timings}
@@
     \includegraphics[scale=0.66]{test_scenes_gpu_timings}
-    \caption{Mean GPU execution times of the compute shader for the stress test scenes.}
+    \caption{Mean GPU execution times of the compute shader that performs stroke expansion and flattening when run on the stress test scenes. The timings are shown in milliseconds.}
     \Description{A stacked bar chart of the GPU timings for stroke expansion from our test scenes}
@@
 
-TODO
 % TODO: line soup is bump allocated. talk about allocation and estimation.
@@
 
-%TODO: \ref{fig:renderings}
+\begin{figure*}
+    \centering
+    \includegraphics[scale=0.2]{test_renderings}
+    \caption{Renderings of some of our test scenes created using our algorithm and associated frame rates.
+All images were rendered on an Apple M1 Max at 2088x1600 resolution using our GPU pipeline.
+\textbf{(a)} \emph{waves.svg} from the Nehab2020 timings data set (340fps); \textbf{(b)} The \emph{mmark-60k} scene with ~180k input segments (72fps); \textbf{(c)} The \emph{long path} scene with ~500k input segments and round caps (40fps with CPU-side dashing - the GPU time to render the scene is 9ms); \textbf{(d)}; A simple test scene showcasing various cap and join styles (950fps); \textbf{(e)} The famous \emph{Ghostscript Tiger} containing both stroked and filled paths (810fps).}
+    \Description{Renderings of the test scenes}
+    \label{fig:renderings}
+\end{figure*}
 

### 2024-04-28 paper.tex
Add evolute figure
Adds a figure and a bit of explanation showing how the addition of evolutes creates strongly correct rendering of high-curvature strokes. Also makes minor tweaks to other figures.

--- 2024-04-27 paper.tex
@@
 
-TODO: figure showing ES, its evolute, and subdivision points for flattening.
+\begin{figure}
+    \includegraphics[scale=0.6]{evolute_fig}
+    \caption{Weakly and strongly correct stroke outlines.
+The top row shows weakly correct stroke outlines.
+In (a) the curvature does not exceed the reciprocal half-width, and the stroke is rendered correctly.
+In (b) the curvature of the bottom outline consistently exceeds the reciprocal half-width, and a section of the outline incorrectly has negative winding number.
+In (c) there is a cusp.
+The bottom row shows corresponding strongly correct renders; (d) is the same as (a), while the other two show additional segments for the evolute and connecting lines (in purple).
+Note that the sections of parallel curve outline with above-threshold curvature are reversed, and that the section enclosed by the evolute has a total winding number of +2.}
+    \Description{Two rows of rendered strokes, the top being weakly correct, only drawing parallel curves, the bottom being strongly correct, with additional evolute segments.}
+    \label{fig:evolutes}
+\end{figure}
+
+The effect of adding evolutes to achieve strong correctness is shown in Figure~\ref{fig:evolutes}.
+The additional evolute segments and connecting lines are output twice, to make the winding numbers consistent and produce a watertight outline.
+All winding numbers are positive, so rendering with the nonzero winding number yields a correct final render.
 

### 2024-04-28 paper.tex
address comments

--- 2024-04-27 paper.tex
@@
 
-\begin{figure*}
-    \centering
-    \includegraphics[scale=0.2]{test_renderings}
-    \caption{Renderings of some of our test scenes created using our algorithm and associated frame rates.
-All images were rendered on an Apple M1 Max at 2088x1600 resolution using our GPU pipeline.
-\textbf{(a)} \emph{waves.svg} from the Nehab2020 timings data set (340fps); \textbf{(b)} The \emph{mmark-60k} scene with ~180k input segments (72fps); \textbf{(c)} The \emph{long path} scene with ~500k input segments and round caps (40fps with CPU-side dashing - the GPU time to render the scene is 9ms); \textbf{(d)}; A simple test scene showcasing various cap and join styles (950fps); \textbf{(e)} The famous \emph{Ghostscript Tiger} containing both stroked and filled paths (810fps).}
-    \Description{Renderings of the test scenes}
-    \label{fig:renderings}
-\end{figure*}
+%TODO: \ref{fig:renderings}
 

### 2024-04-28 gpu_plots.py
address comments

@@ -9,7 +9,12 @@ import sys
 
 matplotlib.style.use('seaborn-v0_8-colorblind')
 
-if len(sys.argv) > 1 and sys.argv[1] == 'tests':
+do_test_scenes = len(sys.argv) > 1 and sys.argv[1] == 'tests' or \
+                 len(sys.argv) > 2 and sys.argv[2] == 'tests'
+save_to_file = len(sys.argv) > 1 and sys.argv[1] == 'save' or \
+               len(sys.argv) > 2 and sys.argv[2] == 'save'
+
+if do_test_scenes:
 	arcs = pd.DataFrame({
 		"long path (arcs)": [8.11, 2.90, 0.559, 0.08245],
 		"mmark-35k (arcs)": [7.50, 0.943, 1.67, 0.13434],
@@ -54,4 +59,10 @@ stacked_data2.plot(kind="bar", stacked=True, width=0.3,
 ax.set_xlim(right=len(stacked_data)-0.5)
 ax.set_ylabel('Time (ms)')
 
-plt.show()
+fig.set_size_inches(5, 4, forward=True)
+
+if save_to_file:
+    filename = "test_scenes_gpu_timings.eps" if do_test_scenes else "nehab_gpu_timings.eps"
+    plt.savefig(filename)
+else:
+    plt.show()

### 2024-04-28 gpu_plots.py
Fix unit on one of the timing plots

@@ -57,12 +57,12 @@ stacked_data.plot(kind="bar", stacked=True, width=0.3,
 stacked_data2.plot(kind="bar", stacked=True, width=0.3, 
                    ax=ax, position=1, hatch='/', rot=0)
 ax.set_xlim(right=len(stacked_data)-0.5)
-ax.set_ylabel('Time (ms)')
-
 fig.set_size_inches(5, 4, forward=True)
 
 if save_to_file:
     filename = "test_scenes_gpu_timings.eps" if do_test_scenes else "nehab_gpu_timings.eps"
+    ylabel = 'Time (ms)' if do_test_scenes else 'Time ($\\mu$s)'
+    ax.set_ylabel(ylabel)
     plt.savefig(filename)
 else:
     plt.show()

### 2024-04-28 gpu_plots.py
fix figure formatting and clipped text

@@ -57,12 +57,20 @@ stacked_data.plot(kind="bar", stacked=True, width=0.3,
 stacked_data2.plot(kind="bar", stacked=True, width=0.3, 
                    ax=ax, position=1, hatch='/', rot=0)
 ax.set_xlim(right=len(stacked_data)-0.5)
-fig.set_size_inches(5, 4, forward=True)
+
+if do_test_scenes:
+    ylabel = 'Time (ms)'
+    filename = "test_scenes_gpu_timings.eps"
+    xinches = 5
+else:
+    ylabel = 'Time ($\\mu$s)'
+    filename = "nehab_gpu_timings.eps"
+    xinches = 5.3
+
+ax.set_ylabel(ylabel)
+fig.set_size_inches(xinches, 4, forward=True)
 
 if save_to_file:
-    filename = "test_scenes_gpu_timings.eps" if do_test_scenes else "nehab_gpu_timings.eps"
-    ylabel = 'Time (ms)' if do_test_scenes else 'Time ($\\mu$s)'
-    ax.set_ylabel(ylabel)
     plt.savefig(filename)
 else:
     plt.show()

### 2024-04-28 paper.tex
address comments
fix figure formatting and clipped text
Tweak color of evolutes
Make the evolute segment more visible by using a brighter color.

--- 2024-04-28 paper.tex
@@
 
-An immediate consequence is that flattening an Euler spiral by choosing subdivision points $s_i = a\cdot i^\frac{2}{3}$ produces a near-optimal flattening, as can visually be seen in Figure\ref{fig:es_flatten}.
+An immediate consequence is that flattening an Euler spiral by choosing subdivision points $s_i = a\cdot i^\frac{2}{3}$ produces a near-optimal flattening, as can visually be seen in Figure~\ref{fig:es_flatten}.
 
@@
 The subdivision density for the parallel curve of an Euler spiral, normalized so that its inflection point is at -1 and the cusp of the parallel curve is at 1, is simply $1 - \sqrt{|1-s^2|}$.
-This function is plotted in Figure\ref{fig:subdiv_density}.
+This function is plotted in Figure~\ref{fig:subdiv_density}.
 

### 2024-04-28 paper.tex

--- 2024-04-28 paper.tex
@@
 \begin{figure}
-    \includegraphics[scale=0.65]{nehab_gpu_timings}
+    \includegraphics[scale=0.69]{nehab_gpu_timings}
     \caption{Mean GPU execution times of the compute shader that performs stroke expansion and flattening when run on the Nehab\cite{Nehab2020} \emph{timings} data set, in microseconds.
@@
 \begin{figure}
-    \includegraphics[scale=0.66]{test_scenes_gpu_timings}
+    \includegraphics[scale=0.73]{test_scenes_gpu_timings}
     \caption{Mean GPU execution times of the compute shader that performs stroke expansion and flattening when run on the stress test scenes. The timings are shown in milliseconds.}
@@
 \end{figure}
+
+\begin{figure*}
+    \centering
+    \includegraphics[scale=0.2]{test_renderings}
+    \caption{Renderings of some of our test scenes created using our algorithm and associated frame rates.
+All images were rendered on an Apple M1 Max at 2088x1600 resolution using our GPU pipeline.
+\textbf{(a)} \emph{waves.svg} from the Nehab2020 timings data set (340fps); \textbf{(b)} The \emph{mmark-60k} scene with ~180k input segments (72fps); \textbf{(c)} The \emph{long path} scene with ~500k input segments and round caps (40fps with CPU-side dashing - the GPU time to render the scene is 9ms); \textbf{(d)}; A simple test scene showcasing various cap and join styles (950fps); \textbf{(e)} The famous \emph{Ghostscript Tiger} containing both stroked and filled paths (810fps).}
+    \Description{Renderings of the test scenes}
+    \label{fig:renderings}
+\end{figure*}
 
@@
 %         * Parallel lines per ES makes sense but parallel arcs per ES much less so, just because of the prim count. Thus, such detailed discussion may make sense downstream of the prim count results.
-
-%TODO: \ref{fig:renderings}
 

### 2024-04-28 bench/src/lib.rs
[bench] code to measure input processing

@@ -48,6 +48,7 @@ struct BenchResult {
     prep_time: f64,
     end_to_end: Stats,
     pipeline_stage: Option<Stats>,
+    input_processing: Stats,
 }
 
 impl Bench {
@@ -112,7 +113,7 @@ impl Bench {
             resolution: None,
             base_color: None,
             interactive: false,
-            complexity: 15,
+            complexity: 20,
         };
 
         let prep_start_time = Instant::now();
@@ -191,38 +192,58 @@ impl Bench {
 
 impl SceneQueryResults {
     fn analyze(&self, stage: &Option<String>) -> BenchResult {
-        let stage_deltas = stage.as_ref().map(|label| {
-            let mut deltas = vec![];
-            for sample in &self.gpu_samples {
-                //println!("{sample:?}");
-                for query in sample {
-                    // When TIMESTAMP_QUERY_INSIDE_PASSES is supported:
-                    let query = if !query.nested_queries.is_empty() {
-                        let mut stage = None;
-                        for nq in &query.nested_queries {
+        let mut stage_deltas = vec![];
+        let mut input_deltas = vec![];
+        for sample in &self.gpu_samples {
+            //println!("{sample:?}");
+            let mut input_delta = 0.;
+            for query in sample {
+                // When TIMESTAMP_QUERY_INSIDE_PASSES is supported:
+                let query = if !query.nested_queries.is_empty() {
+                    let mut found_stage = None;
+                    for nq in &query.nested_queries {
+                        if let Some(label) = stage {
                             if nq.label == *label {
-                                stage = Some(nq);
+                                found_stage = Some(nq);
                             }
                         }
-                        stage
-                    } else if query.label == *label {
-                        Some(query)
-                    } else {
-                        None
-                    };
-                    let Some(query) = query else {
-                        continue;
-                    };
-                    deltas.push(query.time.end - query.time.start);
-                }
+                        if nq.label.starts_with("pathtag_reduce") ||
+                           nq.label.starts_with("pathtag_scan") {
+                            input_delta += nq.time.end - nq.time.start;
+                        }
+                    }
+                    found_stage
+                } else {
+                    if query.label.starts_with("pathtag_reduce") ||
+                       query.label.starts_with("pathtag_scan") {
+                        input_delta += query.time.end - query.time.start;
+                    }
+                    let mut found_stage = None;
[…]

### 2024-04-28 paper.tex
GPU encoding perf section

@@
 Like these other approaches, we also use an intermediate curve, but our choice is an Euler spiral.
-In some ways it is similar to quadratic Béziers -- it also has $O(n^4)$ scaling and is best computed using geometric Hermite interpoliation -- but differs in others.
+In some ways it is similar to quadratic Béziers -- it also has $O(n^4)$ scaling and is best computed using geometric Hermite interpolation -- but differs in others.
 It has no difficulty modeling an inflection point.
@@
 
-The geometry of a stroke outline consists of joins, caps, and the two parallel curves on either side of the input path segemnts, offset by the half linewidth.
+The geometry of a stroke outline consists of joins, caps, and the two parallel curves on either side of the input path segments, offset by the half linewidth.
 The former are not particularly difficult to calculate, but parallel curves of cubic Béziers are notoriously tricky.
@@
 \begin{figure}
-    \includegraphics{espc}
+    \includegraphics[scale=0.8]{espc}
     \caption{Integral of subdivision density for Euler spiral parallel curve, and its approximation}
@@
     \includegraphics[scale=0.75]{cubic_err_measured} \includegraphics[scale=0.75]{cubic_err_approx}
-    \caption{Comparision of measured (top) and approximated (bottom) error for cubic Bézier to Euler spiral conversion. Scale is log (base 10) of the error.}
+    \caption{Comparison of measured (top) and approximated (bottom) error for cubic Bézier to Euler spiral conversion. Scale is log (base 10) of the error.}
     \Description{TODO}
@@
 Each thread in the global grid is assigned to process a single path segment.
-We encode the individual path segments contiguously in a GPU storage buffer such that each element has 1:1 correspondance to a global thread ID in the dispatch.
+We encode the individual path segments contiguously in a GPU storage buffer such that each element has 1:1 correspondence to a global thread ID in the dispatch.
 This layout easily scales to hundreds of thousands of input path segments, which can be distributed across any number of paths and subpaths.
@@
         mmark-60k & 179,929 & 1,784,896 & 858,162 \\
-        long dash (butt) & 503,304 & 1,060,476 & 1,060,476 \\
-        long dash (round) & 503,304 & 2,013,216 & 1,060,476 \\
+        long path (butt) & 503,304 & 1,060,476 & 1,060,476 \\
+        long path (round) & 503,304 & 2,013,216 & 1,060,476 \\
     \hline
@@
 
-Figures \ref{fig:nehab-gpu-timings} and \ref{fig:test-scenes-gpu-timings} show the execution times of our compute shader.
-TODO: brief assessment.
+Figures \ref{fig:nehab-gpu-timings} and \ref{fig:test-scenes-gpu-timings} show the execution times for the two data sets.
+The entire Nehab\ref{2020} timings set adds up to less than 1 ms on all GPUs except the mobile unit.
+Our algorithm is ~14x slower on the Mali-G78 compared to the M1 Max but it is still capable of rendering \emph{waves} in 3 ms.
+We observed that performance of our kernel scales well with increasing GPU ability even at very large workloads, as demonstrated by the order-of-magnitude difference in run time between the mobile, laptop, and high-end desktop form factors we tested with.
+We also observed that outputting arcs instead of lines can lead to a 2x decrease in execution time (particularly in scenes with high curve content, like \emph{waves} and \emph{mmark}) as predicted by the lower overall primitive count.
 
-% TODO: line soup is bump allocated. talk about allocation and estimation.
+\subsection{GPU: Encoding} \label{subsection:encoding-results}
+
+Figure~\ref{fig:tag-monoid-gpu-timings} shows the run time of the tag monoid parallel prefix sums with increasing input size.
+The prefix sums scale better at large input sizes compared to our stroke expansion kernel.
+This can be attributed to the lack of expensive computations and divergence the tag monoid kernel.
+
+Our GPU implementation targets graphics APIs that do not support dynamic memory allocations in shader code.
+This makes it necessary to allocate the storage buffer in advance with enough memory to store the output, however, the precise memory requirement is unknown prior to running the shader.
+
+We augmented our encoding logic to compute a conservative estimate of the required output buffer size based on an upper bound on the number of line primitives per input path segment.
+We found that Wang's formula works well in practice, as it is relatively inexpensive to evaluate, though it results in a higher memory estimate than required (see Table~\ref{table:bump-estimates}).
+For parallel curves, Wang's theorem does not guarantee an accurate upper bound on the required subdivisions, which we worked around by inflating the estimate for the source by a scale factor derived from the linewidth.
+We observed that this increases our CPU encoding time by 2-3x on the larger scenes.
+
+\begin{figure}
+    \includegraphics[scale=0.69]{tag_monoid_gpu_timings}
+    \caption{Mean GPU execution time of the tag monoid parallel prefix sums with increasing number of input path segments, measured on Apple M1 Max and NVIDIA RTX 4090.
+The figure also shows the execution time of the stroke expansion and flattening kernel and compares their scaling with input size}
+    \Description{Mean GPU execution time of the tag monoid kernel}
+    \label{fig:tag-monoid-gpu-timings}
+\end{figure}
+
+\begin{table}
+    \caption{Comparison of CPU encoding time with and without size estimation on the largest scenes. All measurements were computed on an Apple M1 Max CPU.}
+    \begin{tabular}{|l|c|c|}
+    \hline
+        \multicolumn{1}{|c|}{\textbf{Test}} & \textbf{Encoding (ms)} & \textbf{w/ Estimation (ms)} \\
+    \hline
+        waves.svg & 0.244 & 0.460 \\
+        mmark-35k & 2.45 & 6.7 \\
+        mmark-60k & 4.17 & 11.8 \\
+        long path & 4.14 & 6.28 \\
+    \hline
+    \end{tabular}
+    \label{table:bump-estimates}
+\end{table}
 

### 2024-04-28 input_processing_plot.py

@@ -1,0 +1,31 @@
+#!/usr/bin/python3
+# Copyright 2024 the Vello Authors
+# SPDX-License-Identifier: Apache-2.0
+
+import matplotlib.pyplot as plt
+import matplotlib
+import numpy as np
+
+xlabels = ['spiral', 'lorenz', 'spirograph', 'blender', 'roads', 'waves', 'mmark-35k', 'mmark-60k', 'longpathdash']
+segments = np.array([561, 2111, 2783, 2971, 7842, 13308, 105022, 179929, 503304])
+mali_ip = np.array([635.59, 545.17, 588.20, 603.49, 593.95, 556.22, 2270, 4100, 5230])
+mali_flatten = np.array([313.15, 342.61, 335.75, 375.86, 736.16, 1250, 7480, 13080, 8120])
+m1_ip = np.array([69.47, 86.11, 120.83, 135.72, 177.21, 105.26, 89.63, 201.19, 307.82])
+m1_flatten = np.array([99.74, 63.77, 87.28, 99.03, 177.21, 136.48, 950.78, 1540, 2910])
+rtx4090_ip = np.array([13.95, 13.78, 13.94, 13.92, 13.96, 14.30, 15.91, 32.42, 34.41])
+rtx4090_flatten = np.array([13.35, 18.38, 13.64, 20.89, 31.72, 30.97, 131.03, 196.31, 279.69])
+
+
+plt.figure(figsize=(5, 4), layout='constrained')
+#plt.plot(segments, mali_ip, label='Mali-G78 (tag monoid)')
+#plt.plot(segments, mali_flatten, label='Mali-G78 (flatten)')
+plt.plot(segments, m1_ip, label='M1 Max (tag monoid)', linewidth=2)
+plt.plot(segments, m1_flatten, label='M1 Max (flatten)', linewidth=2)
+plt.plot(segments, rtx4090_ip, label='RTX 4090 (tag monoid)', linewidth=2)
+plt.plot(segments, rtx4090_flatten, label='RTX 4090 (tag flatten)', linewidth=2)
+plt.xlabel('Input Segments')
+plt.ylabel('Time ($\\mu$s)')
+plt.legend()
+
+plt.savefig('tag_monoid_gpu_timings.eps')
+plt.show()

### 2024-04-28 paper.tex
Rebase & make paper
Redo timings and primitive count measurements
The script for gathering timings had a bug, where the times were cumulative over the scenes. Now fixed. We also include Nehab in the timings.

As part of the work to add Nehab, there's a new utility to convert scenes into RVG format. It's crude (no colors, viewport is hardcoded), but sufficient for measuring timing and primitive count.

The primitive count measurements were redone, striking waves (as it triggers a Skia bug) and adding mmark-70k.

--- 2024-04-28 paper.tex
@@
 
-We measured the timing of our CPU implementation, generating both line and arc primitives, against Skia's stroker, which outputs a mix of straight lines and quadratic Bézier primitives.
+We measured the timing of our CPU implementation, generating both line and arc primitives, against the Nehab and Skia strokers, which both output a mix of straight lines and quadratic Bézier primitives.
 The results are shown in Figure~\ref{fig:timings}.
-The time shown is the total for test files from the \emph{timings} dataset of \cite{Nehab2020}, and the tolerance in all cases was 0.25.
-TODO: measure against other implementations.
+The time shown is the total for test files from the \emph{timings} dataset of \cite{Nehab2020}, and the tolerance was 0.25 when adjustable as a parameter.
 
@@
     \includegraphics[scale=0.6]{timings}
-    \caption{CPU timings for stroke expansion, comparing two versions of our stroker (outputting line and arc primitives) against Skia's curve-to-curve stroker. Measured on an Apple M1 Max}
+    \caption{CPU timings for stroke expansion, comparing two versions of our stroker (outputting line and arc primitives) against Skia's curve-to-curve stroker. Measured on an AMD 3970}
     \Description{A stacked bar chart of the timings for stroke expansion}
@@
 The number of arcs and quadratic Bézier primitives is comparable at practical tolerances, but the count of quadratic Béziers becomes more favorable at finer tolerances.
-These vertical axis shows the sum of output segment counts for test files from the \emph{timings} dataset.
+These vertical axis shows the sum of output segment counts for test files from the \emph{timings} dataset, plus mmark-70k, and omitting the \emph{waves} example\footnote{The \emph{waves} example triggers triggers Skia issue 336617138 at finer tolerance.}.
 

### 2024-04-28 flatten/src/perf_graph.rs

@@ -91,6 +91,7 @@ pub fn perf_graph(args: PrimCountArgs) {
                 for _ in 0..n_iter {
                     let start = Instant::now();
                     if graph_type == GraphType::Bar {
+                        duration = Duration::ZERO;
                         n_segs = 0;
                     }
                     match prim_type {

### 2024-04-29 paper.tex
Fixup mmark names; more content; future work

--- 2024-04-28 paper.tex
@@
 
-\subsection{Flattening Algorithm}
-
-% TODO: explain flatten_euler
-% TODO: visualize per-segment flattened output for a fill and for a stroke from the perspective of one thread.
-
-% TODO: memory usage discussion: the impact of lines and how arcs might help.
-or work graphs, or maybe mesh shaders.
-Limitation: memory must be allocated up-front.
-bump estimation, violates the no pre-curve-evaluation criterium but simpler techniques (like Wang's) exist and we use it but way over-estimates.
-This is a limitation of the GPU execution model.
-Explain this trade-off.
-  % -> TODO: Memory usage graphs for input scene buffers and LineSoup.
-% TODO: pathological cases: our implementation doesn't handle culling.
+% TODO: flatten_euler.
 
@@
 We focused our evaluation on the number of output primitives and execution time, using the \emph{timings} dataset provided by Nehab2020\cite{Nehab2020}, in addition to our own curve-intensive scenes to stress the GPU implementation.
-We presents our findings in the remainder of this section.
+We present our findings in the remainder of this section.
 
@@
 
-We evaluated the runtime performance of our compute shader on 4 GPU models and multiple form factors: Google Pixel 6 equipped with the Arm \emph{Mali-G78 MP20} mobile GPU, the Apple \emph{M1 Max} integrated laptop GPU, as well as two discrete desktop GPUs: the mid-2015 NVIDIA {GeForce GTX 980Ti} and the late-2022 NVIDIA {GeForce RTX 4090}.
-We authored our entire GPU pipeline in the \emph{WebGPU Shading Language} (WGSL) and used the \emph{wgpu}\cite{wgpu} framework to natively interoperate with the native graphics APIs (we tested the M1 Max on macOS via the Metal API; the mobile and desktop GPUs were tested via the Vulkan API on Android and Ubuntu systems).
+We evaluated the runtime performance of our compute shader on 4 GPU models and multiple form factors: Google Pixel 6 equipped with the Arm \emph{Mali-G78 MP20} mobile GPU, the Apple \emph{M1 Max} integrated laptop GPU, and two discrete desktop GPUs: the mid-2015 NVIDIA {GeForce GTX 980Ti} and the late-2022 NVIDIA {GeForce RTX 4090}.
+We authored our entire GPU pipeline in the \emph{WebGPU Shading Language} (WGSL) and used the \emph{wgpu}\cite{wgpu} framework to interoperate with the native graphics APIs (we tested the M1 Max on macOS via the Metal API; the mobile and desktop GPUs were tested via the Vulkan API on Android and Ubuntu systems).
 
 We instrumented our pipeline to obtain fine-grained pipeline stage execution times using the GPU \emph{timestamp query} feature supported by both Metal and Vulkan.
-This allowed us to collect measurements for the curve flattening stage in isolation.
+This allowed us to collect measurements for the curve flattening and input processing stages in isolation.
 We ran the compute pipeline on the \emph{timings} SVG files provided by Nehab\cite{Nehab2020}.
@@
 
-In order to stress the GPU with a higher workload, we made two additional test scenes.
-The first is a very large stroked path with 503,304 individually styled dashed segments which we adapted from Skia's open-source test corpus.
+We authored two additional test scenes in order stress the GPU with a higher workload.
+The first is a very large stroked path with ~500,000 individually styled dashed segments which we adapted from Skia's open-source test corpus.
 We used two variants of this scene with \emph{butt} and \emph{round} cap styles.
-For the second scene we adapted the \emph{Canvas Paths} test case from MotionMark\cite{todo:mmark}, which renders a large number of randomly generated stroked line, quadratic, and cubic Bézier paths.
-We made two versions with different sizes: mmark-35k and mmark-60k with 30,000 and 60,000 paths respectively.
+For the second scene we adapted the \emph{Canvas Paths} test case from MotionMark\cite{MotionMark}, which renders a large number of randomly generated stroked line, quadratic, and cubic Bézier paths.
+We made two versions with different sizes: mmark-70k and mmark-120k with 70,000 and 120,000 path segments, respectively.
 Table \ref{table:bump-counts} shows the payload sizes of our largest test scenes.
@@
         waves.svg & 13,308 & 406,059 & 111,965 \\
-        mmark-35k & 105,022 & 1,039,899 & 500,351 \\
-        mmark-60k & 179,929 & 1,784,896 & 858,162 \\
+        mmark-70k & 70,000 & 1,039,899 & 500,351 \\
+        mmark-120k & 120,000 & 1,784,896 & 858,162 \\
         long path (butt) & 503,304 & 1,060,476 & 1,060,476 \\
@@
 All images were rendered on an Apple M1 Max at 2088x1600 resolution using our GPU pipeline.
-\textbf{(a)} \emph{waves.svg} from the Nehab2020 timings data set (340fps); \textbf{(b)} The \emph{mmark-60k} scene with ~180k input segments (72fps); \textbf{(c)} The \emph{long path} scene with ~500k input segments and round caps (40fps with CPU-side dashing - the GPU time to render the scene is 9ms); \textbf{(d)}; A simple test scene showcasing various cap and join styles (950fps); \textbf{(e)} The famous \emph{Ghostscript Tiger} containing both stroked and filled paths (810fps).}
+\textbf{(a)} \emph{waves.svg} from the Nehab2020 timings data set (340fps); \textbf{(b)} The \emph{mmark-120k} scene with 120,000 input segments (72fps); \textbf{(c)} The \emph{long path} scene with ~500k input segments and round caps (40fps with CPU-side dashing - the GPU time to render the scene is 9ms); \textbf{(d)}; A simple test scene showcasing various cap and join styles (950fps); \textbf{(e)} The famous \emph{Ghostscript Tiger} containing both stroked and filled paths (810fps).}
     \Description{Renderings of the test scenes}
@@
 Figures \ref{fig:nehab-gpu-timings} and \ref{fig:test-scenes-gpu-timings} show the execution times for the two data sets.
-The entire Nehab\ref{2020} timings set adds up to less than 1 ms on all GPUs except the mobile unit.
-Our algorithm is ~14x slower on the Mali-G78 compared to the M1 Max but it is still capable of rendering \emph{waves} in 3 ms.
-We observed that performance of our kernel scales well with increasing GPU ability even at very large workloads, as demonstrated by the order-of-magnitude difference in run time between the mobile, laptop, and high-end desktop form factors we tested with.
-We also observed that outputting arcs instead of lines can lead to a 2x decrease in execution time (particularly in scenes with high curve content, like \emph{waves} and \emph{mmark}) as predicted by the lower overall primitive count.
+The entire Nehab\ref{Nehab2020} timings set adds up to less than 1 ms on all GPUs except the mobile unit.
+Our algorithm is ~14x slower on the Mali-G78 compared to the M1 Max but it is still capable of processing \emph{waves} in 3 ms.
+We observed that the performance of the kernel scales well with increasing GPU ability even at very large workloads, as demonstrated by the order-of-magnitude difference in run time between the mobile, laptop, and high-end desktop form factors we tested.
+We also confirmed that outputting arcs instead of lines can lead to a 2x decrease in execution time (particularly in scenes with high curve content, like \emph{waves} and \emph{mmark}) as predicted by the lower overall primitive count.
 
-\subsection{GPU: Encoding} \label{subsection:encoding-results}
+\subsection{GPU: Input Processing} \label{subsection:encoding-results}
 
@@
 The prefix sums scale better at large input sizes compared to our stroke expansion kernel.
-This can be attributed to the lack of expensive computations and divergence the tag monoid kernel.
+This can be attributed to a lack of expensive computations and high control-flow uniformity in the tag monoid kernel.
 
@@
 For parallel curves, Wang's theorem does not guarantee an accurate upper bound on the required subdivisions, which we worked around by inflating the estimate for the source by a scale factor derived from the linewidth.
-We observed that this increases our CPU encoding time by 2-3x on the larger scenes.
+We observed that enabling estimation increases our CPU pre-processing time by 2-3x on the largest scenes but the impact is overall negligible when the scene size is modest.
 
 \begin{figure}
-    \includegraphics[scale=0.69]{tag_monoid_gpu_timings}
+    \includegraphics[scale=0.66]{tag_monoid_gpu_timings}
     \caption{Mean GPU execution time of the tag monoid parallel prefix sums with increasing number of input path segments, measured on Apple M1 Max and NVIDIA RTX 4090.
@@
 \begin{table}
-    \caption{Comparison of CPU encoding time with and without size estimation on the largest scenes. All measurements were computed on an Apple M1 Max CPU.}
+    \caption{Comparison of CPU encoding time with and without size estimation on the largest scenes.
+All measurements were computed on an Apple M1 Max.
+The additional cost is negligible when the scene size is modest but tangible with increased complexity.}
     \begin{tabular}{|l|c|c|}
     \hline
-        \multicolumn{1}{|c|}{\textbf{Test}} & \textbf{Encoding (ms)} & \textbf{w/ Estimation (ms)} \\
+        \multicolumn{1}{|c|}{\textbf{Test}} & \textbf{Encoding} & \textbf{Encoding + Estimation} \\
     \hline
-        waves.svg & 0.244 & 0.460 \\
-        mmark-35k & 2.45 & 6.7 \\
-        mmark-60k & 4.17 & 11.8 \\
-        long path & 4.14 & 6.28 \\
+        waves.svg  & 6 $\mu$s & 6.13 $\mu$s \\
+        mmark-70k  & 2.32 ms  & 6.59 ms     \\
+        mmark-120k & 4.17 ms  & 11.26 ms    \\
+        long path  & 4.14 ms  & 6.28 ms     \\
     \hline
@@
 \end{table}
-
-\section{Discussion}
-\subsection{Limitations}
 
@@
 
+    \item The pre-allocation requirement for bump-allocated line soup buffer is a limitation due to today's graphics APIs.
+An interesting future exploration is to look at newer GPU API features, such as mesh shaders and work graphs, for GPU-driven scheduling of the compute workload under a bounded memory footprint.
+
     \item Many of the error metrics were empirically determined.
 The mathematical theory behind them should be developed more rigorously, and that process will likely uncover opportunities to fine-tune the technique.
+
+    \item Dashing stroked paths on the GPU was left out of scope for this paper. Given its parameterization by arc length, we believe the Euler spiral representation is highly suitable for dashing.
 \end{itemize}
-
-% TODO: mention dashing as future work?
-raph@: I think so.
-When we describe dashing, it would be good to say that the ES representation is already optimal for dashing, as it's parametrized by arc length.
-% TODO: triangles
 

### 2024-04-29 bench/src/lib.rs

@@ -116,9 +116,18 @@ impl Bench {
             complexity: 20,
         };
 
-        let prep_start_time = Instant::now();
-        let mut fragment = vello::Scene::new();
-        scene.function.render(&mut fragment, &mut scene_params);
+        let (fragment, prep_time) = {
+            let mut fragment = vello::Scene::new();
+            let mut prep_time = Default::default();
+            let n_warmup_loops = 3;
+            for _ in 0..n_warmup_loops {
+                fragment.reset();
+                let prep_start_time = Instant::now();
+                scene.function.render(&mut fragment, &mut scene_params);
+                prep_time = prep_start_time.elapsed();
+            }
+            (fragment, prep_time)
+        };
 
         let transform = match scene_params.resolution {
             Some(res) => {
@@ -141,10 +150,9 @@ impl Bench {
             antialiasing_method: vello::AaConfig::Area,
         };
 
-        let prep_end_time = Instant::now();
         let (e2e_samples, gpu_samples) = self.sample_scene(&scene, &render_params, count)?;
         Ok(SceneQueryResults {
-            prep_time: prep_end_time - prep_start_time,
+            prep_time,
             e2e_samples,
             gpu_samples,
         })

### 2024-04-29 paper.tex
Change nehab_gpu_timings y-axis units to milliseconds

@@
     \includegraphics[scale=0.69]{nehab_gpu_timings}
-    \caption{Mean GPU execution times of the compute shader that performs stroke expansion and flattening when run on the Nehab\cite{Nehab2020} \emph{timings} data set, in microseconds.
+    \caption{Mean GPU execution times of the compute shader that performs stroke expansion and flattening when run on the Nehab\cite{Nehab2020} \emph{timings} data set.
+The timings are shown in milliseconds.
 The hatched bars show the timings for the version that outputs line primitives while the solid bars show that for arcs.

### 2024-04-29 gpu_plots.py

@@ -5,6 +5,7 @@
 import pandas as pd
 import matplotlib.pyplot as plt
 import matplotlib
+import matplotlib.ticker as tkr
 import sys
 
 matplotlib.style.use('seaborn-v0_8-colorblind')
@@ -58,12 +59,19 @@ stacked_data2.plot(kind="bar", stacked=True, width=0.3,
                    ax=ax, position=1, hatch='/', rot=0)
 ax.set_xlim(right=len(stacked_data)-0.5)
 
+def numfmt(x, pos):
+    s = f'{x/1000:,.0f}'
+    return s
+
+yfmt = tkr.FuncFormatter(numfmt)
+
 if do_test_scenes:
     ylabel = 'Time (ms)'
     filename = "test_scenes_gpu_timings.eps"
     xinches = 5
 else:
-    ylabel = 'Time ($\\mu$s)'
+    ylabel = 'Time (ms)'
+    ax.yaxis.set_major_formatter(yfmt)
     filename = "nehab_gpu_timings.eps"
     xinches = 5.3
 

### 2024-04-29 input_processing_plot.py
fix tag_monoid_gpu_timings axis labels

@@ -4,6 +4,7 @@
 
 import matplotlib.pyplot as plt
 import matplotlib
+import matplotlib.ticker as tkr
 import numpy as np
 
 xlabels = ['spiral', 'lorenz', 'spirograph', 'blender', 'roads', 'waves', 'mmark-35k', 'mmark-60k', 'longpathdash']
@@ -16,16 +17,33 @@ rtx4090_ip = np.array([13.95, 13.78, 13.94, 13.92, 13.96, 14.30, 15.91, 32.42, 3
 rtx4090_flatten = np.array([13.35, 18.38, 13.64, 20.89, 31.72, 30.97, 131.03, 196.31, 279.69])
 
 
-plt.figure(figsize=(5, 4), layout='constrained')
+fig = plt.figure(figsize=(5, 4), layout='constrained')
+ax = fig.add_subplot()
 #plt.plot(segments, mali_ip, label='Mali-G78 (tag monoid)')
 #plt.plot(segments, mali_flatten, label='Mali-G78 (flatten)')
-plt.plot(segments, m1_ip, label='M1 Max (tag monoid)', linewidth=2)
-plt.plot(segments, m1_flatten, label='M1 Max (flatten)', linewidth=2)
-plt.plot(segments, rtx4090_ip, label='RTX 4090 (tag monoid)', linewidth=2)
-plt.plot(segments, rtx4090_flatten, label='RTX 4090 (tag flatten)', linewidth=2)
+ax.plot(segments, m1_ip, label='M1 Max (tag monoid)', linewidth=2)
+ax.plot(segments, m1_flatten, label='M1 Max (flatten)', linewidth=2)
+ax.plot(segments, rtx4090_ip, label='RTX 4090 (tag monoid)', linewidth=2)
+ax.plot(segments, rtx4090_flatten, label='RTX 4090 (flatten)', linewidth=2)
 plt.xlabel('Input Segments')
-plt.ylabel('Time ($\\mu$s)')
 plt.legend()
+
+def msfmt(x, pos):
+    s = f'{x/1000:,.2f}'
+    return s
+
+xfmt = tkr.FuncFormatter(lambda x, p: format(int(x), ','))
+yfmt = tkr.FuncFormatter(lambda x, p: f'{x/1000:,.1f}')
+ax.xaxis.set_major_formatter(xfmt)
+
+log_scale = False
+if log_scale:
+    plt.yscale('log')
[…]

### 2024-04-29 paper.tex
use log scale for the tag monoid plot

@@
     \caption{Mean GPU execution time of the tag monoid parallel prefix sums with increasing number of input path segments, measured on Apple M1 Max and NVIDIA RTX 4090.
-The figure also shows the execution time of the stroke expansion and flattening kernel and compares their scaling with input size}
+The figure also shows the execution time of the stroke expansion and flattening kernel and compares their scaling with input size.
+The y-axis is shown in logarithmic scale to highlight the similar scaling trend on both GPUs despite the order-of-magnitude difference in performance.}
     \Description{Mean GPU execution time of the tag monoid kernel}

### 2024-04-29 input_processing_plot.py

@@ -36,10 +36,10 @@ xfmt = tkr.FuncFormatter(lambda x, p: format(int(x), ','))
 yfmt = tkr.FuncFormatter(lambda x, p: f'{x/1000:,.1f}')
 ax.xaxis.set_major_formatter(xfmt)
 
-log_scale = False
+log_scale = True
 if log_scale:
     plt.yscale('log')
-    plt.yticks([1, 100, 500, 1000, 1500, 3000], ['1', '100', '500', '1000', '1500', '3000'])
+    plt.yticks([1, 100, 300, 1000, 3000], ['1', '100', '300', '1000', '3000'])
     plt.ylabel('Time ($\\mu$s)')
 else:
     ax.yaxis.set_major_formatter(yfmt)

### 2024-04-29 paper.tex
fix LaTeX

@@
 We made two versions with different sizes: mmark-70k and mmark-120k with 70,000 and 120,000 path segments, respectively.
-Table \ref{table:bump-counts} shows the payload sizes of our largest test scenes.
+Table~\ref{table:bump-counts} shows the payload sizes of our largest test scenes.
 
@@
 
-Figures \ref{fig:nehab-gpu-timings} and \ref{fig:test-scenes-gpu-timings} show the execution times for the two data sets.
-The entire Nehab\ref{Nehab2020} timings set adds up to less than 1 ms on all GPUs except the mobile unit.
+Figures~\ref{fig:nehab-gpu-timings} and \ref{fig:test-scenes-gpu-timings} show the execution times for the two data sets.
+The entire Nehab\cite{Nehab2020} timings set adds up to less than 1 ms on all GPUs except the mobile unit.
 Our algorithm is ~14x slower on the Mali-G78 compared to the M1 Max but it is still capable of processing \emph{waves} in 3 ms.

### 2024-04-29 paper.tex
Add a actual vs estimated memory table (which I forgot to do earlier) and expand discussion

@@
         mmark-120k & 120,000 & 1,784,896 & 858,162 \\
-        long path (butt) & 503,304 & 1,060,476 & 1,060,476 \\
-        long path (round) & 503,304 & 2,013,216 & 1,060,476 \\
+        long dash (butt) & 503,304 & 1,060,476 & 1,060,476 \\
+        long dash (round) & 503,304 & 2,013,216 & 1,060,476 \\
     \hline
@@
 This can be attributed to a lack of expensive computations and high control-flow uniformity in the tag monoid kernel.
+The CPU encoding times (prior to the GPU prefix sums) for some of the large scenes is shown in Table~\ref{table:bump-estimate-cost}.
+Dashing on the GPU was out of scope, so the dash style applied to the \emph{long dash} scene had to be processed on the CPU and dashes were sequantially encoded as individual strokes.
+Including CPU-side dashing, the time to encode \emph{long dash} on the Apple M1 Max was approximately 25 ms on average, while the average time to encode the path after dashing was 4.14 ms.
 
@@
 We augmented our encoding logic to compute a conservative estimate of the required output buffer size based on an upper bound on the number of line primitives per input path segment.
-We found that Wang's formula works well in practice, as it is relatively inexpensive to evaluate, though it results in a higher memory estimate than required (see Table~\ref{table:bump-estimates}).
+We found that Wang's formula works well in practice, as it is relatively inexpensive to evaluate, though it results in a higher memory estimate than required (see Table~\ref{table:bump-estimate-mem}).
 For parallel curves, Wang's theorem does not guarantee an accurate upper bound on the required subdivisions, which we worked around by inflating the estimate for the source by a scale factor derived from the linewidth.
@@
     \caption{Mean GPU execution time of the tag monoid parallel prefix sums with increasing number of input path segments, measured on Apple M1 Max and NVIDIA RTX 4090.
-The figure also shows the execution time of the stroke expansion and flattening kernel and compares their scaling with input size.
+The figure also shows the execution time of the stroke expansion and flattening kernel.
 The y-axis is shown in logarithmic scale to highlight the similar scaling trend on both GPUs despite the order-of-magnitude difference in performance.}
@@
     \hline
-        waves.svg  & 6 $\mu$s & 6.13 $\mu$s \\
-        mmark-70k  & 2.32 ms  & 6.59 ms     \\
-        mmark-120k & 4.17 ms  & 11.26 ms    \\
-        long path  & 4.14 ms  & 6.28 ms     \\
+        roads.svg  & 398.56 $\mu$s & 618.24 $\mu$s \\
+        waves.svg  & 197.53 $\mu$s & 418.96 $\mu$s \\
+        mmark-70k  & 2.69 ms       & 6.89 ms       \\
+        mmark-120k & 4.28 ms       & 11.59 ms      \\
     \hline
     \end{tabular}
-    \label{table:bump-estimates}
+    \label{table:bump-estimate-cost}
+\end{table}
+
+\begin{table}
+    \caption{The estimated and actual \emph{line soup} buffer sizes for scenes rendered at 2088x1600 resolution.
+Our estimation uses Wang's formula combined with a scaling heuristic based on the linewidth.
+In the worst case, our buffer size estimator overshoots by a factor ~4.}
+    \begin{tabular}{|l|c|c|}
+    \hline
+        \multicolumn{1}{|c|}{\textbf{Test}} & Actual Size & Estimated Size \\
+    \hline
+        roads.svg  & 958.78 KB & 2.45 MB  \\
+        waves.svg  & 9.37 MB   & 10.44 MB \\
+        mmark-70k  & 23.81 MB  & 100.79   \\
+        mmark-120k & 40.86 MB  & 173.56   \\
+    \hline
+    \end{tabular}
+    \label{table:bump-estimate-mem}
 \end{table}
@@
     \item The pre-allocation requirement for bump-allocated line soup buffer is a limitation due to today's graphics APIs.
-An interesting future exploration is to look at newer GPU API features, such as mesh shaders and work graphs, for GPU-driven scheduling of the compute workload under a bounded memory footprint.
+A more accurate and optimized buffer size estimation heuristic is considered future work.
+Another interesting future exploration is to employ newer GPU API features, such as mesh shaders and work graphs, for GPU-driven scheduling of the compute workload under a bounded memory footprint.
 

### 2024-04-29 gpu_plots.py

@@ -17,13 +17,13 @@ save_to_file = len(sys.argv) > 1 and sys.argv[1] == 'save' or \
 
 if do_test_scenes:
 	arcs = pd.DataFrame({
-		"long path (arcs)": [8.11, 2.90, 0.559, 0.08245],
+		"long dash (arcs)": [8.11, 2.90, 0.559, 0.08245],
 		"mmark-35k (arcs)": [7.50, 0.943, 1.67, 0.13434],
 		"mmark-60k (arcs)": [13.17, 1.54, 2.38, 0.19562]
 		}, index=["Mali-G78", "M1 Max", "GTX 980Ti", "RTX 4090"]
 	)
 	lines = pd.DataFrame({
-		"long path (lines)": [13.06, 3.11, 1.35, 0.31733],
+		"long dash (lines)": [13.06, 3.11, 1.35, 0.31733],
 		"mmark-35k (lines)": [12.45, 1.73, 2.95, 0.22960],
 		"mmark-60k (lines)": [22.52, 2.87, 4.34, 0.38593],
 		}, index=["Mali-G78", "M1 Max", "GTX 980Ti", "RTX 4090"]

### 2024-04-29 paper.tex
add missing units
Fix citations
Changed the format to "authoryear". The text now uses \citet for
author[year] type in-sentence citations and \citenum for parenthetical
citations. This conforms to the ACM style rules described on
https://www.acm.org/publications/authors/reference-formatting.

@@
 \usepackage{booktabs,xcolor}
+\usepackage{natbib}
+\setcitestyle{authoryear}
 \begin{document}
@@
 Informally, a line of the same width as the stroke is swept along the path, normal to it, and also decorated with joins and caps.
-Though stroke rendering is referenced in many 2D graphics standards, it has lacked a precise formal definition until recent work, primarily \cite{Nehab2020} and \cite{Kilgard2020}.
+Though stroke rendering is referenced in many 2D graphics standards, it has lacked a precise formal definition until recent work, primarily \citet{Nehab2020} and \citet{Kilgard2020}.
 
-The \cite{Nehab2020} paper gives a comprehensive survey of techniques for stroke expansion, and an algorithm which, for a number of reasons, is only suitable for implementation on CPU (though adapting it to GPU is listed as promising future work).
+The \citet{Nehab2020} paper gives a comprehensive survey of techniques for stroke expansion, and an algorithm which, for a number of reasons, is only suitable for implementation on CPU (though adapting it to GPU is listed as promising future work).
 It classifies techniques into \emph{local,} where each path segment generates closed geometry (which may be triangles or other primitives), and \emph{global,} where the overall result is a closed outline of the stroked path, but the partial result from each segment is in general open.
@@
 The two notions are equivalent for sufficiently well-behaved input, in particular when the curvature measured at endpoints of path segments does not exceed the reciprocal of the half linewidth.
-As described in detail in \cite{Nehab2020}, very few existing implementations actually implement the strong version, so document authors have become accustomed to not depending on behavior at endpoints.
-Standards for graphics formats, such as SVG\cite{Svg2}, also provide an ``out,'' enabling the weaker behavior.
+As described in detail in \citet{Nehab2020}, very few existing implementations actually implement the strong version, so document authors have become accustomed to not depending on behavior at endpoints.
+Standards for graphics formats, such as SVG[\citenum{Svg2}], also provide an ``out,'' enabling the weaker behavior.
 While our current implementation emphasizes weak correctness, we believe it can be extended to the stronger sense, by implementing evolutes and inner join contours.
@@
 Some are \emph{local,} in that they break the stroke up into smaller closed pieces, the union of which forms an approximation to the stroke.
-Included in this category is polar stroking \cite{Kilgard2020}, which proposes a GPU-friendly subdivision scheme based on an angle step error.
+Included in this category is polar stroking[\citenum{Kilgard2020}], which proposes a GPU-friendly subdivision scheme based on an angle step error.
 Others are based on stroke expansion, generating an explicit outline, which is then filled to produce the stroke rendering.
-Such techniques are further divided into \emph{flattening} approaches, where the outline is approximated by a polyline, and \emph{curve-based} approaches, exemplified by the approach in \cite{Nehab2020}.
-In both cases, the outline contains approximations of parallel curves of the input curve segments, and, as proposed in \cite{Nehab2020}, their evolutes when targeting strong correctness.
-For a more detailed survey of existing stroke rendering techniques, the reader is directed to Section 4 of \cite{Nehab2020}.
+Such techniques are further divided into \emph{flattening} approaches, where the outline is approximated by a polyline, and \emph{curve-based} approaches, exemplified by the approach in \citet{Nehab2020}.
+In both cases, the outline contains approximations of parallel curves of the input curve segments, and, as proposed in \citet{Nehab2020}, their evolutes when targeting strong correctness.
+For a more detailed survey of existing stroke rendering techniques, the reader is directed to Section 4 of \citet{Nehab2020}.
 
 The subproblem of approximating parallel curves has also spawned extensive literature, none of which is entirely satisfying, especially when it comes to algorithms that can be efficiently evaluated on GPU.
-An example of a reasonably good algorithm that computes flattened parallel curves is \cite{Yzerman2020}, as used in the Blend2D rendering library.
-For producing curved outlines, the Tiller and Hanson algorithm\cite{Tiller1984} is commonly implemented and cited, but its performance is quite poor when applied to cubic Béziers.
-The quadratic Bézier version is adequate, and forms the basis of the algorithm in \cite{Nehab2020}.
-A variant is also used in Skia\cite{Skia}; instead of lowering the cubic Bézier to a quadratic approximation and then computing the parallel curve, Skia approximates the parallel curve directly with a quadratic Bézier, then measures the error to determine whether further subdivision is necessary.
+An example of a reasonably good algorithm that computes flattened parallel curves is \citet{Yzerman2020}, as used in the Blend2D rendering library.
+For producing curved outlines, the Tiller and Hanson[\citenum{Tiller1984}] algorithm is commonly implemented and cited, but its performance is quite poor when applied to cubic Béziers.
+The quadratic Bézier version is adequate, and forms the basis of the algorithm in \citet{Nehab2020}.
+A variant is also used in Skia[\citenum{Skia}]; instead of lowering the cubic Bézier to a quadratic approximation and then computing the parallel curve, Skia approximates the parallel curve directly with a quadratic Bézier, then measures the error to determine whether further subdivision is necessary.
 
@@
 Otherwise, the curve is subdivided (usually at $t = 0.5$) and each half is recursively approximated.
-A substantial fraction of all curve approximation methods in the literature are of this form, including \cite{Nehab2020}.
+A substantial fraction of all curve approximation methods in the literature are of this form, including \citet{Nehab2020}.
 The main disadvantage is the cost of computing the error metric.
@@
 If the error metric is accurate, then approximation is near-optimal.
-One example of an invertible error metric is angle step, used in polar stroking\cite{Kilgard2020}; the number of subdivisions is the total angle subsumed by the curve divided by the angle step size, and the parameter value for each subdivision is the result of solving for a tangent direction.
-Another widely used invertible error metric is Wang's formula (\cite{Goldman2003}, Section 5.6.3), which gives a bound on the flattening error based on the second derivative of the curve.
+One example of an invertible error metric is angle step, used in polar stroking (\citet{Kilgard2020}); the number of subdivisions is the total angle subsumed by the curve divided by the angle step size, and the parameter value for each subdivision is the result of solving for a tangent direction.
+Another widely used invertible error metric is Wang's formula (\citet{Goldman2003}, Section 5.6.3), which gives a bound on the flattening error based on the second derivative of the curve.
 This metric is conservative but works well in practice; among other applications, it is used in Skia for path flattening.
@@
 It is common to approximate cubic Béziers to some intermediate curve format more conducive to offsetting and flattening.
-A number of published solutions (\cite{Yzerman2020}, \cite{Nehab2020}) use quadratic Béziers, as it is well suited for computation of parallel curves.
+A number of published solutions (\citet{Yzerman2020}, \citet{Nehab2020}) use quadratic Béziers, as it is well suited for computation of parallel curves.
 Even so, this curve has some disadvantages.
@@
 
-The parallel curve of the Euler spiral (also known as ``clothoid'') was characterized by Wieleitner well over a hundred years ago\cite{Wieleitner1907}, and has a straightforward closed-form Cesàro representation, curvature as a function of arc length.
+The parallel curve of the Euler spiral (also known as ``clothoid'') was characterized by \citet{Wieleitner1907} well over a hundred years ago[\citenum{Wieleitner1907}], and has a straightforward closed-form Cesàro representation, curvature as a function of arc length.
 TODO: present closed-form solution.
@@
 Using a subpixel Fréchet distance bound guarantees that the rendered image does not deviate visibly from the exact rendering.
-Another choice would be uniform steps in tangent angle, as chosen by polar stroking\cite{Kilgard2020}.
+Another choice would be uniform steps in tangent angle, as chosen by polar stroking[\citenum{Kilgard2020}].
 However, at small curvature, the stroked path can be off by several pixels, and at large curvature there may be considerably more subdivision than needed for faithful rendering.
-It should be noted, to avoid the undershoot at low curvature, both the Skia\cite{Skia} and Rive\cite{Rive} renderers use a hybrid of the Wang and polar stroking error metrics.
+It should be noted, to avoid the undershoot at low curvature, both the Skia[\citenum{Skia}] and Rive[\citenum{Rive}] renderers use a hybrid of the Wang and polar stroking error metrics.
 
@@
 When a curve is monotonic, error is also monotonic with respect to both subdivision points.
-Thus, any standard root finding technique can find a subdivision with a given error (bisection is the best known, but the ITP method \cite{Oliveira2020} is better).
+Thus, any standard root finding technique can find a subdivision with a given error (bisection is the best known, but the ITP method \citet{Oliveira2020} is better).
 An similar outer loop with a root finding technique can find the error which minimizes the global error for the entire flattening, which, due to monotonicity, is that for which the error of the final segment equals that of the previous segments.
@@
 The problem of approximating a curve by a sequence of arc segments has extensive literature, but none of the published solutions are quite suitable for our application.
-The specific problem of approximating an Euler spiral by arcs is considered in \cite{Meek2004} using a ``cut then measure'' adaptive subdivision scheme, but their solution has poor quality; it scales as $O(1/n^2)$, while $O(1/n^3)$ is attainable.
-The result was improved ``slightly'' by \cite{Narayan2014}.
-The literature also contains optimal results, namely \cite{Maier2014} and \cite{Nuntawisuttiwong2021}, but at considerable cost; both approaches claim $O(n^2)$ time complexity.
+The specific problem of approximating an Euler spiral by arcs is considered in \citet{Meek2004} using a ``cut then measure'' adaptive subdivision scheme, but their solution has poor quality; it scales as $O(1/n^2)$, while $O(1/n^3)$ is attainable.
+The result was improved ``slightly'' by \citet{Narayan2014}.
+The literature also contains optimal results, namely \citet{Maier2014} and \citet{Nuntawisuttiwong2021}, but at considerable cost; both approaches claim $O(n^2)$ time complexity.
 The through-line for all these results is that they are solving a harder problem: adopting the constraint that the generated arc sequence is $G^1$ continuous.
@@
 
-In the principled, correct specification for stroking\cite{Nehab2020}, parallel curves are sufficient only for segments in which the curvature does not exceed the reciprocal half width.
+In the principled, correct specification for stroking[\citenum{Nehab2020}], parallel curves are sufficient only for segments in which the curvature does not exceed the reciprocal half width.
 When it does, additional segments must be drawn, including evolutes of the original curve.
 In general, the evolute of a cubic Bézier is a very complex curve, requiring approximation techniques.
-By contrast, the evolute of an Euler spiral ($\kappa = as$) is another spiral with a simple Cesàro equation, namely $\kappa = -a^{-1}s^{-3}$, an instance of the general result that the evolute of a log-aesthetic curve is another log-aesthetic curve\cite{Yoshida2012}.
+By contrast, the evolute of an Euler spiral ($\kappa = as$) is another spiral with a simple Cesàro equation, namely $\kappa = -a^{-1}s^{-3}$, an instance of the general result that the evolute of a log-aesthetic curve is another log-aesthetic curve[\citenum{Yoshida2012}].
 
@@
 Computing parallel curves directly on cubic Bézier curve segments is not very tractable.
-In particular, the widely cited Tiller-Hanson algorithm\cite{Tiller1984} performs well for quadratic Béziers but significantly worse for cubics.
+In particular, the widely cited Tiller-Hanson algorithm[\citenum{Tiller1984}] performs well for quadratic Béziers but significantly worse for cubics.
 
@@
 Subdivisions are also indicated at special points; for example, since quadratic Béziers cannot represent inflection points, and geometric Hermite interpolation is numerically unstable if the input curve is not convex, lowering to quadratic Béziers also requires calculation of inflection points and subdividing there.
-A good example of this pattern is Nehab\cite{Nehab2020}.
+A good example of this pattern is \citet{Nehab2020}.
 One advantage of Euler spirals over quadratic Béziers is that they can represent inflection points just fine, so it is not necessary to compute those for additional subdivision.
@@
 A key step in approximating one curve with another is evaluating the error of the approximation.
-A common approach (used in \cite{Nehab2020} among others) is to generate the approximate curve, the measure the distance, often by sampling at multiple points on the curve.
+A common approach (used in \citet{Nehab2020} among others) is to generate the approximate curve, the measure the distance, often by sampling at multiple points on the curve.
 All this is potentially slow, with the additional risk of underestimating the error due to undersampling.
@@
 Given tangent angles relative to the chord, finding the Euler spiral segment that minimizes total curvature variation is a form of geometric Hermite interpolation.
-There are a number of published solutions to this problem, involving nontrivial numerical solving techniques: gradient descent\cite{Kimia2003}, bisection\cite{Walton2009}, or Newton iteration\cite{Connor2014}.
+There are a number of published solutions to this problem, involving nontrivial numerical solving techniques: gradient descent[\citenum{Kimia2003}], bisection[\citenum{Walton2009}], or Newton iteration[\citenum{Connor2014}].
 A more direct approach is to approximate the function as a reasonably low-order polynomial in terms of the endpoint angles.
-Our approach is to use a 7th order polynomial, which is more precise than the 3rd order polynomial proposed in \cite{Reif2021}, at only modest increased cost.
+Our approach is to use a 7th order polynomial, which is more precise than the 3rd order polynomial proposed in \citet{Reif2021}, at only modest increased cost.
 The exact coefficients used are presented in Appendix \ref{appendix:gh}.
@@
 
-As recommended in both \cite{Nehab2020} and \cite{Kilgard2020}, the rendering of a cusp with infinite curvature matches that of a near-cusp.
-The code to detect near-zero derivatives and re-evaluate is a small amount of non-divergent logic, unlike the additional ``regularization'' pass proposed in Section 3.1 of \cite{Nehab2020} to handle special cases.
+As recommended in both \citet{Nehab2020} and \citet{Kilgard2020}, the rendering of a cusp with infinite curvature matches that of a near-cusp.
+The code to detect near-zero derivatives and re-evaluate is a small amount of non-divergent logic, unlike the additional ``regularization'' pass proposed in Section 3.1 of \citet{Nehab2020} to handle special cases.
 
@@
 In particular, detecting locations in the cubic Bézier where the curvature crosses a given quantity is a medium-degree polynomial, and in general requires numerical techniques for root finding.
-In the approach of \cite{Nehab2020}, the root finder not only requires a hybrid Newton/bisection method, which requires iteration, but is also recursive in that it uses the roots of a polynomial of one degree lower as a subroutine.
+In the approach of \citet{Nehab2020}, the root finder not only requires a hybrid Newton/bisection method, which requires iteration, but is also recursive in that it uses the roots of a polynomial of one degree lower as a subroutine.
 In general, polynomials up to cubic can be considered GPU-friendly, while finding cusps in cubic Béziers requires a higher degree.
@@
 Both versions can generate stroked outlines with line and arc primitives.
-We focused our evaluation on the number of output primitives and execution time, using the \emph{timings} dataset provided by Nehab2020\cite{Nehab2020}, in addition to our own curve-intensive scenes to stress the GPU implementation.
+We focused our evaluation on the number of output primitives and execution time, using the \emph{timings} dataset provided by \citet{Nehab2020}, in addition to our own curve-intensive scenes to stress the GPU implementation.
 We present our findings in the remainder of this section.
@@
 The results are shown in Figure~\ref{fig:timings}.
-The time shown is the total for test files from the \emph{timings} dataset of \cite{Nehab2020}, and the tolerance in all cases was 0.25.
+The time shown is the total for test files from the \emph{timings} dataset of \citet{Nehab2020}, and the tolerance in all cases was 0.25.
 TODO: measure against other implementations.
 
-Similarly to \cite{Nehab2020}'s observations, Skia is the fastest stroke expansion algorithm we measured.
+Similarly to \citet{Nehab2020}'s observations, Skia is the fastest stroke expansion algorithm we measured.
 The speed is attained with some compromises, in particular an imprecise error estimation that can sometimes generate outlines exceeding the given tolerance.
@@
 We evaluated the runtime performance of our compute shader on 4 GPU models and multiple form factors: Google Pixel 6 equipped with the Arm \emph{Mali-G78 MP20} mobile GPU, the Apple \emph{M1 Max} integrated laptop GPU, and two discrete desktop GPUs: the mid-2015 NVIDIA {GeForce GTX 980Ti} and the late-2022 NVIDIA {GeForce RTX 4090}.
-We authored our entire GPU pipeline in the \emph{WebGPU Shading Language} (WGSL) and used the \emph{wgpu}\cite{wgpu} framework to interoperate with the native graphics APIs (we tested the M1 Max on macOS via the Metal API; the mobile and desktop GPUs were tested via the Vulkan API on Android and Ubuntu systems).
+We authored our entire GPU pipeline in the \emph{WebGPU Shading Language} (WGSL) and used the \emph{wgpu}[\citenum{wgpu}] framework to interoperate with the native graphics APIs (we tested the M1 Max on macOS via the Metal API; the mobile and desktop GPUs were tested via the Vulkan API on Android and Ubuntu systems).
 
@@
 This allowed us to collect measurements for the curve flattening and input processing stages in isolation.
-We ran the compute pipeline on the \emph{timings} SVG files provided by Nehab\cite{Nehab2020}.
+We ran the compute pipeline on the \emph{timings} SVG files provided by \citet{Nehab2020}.
 The largest of these SVGs is \emph{waves} (see rendering \emph{(a)} in Figure \ref{fig:renderings}) which contains 42 stroked paths and a total of 13,308 path segments.
@@
 We used two variants of this scene with \emph{butt} and \emph{round} cap styles.
-For the second scene we adapted the \emph{Canvas Paths} test case from MotionMark\cite{MotionMark}, which renders a large number of randomly generated stroked line, quadratic, and cubic Bézier paths.
+For the second scene we adapted the \emph{Canvas Paths} test case from MotionMark[\citenum{MotionMark}], which renders a large number of randomly generated stroked line, quadratic, and cubic Bézier paths.
 We made two versions with different sizes: mmark-70k and mmark-120k with 70,000 and 120,000 path segments, respectively.
@@
     \includegraphics[scale=0.69]{nehab_gpu_timings}
-    \caption{Mean GPU execution times of the compute shader that performs stroke expansion and flattening when run on the Nehab\cite{Nehab2020} \emph{timings} data set.
+    \caption{Mean GPU execution times of the compute shader that performs stroke expansion and flattening when run on the \citet{Nehab2020} \emph{timings} data set.
 The timings are shown in milliseconds.
@@
 The compute shader executes at least an order of magnitude faster on the other GPUs.}
-    \Description{A stacked bar chart of the GPU timings for stroke expansion from Nehab\cite{Nehab2020} timings SVGs}
+    \Description{A stacked bar chart of the GPU timings for stroke expansion from \citet{Nehab2020} timings SVGs}
     \label{fig:nehab-gpu-timings}
@@
 Figures~\ref{fig:nehab-gpu-timings} and \ref{fig:test-scenes-gpu-timings} show the execution times for the two data sets.
-The entire Nehab\cite{Nehab2020} timings set adds up to less than 1 ms on all GPUs except the mobile unit.
+The entire \citet{Nehab2020} timings set adds up to less than 1 ms on all GPUs except the mobile unit.
 Our algorithm is ~14x slower on the Mali-G78 compared to the M1 Max but it is still capable of processing \emph{waves} in 3 ms.
@@
 
-Our GPU implementation targets graphics APIs that do not support dynamic memory allocations in shader code.
+Our GPU implementation targets graphics APIs that do not support dynamic memory allocation in shader code (in contrast to CUDA, which provides the equivalent of \emph{malloc}).
 This makes it necessary to allocate the storage buffer in advance with enough memory to store the output, however, the precise memory requirement is unknown prior to running the shader.
@@
 We augmented our encoding logic to compute a conservative estimate of the required output buffer size based on an upper bound on the number of line primitives per input path segment.
-We found that Wang's formula works well in practice, as it is relatively inexpensive to evaluate, though it results in a higher memory estimate than required (see Table~\ref{table:bump-estimate-mem}).
-For parallel curves, Wang's theorem does not guarantee an accurate upper bound on the required subdivisions, which we worked around by inflating the estimate for the source by a scale factor derived from the linewidth.
+We found that Wang's formula[\citenum{Goldman2003}] works well in practice, as it is relatively inexpensive to evaluate, though it results in a higher memory estimate than required (see Table~\ref{table:bump-estimate-mem}).
+For parallel curves, Wang's formula does not guarantee an accurate upper bound on the required subdivisions, which we worked around by inflating the estimate for the source by a scale factor derived from the linewidth.
 We observed that enabling estimation increases our CPU pre-processing time by 2-3x on the largest scenes but the impact is overall negligible when the scene size is modest.
@@
         waves.svg  & 9.37 MB   & 10.44 MB   \\
-        mmark-70k  & 23.81 MB  & 100.79   \\
-        mmark-120k & 40.86 MB  & 173.56   \\
+        mmark-70k  & 23.81 MB  & 100.79 MB  \\
+        mmark-120k & 40.86 MB  & 173.56 MB  \\
     \hline
@@
 
+\section{Conclusion}
+
+TODO:
+
 \section{Future work}
@@
     \item The GPU implementation sequentializes some work that could be parallel, and is thus not as well load-balanced as it might be.
-An appealing future direction is to split the pipeline into separate stages, executed by the GPU as nodes in a work graph\cite{Patel2024}.
+An appealing future direction is to split the pipeline into separate stages, executed by the GPU as nodes in a work graph[\citenum{Patel2024}].
 This structure is appealing because load balancing is done by hardware.
@@
 A more accurate and optimized buffer size estimation heuristic is considered future work.
-Another interesting future exploration is to employ newer GPU API features, such as mesh shaders and work graphs, for GPU-driven scheduling of the compute workload under a bounded memory footprint.
+It may also be interesting to explore whether graphics APIs can be extended to facilitate GPU-driven scheduling of the compute workload under a bounded memory footprint.
 
@@
 \bibliographystyle{ACM-Reference-Format}
+\setcitestyle{numbers}
 \bibliography{paper}

### 2024-04-29 paper.tex
Fix typos; add a keywords section with TODO

@@
 \end{abstract}
+
+\keywords{Vector Graphics, GPU, TODO}
 
@@
 The error metric for flattening an arc is exact.
-It always yields the minimum number of subdivisions needed to flatten the curve, and the flattening error is the least possible with given that number of subdivisions.
+It always yields the minimum number of subdivisions needed to flatten the curve, and the flattening error is the least possible given that number of subdivisions.
 For general curves, an exact error bound is not feasible, and we resort to an approximation.
@@
 
-An immediate consequence is that flattening an Euler spiral by choosing subdivision points $s_i = a\cdot i^\frac{2}{3}$ produces a near-optimal flattening, as can visually be seen in Figure~\ref{fig:es_flatten}.
+An immediate consequence is that flattening an Euler spiral by choosing subdivision points $s_i = a\cdot i^\frac{2}{3}$ produces a near-optimal flattening, as visualized in Figure~\ref{fig:es_flatten}.
 
@@
 The geometry of a stroke outline consists of joins, caps, and the two parallel curves on either side of the input path segments, offset by the half linewidth.
-The former are not particularly difficult to calculate, but parallel curves of cubic Béziers are notoriously tricky.
+The joins and caps are not particularly difficult to calculate, but parallel curves of cubic Béziers are notoriously tricky.
 Analytically, it is a tenth order algebraic curve, which is not particularly feasible to compute directly.
@@
 
-Proposed image: show circular arcs with polar stroking vs near-optimal flattening.
+TODO: Proposed image: show circular arcs with polar stroking vs near-optimal flattening.
 

### 2024-04-29 flatten/src/euler.rs
Fix typos.

@@ -443,7 +443,7 @@ fn integ_euler_12n(mut k0: f64, mut k1: f64, n: usize) -> (f64, f64) {
     (x * ds, y * ds)
 }
 
-/// Evaulate the Euler spiral integral.
+/// Evaluate the Euler spiral integral.
 ///
 /// Compute the following integral to the desired accuracy.
 ///

### 2024-04-29 paper.tex

--- 2024-04-29 paper.tex
@@
 \begin{figure}
-    \includegraphics[scale=0.8]{espc}
+    \includegraphics{espc}
     \caption{Integral of subdivision density for Euler spiral parallel curve, and its approximation}
@@
 
-% TODO: flatten_euler.
+\subsection{Flattening Algorithm}
+
+% TODO: explain flatten_euler
+% TODO: visualize per-segment flattened output for a fill and for a stroke from the perspective of one thread.
+
+% TODO: memory usage discussion: the impact of lines and how arcs might help.
+or work graphs, or maybe mesh shaders.
+Limitation: memory must be allocated up-front.
+bump estimation, violates the no pre-curve-evaluation criterium but simpler techniques (like Wang's) exist and we use it but way over-estimates.
+This is a limitation of the GPU execution model.
+Explain this trade-off.
+  % -> TODO: Memory usage graphs for input scene buffers and LineSoup.
+% TODO: pathological cases: our implementation doesn't handle culling.
 
@@
 We focused our evaluation on the number of output primitives and execution time, using the \emph{timings} dataset provided by Nehab2020\cite{Nehab2020}, in addition to our own curve-intensive scenes to stress the GPU implementation.
-We present our findings in the remainder of this section.
+We presents our findings in the remainder of this section.
 
@@
 
-We evaluated the runtime performance of our compute shader on 4 GPU models and multiple form factors: Google Pixel 6 equipped with the Arm \emph{Mali-G78 MP20} mobile GPU, the Apple \emph{M1 Max} integrated laptop GPU, and two discrete desktop GPUs: the mid-2015 NVIDIA {GeForce GTX 980Ti} and the late-2022 NVIDIA {GeForce RTX 4090}.
-We authored our entire GPU pipeline in the \emph{WebGPU Shading Language} (WGSL) and used the \emph{wgpu}\cite{wgpu} framework to interoperate with the native graphics APIs (we tested the M1 Max on macOS via the Metal API; the mobile and desktop GPUs were tested via the Vulkan API on Android and Ubuntu systems).
+We evaluated the runtime performance of our compute shader on 4 GPU models and multiple form factors: Google Pixel 6 equipped with the Arm \emph{Mali-G78 MP20} mobile GPU, the Apple \emph{M1 Max} integrated laptop GPU, as well as two discrete desktop GPUs: the mid-2015 NVIDIA {GeForce GTX 980Ti} and the late-2022 NVIDIA {GeForce RTX 4090}.
+We authored our entire GPU pipeline in the \emph{WebGPU Shading Language} (WGSL) and used the \emph{wgpu}\cite{wgpu} framework to natively interoperate with the native graphics APIs (we tested the M1 Max on macOS via the Metal API; the mobile and desktop GPUs were tested via the Vulkan API on Android and Ubuntu systems).
 
 We instrumented our pipeline to obtain fine-grained pipeline stage execution times using the GPU \emph{timestamp query} feature supported by both Metal and Vulkan.
-This allowed us to collect measurements for the curve flattening and input processing stages in isolation.
+This allowed us to collect measurements for the curve flattening stage in isolation.
 We ran the compute pipeline on the \emph{timings} SVG files provided by Nehab\cite{Nehab2020}.
@@
 
-We authored two additional test scenes in order stress the GPU with a higher workload.
-The first is a very large stroked path with ~500,000 individually styled dashed segments which we adapted from Skia's open-source test corpus.
+In order to stress the GPU with a higher workload, we made two additional test scenes.
+The first is a very large stroked path with 503,304 individually styled dashed segments which we adapted from Skia's open-source test corpus.
 We used two variants of this scene with \emph{butt} and \emph{round} cap styles.
-For the second scene we adapted the \emph{Canvas Paths} test case from MotionMark\cite{MotionMark}, which renders a large number of randomly generated stroked line, quadratic, and cubic Bézier paths.
-We made two versions with different sizes: mmark-70k and mmark-120k with 70,000 and 120,000 path segments, respectively.
-Table~\ref{table:bump-counts} shows the payload sizes of our largest test scenes.
+For the second scene we adapted the \emph{Canvas Paths} test case from MotionMark\cite{todo:mmark}, which renders a large number of randomly generated stroked line, quadratic, and cubic Bézier paths.
+We made two versions with different sizes: mmark-35k and mmark-60k with 30,000 and 60,000 paths respectively.
+Table \ref{table:bump-counts} shows the payload sizes of our largest test scenes.
 
@@
         waves.svg & 13,308 & 406,059 & 111,965 \\
-        mmark-70k & 70,000 & 1,039,899 & 500,351 \\
-        mmark-120k & 120,000 & 1,784,896 & 858,162 \\
+        mmark-35k & 105,022 & 1,039,899 & 500,351 \\
+        mmark-60k & 179,929 & 1,784,896 & 858,162 \\
         long dash (butt) & 503,304 & 1,060,476 & 1,060,476 \\
@@
     \includegraphics[scale=0.69]{nehab_gpu_timings}
-    \caption{Mean GPU execution times of the compute shader that performs stroke expansion and flattening when run on the Nehab\cite{Nehab2020} \emph{timings} data set.
-The timings are shown in milliseconds.
+    \caption{Mean GPU execution times of the compute shader that performs stroke expansion and flattening when run on the Nehab\cite{Nehab2020} \emph{timings} data set, in microseconds.
 The hatched bars show the timings for the version that outputs line primitives while the solid bars show that for arcs.
@@
 All images were rendered on an Apple M1 Max at 2088x1600 resolution using our GPU pipeline.
-\textbf{(a)} \emph{waves.svg} from the Nehab2020 timings data set (340fps); \textbf{(b)} The \emph{mmark-120k} scene with 120,000 input segments (72fps); \textbf{(c)} The \emph{long path} scene with ~500k input segments and round caps (40fps with CPU-side dashing - the GPU time to render the scene is 9ms); \textbf{(d)}; A simple test scene showcasing various cap and join styles (950fps); \textbf{(e)} The famous \emph{Ghostscript Tiger} containing both stroked and filled paths (810fps).}
+\textbf{(a)} \emph{waves.svg} from the Nehab2020 timings data set (340fps); \textbf{(b)} The \emph{mmark-60k} scene with ~180k input segments (72fps); \textbf{(c)} The \emph{long path} scene with ~500k input segments and round caps (40fps with CPU-side dashing - the GPU time to render the scene is 9ms); \textbf{(d)}; A simple test scene showcasing various cap and join styles (950fps); \textbf{(e)} The famous \emph{Ghostscript Tiger} containing both stroked and filled paths (810fps).}
     \Description{Renderings of the test scenes}
@@
 
-Figures~\ref{fig:nehab-gpu-timings} and \ref{fig:test-scenes-gpu-timings} show the execution times for the two data sets.
-The entire Nehab\cite{Nehab2020} timings set adds up to less than 1 ms on all GPUs except the mobile unit.
-Our algorithm is ~14x slower on the Mali-G78 compared to the M1 Max but it is still capable of processing \emph{waves} in 3 ms.
-We observed that the performance of the kernel scales well with increasing GPU ability even at very large workloads, as demonstrated by the order-of-magnitude difference in run time between the mobile, laptop, and high-end desktop form factors we tested.
-We also confirmed that outputting arcs instead of lines can lead to a 2x decrease in execution time (particularly in scenes with high curve content, like \emph{waves} and \emph{mmark}) as predicted by the lower overall primitive count.
+Figures \ref{fig:nehab-gpu-timings} and \ref{fig:test-scenes-gpu-timings} show the execution times of our compute shader.
+TODO: brief assessment.
 
-\subsection{GPU: Input Processing} \label{subsection:encoding-results}
+% TODO: line soup is bump allocated. talk about allocation and estimation.
 
-Figure~\ref{fig:tag-monoid-gpu-timings} shows the run time of the tag monoid parallel prefix sums with increasing input size.
-The prefix sums scale better at large input sizes compared to our stroke expansion kernel.
-This can be attributed to a lack of expensive computations and high control-flow uniformity in the tag monoid kernel.
-The CPU encoding times (prior to the GPU prefix sums) for some of the large scenes is shown in Table~\ref{table:bump-estimate-cost}.
-Dashing on the GPU was out of scope, so the dash style applied to the \emph{long dash} scene had to be processed on the CPU and dashes were sequantially encoded as individual strokes.
-Including CPU-side dashing, the time to encode \emph{long dash} on the Apple M1 Max was approximately 25 ms on average, while the average time to encode the path after dashing was 4.14 ms.
-
-Our GPU implementation targets graphics APIs that do not support dynamic memory allocations in shader code.
-This makes it necessary to allocate the storage buffer in advance with enough memory to store the output, however, the precise memory requirement is unknown prior to running the shader.
-
-We augmented our encoding logic to compute a conservative estimate of the required output buffer size based on an upper bound on the number of line primitives per input path segment.
-We found that Wang's formula works well in practice, as it is relatively inexpensive to evaluate, though it results in a higher memory estimate than required (see Table~\ref{table:bump-estimate-mem}).
-For parallel curves, Wang's theorem does not guarantee an accurate upper bound on the required subdivisions, which we worked around by inflating the estimate for the source by a scale factor derived from the linewidth.
-We observed that enabling estimation increases our CPU pre-processing time by 2-3x on the largest scenes but the impact is overall negligible when the scene size is modest.
-
-\begin{figure}
-    \includegraphics[scale=0.66]{tag_monoid_gpu_timings}
-    \caption{Mean GPU execution time of the tag monoid parallel prefix sums with increasing number of input path segments, measured on Apple M1 Max and NVIDIA RTX 4090.
-The figure also shows the execution time of the stroke expansion and flattening kernel.
-The y-axis is shown in logarithmic scale to highlight the similar scaling trend on both GPUs despite the order-of-magnitude difference in performance.}
-    \Description{Mean GPU execution time of the tag monoid kernel}
-    \label{fig:tag-monoid-gpu-timings}
-\end{figure}
-
-\begin{table}
-    \caption{Comparison of CPU encoding time with and without size estimation on the largest scenes.
-All measurements were computed on an Apple M1 Max.
-The additional cost is negligible when the scene size is modest but tangible with increased complexity.}
-    \begin{tabular}{|l|c|c|}
-    \hline
-        \multicolumn{1}{|c|}{\textbf{Test}} & \textbf{Encoding} & \textbf{Encoding + Estimation} \\
-    \hline
-        roads.svg  & 398.56 $\mu$s & 618.24 $\mu$s \\
-        waves.svg  & 197.53 $\mu$s & 418.96 $\mu$s \\
-        mmark-70k  & 2.69 ms       & 6.89 ms       \\
-        mmark-120k & 4.28 ms       & 11.59 ms      \\
-    \hline
-    \end{tabular}
-    \label{table:bump-estimate-cost}
-\end{table}
-
-\begin{table}
-    \caption{The estimated and actual \emph{line soup} buffer sizes for scenes rendered at 2088x1600 resolution.
-Our estimation uses Wang's formula combined with a scaling heuristic based on the linewidth.
-In the worst case, our buffer size estimator overshoots by a factor ~4.}
-    \begin{tabular}{|l|c|c|}
-    \hline
-        \multicolumn{1}{|c|}{\textbf{Test}} & Actual Size & Estimated Size \\
-    \hline
-        roads.svg  & 958.78 KB & 2.45 MB  \\
-        waves.svg  & 9.37 MB   & 10.44 MB \\
-        mmark-70k  & 23.81 MB  & 100.79   \\
-        mmark-120k & 40.86 MB  & 173.56   \\
-    \hline
-    \end{tabular}
-    \label{table:bump-estimate-mem}
-\end{table}
+\section{Discussion}
+\subsection{Limitations}
 
@@
 
-    \item The pre-allocation requirement for bump-allocated line soup buffer is a limitation due to today's graphics APIs.
-A more accurate and optimized buffer size estimation heuristic is considered future work.
-Another interesting future exploration is to employ newer GPU API features, such as mesh shaders and work graphs, for GPU-driven scheduling of the compute workload under a bounded memory footprint.
-
     \item Many of the error metrics were empirically determined.
 The mathematical theory behind them should be developed more rigorously, and that process will likely uncover opportunities to fine-tune the technique.
+\end{itemize}
 
-    \item Dashing stroked paths on the GPU was left out of scope for this paper. Given its parameterization by arc length, we believe the Euler spiral representation is highly suitable for dashing.
-\end{itemize}
+% TODO: mention dashing as future work?
+raph@: I think so.
+When we describe dashing, it would be good to say that the ES representation is already optimal for dashing, as it's parametrized by arc length.
+% TODO: triangles
 

### 2024-04-29 paper.tex
Tweak timings figure
Clarify caption, use milliseconds.

@@
 Like these other approaches, we also use an intermediate curve, but our choice is an Euler spiral.
-In some ways it is similar to quadratic Béziers -- it also has $O(n^4)$ scaling and is best computed using geometric Hermite interpolation -- but differs in others.
+In some ways it is similar to quadratic Béziers -- it also has $O(n^4)$ scaling and is best computed using geometric Hermite interpoliation -- but differs in others.
 It has no difficulty modeling an inflection point.
@@
 
-The geometry of a stroke outline consists of joins, caps, and the two parallel curves on either side of the input path segments, offset by the half linewidth.
+The geometry of a stroke outline consists of joins, caps, and the two parallel curves on either side of the input path segemnts, offset by the half linewidth.
 The former are not particularly difficult to calculate, but parallel curves of cubic Béziers are notoriously tricky.
@@
     \includegraphics[scale=0.75]{cubic_err_measured} \includegraphics[scale=0.75]{cubic_err_approx}
-    \caption{Comparison of measured (top) and approximated (bottom) error for cubic Bézier to Euler spiral conversion. Scale is log (base 10) of the error.}
+    \caption{Comparision of measured (top) and approximated (bottom) error for cubic Bézier to Euler spiral conversion. Scale is log (base 10) of the error.}
     \Description{TODO}
@@
 Each thread in the global grid is assigned to process a single path segment.
-We encode the individual path segments contiguously in a GPU storage buffer such that each element has 1:1 correspondence to a global thread ID in the dispatch.
+We encode the individual path segments contiguously in a GPU storage buffer such that each element has 1:1 correspondance to a global thread ID in the dispatch.
 This layout easily scales to hundreds of thousands of input path segments, which can be distributed across any number of paths and subpaths.
@@
 The number of arcs and quadratic Bézier primitives is comparable at practical tolerances, but the count of quadratic Béziers becomes more favorable at finer tolerances.
-These vertical axis shows the sum of output segment counts for test files from the \emph{timings} dataset.
+The vertical axis shows the sum of output segment counts for test files from the \emph{timings} dataset, plus mmark-70k (as described in more detail in Section~\ref{subsection:gpu-results}), and omitting the \emph{waves} example\footnote{The \emph{waves} example triggers triggers Skia issue 336617138 at finer tolerance.}.
 

### 2024-04-29 paper.tex
rebase and address comments

--- 2024-04-29 paper.tex
@@
 
-In order to stress the GPU with a higher workload, we made two additional test scenes.
-The first is a very large stroked path with 503,304 individually styled dashed segments which we adapted from Skia's open-source test corpus.
+We authored two additional test scenes in order stress the GPU with a higher workload.
+The first is a very large stroked path with ~500,000 individually styled dashed segments which we adapted from Skia's open-source test corpus.
 We used two variants of this scene with \emph{butt} and \emph{round} cap styles.
-For the second scene we adapted the \emph{Canvas Paths} test case from MotionMark\cite{todo:mmark}, which renders a large number of randomly generated stroked line, quadratic, and cubic Bézier paths.
-We made two versions with different sizes: mmark-35k and mmark-60k with 30,000 and 60,000 paths respectively.
-Table \ref{table:bump-counts} shows the payload sizes of our largest test scenes.
+For the second scene we adapted the \emph{Canvas Paths} test case from MotionMark\cite{MotionMark}, which renders a large number of randomly generated stroked line, quadratic, and cubic Bézier paths.
+We made two versions with different sizes: mmark-70k and mmark-120k with 70,000 and 120,000 path segments, respectively.
+Table~\ref{table:bump-counts} shows the payload sizes of our largest test scenes.
 
@@
         waves.svg & 13,308 & 406,059 & 111,965 \\
-        mmark-35k & 105,022 & 1,039,899 & 500,351 \\
-        mmark-60k & 179,929 & 1,784,896 & 858,162 \\
+        mmark-70k & 70,000 & 1,039,899 & 500,351 \\
+        mmark-120k & 120,000 & 1,784,896 & 858,162 \\
         long dash (butt) & 503,304 & 1,060,476 & 1,060,476 \\
@@
 
-Figures \ref{fig:nehab-gpu-timings} and \ref{fig:test-scenes-gpu-timings} show the execution times of our compute shader.
-TODO: brief assessment.
+Figures~\ref{fig:nehab-gpu-timings} and \ref{fig:test-scenes-gpu-timings} show the execution times for the two data sets.
+The entire Nehab\cite{Nehab2020} timings set adds up to less than 1 ms on all GPUs except the mobile unit.
+Our algorithm is ~14x slower on the Mali-G78 compared to the M1 Max but it is still capable of processing \emph{waves} in 3 ms.
+We observed that the performance of the kernel scales well with increasing GPU ability even at very large workloads, as demonstrated by the order-of-magnitude difference in run time between the mobile, laptop, and high-end desktop form factors we tested.
+We also confirmed that outputting arcs instead of lines can lead to a 2x decrease in execution time (particularly in scenes with high curve content, like \emph{waves} and \emph{mmark}) as predicted by the lower overall primitive count.
 
-% TODO: line soup is bump allocated. talk about allocation and estimation.
+\subsection{GPU: Input Processing} \label{subsection:encoding-results}
 
-\section{Discussion}
-\subsection{Limitations}
+Figure~\ref{fig:tag-monoid-gpu-timings} shows the run time of the tag monoid parallel prefix sums with increasing input size.
+The prefix sums scale better at large input sizes compared to our stroke expansion kernel.
+This can be attributed to a lack of expensive computations and high control-flow uniformity in the tag monoid kernel.
+The CPU encoding times (prior to the GPU prefix sums) for some of the large scenes is shown in Table~\ref{table:bump-estimate-cost}.
+Dashing on the GPU was out of scope, so the dash style applied to the \emph{long dash} scene had to be processed on the CPU and dashes were sequantially encoded as individual strokes.
+Including CPU-side dashing, the time to encode \emph{long dash} on the Apple M1 Max was approximately 25 ms on average, while the average time to encode the path after dashing was 4.14 ms.
+
+Our GPU implementation targets graphics APIs that do not support dynamic memory allocation in shader code (unlike CUDA which provides the equivalent of \emph{malloc}).
+This makes it necessary to allocate the storage buffer in advance with enough memory to store the output, however, the precise memory requirement is unknown prior to running the shader.
+
+We augmented our encoding logic to compute a conservative estimate of the required output buffer size based on an upper bound on the number of line primitives per input path segment.
+We found that Wang's formula[\citenum{Goldman2003}] works well in practice, as it is relatively inexpensive to evaluate, though it results in a higher memory estimate than required (see Table~\ref{table:bump-estimate-mem}).
+For parallel curves, Wang's formula does not guarantee an accurate upper bound on the required subdivisions, which we worked around by inflating the estimate for the source by a scale factor derived from the linewidth.
+We observed that enabling estimation increases our CPU pre-processing time by 2-3x on the largest scenes but the impact is overall negligible when the scene size is modest.
+
+\begin{figure}
+    \includegraphics[scale=0.66]{tag_monoid_gpu_timings}
+    \caption{Mean GPU execution time of the tag monoid parallel prefix sums with increasing number of input path segments, measured on Apple M1 Max and NVIDIA RTX 4090.
+The figure also shows the execution time of the stroke expansion and flattening kernel.
+The y-axis is shown in logarithmic scale to highlight the similar scaling trend on both GPUs despite the order-of-magnitude difference in performance.}
+    \Description{Mean GPU execution time of the tag monoid kernel}
+    \label{fig:tag-monoid-gpu-timings}
+\end{figure}
+
+\begin{table}
+    \caption{Comparison of CPU encoding time with and without size estimation on the largest scenes.
+All measurements were computed on an Apple M1 Max.
+The additional cost is negligible when the scene size is modest but tangible with increased complexity.}
+    \begin{tabular}{|l|c|c|}
+    \hline
+        \multicolumn{1}{|c|}{\textbf{Test}} & \textbf{Encoding} & \textbf{Encoding + Estimation} \\
+    \hline
+        roads.svg  & 398.56 $\mu$s & 618.24 $\mu$s \\
+        waves.svg  & 197.53 $\mu$s & 418.96 $\mu$s \\
+        mmark-70k  & 2.69 ms       & 6.89 ms       \\
+        mmark-120k & 4.28 ms       & 11.59 ms      \\
+    \hline
+    \end{tabular}
+    \label{table:bump-estimate-cost}
+\end{table}
+
+\begin{table}
+    \caption{The estimated and actual \emph{line soup} buffer sizes for scenes rendered at 2088x1600 resolution.
+Our estimation uses Wang's formula combined with a scaling heuristic based on the linewidth.
+In the worst case, our buffer size estimator overshoots by a factor ~4.}
+    \begin{tabular}{|l|c|c|}
+    \hline
+        \multicolumn{1}{|c|}{\textbf{Test}} & Actual Size & Estimated Size \\
+    \hline
+        roads.svg  & 958.78 KB & 2.45 MB    \\
+        waves.svg  & 9.37 MB   & 10.44 MB   \\
+        mmark-70k  & 23.81 MB  & 100.79 MB  \\
+        mmark-120k & 40.86 MB  & 173.56 MB  \\
+    \hline
+    \end{tabular}
+    \label{table:bump-estimate-mem}
+\end{table}
+
+\section{Conclusion}
+
+TODO:
 
@@
 
+    \item The pre-allocation requirement for bump-allocated line soup buffer is a limitation due to today's graphics APIs.
+A more accurate and optimized buffer size estimation heuristic is considered future work.
+It may also be interesting to explore whether graphics APIs can be extended to facilitate GPU-driven scheduling of the compute workload under a bounded memory footprint.
+
     \item Many of the error metrics were empirically determined.
 The mathematical theory behind them should be developed more rigorously, and that process will likely uncover opportunities to fine-tune the technique.
+
+    \item Dashing stroked paths on the GPU was left out of scope for this paper. Given its parameterization by arc length, we believe the Euler spiral representation is highly suitable for dashing.
 \end{itemize}
-
-% TODO: mention dashing as future work?
-raph@: I think so.
-When we describe dashing, it would be good to say that the ES representation is already optimal for dashing, as it's parametrized by arc length.
-% TODO: triangles
 

### 2024-04-29 paper.tex
fix phrasing

@@
 
-Our GPU implementation targets graphics APIs that do not support dynamic memory allocation in shader code (unlike CUDA which provides the equivalent of \emph{malloc}).
+Our GPU implementation targets graphics APIs that do not support dynamic memory allocation in shader code (in contrast to CUDA, which provides the equivalent of \emph{malloc}).
 This makes it necessary to allocate the storage buffer in advance with enough memory to store the output, however, the precise memory requirement is unknown prior to running the shader.

### 2024-04-29 paper.tex
fix keywords; inline comment

@@
 
+\keywords{Vector Graphics, Stroke, Offset Curve, Path Rendering, GPU}
+
 \maketitle
@@
 Using a subpixel Fréchet distance bound guarantees that the rendered image does not deviate visibly from the exact rendering.
-Another choice would be uniform steps in tangent angle, as chosen by polar stroking\cite{Kilgard2020}.
+Another choice would be uniform steps in tangent angle, as chosen by polar stroking[\citenum{Kilgard2020}].
 However, at small curvature, the stroked path can be off by several pixels, and at large curvature there may be considerably more subdivision than needed for faithful rendering.
-It should be noted, to avoid the undershoot at low curvature, both the Skia\cite{Skia} and Rive\cite{Rive} renderers use a hybrid of the Wang and polar stroking error metrics.
+It should be noted, to avoid the undershoot at low curvature, both the Skia[\citenum{Skia}] and Rive[\citenum{Rive}] renderers use a hybrid of the Wang and polar stroking error metrics.
 
-Proposed image: show circular arcs with polar stroking vs near-optimal flattening.
+TODO: Proposed image: show circular arcs with polar stroking vs near-optimal flattening.
 
@@
 \bibliographystyle{ACM-Reference-Format}
+\setcitestyle{numbers} % Switch back to numerical mode to show citation numbers in bibliography
 \bibliography{paper}

### 2024-04-29 paper.tex
Add g comparison figure
WIP

--- 2024-04-29 paper.tex
@@
 
-TODO: image. This will likely strongly resemble Fig. 8a from Narayan2014.
-
 Remarkably, the approximation of an Euler spiral parallel curve by arc segments is almost as simple as that for Euler spirals to arcs.
@@
 
-% Make improved quality claim here? I think it may be too subtle.
+See Figure~\ref{fig:g_comparison} for a comparison between flattening and approximation with arc segments.
+The arc segment version has many fewer segments at the same tolerance, while preserving very high visual quality.
+
+\begin{figure}
+    \includegraphics[scale=0.65]{g_comparison}
+    \caption{A lowercase `g' glyph from Nimbus Roman (left, constructed from cubic Bézier segments), its flattening to lines (center), and its approximation by arc segments (right), both with a tolerance of 2.0.}
+    \Description{TODO}
+    \label{fig:g_comparison}
+\end{figure}
 
@@
 
-Figures~\ref{fig:nehab-gpu-timings} and \ref{fig:test-scenes-gpu-timings} show the execution times for the two data sets.
-The entire Nehab\cite{Nehab2020} timings set adds up to less than 1 ms on all GPUs except the mobile unit.
-Our algorithm is ~14x slower on the Mali-G78 compared to the M1 Max but it is still capable of processing \emph{waves} in 3 ms.
-We observed that the performance of the kernel scales well with increasing GPU ability even at very large workloads, as demonstrated by the order-of-magnitude difference in run time between the mobile, laptop, and high-end desktop form factors we tested.
-We also confirmed that outputting arcs instead of lines can lead to a 2x decrease in execution time (particularly in scenes with high curve content, like \emph{waves} and \emph{mmark}) as predicted by the lower overall primitive count.
+Figures \ref{fig:nehab-gpu-timings} and \ref{fig:test-scenes-gpu-timings} show the execution times of our compute shader.
+TODO: brief assessment.
 
-\subsection{GPU: Input Processing} \label{subsection:encoding-results}
+% TODO: line soup is bump allocated. talk about allocation and estimation.
 
-Figure~\ref{fig:tag-monoid-gpu-timings} shows the run time of the tag monoid parallel prefix sums with increasing input size.
-The prefix sums scale better at large input sizes compared to our stroke expansion kernel.
-This can be attributed to a lack of expensive computations and high control-flow uniformity in the tag monoid kernel.
-The CPU encoding times (prior to the GPU prefix sums) for some of the large scenes is shown in Table~\ref{table:bump-estimate-cost}.
-Dashing on the GPU was out of scope, so the dash style applied to the \emph{long dash} scene had to be processed on the CPU and dashes were sequantially encoded as individual strokes.
-Including CPU-side dashing, the time to encode \emph{long dash} on the Apple M1 Max was approximately 25 ms on average, while the average time to encode the path after dashing was 4.14 ms.
-
-Our GPU implementation targets graphics APIs that do not support dynamic memory allocation in shader code (unlike CUDA which provides the equivalent of \emph{malloc}).
-This makes it necessary to allocate the storage buffer in advance with enough memory to store the output, however, the precise memory requirement is unknown prior to running the shader.
-
-We augmented our encoding logic to compute a conservative estimate of the required output buffer size based on an upper bound on the number of line primitives per input path segment.
-We found that Wang's formula[\citenum{Goldman2003}] works well in practice, as it is relatively inexpensive to evaluate, though it results in a higher memory estimate than required (see Table~\ref{table:bump-estimate-mem}).
-For parallel curves, Wang's formula does not guarantee an accurate upper bound on the required subdivisions, which we worked around by inflating the estimate for the source by a scale factor derived from the linewidth.
-We observed that enabling estimation increases our CPU pre-processing time by 2-3x on the largest scenes but the impact is overall negligible when the scene size is modest.
-
-\begin{figure}
-    \includegraphics[scale=0.66]{tag_monoid_gpu_timings}
-    \caption{Mean GPU execution time of the tag monoid parallel prefix sums with increasing number of input path segments, measured on Apple M1 Max and NVIDIA RTX 4090.
-The figure also shows the execution time of the stroke expansion and flattening kernel.
-The y-axis is shown in logarithmic scale to highlight the similar scaling trend on both GPUs despite the order-of-magnitude difference in performance.}
-    \Description{Mean GPU execution time of the tag monoid kernel}
-    \label{fig:tag-monoid-gpu-timings}
-\end{figure}
-
-\begin{table}
-    \caption{Comparison of CPU encoding time with and without size estimation on the largest scenes.
-All measurements were computed on an Apple M1 Max.
-The additional cost is negligible when the scene size is modest but tangible with increased complexity.}
-    \begin{tabular}{|l|c|c|}
-    \hline
-        \multicolumn{1}{|c|}{\textbf{Test}} & \textbf{Encoding} & \textbf{Encoding + Estimation} \\
-    \hline
-        roads.svg  & 398.56 $\mu$s & 618.24 $\mu$s \\
-        waves.svg  & 197.53 $\mu$s & 418.96 $\mu$s \\
-        mmark-70k  & 2.69 ms       & 6.89 ms       \\
-        mmark-120k & 4.28 ms       & 11.59 ms      \\
-    \hline
-    \end{tabular}
-    \label{table:bump-estimate-cost}
-\end{table}
-
-\begin{table}
-    \caption{The estimated and actual \emph{line soup} buffer sizes for scenes rendered at 2088x1600 resolution.
-Our estimation uses Wang's formula combined with a scaling heuristic based on the linewidth.
-In the worst case, our buffer size estimator overshoots by a factor ~4.}
-    \begin{tabular}{|l|c|c|}
-    \hline
-        \multicolumn{1}{|c|}{\textbf{Test}} & Actual Size & Estimated Size \\
-    \hline
-        roads.svg  & 958.78 KB & 2.45 MB    \\
-        waves.svg  & 9.37 MB   & 10.44 MB   \\
-        mmark-70k  & 23.81 MB  & 100.79 MB  \\
-        mmark-120k & 40.86 MB  & 173.56 MB  \\
-    \hline
-    \end{tabular}
-    \label{table:bump-estimate-mem}
-\end{table}
-
-\section{Conclusion}
-
-TODO:
+\section{Discussion}
+\subsection{Limitations}
 
@@
 
-    \item The pre-allocation requirement for bump-allocated line soup buffer is a limitation due to today's graphics APIs.
-A more accurate and optimized buffer size estimation heuristic is considered future work.
-It may also be interesting to explore whether graphics APIs can be extended to facilitate GPU-driven scheduling of the compute workload under a bounded memory footprint.
-
     \item Many of the error metrics were empirically determined.
 The mathematical theory behind them should be developed more rigorously, and that process will likely uncover opportunities to fine-tune the technique.
+\end{itemize}
 
-    \item Dashing stroked paths on the GPU was left out of scope for this paper. Given its parameterization by arc length, we believe the Euler spiral representation is highly suitable for dashing.
-\end{itemize}
+% TODO: mention dashing as future work?
+raph@: I think so.
+When we describe dashing, it would be good to say that the ES representation is already optimal for dashing, as it's parametrized by arc length.
+% TODO: triangles
 

### 2024-04-29 flatten/src/svg.rs

@@ -228,6 +228,7 @@ impl SvgScene {
         &self,
         out: &mut impl Write,
         paths: &[LoweredPath<impl Lowering>],
+        outline: bool,
     ) -> Result<(), Box<dyn Error>> {
         // these should probably be fields in the scene
         let width = 1024;
@@ -235,7 +236,19 @@ impl SvgScene {
         writeln!(out, "<svg width=\"{width}\" height=\"{height}\" viewBox=\"0 0 {width} {height}\" xmlns=\"http://www.w3.org/2000/svg\">")?;
         for path in paths {
             let svg_path = path.to_svg();
-            writeln!(out, "  <path d='{svg_path}' />")?;
+            let style = if outline {
+                "stroke='#000' fill='#ccc' "
+            } else {
+                ""
+            };
+            writeln!(out, "  <path d='{svg_path}' {style}/>")?;
+            if outline {
+                for segment in &path.path {
+                    let p = segment.end_point();
+                    let (x, y) = (p.x, p.y);
+                    writeln!(out, "  <circle cx='{x}' cy='{y}' r='3' />")?;
+                }
+            }
         }
         writeln!(out, "</svg>")?;
         Ok(())
@@ -253,6 +266,8 @@ pub struct SvgArgs {
     primitive: Option<String>,
     #[arg(short, long)]
     output_file: Option<String>,
+    #[arg(short, long)]
+    figure_outline: bool,
 }
 
 #[derive(Clone, Parser)]
@@ -296,11 +311,11 @@ pub fn svg_main(args: SvgArgs) {
         match prim_type {
             PrimType::Line => {
                 let paths: Vec<LoweredPath<Line>> = scene.expand(tolerance);
-                scene.to_svg(&mut f, &paths).unwrap();
+                scene.to_svg(&mut f, &paths, args.figure_outline).unwrap();
             }
             PrimType::Arc => {
                 let paths: Vec<LoweredPath<ArcSegment>> = scene.expand(tolerance);
-                scene.to_svg(&mut f, &paths).unwrap();
+                scene.to_svg(&mut f, &paths, args.figure_outline).unwrap();
             }
         }
     }

### 2024-04-29 paper.tex

--- 2024-04-29 paper.tex
@@
 
-We measured the timing of our CPU implementation, generating both line and arc primitives, against Skia's stroker, which outputs a mix of straight lines and quadratic Bézier primitives.
+We measured the timing of our CPU implementation, generating both line and arc primitives, against the Nehab and Skia strokers, which both output a mix of straight lines and quadratic Bézier primitives.
 The results are shown in Figure~\ref{fig:timings}.
-The time shown is the total for test files from the \emph{timings} dataset of \citet{Nehab2020}, and the tolerance in all cases was 0.25.
-TODO: measure against other implementations.
+The time shown is the total for test files from the \emph{timings} dataset of \citet{Nehab2020}, and the tolerance was 0.25 when adjustable as a parameter.
 

### 2024-04-29 paper.tex

@@
 The number of arcs and quadratic Bézier primitives is comparable at practical tolerances, but the count of quadratic Béziers becomes more favorable at finer tolerances.
-The vertical axis shows the sum of output segment counts for test files from the \emph{timings} dataset, plus mmark-70k (as described in more detail in Section~\ref{subsection:gpu-results}), and omitting the \emph{waves} example\footnote{The \emph{waves} example triggers triggers Skia issue 336617138 at finer tolerance.}.
+The vertical axis shows the sum of output segment counts for test files from the \emph{timings} dataset, plus mmark-70k (as described in more detail in Section~\ref{subsection:gpu-results}), and omitting the \emph{waves} example\footnote{The \emph{waves} example triggers Skia issue https://issues.skia.org/issues/336617138 at finer tolerance.}.
 
@@
 \end{figure}
-
-NOTE: the increase in Skia numbers at tolerances below 0.01 or so is the result of a precision bug in Skia, uncovered during preparation of this chart.
-We expect that to be fixed, and will revise the chart.
 

### 2024-04-29 paper.tex

@@
 The parallel curve of the Euler spiral (also known as ``clothoid'') was characterized by \citet{Wieleitner1907} well over a hundred years ago[\citenum{Wieleitner1907}], and has a straightforward closed-form Cesàro representation, curvature as a function of arc length.
-TODO: present closed-form solution.
 
@@
 However, at small curvature, the stroked path can be off by several pixels, and at large curvature there may be considerably more subdivision than needed for faithful rendering.
+
+The limitation of the angle step error metric is shown in Figure~\ref{fig:angle_err}.
+The top row shows the use of a distance-based error metric, as is used in our approach, which is visually consistent at varying curvature (in practice, a tolerance value of 0.25 of a pixel is below the threshold of perceptibility).
+The bottom row shows a consistent angle step, as implemented in polar stroking, but has excessive distance error at low curvature, and excessive subdivision at high curvature.
 It should be noted, to avoid the undershoot at low curvature, both the Skia[\citenum{Skia}] and Rive[\citenum{Rive}] renderers use a hybrid of the Wang and polar stroking error metrics.
 
-TODO: Proposed image: show circular arcs with polar stroking vs near-optimal flattening.
+\begin{figure}
+    \includegraphics[scale=0.5]{angle_err}
+    \caption{Comparison of error metrics. Top row (a) shows a distance based error metric. Bottom row (b) shows the angle step error metric.}
+    \Description{Four stroked arcs with different amount of subdivision. The top row shows a low-curvature and high-curvature arc with a consistent distance error}
+    \label{fig:angle_err}
+\end{figure}
 
@@
 \end{figure}
+
+\begin{figure*}
+    \includegraphics[scale=0.2]{tricky_strokes}
+    \caption{Stroked cubic Béziers with cusps and near-cusps, rendered using our algorithm. The individual curves were adapted from the Skia test corpus.}
+    \Description{Stroked cubic Béziers with cusps and near-cusps, rendered using our algorithm.}
+    \label{fig:tricky_strokes}
+\end{figure*}
 

### 2024-04-29 paper.tex
some textual fixes

@@
 
-The optimal flattening of a curve, with a maximum error constraint, is as follows.
-An optimal flattening contains the minimum number of subdivisions required to meet that constraint.
-Further, the error is the minimum possible with that number of subdivisions.
+The optimal flattening of a curve contains the minimum number of subdivisions required to meet a maximum error constraint, in which the error is the minimum possible with that number of subdivisions.
 
@@
 Thus, any standard root finding technique can find a subdivision with a given error (bisection is the best known, but the ITP method \citet{Oliveira2020} is better).
-An similar outer loop with a root finding technique can find the error which minimizes the global error for the entire flattening, which, due to monotonicity, is that for which the error of the final segment equals that of the previous segments.
+A similar outer loop with a root finding technique can find the error which minimizes the global error for the entire flattening, which, due to monotonicity, is that for which the error of the final segment equals that of the previous segments.
 Subdividing at the inflection point adds at most one additional point; it is possible to optimize further, but we do not consider that (though the near-optimal flattening we will later present is not subject to this constraint).
 
-\subsection{The subdivision density integral}
+\subsection{The subdivision density integral} \label{subsection:subdiv-density-int}
 
@@
 
-The subdivision density integral [TODO: probably should use section or equation number reference] is fairly straightforward to compute in the forward direction, but not invertible using a straightforward closed-form equation.
+The subdivision density integral (Section~\ref{subsection:subdiv-density-int}) is fairly straightforward to compute in the forward direction, but not invertible using a straightforward closed-form equation.
 Numerical techniques are possible, but require multiple iterations to achieve sufficient accuracy, so are slower.
-In this subsection, we present a straightforward and accurate approximation, based on piecewise easily invertible functions.
+In this subsection, we present a straightforward and accurate approximation, based on easily invertible piecewise functions.
 If higher flattening quality is desired at the expense of slower computation, this approximation can be used to determine a good initial value for numeric techniques; two iterations of Newton solving are enough to refine this guess to within 32-bit floating point accuracy.
@@
 
-This formula was determined empirically by curve-fitting measured error values from fitting Euler spiral parallel curves to arcs, but was also inspired by applying the general error metric formula to the analytical equations for Euler spiral parallel curve, and dropping higher order terms.
+This formula was determined empirically by curve-fitting measured error values from approximating Euler spiral parallel curves to arcs, but was also inspired by applying the general error metric formula to the analytical equations for Euler spiral parallel curve, and dropping higher order terms.
 A more rigorous derivation, ideally with firm error bounds, remains as future work.
@@
 
-A typical pattern for converting from one curve type to another is \emph{adaptive subdivision.} (TODO: is there a good general reference for this?) An approximate curve is found in the parameter space of the target curve family.
+A typical pattern for converting from one curve type to another is \emph{adaptive subdivision.} An approximate curve is found in the parameter space of the target curve family.
 The error of the approximation is measured.
@@
 A key step in approximating one curve with another is evaluating the error of the approximation.
-A common approach (used in \citet{Nehab2020} among others) is to generate the approximate curve, the measure the distance, often by sampling at multiple points on the curve.
+A common approach (used in \citet{Nehab2020} among others) is to generate the approximate curve, then measure the distance, often by sampling at multiple points on the curve.
 All this is potentially slow, with the additional risk of underestimating the error due to undersampling.
@@
 
-The cubic Bézier approximating the Euler spiral is that with the distance of each control point from the endpoint is $d = \frac{2}{3(1 + \cos \theta)}$, where $\theta$ is the angle of the endpoint tangent relative to the chord.
+The cubic Bézier approximating the Euler spiral is one in which the distance of each control point from the endpoint is $d = \frac{2}{3(1 + \cos \theta)}$, where $\theta$ is the angle of the endpoint tangent relative to the chord.
 This is a generalization of the standard approximation of an arc.
@@
 We have validated this error metric in randomized testing.
-For values of $\theta$ between $0$ and $0.5$, and values of $d$ between $0$ and $0.6$, this estimate is always conservative, and it is also tight: over Bézier curves generated randomly with parameters drawn from a uniform distribution this range, the mean ratio of estimated to true error is 1.656.
+For values of $\theta$ between $0$ and $0.5$, and values of $d$ between $0$ and $0.6$, this estimate is always conservative, and it is also tight: over Bézier curves generated randomly with parameters drawn from a uniform distribution in this range, the mean ratio of estimated to true error is 1.656.
 
@@
 
-Fortunately, finding the cusp in the parallel curve of an Euler spiral parallel curve is a simple linear equation, and there is at most one cusp in any such segment.
+Fortunately, finding the cusp in the parallel curve of an Euler spiral is a simple linear equation, and there is at most one cusp in any such segment.
 Euler spirals are thus a good solution for determining cusps as a piecewise linear approximation to curvature.
@@
 Notably, our implementation is fully data-parallel on input path segments and does not require any computationally expensive curve evaluation on the CPU.
-We implemented our shader on general-purpose GPU compute primitives supported by all modern graphics APIs, particularly Metal, Vulkan, D3D12, and WebGPU[?
-citation?].
+We implemented our shader on general-purpose GPU compute primitives supported by all modern graphics APIs, particularly Metal, Vulkan, D3D12, and WebGPU[\citenum{WebGPU}].
 As such, the ideas presented in this section are portable to various GPU platforms.
@@
 A path tag consists of 8 bits that store \emph{stream offset increments}, such that the inclusive prefix sums of all increments for every element in the tag stream yield the stream offsets.
-We call this structure the \emph{tag monoid} and have implemented the computation as a general parallel prefix scan.[TODO: citation?]
+We call this structure the \emph{tag monoid} and have implemented the computation as a parallel prefix scan.
 

### 2024-04-29 paper.tex

@@
 
-The optimal flattening of a curve contains the minimum number of subdivisions required to meet a maximum error constraint, in which the error is the minimum possible with that number of subdivisions.
+The optimal flattening of a curve contains the minimum number of subdivisions required to satisfy a maximum error constraint, in which the error is the minimum possible with that number of subdivisions.
 
@@
 Numerical techniques are possible, but require multiple iterations to achieve sufficient accuracy, so are slower.
-In this subsection, we present a straightforward and accurate approximation, based on easily invertible piecewise functions.
+In this subsection, we present a straightforward and accurate approximation, constructed piecewise from easily invertible functions.
 If higher flattening quality is desired at the expense of slower computation, this approximation can be used to determine a good initial value for numeric techniques; two iterations of Newton solving are enough to refine this guess to within 32-bit floating point accuracy.

### 2024-04-29 paper.tex

@@
-\documentclass[sigconf, authordraft]{acmart}
+\documentclass[sigconf]{acmart}
 \usepackage[linesnumbered]{algorithm2e}
@@
 \title{GPU-friendly Stroke Expansion}
-\author{Raph Levien}
-\author{Arman Uguray}
-\affiliation{
-    \institution{Google}
-    \city{San Francisco}
-    \state{CA}
-    \country{USA}
-}
+\author{Anonymous Author 1}
+\author{Anonymous Author 2}
+%\affiliation{
+%    \institution{Google}
+%    \city{San Francisco}
+%    \state{CA}
+%    \country{USA}
+%}
+\begin{CCSXML}
+    <ccs2012>
+       <concept>
+           <concept_id>10010147.10010371.10010372</concept_id>
+           <concept_desc>Computing methodologies~Rendering</concept_desc>
+           <concept_significance>500</concept_significance>
+           </concept>
+       <concept>
+           <concept_id>10010147.10010371.10010396.10010399</concept_id>
+           <concept_desc>Computing methodologies~Parametric curve and surface models</concept_desc>
+           <concept_significance>500</concept_significance>
+           </concept>
+     </ccs2012>
+\end{CCSXML}
+
+\ccsdesc[500]{Computing methodologies~Rendering}
+\ccsdesc[500]{Computing methodologies~Parametric curve and surface models}
 

### 2024-04-29 paper.tex

@@
 
-\section{Conclusion}
+\section{Conclusions}
 
-TODO:
+Path stroking has received attention in recent years, with \citet{Nehab2020} and \citet{Kilgard2020} both having proposed a complete theory of correct stroking in vector graphics.
+While there are a number of implementations of path filling on the GPU, the goal of also implementing stroke expansion on the GPU had not yet been realized.
+Building upon the theory proposed by \citet{Nehab2020}, we implemented an approach to stroke expansion that is GPU-friendly, and achieves high performance on commodity hardware.
+
+We present the Euler spiral as an intermediate representation for a fast and precise approximation of both filled and stroked Bézier paths.
+Our method for lowering to both line and arc primitives avoids recursion and minimizes divergence, potentially serious problems for parallel evaluation on the GPU.
+We propose a novel error metric to directly estimate approximation error as opposed to the commonly employed measure-then-cut approaches which often consume the lion's share of computational expense.
+
+Our approach includes an efficient encoding scheme that alleviates the need for expensive CPU pre-computations and unlocks fully GPU-driven rendering of the entire vector graphics model.
 

### 2024-04-29 gpu_plots.py
fix plot

@@ -18,14 +18,14 @@ save_to_file = len(sys.argv) > 1 and sys.argv[1] == 'save' or \
 if do_test_scenes:
 	arcs = pd.DataFrame({
 		"long dash (arcs)": [8.11, 2.90, 0.559, 0.08245],
-		"mmark-35k (arcs)": [7.50, 0.943, 1.67, 0.13434],
-		"mmark-60k (arcs)": [13.17, 1.54, 2.38, 0.19562]
+		"mmark-70k (arcs)": [7.50, 0.943, 1.67, 0.13434],
+		"mmark-120k (arcs)": [13.17, 1.54, 2.38, 0.19562]
 		}, index=["Mali-G78", "M1 Max", "GTX 980Ti", "RTX 4090"]
 	)
 	lines = pd.DataFrame({
 		"long dash (lines)": [13.06, 3.11, 1.35, 0.31733],
-		"mmark-35k (lines)": [12.45, 1.73, 2.95, 0.22960],
-		"mmark-60k (lines)": [22.52, 2.87, 4.34, 0.38593],
+		"mmark-70k (lines)": [12.45, 1.73, 2.95, 0.22960],
+		"mmark-120k (lines)": [22.52, 2.87, 4.34, 0.38593],
 		}, index=["Mali-G78", "M1 Max", "GTX 980Ti", "RTX 4090"]
 	)
 else:

### 2024-04-30 paper.tex
Restore authors and remove copyright templating

@@
-\documentclass[sigconf]{acmart}
+\documentclass[sigconf, nonacm]{acmart}
 \usepackage[linesnumbered]{algorithm2e}
@@
 \title{GPU-friendly Stroke Expansion}
-\author{Anonymous Author 1}
-\author{Anonymous Author 2}
-%\affiliation{
-%    \institution{Google}
-%    \city{San Francisco}
-%    \state{CA}
-%    \country{USA}
-%}
+\author{Raph Levien}
+\author{Arman Uguray}
+\affiliation{
+    \institution{Google}
+    \city{San Francisco}
+    \state{CA}
+    \country{USA}
+}
 \begin{CCSXML}
@@
 \ccsdesc[500]{Computing methodologies~Parametric curve and surface models}
+
+% Re-enable these if/when published, also remove the `nonacm` version in \documentclass above.
+%\copyrightyear{2024}
+%\setcopyright{acmcopyright}
+%\conferenceinfo{Foo}{Month Day-Day, Year, Location}
+%\isbn{FOO}\acmPrice{BAR}
+%\doi{FOO}
+
+%\setcopyright{none}
 
@@
 The additional evolute segments and connecting lines are output twice, to make the winding numbers consistent and produce a watertight outline.
-All winding numbers are positive, so rendering with the nonzero winding number yields a correct final render.
+All winding numbers are positive, so rendering with the nonzero winding rule yields a correct final render.
 
@@
 Initial values of 1 and 0 represent the range (0, 1).
-Pushing the stack, subdividing the range in half, is representing by halving dt and doubling scaled\_t0, which leaves the start of the range invariant but its size halved.
+Pushing the stack, subdividing the range in half, is represented by halving dt and doubling scaled\_t0, which leaves the start of the range invariant but its size halved.
 After accepting an approximation, the next range is determined by incrementing scaled\_t0, then repeatedly popping the stack by doubling dt and halving scaled\_t0, as long as the latter is odd.
@@
 We found that Wang's formula[\citenum{Goldman2003}] works well in practice, as it is relatively inexpensive to evaluate, though it results in a higher memory estimate than required (see Table~\ref{table:bump-estimate-mem}).
-For parallel curves, Wang's formula does not guarantee an accurate upper bound on the required subdivisions, which we worked around by inflating the estimate for the source by a scale factor derived from the linewidth.
+For parallel curves, Wang's formula does not guarantee an accurate upper bound on the required subdivisions, which we worked around by inflating the estimate for the source segment by a scale factor derived from the linewidth.
 We observed that enabling estimation increases our CPU pre-processing time by 2-3x on the largest scenes but the impact is overall negligible when the scene size is modest.
@@
 Our method for lowering to both line and arc primitives avoids recursion and minimizes divergence, potentially serious problems for parallel evaluation on the GPU.
-We propose a novel error metric to directly estimate approximation error as opposed to the commonly employed measure-then-cut approaches which often consume the lion's share of computational expense.
+We propose a novel error metric to directly estimate approximation error as opposed to the commonly employed cut-then-measure approaches which often consume the lion's share of computational expense.
 

### 2024-06-12 paper.tex
Move final rendering after the measurement tables in markup
Re-anonymize authors
Submission of the revised version is also anonymous, so hide authors and affiliations. We'll revert this later.

Only including the .tex change, including the pdf would make it harder to revert.
Expand on evolutes in Section 8
- Added a new stroke renderings figure demonstrating strong and weak
  correctness and outlines.
- Added a new Algorithm block with the outline of the WGSL
  `flatten_euler` function.
- Added a couple paragraphs that summarizes the logic.

@@
 \setcitestyle{authoryear}
+\SetAlFnt{\small\sffamily}
+
 \begin{document}
 \title{GPU-friendly Stroke Expansion}
-\author{Raph Levien}
-\author{Arman Uguray}
-\affiliation{
-    \institution{Google}
-    \city{San Francisco}
-    \state{CA}
-    \country{USA}
-}
+\author{Anonymous Author 1}
+\author{Anonymous Author 2}
+%\affiliation{
+%    \institution{Google}
+%    \city{San Francisco}
+%    \state{CA}
+%    \country{USA}
+%}
 \begin{CCSXML}
@@
 \begin{figure*}
-    \includegraphics[scale=0.2]{tricky_strokes}
-    \caption{Stroked cubic Béziers with cusps and near-cusps, rendered using our algorithm. The individual curves were adapted from the Skia test corpus.}
-    \Description{Stroked cubic Béziers with cusps and near-cusps, rendered using our algorithm.}
-    \label{fig:tricky_strokes}
+    \includegraphics[scale=0.19]{stroke_stills}
+    \caption{Renderings of stroked cubic Béziers exhibiting strong and weak correctness.
+The outlines generated by our GPU stroking compute shader are visualized in green and blue.
+Segments with a winding direction that points \emph{up} are colored blue while those pointing \emph{down} are colored green.
+The curves were adapted to our GPU input format from the \emph{stills} dataset by \citet{Nehab2020} and the Skia test corpus.}
+    \Description{Selection of stroked cubic Béziers exhibiting strong and weak correctness}
+    \label{fig:stroke_stills}
 \end{figure*}
@@
   \item The implementation must be able to handle a large number of inputs at real-time rates. 
-  \item The stroke outlines must meet the weak correctness definition.
+  \item The stroke outlines must meet the strong correctness definition.
   \item The implementation must support the standard SVG stroke end cap (\emph{butt}, \emph{square}, \emph{round}) and join styles (\emph{bevel}, \emph{miter}, \emph{round}).
@@
   \eIf{$style$ is $fill$}{
-      output source curve\;
+      \tcp{Output source curve}
+      flatten\_cubic(segment, $0$)\;
   }{
@@
       }{
-          output both offset curves\;
+          \tcp{Output both offset curves}
+          flatten\_cubic(segment,  style.offset)\;
+          flatten\_cubic(segment, -style.offset)\;
+
           $neighbor \gets readScene(tid + 1)$\;
@@
 For a filled primitive, a thread only outputs the flattened approximation of the path segment.
-For a stroke, a thread outputs the polylines approximating multiple curves: a) the two parallel curves on both sides of the source segment, offset by the desired stroke \emph{halfwidth}, and b) the joins that connect the path segment to the next adjacent segment.
+For a stroke, a thread outputs the polylines approximating multiple curves: a) the two parallel curves on both sides of the source segment, offset by the desired stroke \emph{halfwidth}, and b) the joins that connect the path segment to the next adjacent segment, and c) evolute patches in regions of high curvature.
 We also store metadata in our input encoding that marks the position of a path segment within the containing subpath.
@@
 We require that the input segments within a subpath are stored in order which simplifies the access to the adjacent segment down to a buffer read at the next array offset.
-While we take care to output the \emph{outer} join segment (where the offset curves meet at a reflex angle) according to the join style, we output the inner join simply as a line segment.
-
-A property of this data-parallel structure is that the output is unordered.
+
+\newcommand{\nosemic}{\SetEndCharOfAlgoLine{\relax}}% Drop semi-colon ;
+\newcommand{\dosemic}{\SetEndCharOfAlgoLine{\string;}}% Reinstate semi-colon ;
+\newcommand{\pushline}{\Indp}% Indent
+\newcommand{\popline}{\Indm\dosemic}% Undent
+
+\begin{algorithm}
+\SetKwProg{Fn}{fn}{}{end}
+\SetKwBlock{Loop}{loop}{end}
+\Fn{flatten\_cubic(cubic: Cubic, offset: f32)}{
+    TOL $\gets$ Euler spiral fit tolerance\;
+    \tcp{'t0\_u' and 'dt' track the recursion stack}
+    t0\_u $\gets$ 0u\;
+    dt $\gets$ 1.0\;
+    last\_p $\gets$ cubic.p0\;
+    last\_q $\gets$ cubic.p1 - cubic.p0\;
+    \Loop{
+        t0 $\gets$ f32(t0\_u) * dt\;
+        \If{dt == 1.0}{break\;}
+        t1 $\gets$ t0 + dt\;
+        (next\_p, next\_q) $\gets$ eval\_cubic\_and\_deriv(cubic, t1)\;
+        \tcp{Estimate error of geometric}
+        \tcp{Hermite interpolation to Euler spiral}
+        \nosemic(th0, th1, chord\_len, err) $\gets$ fit\_cubic\_to\_es(\;
+        \pushline\dosemic last\_p, next\_p, last\_q, next\_q, dt)\;
+        \popline \eIf{err <= TOL}{
+            es $\gets$ euler\_segment(th0, th1, last\_p, next\_p)\;
+            \tcp{Find location of any cusp in outline.}
+            \tcp{If $0 \leq t < 1$, draw evolute at 't'.}
+            cusp0 $\gets$ es\_curvature(es, 0) * offset + chord\_len\;
+            cusp1 $\gets$ es\_curvature(es, 1) * offset + chord\_len\;
+            t $\gets$ cusp0 * cusp1 >= 0 ? 1 : cusp0 / (cusp0 - cusp1)\;
+            \If{cusp0 >= 0}{
+                evolute\_finalize()\;
+                es\_seg\_flatten\_offset(es, offset, vec2(0, t))\;
+            }
+            \If{cusp0 < 0 OR t < 1}{
+                evolute\_start(es, offset)\;
+                range $\gets$ cusp0 >= 0 ? vec2(t, 1) : vec2(0, t)\;
+                es\_seg\_flatten\_evolute(es, range)\;
+                es\_seg\_flatten\_offset\_reverse(es, offset, range)\;
+            }
+            \If{cusp0 < 0 AND t < 1}{
+                evolute\_finalize()\;
+                es\_seg\_flatten\_offset(es, offset, vec2(t, 1))\;
+            }
+            \tcp{Pop the stack}
+            t0\_u $\gets$ t0\_u + 1u\;
+            shift $\gets$ countTrailingZeros(t0\_u)\;
+            t0\_u $\gets$ t0\_u >> shift;
+            dt $\gets$ dt * f32(1u << shift)\;
+            last\_p $\gets$ next\_p\;
+            last\_q $\gets$ next\_q\;
+        }{
+            \tcp{Error is above the tolerance.}
+            \tcp{Push the stack; Continue subdivision}
+            t0\_u $\gets$ t0\_u * 2u\;
+            dt $\gets$ dt * 0.5\;
+        }
+    }
+}
+\caption{Outline of the iterative cubic Bézier flattening routine invoked in our compute shader. A working WGSL implementation is provided in the supplemental materials.}
+\label{alg:flatten-cubic}
+\end{algorithm}
+
+All input segments first get converted to a cubic Bézier by degree elevation.
+Our flattening routine (see Algorithm~\ref{alg:flatten-cubic}) iteratively subdivides the input cubic segment in half, and at each step, computes an error estimate for geometric Hermite interpolation to an Euler spiral segment.
+Once the error satisfies a desired tolerance threshold, the Euler spiral segment is directly flattened to a polyline (or, alternately, converted to a sequence of arc segments) by invoking a subroutine that is parameterized by the stroke halfwidth (see \texttt{es\_seg\_flatten\_offset} in Algoritm~\ref{alg:flatten-cubic}).
+For a regular fill, this parameter is set to $0$.
+For strokes, the sign of the parameter determines the position of the parallel curve relative to the source segment as well as the winding direction of the output lines.
+
+This procedure is sufficient to generate weakly correct stroke outlines.
+If the outline contains a cusp -- which can be determined by evaluating the curvature variation along the Euler spiral segment -- we output an evolute patch (as shown in Figure~\ref{fig:evolutes}) in order to achieve strong correctness.
+
+A property of the data-parallel structure is that the output is unordered.
 The cubic-Bézier-to-Euler-spiral conversion is sequential within a single thread, so it is possible to preserve ordering at the scope of an individual path segment, if desired.
@@
 \end{figure}
+
+Figures~\ref{fig:nehab-gpu-timings} and \ref{fig:test-scenes-gpu-timings} show the execution times for the two data sets.
+The entire \citet{Nehab2020} timings set adds up to less than 1 ms on all GPUs except the mobile unit.
+Our algorithm is ~14x slower on the Mali-G78 compared to the M1 Max but it is still capable of processing \emph{waves} in 3 ms.
+We observed that the performance of the kernel scales well with increasing GPU ability even at very large workloads, as demonstrated by the order-of-magnitude difference in run time between the mobile, laptop, and high-end desktop form factors we tested.
+We also confirmed that outputting arcs instead of lines can lead to a 2x decrease in execution time (particularly in scenes with high curve content, like \emph{waves} and \emph{mmark}) as predicted by the lower overall primitive count.
+
+\subsection{GPU: Input Processing} \label{subsection:encoding-results}
+
+Figure~\ref{fig:tag-monoid-gpu-timings} shows the run time of the tag monoid parallel prefix sums with increasing input size.
+The prefix sums scale better at large input sizes compared to our stroke expansion kernel.
+This can be attributed to a lack of expensive computations and high control-flow uniformity in the tag monoid kernel.
+The CPU encoding times (prior to the GPU prefix sums) for some of the large scenes is shown in Table~\ref{table:bump-estimate-cost}.
+Dashing on the GPU was out of scope, so the dash style applied to the \emph{long dash} scene had to be processed on the CPU and dashes were sequantially encoded as individual strokes.
+Including CPU-side dashing, the time to encode \emph{long dash} on the Apple M1 Max was approximately 25 ms on average, while the average time to encode the path after dashing was 4.14 ms.
+
+Our GPU implementation targets graphics APIs that do not support dynamic memory allocation in shader code (in contrast to CUDA, which provides the equivalent of \emph{malloc}).
+This makes it necessary to allocate the storage buffer in advance with enough memory to store the output, however, the precise memory requirement is unknown prior to running the shader.
+
+We augmented our encoding logic to compute a conservative estimate of the required output buffer size based on an upper bound on the number of line primitives per input path segment.
+We found that Wang's formula[\citenum{Goldman2003}] works well in practice, as it is relatively inexpensive to evaluate, though it results in a higher memory estimate than required (see Table~\ref{table:bump-estimate-mem}).
+For parallel curves, Wang's formula does not guarantee an accurate upper bound on the required subdivisions, which we worked around by inflating the estimate for the source segment by a scale factor derived from the linewidth.
+We observed that enabling estimation increases our CPU pre-processing time by 2-3x on the largest scenes but the impact is overall negligible when the scene size is modest.
+
+\begin{figure}
+    \includegraphics[scale=0.66]{tag_monoid_gpu_timings}
+    \caption{Mean GPU execution time of the tag monoid parallel prefix sums with increasing number of input path segments, measured on Apple M1 Max and NVIDIA RTX 4090.
+The figure also shows the execution time of the stroke expansion and flattening kernel.
+The y-axis is shown in logarithmic scale to highlight the similar scaling trend on both GPUs despite the order-of-magnitude difference in performance.}
+    \Description{Mean GPU execution time of the tag monoid kernel}
+    \label{fig:tag-monoid-gpu-timings}
+\end{figure}
+
+\begin{table}
+    \caption{Comparison of CPU encoding time with and without size estimation on the largest scenes.
+All measurements were computed on an Apple M1 Max.
+The additional cost is negligible when the scene size is modest but tangible with increased complexity.}
+    \begin{tabular}{|l|c|c|}
+    \hline
+        \multicolumn{1}{|c|}{\textbf{Test}} & \textbf{Encoding} & \textbf{Encoding + Estimation} \\
+    \hline
+        roads.svg  & 398.56 $\mu$s & 618.24 $\mu$s \\
+        waves.svg  & 197.53 $\mu$s & 418.96 $\mu$s \\
+        mmark-70k  & 2.69 ms       & 6.89 ms       \\
+        mmark-120k & 4.28 ms       & 11.59 ms      \\
+    \hline
+    \end{tabular}
+    \label{table:bump-estimate-cost}
+\end{table}
+
+\begin{table}
+    \caption{The estimated and actual \emph{line soup} buffer sizes for scenes rendered at 2088x1600 resolution.
+Our estimation uses Wang's formula combined with a scaling heuristic based on the linewidth.
+In the worst case, our buffer size estimator overshoots by a factor ~4.}
+    \begin{tabular}{|l|c|c|}
+    \hline
+        \multicolumn{1}{|c|}{\textbf{Test}} & Actual Size & Estimated Size \\
+    \hline
+        roads.svg  & 958.78 KB & 2.45 MB    \\
+        waves.svg  & 9.37 MB   & 10.44 MB   \\
+        mmark-70k  & 23.81 MB  & 100.79 MB  \\
+        mmark-120k & 40.86 MB  & 173.56 MB  \\
+    \hline
+    \end{tabular}
+    \label{table:bump-estimate-mem}
+\end{table}
 
@@
 \end{figure*}
-
-Figures~\ref{fig:nehab-gpu-timings} and \ref{fig:test-scenes-gpu-timings} show the execution times for the two data sets.
-The entire \citet{Nehab2020} timings set adds up to less than 1 ms on all GPUs except the mobile unit.
-Our algorithm is ~14x slower on the Mali-G78 compared to the M1 Max but it is still capable of processing \emph{waves} in 3 ms.
-We observed that the performance of the kernel scales well with increasing GPU ability even at very large workloads, as demonstrated by the order-of-magnitude difference in run time between the mobile, laptop, and high-end desktop form factors we tested.
-We also confirmed that outputting arcs instead of lines can lead to a 2x decrease in execution time (particularly in scenes with high curve content, like \emph{waves} and \emph{mmark}) as predicted by the lower overall primitive count.
-
-\subsection{GPU: Input Processing} \label{subsection:encoding-results}
-
-Figure~\ref{fig:tag-monoid-gpu-timings} shows the run time of the tag monoid parallel prefix sums with increasing input size.
-The prefix sums scale better at large input sizes compared to our stroke expansion kernel.
-This can be attributed to a lack of expensive computations and high control-flow uniformity in the tag monoid kernel.
-The CPU encoding times (prior to the GPU prefix sums) for some of the large scenes is shown in Table~\ref{table:bump-estimate-cost}.
-Dashing on the GPU was out of scope, so the dash style applied to the \emph{long dash} scene had to be processed on the CPU and dashes were sequantially encoded as individual strokes.
-Including CPU-side dashing, the time to encode \emph{long dash} on the Apple M1 Max was approximately 25 ms on average, while the average time to encode the path after dashing was 4.14 ms.
-
-Our GPU implementation targets graphics APIs that do not support dynamic memory allocation in shader code (in contrast to CUDA, which provides the equivalent of \emph{malloc}).
-This makes it necessary to allocate the storage buffer in advance with enough memory to store the output, however, the precise memory requirement is unknown prior to running the shader.
-
-We augmented our encoding logic to compute a conservative estimate of the required output buffer size based on an upper bound on the number of line primitives per input path segment.
-We found that Wang's formula[\citenum{Goldman2003}] works well in practice, as it is relatively inexpensive to evaluate, though it results in a higher memory estimate than required (see Table~\ref{table:bump-estimate-mem}).
-For parallel curves, Wang's formula does not guarantee an accurate upper bound on the required subdivisions, which we worked around by inflating the estimate for the source segment by a scale factor derived from the linewidth.
-We observed that enabling estimation increases our CPU pre-processing time by 2-3x on the largest scenes but the impact is overall negligible when the scene size is modest.
-
-\begin{figure}
-    \includegraphics[scale=0.66]{tag_monoid_gpu_timings}
-    \caption{Mean GPU execution time of the tag monoid parallel prefix sums with increasing number of input path segments, measured on Apple M1 Max and NVIDIA RTX 4090.
-The figure also shows the execution time of the stroke expansion and flattening kernel.
-The y-axis is shown in logarithmic scale to highlight the similar scaling trend on both GPUs despite the order-of-magnitude difference in performance.}
-    \Description{Mean GPU execution time of the tag monoid kernel}
-    \label{fig:tag-monoid-gpu-timings}
-\end{figure}
-
-\begin{table}
-    \caption{Comparison of CPU encoding time with and without size estimation on the largest scenes.
-All measurements were computed on an Apple M1 Max.
-The additional cost is negligible when the scene size is modest but tangible with increased complexity.}
-    \begin{tabular}{|l|c|c|}
-    \hline
-        \multicolumn{1}{|c|}{\textbf{Test}} & \textbf{Encoding} & \textbf{Encoding + Estimation} \\
-    \hline
-        roads.svg  & 398.56 $\mu$s & 618.24 $\mu$s \\
-        waves.svg  & 197.53 $\mu$s & 418.96 $\mu$s \\
-        mmark-70k  & 2.69 ms       & 6.89 ms       \\
-        mmark-120k & 4.28 ms       & 11.59 ms      \\
-    \hline
-    \end{tabular}
-    \label{table:bump-estimate-cost}
-\end{table}
-
-\begin{table}
-    \caption{The estimated and actual \emph{line soup} buffer sizes for scenes rendered at 2088x1600 resolution.
-Our estimation uses Wang's formula combined with a scaling heuristic based on the linewidth.
-In the worst case, our buffer size estimator overshoots by a factor ~4.}
-    \begin{tabular}{|l|c|c|}
-    \hline
-        \multicolumn{1}{|c|}{\textbf{Test}} & Actual Size & Estimated Size \\
-    \hline
-        roads.svg  & 958.78 KB & 2.45 MB    \\
-        waves.svg  & 9.37 MB   & 10.44 MB   \\
-        mmark-70k  & 23.81 MB  & 100.79 MB  \\
-        mmark-120k & 40.86 MB  & 173.56 MB  \\
-    \hline
-    \end{tabular}
-    \label{table:bump-estimate-mem}
-\end{table}
 

### 2024-06-12 paper.tex
Address feedback

@@
 Segments with a winding direction that points \emph{up} are colored blue while those pointing \emph{down} are colored green.
-The curves were adapted to our GPU input format from the \emph{stills} dataset by \citet{Nehab2020} and the Skia test corpus.}
+The curves were adapted to from the \emph{stills} dataset by \citet{Nehab2020} and the Skia test corpus.}
     \Description{Selection of stroked cubic Béziers exhibiting strong and weak correctness}
@@
 All input segments first get converted to a cubic Bézier by degree elevation.
-Our flattening routine (see Algorithm~\ref{alg:flatten-cubic}) iteratively subdivides the input cubic segment in half, and at each step, computes an error estimate for geometric Hermite interpolation to an Euler spiral segment.
-Once the error satisfies a desired tolerance threshold, the Euler spiral segment is directly flattened to a polyline (or, alternately, converted to a sequence of arc segments) by invoking a subroutine that is parameterized by the stroke halfwidth (see \texttt{es\_seg\_flatten\_offset} in Algoritm~\ref{alg:flatten-cubic}).
+Our flattening routine (see Algorithm~\ref{alg:flatten-cubic}) starts by computing the error estimate for geometric Hermite interpolation from the cubic Bézier to an Euler spiral segment.
+If the error exceeds the desired tolerance threshold, the cubic segment is iteratively subdivided in half and a new error estimate is computed.
+Once the error satisfies the tolerance, the Euler spiral segment is directly flattened to a polyline (or, alternately, converted to a sequence of arc segments) by invoking a subroutine that is parameterized by the stroke halfwidth (see \texttt{es\_seg\_flatten\_offset} in Algoritm~\ref{alg:flatten-cubic}).
 For a regular fill, this parameter is set to $0$.
@@
 If the outline contains a cusp -- which can be determined by evaluating the curvature variation along the Euler spiral segment -- we output an evolute patch (as shown in Figure~\ref{fig:evolutes}) in order to achieve strong correctness.
+A selection of renderings produced by our algorithm that demonstrate both strong and weak correctness are shown in Figure~\ref{fig:stroke_stills}.
 

### 2024-06-12 paper.tex
Rewrite of introduction
The introduction is rewritten, both to address reviewer feedback and also to strenghten our statement of strong correctness.

Also redo the transition from circular arc to the sqrt-kappa error metric, making the statement stronger, and also including the monotonic curvature condition.

More is to be done, basically making things flow more smoothly in subsequent sections, but I wanted to get this part up for review, as other revision will depend on it.

@@
 
-Rendering of vector graphics documents requires handling both filled and stroked primitives.
-There is substantial literature on GPU rendering of filled paths, but many fewer published techniques for strokes.
-It is a more challenging problem, especially for parallel computation, because path segments cannot be processed independently of each other; the \emph{joins} between adjacent path segments depend on context.
-In addition, the path topology affects the rendered result, in particular whether subpaths are open or closed.
-In the former case, the endpoints are rendered with \emph{caps}.
+Stroke rendering is an essential part of the standard vector graphics imaging model, and is thus widely used.
+The standard representation of paths is a sequence of cubic Bézier segments.
+While there are a number of published techniques for GPU rendering of filled paths, stroke rendering is more challenging to parallelize, largely because path segments cannot be processed independently of each other; rendering of the \emph{joins} between path segments depends on both adjoining segments, and the endpoints of paths (when not closed) are rendered with \emph{caps}.
 Joins and caps can have multiple styles, including different styles within the same document.
 
-Stroke expansion is the process of generating an outline that, when filled, produces the \emph{stroke} of a given input path.
-Informally, a line of the same width as the stroke is swept along the path, normal to it, and also decorated with joins and caps.
-Though stroke rendering is referenced in many 2D graphics standards, it has lacked a precise formal definition until recent work, primarily \citet{Nehab2020} and \citet{Kilgard2020}.
-
-The \citet{Nehab2020} paper gives a comprehensive survey of techniques for stroke expansion, and an algorithm which, for a number of reasons, is only suitable for implementation on CPU (though adapting it to GPU is listed as promising future work).
-It classifies techniques into \emph{local,} where each path segment generates closed geometry (which may be triangles or other primitives), and \emph{global,} where the overall result is a closed outline of the stroked path, but the partial result from each segment is in general open.
-Our technique is considered global in this scheme, yet allows independent processing of each path segment.
-
-A number of factors contribute to an algorithm being ``GPU-friendly.'' In addition to simply being able to process the input segments in parallel, such an algorithm also avoids divergent control flow (avoiding the need for explicit subdivision at special events such as inflection points and cusps) and uses robust numerical techniques not subject to particular problems when evaluated using 32-bit floating point numbers.
-
+A typical vector graphics document may have thousands of stroked paths.
+A map or CAD drawing may have tens of thousands of paths, not unusually a million path segments.
+At this scale, the computational cost of stroke rendering on CPU may then result in loss of interactive rendering performance.
+Thus, we desire to improve performance by offloading stroke rendering to the GPU.
+To realize performance gains, the algorithm needs to be GPU-friendly, meaning among other things that it exploits the available parallelism, is work efficient, and avoids unnecessarily complex (and potentially divergent) control flow.
+In addition, it must remain numerically robust even when computed with 32 bit floating point numbers.
+
+There are a number of strategies for rendering strokes.
+Among them are \emph{local} techniques that break down the stroke into individual closed primitives, distance field techniques (or similar techniques based on point inclusion).
+In this paper, we focus on \emph{stroke expansion,} the generation of an outline that, when filled, represents the rendered stroke.
+Implementing stroke rendering using stroke expansion enables a unified approach to path rendering downstream, rendering both filled and stroked paths.
+Among other benefits, such an approach avoids a large number of draw calls when stroked and filled paths are finely interleaved, as is often the case.
+
+Correctness is another challenging aspect of stroke rendering.
 We propose that the correctness of stroke outlines be divided into \emph{weak correctness} and \emph{strong correctness.} We define strong correctness as the computation of the outline of a line swept along the segment, maintaining normal orientation, combined with stroke caps and joins.
-Weak correctness, by contrast, only requires the parallel curves of stroke segments, combined with caps and the outer contours of joins.
-The two notions are equivalent for sufficiently well-behaved input, in particular when the curvature measured at endpoints of path segments does not exceed the reciprocal of the half linewidth.
-As described in detail in \citet{Nehab2020}, very few existing implementations actually implement the strong version, so document authors have become accustomed to not depending on behavior at endpoints.
-Standards for graphics formats, such as SVG[\citenum{Svg2}], also provide an ``out,'' enabling the weaker behavior.
-While our current implementation emphasizes weak correctness, we believe it can be extended to the stronger sense, by implementing evolutes and inner join contours.
-
-A great number of rendering techniques for strokes have been proposed.
-Some are \emph{local,} in that they break the stroke up into smaller closed pieces, the union of which forms an approximation to the stroke.
-Included in this category is polar stroking[\citenum{Kilgard2020}], which proposes a GPU-friendly subdivision scheme based on an angle step error.
-Others are based on stroke expansion, generating an explicit outline, which is then filled to produce the stroke rendering.
-Such techniques are further divided into \emph{flattening} approaches, where the outline is approximated by a polyline, and \emph{curve-based} approaches, exemplified by the approach in \citet{Nehab2020}.
-In both cases, the outline contains approximations of parallel curves of the input curve segments, and, as proposed in \citet{Nehab2020}, their evolutes when targeting strong correctness.
-For a more detailed survey of existing stroke rendering techniques, the reader is directed to Section 4 of \citet{Nehab2020}.
-
-The subproblem of approximating parallel curves has also spawned extensive literature, none of which is entirely satisfying, especially when it comes to algorithms that can be efficiently evaluated on GPU.
+Weak correctness, by contrast, only requires the parallel curves of path segments, combined with caps and the outer contours of joins.
+Both \citet{Nehab2020} and \citet{Kilgard2020} provide useful definitions of strongly correct stroke rendering, and Nehab in particular describes how to achieve it in stroke expansion.
+Briefly, when the path curvature exceeds the reciprocal of the stroke half-width, then addition \emph{evolute} segments need to be added, in addition to the parallel curves required by weak correctness, and also the inner contours of joins are required in general.
+The technique described in this paper is intended for strongly correct stroke rendering; it produces outlines that render the same shape as \citet{Nehab2020}.
+
+Correct stroke rendering also relies on accurate approximation of the parallel curves of the input paths.
+Determining the parallel curve of a cubic Bézier is not analytically tractable, as it is a tenth order algebraic formula\citenum{Farouki1990}.
+Thus, all practical stroke rendering techniques employ an approximation, with error tolerance as a tunable parameter, typically representing the maximum allowable Fréchet distance between the true curve and its approximation.
+A typical value is 0.25 device pixel, as that is close to the threshold of visibility.
+A correct algorithm will produce an approximate curve within the error tolerance, but an efficient algorithm will not subdivide significantly more than is necessary to meet the error bound.
+Some specifications for vector graphics, including PDF\citenum{PDF2008}, specify the error tolerance explicitly.
+
+The subproblem of approximating parallel curves has spawned extensive literature, none of which is entirely satisfying, especially when it comes to algorithms that can be efficiently evaluated on GPU.
 An example of a reasonably good algorithm that computes flattened parallel curves is \citet{Yzerman2020}, as used in the Blend2D rendering library.
@@
 A variant is also used in Skia[\citenum{Skia}]; instead of lowering the cubic Bézier to a quadratic approximation and then computing the parallel curve, Skia approximates the parallel curve directly with a quadratic Bézier, then measures the error to determine whether further subdivision is necessary.
+
+Previous work on stroke rendering does not fully solve the problem of fast and correct stroke rendering.
+Most existing implementations have correctness problems, in particular failing to draw the evolutes and inner joins as required for strong correctness; \citet{Nehab2020} provides a detailed survey.
+In addition, most implementations are not suitable for running on GPU, requiring sequential execution to produce watertight outlines.
+One GPU implementation is polar stroking (\cite{Kilgard2020}), a local technique, but among other limitations, it does not bound the Fréchet distance error (rather, it bounds the angle error).
+
+Our central approach bounding the error tolerance is \emph{error metrics,} which predict the Fréchet distance error.
+The error metrics presented in this paper are both straightforward to compute, and also \emph{invertible}, which avoids excessive subdivision and also increases the amount of available parallelism to exploit.
+Previous approaches to error measurement generally follow a ``cut then measure'' approach, where the approximate curve is generated, then error measured by sampling.
+Such sampling is often time consuming, involving iteration, and even then risks underestimating the error due to inadequate sampling.
 
@@
 
-We are of course concerned with the flattening of more general curves (ultimately the parallel curve of a cubic Bézier), not simply circular arcs.
-It is tempting to sample the curvature and plug it into the formula above, but this can both dramatically undershoot the error (predicting that no subdivision is needed at an inflection point) and overshoot it (requiring infinite subdivision at a cusp; and cusps).
-Apparently, some kind of average curvature is needed.
-
-We propose the following error metric to estimate the maximum distance between an arbitrary curve of length $\hat{s}$ and its chord.
+We are of course concerned with the flattening of more general curves (ultimately the parallel curve of a cubic Bézier), not simply circular arcs, so the curvature is not constant.
+An obvious approach to try would be using the maximum error.
+This would be conservative with respect to error tolerance, but also generate too much subdivision when the curve has a small region of high curvature.
+In the limit, the curve can contain a cusp of infinite curvature, which would entail infinite subdivision.
+Similarly, sampling the curvature at a single point can either underestimate or overestimate the error.
+The ideal approach would be a form of average curvature that is tractable to compute, and accurately predicts the flattening error.
+To that end, we propose the following error metric to estimate the maximum distance between an arbitrary curve of length $\hat{s}$ and its chord.
 
@@
 This formula is the same as the approximate error metric for circular arcs, except that instead of a constant curvature value, a norm-like average with an exponent of 1/2 is used (it is not considered a true norm because the triangle inequality does not hold).
-We chose this particular formulation because it tends to produce invertible error metrics, and because it works well in practice.
+This particular formulation has two strong advantages.
+First, it has been empirically validated to accurately predict the distance between chord and curve for a variety of curves.
+In particular, we believe that when curvature is monotonic, it is a tight bound both above and below (and later we will see that in our algorithm it is always evaluated on curve segments with such monotonic curvature).
+Second, this formula lends itself to invertible error metrics.
 

### 2024-06-13 paper.tex
Address reviewer #1's feedback on Section 8
This largely addresses the first reviewer's feedback on Section 8,
particularly the following asks:

1. Section 8.1 expands on the broader renderer architecture that the
   flatten stage was integrated into, clarifying the sorting
   requirements on line soup.

2. Added a new figure to illustrate the pipeline with broad strokes,
   breaking it into "input processing", "flattening", and
   "rasterization" blocks.

--- 2024-06-12 paper.tex
@@
   \item The implementation must support the standard SVG stroke end cap (\emph{butt}, \emph{square}, \emph{round}) and join styles (\emph{bevel}, \emph{miter}, \emph{round}).
-  \item The CPU must do no work beyond the basic preparation of the input path data for GPU consumption.
+  \item The CPU must do no work beyond the basic preparation of the input path data for GPU consumption, in order to maximize the exploited parallelism.
 \end{enumerate}
@@
 The rest of this section describes the design of our GPU pipeline and input encoding scheme.
-
-% TODO: - Watertightness
 
@@
 
-A property of the data-parallel structure is that the output is unordered.
+A property of the data-parallel structure is that the output of the compute stage is unordered.
 The cubic-Bézier-to-Euler-spiral conversion is sequential within a single thread, so it is possible to preserve ordering at the scope of an individual path segment, if desired.
@@
 We refer to this output data structure as \emph{line soup}.
-We integrated the compute shader as a stage in a larger compute shader pipeline (discussed in Section \ref{subsection:gpu-results}) that spatially sorts the line soup in subsequent binning and tiling stages prior to rasterization.
+An application may want to sort the line soup, depending on the requirements of the chosen rendering algorithm.
+For instance, our own renderer employs tile-based scanline rasterization in a compute shader, with a pipeline architecture that resembles \emph{cudaraster} (\citet{Laine2011}) and \emph{MPVG} (\citet{Ganacim2014}).
+In this design, the line soup elements are spatially sorted by subsequent compute stages into a grid of 16x16 pixel tiles.
+The final compute stage of our renderer (called \emph{fine rasterizer}) operates at tile granularity and computes pixel coverage for filled polygons outlined by the line soup (the details of which are beyond the scope of this paper).
+
+\begin{figure*}
+    \includegraphics[scale=0.275]{pipeline}
+    \caption{We integrated our curve flattening shader into our own renderer implemented entirely in compute shaders.
+The flattening stage (described in Section~\ref{section:gpu-impl}) converts stroked paths to filled outlines and flattens them into an unordered soup of line segments.
+Our renderer spatially sorts the line soup into tiles before computing pixel coverage using a highly parallel scanline rasterizer.}
+    \Description{Our GPU pipeline architecture in broad strokes}
+    \label{fig:pipeline}
+\end{figure*}
 

### 2024-06-13 paper.tex
Fix citation formatting and typos
This commit applies nearly almost all of the typos and suggested wording changes of reviewer 1.

--- 2024-06-12 paper.tex
@@
 Correct stroke rendering also relies on accurate approximation of the parallel curves of the input paths.
-Determining the parallel curve of a cubic Bézier is not analytically tractable, as it is a tenth order algebraic formula\citenum{Farouki1990}.
+Determining the parallel curve of a cubic Bézier is not analytically tractable, as it is a tenth order algebraic formula[\citenum{Farouki1990}].
 Thus, all practical stroke rendering techniques employ an approximation, with error tolerance as a tunable parameter, typically representing the maximum allowable Fréchet distance between the true curve and its approximation.
@@
 A correct algorithm will produce an approximate curve within the error tolerance, but an efficient algorithm will not subdivide significantly more than is necessary to meet the error bound.
-Some specifications for vector graphics, including PDF\citenum{PDF2008}, specify the error tolerance explicitly.
+Some specifications for vector graphics, including PDF[\citenum{PDF2008}], specify the error tolerance explicitly.
 
@@
 In addition, most implementations are not suitable for running on GPU, requiring sequential execution to produce watertight outlines.
-One GPU implementation is polar stroking (\cite{Kilgard2020}), a local technique, but among other limitations, it does not bound the Fréchet distance error (rather, it bounds the angle error).
+One GPU implementation is polar stroking (\citet{Kilgard2020}), a local technique, but among other limitations, it does not bound the Fréchet distance error (rather, it bounds the angle error).
 
@@
 These segments must be within an error tolerance of the source curve, and ideally close to a minimal number of them.
-We consider a number of source to target pairs, most importantly cubic Béziers to Euler spirals, and Euler spiral parallel curves to either lines or arcs.
+We consider a number of source-to-target pairs, most importantly cubic Béziers to Euler spirals, and Euler spiral parallel curves to either lines or arcs.
 
@@
 This metric is conservative but works well in practice; among other applications, it is used in Skia for path flattening.
-One limitation of Wang's method is that it only applies to the flattening of filled outlines; when applied naively to generation of parallel curves, it can undershoot substantially, especially near cusps.
+The chief limitation of Wang's method is that it computes a subdivision bound for a Bézier curve without accounting for the displacement of the parallel curve due to stroking.
+When applied naively to generation of parallel curves, it can undershoot substantially, especially near cusps.
 
@@
     \includegraphics[scale=0.3]{es_flatten}
-    \caption{Flattening of an Euler spiral using points spaced by power-law}
-    \Description{An Euler spiral along with its flattened representation. The flattened curve has an error}
+    \caption{The flattened approximation to an Euler spiral using line segments is achieved by choosing subdivision points using a simple power law formula with an exponent of 2/3.}
+    \Description{An Euler spiral along with its flattened representation. The flattened curve has a consistent error, and the subdivision points are spaced according to a}
     \label{fig:es_flatten}
@@
 
-The parallel curve of the Euler spiral (also known as ``clothoid'') was characterized by \citet{Wieleitner1907} well over a hundred years ago[\citenum{Wieleitner1907}], and has a straightforward closed-form Cesàro representation, curvature as a function of arc length.
+The parallel curve of the Euler spiral (also known as ``clothoid'') was characterized by \citet{Wieleitner1907} well over a hundred years ago, and has a straightforward closed-form Cesàro representation, curvature as a function of arc length, of the form $\kappa(s) = c_0{s - s_0}^{-1/2} + c_1$.
 
@@
 When a curve is monotonic, error is also monotonic with respect to both subdivision points.
-Thus, any standard root finding technique can find a subdivision with a given error (bisection is the best known, but the ITP method \citet{Oliveira2020} is better).
+Thus, any standard root finding technique can find a subdivision with a given error (bisection is the most popular, but the ITP method[\citenum{Oliveira2020}] is better).
 A similar outer loop with a root finding technique can find the error which minimizes the global error for the entire flattening, which, due to monotonicity, is that for which the error of the final segment equals that of the previous segments.
@@
 
-The subdivision density for the parallel curve of an Euler spiral, normalized so that its inflection point is at -1 and the cusp of the parallel curve is at 1, is simply $1 - \sqrt{|1-s^2|}$.
+The subdivision density for the parallel curve of an Euler spiral, normalized so that its inflection point is at $-1$ and the cusp of the parallel curve is at 1, is simply $1 - \sqrt{|1-s^2|}$.
 This function is plotted in Figure~\ref{fig:subdiv_density}.
@@
 For an Euler spiral segment, $\kappa'(s)$ is constant and thus this error metric becomes nearly trivial.
-With n subdivisions, the estimated distance is simply $\frac{s^3\kappa'}{120n^3}$.
+With $n$ subdivisions, the estimated distance is simply $\frac{s^3\kappa'}{120n^3}$.
 Solving for $n$, we get $n = s\sqrt[3]{\frac{|\kappa'|}{120d}}$ subdivisions, and those are divided evenly by arc length, as the subdivision density is constant across the curve, just as is the case for flattening arcs to lines.
@@
 
-One consequence of this formula is that, since $h$ appears under absolute value, the same arc approximation can be used for both sides of a stroke.
+One consequence of this formula is that, since the error is in terms of the absolute value of $h$, independent of sign, the same arc approximation can be used for both sides of a stroke.
 
@@
 
-Many techniques for stroke expansion described in the literature apply some lowering of cubic Bézier curves to a simpler curve type, more tractable for evaluating parallel curve.
+Many techniques for stroke expansion described in the literature apply some lowering of cubic Bézier curves to a simpler curve type that is more tractable for evaluating parallel curve.
 Computing parallel curves directly on cubic Bézier curve segments is not very tractable.
@@
 
-The entire recursion stack for adaptive subdivision can be represented in two words: dt and scaled\_t0.
-The range is dt * scaled\_t0 ..
-dt * (scaled\_t0 + 1).
-Initial values of 1 and 0 represent the range (0, 1).
-Pushing the stack, subdividing the range in half, is represented by halving dt and doubling scaled\_t0, which leaves the start of the range invariant but its size halved.
-After accepting an approximation, the next range is determined by incrementing scaled\_t0, then repeatedly popping the stack by doubling dt and halving scaled\_t0, as long as the latter is odd.
+The entire recursion stack for adaptive subdivision can be represented in two words: $dt$ and $\mbox{scaled}\_t0$.
+The range is $dt \cdot \mbox{scaled}\_t0 ..
+dt \cdot (\mbox{scaled}\_t0 + 1)$.
+Initial values of $1$ and $0$ represent the range $(0, 1)$.
+Pushing the stack, subdividing the range in half, is represented by halving $dt$ and doubling $\mbox{scaled}\_t0$, which leaves the start of the range invariant but its size halved.
+After accepting an approximation, the next range is determined by incrementing $\mbox{scaled}\_t0$, then repeatedly popping the stack by doubling $dt$ and halving $\mbox{scaled}\_t0$, as long as the latter is odd.
 This can in fact be achieved without iteration by using the ``count trailing zeros'' intrinsic.
@@
 
-Our approach is to leverage the conversion to Euler spiral segments, which have finite curvature.
+Our approach leverages the conversion to Euler spiral segments, which have finite curvature.
 The geometric Hermite interpolation depends on accurate tangents at the endpoints, which in turn are derived from derivatives.
@@
 The \emph{lineto} and \emph{curveto} instructions denote an individual \emph{path segment} consisting of either a straight line or a cubic Bézier.
-A given subsequence of path segments can be bounded by a \emph{moveto} and/or \emph{closepath} instruction which mark the beginning and end of a \emph{subpath}.
+A given subsequence of path segments can be bounded by a \emph{moveto} and/or \emph{closepath} instruction that marks the beginning and end of a \emph{subpath}.
 A subpath must form a closed contour if painted as a fill.
@@
 
-We authored two additional test scenes in order stress the GPU with a higher workload.
+We authored two additional test scenes in order to stress the GPU with a higher workload.
 The first is a very large stroked path with ~500,000 individually styled dashed segments which we adapted from Skia's open-source test corpus.
@@
 The contents of each scene were uniformly scaled up (by encoding a transform) to fit the size of the texture, in order to increase the required number of output primitives.
-As with the CPU timings, we used an error tolerance of 0.25 in all cases.
+As with the CPU timings, we used an error tolerance of 0.25 pixels in all cases.
 We recorded the execution times across 3,000 GPU submissions for each scene.
@@
 The hatched bars show the timings for the version that outputs line primitives while the solid bars show that for arcs.
-Mali-G78 (the lowest end GPU we tested) converts \emph{waves} (the largest scene in the set) to lines in ~3ms.
+Mali-G78 (the lowest-end GPU we tested) converts \emph{waves} (the largest scene in the set) to lines in \~3~ms.
 The compute shader executes at least an order of magnitude faster on the other GPUs.}
@@
 The CPU encoding times (prior to the GPU prefix sums) for some of the large scenes is shown in Table~\ref{table:bump-estimate-cost}.
-Dashing on the GPU was out of scope, so the dash style applied to the \emph{long dash} scene had to be processed on the CPU and dashes were sequantially encoded as individual strokes.
+Our implementation still applies dashing on the CPU.
+It is conceptually fairly straightforwad to implement on GPU; the core algorithm is prefix sum applied to segment length, and the Euler spiral representation is particularly well suited.
+However, there are complex implementatoin tradeoffs, so the dash style applied to the \emph{long dash} scene had to be processed on the CPU and dashes were sequentially encoded as individual strokes.
 Including CPU-side dashing, the time to encode \emph{long dash} on the Apple M1 Max was approximately 25 ms on average, while the average time to encode the path after dashing was 4.14 ms.
@@
 For parallel curves, Wang's formula does not guarantee an accurate upper bound on the required subdivisions, which we worked around by inflating the estimate for the source segment by a scale factor derived from the linewidth.
-We observed that enabling estimation increases our CPU pre-processing time by 2-3x on the largest scenes but the impact is overall negligible when the scene size is modest.
+We observed that enabling estimation increases our CPU pre-processing time by 2--3x on the largest scenes but the overall impact is negligible when the scene size is modest.
 
@@
 Our estimation uses Wang's formula combined with a scaling heuristic based on the linewidth.
-In the worst case, our buffer size estimator overshoots by a factor ~4.}
+In the worst case, our buffer size estimator overshoots by a factor of \~4.}
     \begin{tabular}{|l|c|c|}
@@
 
-The Euler spiral is represented by $\kappa(s) = k_0 + k_1s$, where $s$ ranges from 0 to 1, i.e.
-the arc length is unit normalized.
+The Euler spiral is represented by $\kappa(s) = k_0 + k_1s$, where $s$ ranges from 0 to 1, i.e., the arc length is unit normalized.
 This segment is scaled, rotated, and translated into place so that its endpoints match the desired locations.

### 2024-06-13 paper.tex
Tweaks to flatten algorithm
Small changes to the pseudocode, and fixing up the text to be more consistent with it.

@@
 We propose that the correctness of stroke outlines be divided into \emph{weak correctness} and \emph{strong correctness.} We define strong correctness as the computation of the outline of a line swept along the segment, maintaining normal orientation, combined with stroke caps and joins.
-Weak correctness, by contrast, only requires the parallel curves of path segments, combined with caps and the outer contours of joins.
+Weak correctness, by contrast, only requires the parallel curves (also known as ``offset curves'') of path segments, combined with caps and the outer contours of joins.
 Both \citet{Nehab2020} and \citet{Kilgard2020} provide useful definitions of strongly correct stroke rendering, and Nehab in particular describes how to achieve it in stroke expansion.
@@
 
-\subsection{Optimal flattening}
-
-The optimal flattening of a curve contains the minimum number of subdivisions required to satisfy a maximum error constraint, in which the error is the minimum possible with that number of subdivisions.
-
-For a convex curve, the optimal flattening is tractable to compute, though not necessarily very fast.
+%\subsection{Optimal flattening}
+%
+%The optimal flattening of a curve contains the minimum number of subdivisions required to satisfy a maximum error constraint, in which the error is the minimum possible with that number of subdivisions.
+%
+%For a convex curve, the optimal flattening is tractable to compute, though not necessarily very fast.
 When a curve is monotonic, error is also monotonic with respect to both subdivision points.
@@
 
-The entire recursion stack for adaptive subdivision can be represented in two words: $dt$ and $\mbox{scaled}\_t0$.
-The range is $dt \cdot \mbox{scaled}\_t0 ..
-dt \cdot (\mbox{scaled}\_t0 + 1)$.
+The entire recursion stack for adaptive subdivision can be represented in two words: $dt$ and $t0\_u$ (the latter being an unsigned integer).
+The range is $dt \cdot t0\_u ..
+dt \cdot (t0\_u + 1)$.
 Initial values of $1$ and $0$ represent the range $(0, 1)$.
-Pushing the stack, subdividing the range in half, is represented by halving $dt$ and doubling $\mbox{scaled}\_t0$, which leaves the start of the range invariant but its size halved.
-After accepting an approximation, the next range is determined by incrementing $\mbox{scaled}\_t0$, then repeatedly popping the stack by doubling $dt$ and halving $\mbox{scaled}\_t0$, as long as the latter is odd.
+Pushing the stack, subdividing the range in half, is represented by halving $dt$ and doubling $t0\_u$, which leaves the start of the range invariant but its size halved.
+After accepting an approximation, the next range is determined by incrementing $t0\_u$, then repeatedly popping the stack by doubling $dt$ and halving $t0\_u$, as long as the latter is odd.
 This can in fact be achieved without iteration by using the ``count trailing zeros'' intrinsic.
@@
 We focused primarily on the global stroke-to-fill conversion problem in order to generate polygonal outlines for rasterization.
-Since the Euler spiral approximation can fit any source cubic Bézier and its parallel offset curves, our implementation can be used to render both filled and stroked primitives within a satisfactory error tolerance.
+Since the Euler spiral approximation can fit any source cubic Bézier and its parallel curves, our implementation can be used to render both filled and stroked primitives within a satisfactory error tolerance.
 
@@
 The subpath contains two adjacent cubic Bézier curves rendered as both a stroke and a fill.
-In the stroked version, thread $n$ flattens the offset curves of the first curve (blue) and the join segment with the miter style (dark blue).
-Thread $n + 1$ flattens the offset curves of the second curve (orange) and the end cap with the round style (dark orange).
+In the stroked version, thread $n$ flattens the parallel curves of the first curve (blue) and the join segment with the miter style (dark blue).
+Thread $n + 1$ flattens the parallel curves of the second curve (orange) and the end cap with the round style (dark orange).
 Thread $n + 2$ handles the \emph{subpath end segment} and flattens the start cap (red).
@@
       }{
-          \tcp{Output both offset curves}
+          \tcp{Output both parallel curves}
           flatten\_cubic(segment,  style.offset)\;
@@
     \tcp{'t0\_u' and 'dt' track the recursion stack}
-    t0\_u $\gets$ 0u\;
+    t0\_u $\gets$ 0\;
     dt $\gets$ 1.0\;
@@
         t0 $\gets$ f32(t0\_u) * dt\;
-        \If{dt == 1.0}{break\;}
+        \If{t0 == 1.0}{break\;}
         t1 $\gets$ t0 + dt\;
@@
             \tcp{Pop the stack}
-            t0\_u $\gets$ t0\_u + 1u\;
+            t0\_u $\gets$ t0\_u + 1\;
             shift $\gets$ countTrailingZeros(t0\_u)\;
             t0\_u $\gets$ t0\_u >> shift;
-            dt $\gets$ dt * f32(1u << shift)\;
+            dt $\gets$ dt * f32(1 << shift)\;
             last\_p $\gets$ next\_p\;
@@
             \tcp{Push the stack; Continue subdivision}
-            t0\_u $\gets$ t0\_u * 2u\;
+            t0\_u $\gets$ t0\_u * 2\;
             dt $\gets$ dt * 0.5\;

### 2024-06-13 paper.tex
Address feedback on section 8 edits

--- 2024-06-13 paper.tex
@@
 
-A property of the data-parallel structure is that the output of the compute stage is unordered.
-The cubic-Bézier-to-Euler-spiral conversion is sequential within a single thread, so it is possible to preserve ordering at the scope of an individual path segment, if desired.
-However, ordering cannot be guaranteed across path segments.
+The output of the compute shader is a collection of line segments that gets stored in a GPU storage buffer.
+Every element in the output contains the viewport coordinates of the two endpoints of a single line segment.
+A property of the data-parallel structure is that the line segments are unordered.
+Because the segments are processed in parallel, ordering cannot be guaranteed across path segments.
 We refer to this output data structure as \emph{line soup}.
+
 An application may want to sort the line soup, depending on the requirements of the chosen rendering algorithm.
-For instance, our own renderer employs tile-based scanline rasterization in a compute shader, with a pipeline architecture that resembles \emph{cudaraster} (\citet{Laine2011}) and \emph{MPVG} (\citet{Ganacim2014}).
+For instance, our own renderer employs tile-based scanline rasterization in a compute shader, with a pipeline architecture that resembles \emph{cudaraster}[\citenum{Laine2011}] and \emph{MPVG}[\citenum{Ganacim2014}].
 In this design, the line soup elements are spatially sorted by subsequent compute stages into a grid of 16x16 pixel tiles.
-The final compute stage of our renderer (called \emph{fine rasterizer}) operates at tile granularity and computes pixel coverage for filled polygons outlined by the line soup (the details of which are beyond the scope of this paper).
+The final compute stage of our renderer (called \emph{fine rasterizer}) operates at tile granularity and computes pixel coverage for filled polygons outlined by the line soup (Figure~\ref{fig:pipeline} illustrates the overall structure of our renderer, the details of which are beyond the scope of this paper).
 
@@
 
-The compute shader writes the line soup output to a GPU storage buffer.
-Each line soup entry contains the viewport coordinates of the two endpoints of a singe line segment.
 We also implemented a variant that outputs circular arc segments instead of lines, in which the output entries contain the arc endpoints in addition to its curvature.

### 2024-06-13 paper.tex

--- 2024-06-13 paper.tex
@@
 
-A property of the data-parallel structure is that the output is unordered.
-The cubic-Bézier-to-Euler-spiral conversion is sequential within a single thread, so it is possible to preserve ordering at the scope of an individual path segment, if desired.
-However, ordering cannot be guaranteed across path segments.
+The output of the compute shader is a collection of line segments that gets stored in a GPU storage buffer.
+Every element in the output contains the viewport coordinates of the two endpoints of a single line segment.
+A property of the data-parallel structure is that the line segments are unordered.
+Because the segments are processed in parallel, ordering cannot be guaranteed across path segments.
 We refer to this output data structure as \emph{line soup}.
-We integrated the compute shader as a stage in a larger compute shader pipeline (discussed in Section \ref{subsection:gpu-results}) that spatially sorts the line soup in subsequent binning and tiling stages prior to rasterization.
-
-The compute shader writes the line soup output to a GPU storage buffer.
-Each line soup entry contains the viewport coordinates of the two endpoints of a singe line segment.
+
+An application may want to sort the line soup, depending on the requirements of the chosen rendering algorithm.
+For instance, our own renderer employs tile-based scanline rasterization in a compute shader, with a pipeline architecture that resembles \emph{cudaraster}[\citenum{Laine2011}] and \emph{MPVG}[\citenum{Ganacim2014}].
+In this design, the line soup elements are spatially sorted by subsequent compute stages into a grid of 16x16 pixel tiles.
+The final compute stage of our renderer (called \emph{fine rasterizer}) operates at tile granularity and computes pixel coverage for filled polygons outlined by the line soup (Figure~\ref{fig:pipeline} illustrates the overall structure of our renderer, the details of which are beyond the scope of this paper).
+
+\begin{figure*}
+    \includegraphics[scale=0.275]{pipeline}
+    \caption{We integrated our curve flattening shader into our own renderer implemented entirely in compute shaders.
+The flattening stage (described in Section~\ref{section:gpu-impl}) converts stroked paths to filled outlines and flattens them into an unordered soup of line segments.
+Our renderer spatially sorts the line soup into tiles before computing pixel coverage using a highly parallel scanline rasterizer.}
+    \Description{Our GPU pipeline architecture in broad strokes}
+    \label{fig:pipeline}
+\end{figure*}
+
 We also implemented a variant that outputs circular arc segments instead of lines, in which the output entries contain the arc endpoints in addition to its curvature.

### 2024-06-14 paper.tex
Minor edits and figure tweaks
A bit more detail on the subdivision density integral approximation. Remove mmark from timings graph. Minor edits in response to review feedback.

@@
 
-Stroke rendering is an essential part of the standard vector graphics imaging model, and is thus widely used.
+Stroke rendering is an essential and widely part of the standard vector graphics imaging model.
 The standard representation of paths is a sequence of cubic Bézier segments.
 While there are a number of published techniques for GPU rendering of filled paths, stroke rendering is more challenging to parallelize, largely because path segments cannot be processed independently of each other; rendering of the \emph{joins} between path segments depends on both adjoining segments, and the endpoints of paths (when not closed) are rendered with \emph{caps}.
-Joins and caps can have multiple styles, including different styles within the same document.
+Joins and caps can be rendered with a variety of styles within the same document.
+The style of stroke rendering also optionally includes \emph{dashing}, which applies a pattern of dashes and gaps to the outline of the shape, as illustrated in Figure~\ref{fig:stroke_styles}.
+
+\begin{figure}
+    \vspace{15pt}
+    \includegraphics[scale=0.4]{stroke_styles}
+    \caption{Illustration of stroke styles applied to a source Bézier path (shown in black).}
+    \Description{Illustration of stroke styles applied to a source Bézier path (shown in black).}
+    \label{fig:stroke_styles}
+\end{figure}
 
 A typical vector graphics document may have thousands of stroked paths.
-A map or CAD drawing may have tens of thousands of paths, not unusually a million path segments.
-At this scale, the computational cost of stroke rendering on CPU may then result in loss of interactive rendering performance.
-Thus, we desire to improve performance by offloading stroke rendering to the GPU.
-To realize performance gains, the algorithm needs to be GPU-friendly, meaning among other things that it exploits the available parallelism, is work efficient, and avoids unnecessarily complex (and potentially divergent) control flow.
-In addition, it must remain numerically robust even when computed with 32 bit floating point numbers.
+A map or CAD drawing may have tens of thousands of paths, possibly consisting of millions of path segments.
+At this scale, the computational cost of stroke rendering on a CPU may result in loss of interactive rendering performance.
+We believe the runtime performance can be improved significantly by offloading stroke rendering to a GPU.
+This requires that the algorithm be GPU-friendly: that it exploit the available parallelism, be work efficient, and avoid unnecessarily complex and divergent control flow.
+It must also remain numerically robust when computed with 32-bit floating pointing numbers.
 
@@
 In this paper, we focus on \emph{stroke expansion,} the generation of an outline that, when filled, represents the rendered stroke.
-Implementing stroke rendering using stroke expansion enables a unified approach to path rendering downstream, rendering both filled and stroked paths.
+Implementing stroke rendering using stroke expansion enables a unified approach to path rendering both filled and stroked paths downstream.
 Among other benefits, such an approach avoids a large number of draw calls when stroked and filled paths are finely interleaved, as is often the case.
@@
 Both \citet{Nehab2020} and \citet{Kilgard2020} provide useful definitions of strongly correct stroke rendering, and Nehab in particular describes how to achieve it in stroke expansion.
-Briefly, when the path curvature exceeds the reciprocal of the stroke half-width, then addition \emph{evolute} segments need to be added, in addition to the parallel curves required by weak correctness, and also the inner contours of joins are required in general.
-The technique described in this paper is intended for strongly correct stroke rendering; it produces outlines that render the same shape as \citet{Nehab2020}.
+Briefly, when the path curvature exceeds the reciprocal of the stroke half-width, \emph{evolute} segments and the \emph{inner contours of joins} are required in addition to the parallel curves.
+The technique described in this paper produces strongly correct outlines according to this definition.
+Examples of weakly and strongly correct rendering are shown in Figure~\ref{fig:weak_strong}.
+
+\begin{figure}
+    \includegraphics[scale=0.5]{weak_strong}
+    \caption{Example of weakly correct (top) and strongly correct (bottom) stroke rendering.
+The left example has a curvature cusp and requires evolutes to render correctly, while the right requires the inner join.}
+    \Description{Two examples of strokes.
+The top left is missing an evolute section, making it appear that a half-disc has been carved out.
+The top right is missing inner joins, looking like spokes rather than a connected path.
+The bottom two examples are rendered with the full strokes.}
+    \label{fig:weak_strong}
+\end{figure}
 
 Correct stroke rendering also relies on accurate approximation of the parallel curves of the input paths.
-Determining the parallel curve of a cubic Bézier is not analytically tractable, as it is a tenth order algebraic formula[\citenum{Farouki1990}].
+Determining the parallel curve of a cubic Bézier is not analytically tractable, as it is a tenth order algebraic formula~[\citenum{Farouki1990}].
 Thus, all practical stroke rendering techniques employ an approximation, with error tolerance as a tunable parameter, typically representing the maximum allowable Fréchet distance between the true curve and its approximation.
@@
 A correct algorithm will produce an approximate curve within the error tolerance, but an efficient algorithm will not subdivide significantly more than is necessary to meet the error bound.
-Some specifications for vector graphics, including PDF[\citenum{PDF2008}], specify the error tolerance explicitly.
+Some specifications for vector graphics, including PDF~[\citenum{PDF2008}], specify the error tolerance explicitly.
 
@@
 An example of a reasonably good algorithm that computes flattened parallel curves is \citet{Yzerman2020}, as used in the Blend2D rendering library.
-For producing curved outlines, the Tiller and Hanson[\citenum{Tiller1984}] algorithm is commonly implemented and cited, but its performance is quite poor when applied to cubic Béziers.
+For producing curved outlines, the Tiller and Hanson~[\citenum{Tiller1984}] algorithm is commonly implemented and cited, but its performance is quite poor when applied to cubic Béziers.
 The quadratic Bézier version is adequate, and forms the basis of the algorithm in \citet{Nehab2020}.
-A variant is also used in Skia[\citenum{Skia}]; instead of lowering the cubic Bézier to a quadratic approximation and then computing the parallel curve, Skia approximates the parallel curve directly with a quadratic Bézier, then measures the error to determine whether further subdivision is necessary.
+A variant is also used in Skia~[\citenum{Skia}]; instead of lowering the cubic Bézier to a quadratic approximation and then computing the parallel curve, Skia approximates the parallel curve directly with a quadratic Bézier, then measures the error to determine whether further subdivision is necessary.
 
@@
 The error metrics presented in this paper are both straightforward to compute, and also \emph{invertible}, which avoids excessive subdivision and also increases the amount of available parallelism to exploit.
-Previous approaches to error measurement generally follow a ``cut then measure'' approach, where the approximate curve is generated, then error measured by sampling.
+Previous approaches to error measurement generally follow a ``cut then measure'' approach, where the approximate curve is generated, then the error measured by sampling.
 Such sampling is often time consuming, involving iteration, and even then risks underestimating the error due to inadequate sampling.
 
-This paper presents a comprehensive solution to stroke expansion, well suited to GPU implementation.
+Previous work, particularly \citet{Kilgard2020}, has explained how to break the input path into dash segments of the specified length on GPU using a prefix sum of segment arc lengths.
+There are implementation tradeoffs to GPU-based dash processing, including potentially a larger number of dispatch stages.
+We anticipate that the potential performance benefit of implementing dashing on a GPU depends on the scene and other factors in the overall rendering pipeline.
+While we would expect further performance improvement from GPU-based dashing, we did not fully explore the solution space and we were able to achieve acceptable performance with dash processing on CPU.
+
+This paper presents a comprehensive solution to the core problem of stroke expansion, well suited to GPU implementation.
 It can produce both flattened polylines or an approximation consisting of circular arcs.
@@
 From this, we can derive a precise, invertible error metric.
-Subdividing the arc in to $n$ segments, the distance error for each segment is $\frac{1}{\kappa}(1-\cos \frac{\kappa s}{2n})$.
-Solving for n, we get:
+Subdividing the arc into $n$ segments, the distance error for each segment is $\frac{1}{\kappa}(1-\cos \frac{\kappa s}{2n})$.
+Solving for $n$, we get:
 
@@
 We are of course concerned with the flattening of more general curves (ultimately the parallel curve of a cubic Bézier), not simply circular arcs, so the curvature is not constant.
-An obvious approach to try would be using the maximum error.
+An obvious approach to try would be to use the maximum error.
 This would be conservative with respect to error tolerance, but also generate too much subdivision when the curve has a small region of high curvature.
@@
     \caption{The flattened approximation to an Euler spiral using line segments is achieved by choosing subdivision points using a simple power law formula with an exponent of 2/3.}
-    \Description{An Euler spiral along with its flattened representation. The flattened curve has a consistent error, and the subdivision points are spaced according to a}
+    \Description{An Euler spiral along with its flattened representation. The flattened curve has a consistent error, and the subdivision points are spaced according to a power law.}
     \label{fig:es_flatten}
@@
 Even so, this curve has some disadvantages.
-In particular, it cannot model an inflection point, so the source curve must be subdivided at inflection points.
+For one, it cannot model an inflection point, so the source curve must be subdivided at inflection points.
 
@@
 The parallel curve of the Euler spiral (also known as ``clothoid'') was characterized by \citet{Wieleitner1907} well over a hundred years ago, and has a straightforward closed-form Cesàro representation, curvature as a function of arc length, of the form $\kappa(s) = c_0{s - s_0}^{-1/2} + c_1$.
+Euler spirals as an intermediate curve representation for computing parallel curves was proposed in \citet{Levien2021}, which also expands on the Wieleitner reference (including an English translation) and Cesàro equation.
 
@@
 The geometry of a stroke outline consists of joins, caps, and the two parallel curves on either side of the input path segments, offset by the half linewidth.
-The joins and caps are not particularly difficult to calculate, but parallel curves of cubic Béziers are notoriously tricky.
+The joins and caps are not particularly difficult to calculate, but parallel curves of cubic Béziers are notoriously tricky~[\citenum{Kilgard2020a}].
 Analytically, it is a tenth order algebraic curve, which is not particularly feasible to compute directly.
@@
 Using a subpixel Fréchet distance bound guarantees that the rendered image does not deviate visibly from the exact rendering.
-Another choice would be uniform steps in tangent angle, as chosen by polar stroking[\citenum{Kilgard2020}].
+Another choice would be uniform steps in tangent angle, as chosen by polar stroking~[\citenum{Kilgard2020}].
 However, at small curvature, the stroked path can be off by several pixels, and at large curvature there may be considerably more subdivision than needed for faithful rendering.
@@
 The bottom row shows a consistent angle step, as implemented in polar stroking, but has excessive distance error at low curvature, and excessive subdivision at high curvature.
-It should be noted, to avoid the undershoot at low curvature, both the Skia[\citenum{Skia}] and Rive[\citenum{Rive}] renderers use a hybrid of the Wang and polar stroking error metrics.
+It should be noted, to avoid the undershoot at low curvature, both the Skia~[\citenum{Skia}] and Rive~[\citenum{Rive}] renderers use a hybrid of the Wang and polar stroking error metrics.
 
@@
 In this subsection, we present a straightforward and accurate approximation, constructed piecewise from easily invertible functions.
-If higher flattening quality is desired at the expense of slower computation, this approximation can be used to determine a good initial value for numeric techniques; two iterations of Newton solving are enough to refine this guess to within 32-bit floating point accuracy.
-
-The approximation is given as follows:
+This approximation was found empirically, by interactively tuning candiadate approximation functions in the Desmos graphing calculator.
+The exact integral and the approximation are shown in Figure~\ref{fig:espc}.
+Visually, it is clear that the agreement is close.
+We also built a test suite to exhaustively test the subdivision count using a property testing approach, finding that the worst case discrepancy between approximate and exact results is 6\%.
+If higher flattening quality is desired at the expense of slower computation, this approximation could be used to determine a good initial value for numeric techniques; two iterations of Newton solving are enough to refine this guess to within 32-bit floating point accuracy.
+
+The approximation is as follows:
 
@@
 
-The exact integral and the approximation given above are shown in Figure~\ref{fig:espc}.
-Visually, it is clear that the agreement is close, and in numerical testing the worst case discrepancy between approximate and exact results is approximately 6\%.
-
 \begin{figure}
@@
 
-See Figure~\ref{fig:g_comparison} for a comparison between flattening and approximation with arc segments.
+See Figure~\ref{fig:g_comparison} for a comparison between flattening to a polyline and approximation with arc segments.
 The arc segment version has many fewer segments at the same tolerance, while preserving very high visual quality.
@@
 
-In the principled, correct specification for stroking[\citenum{Nehab2020}], parallel curves are sufficient only for segments in which the curvature does not exceed the reciprocal half width.
+In the principled, correct specification for stroking~[\citenum{Nehab2020}], parallel curves are sufficient only for segments in which the curvature does not exceed the reciprocal half-width.
 When it does, additional segments must be drawn, including evolutes of the original curve.
 In general, the evolute of a cubic Bézier is a very complex curve, requiring approximation techniques.
-By contrast, the evolute of an Euler spiral ($\kappa = as$) is another spiral with a simple Cesàro equation, namely $\kappa = -a^{-1}s^{-3}$, an instance of the general result that the evolute of a log-aesthetic curve is another log-aesthetic curve[\citenum{Yoshida2012}].
+By contrast, the evolute of an Euler spiral ($\kappa = as$) is another spiral with a simple Cesàro equation, namely $\kappa = -a^{-1}s^{-3}$, an instance of the general result that the evolute of a log-aesthetic curve is another log-aesthetic curve~[\citenum{Yoshida2012}].
 
@@
 Computing parallel curves directly on cubic Bézier curve segments is not very tractable.
-In particular, the widely cited Tiller-Hanson algorithm[\citenum{Tiller1984}] performs well for quadratic Béziers but significantly worse for cubics.
+In particular, the widely cited Tiller-Hanson algorithm~[\citenum{Tiller1984}] performs well for quadratic Béziers but significantly worse for cubics.
 
@@
 A good example of this pattern is \citet{Nehab2020}.
-One advantage of Euler spirals over quadratic Béziers is that they can represent inflection points just fine, so it is not necessary to compute those for additional subdivision.
+One advantage of Euler spirals over quadratic Béziers is that they can represent inflection points just fine, so it is not necessary to solve for the inflection points, or do additional subdivision.
+Avoiding these helps make the algorithm a bit more GPU-friendly.
 
@@
 Our approach to the error metric has two major facets.
-First, we obtain a cubic Bézier which is a very good fit to the Euler spiral, then we estimate the distance between that and the source cubic.
+First, we obtain a secondary cubic Bézier which is a very good fit to the Euler spiral, then we estimate the distance between that and the source cubic.
 Due to the triangle inequality, the sum of these is a conservative estimate of the true Fréchet distance between the cubic and the Euler spiral.
@@
 Given tangent angles relative to the chord, finding the Euler spiral segment that minimizes total curvature variation is a form of geometric Hermite interpolation.
-There are a number of published solutions to this problem, involving nontrivial numerical solving techniques: gradient descent[\citenum{Kimia2003}], bisection[\citenum{Walton2009}], or Newton iteration[\citenum{Connor2014}].
+There are a number of published solutions to this problem, involving nontrivial numerical solving techniques: gradient descent~[\citenum{Kimia2003}], bisection~[\citenum{Walton2009}], or Newton iteration~[\citenum{Connor2014}].
 A more direct approach is to approximate the function as a reasonably low-order polynomial in terms of the endpoint angles.
@@
 Pushing the stack, subdividing the range in half, is represented by halving $dt$ and doubling $t0\_u$, which leaves the start of the range invariant but its size halved.
-After accepting an approximation, the next range is determined by incrementing $t0\_u$, then repeatedly popping the stack by doubling $dt$ and halving $t0\_u$, as long as the latter is odd.
-This can in fact be achieved without iteration by using the ``count trailing zeros'' intrinsic.
+After accepting an approximation, the next range is determined by incrementing $t0\_u$.
+At this point, the decision of whether to pop the stack or continue is dependent on whether $t0\_u$ is even or odd, respectively.
+If even, popping the stack is accomplished by doubling $dt$ and halving $t0\_u$, and this is repeated until $t0\_u$ becomes odd.
+The repeated stack popping can be accomplished without iteraction by using the ``count trailing zeros'' intrinsic to determine the number of stack pops $n$, then multiplying $dt$ by $2^n$, and shifting $t0\_u$ right by $n$ bits.
 
@@
 
-A Bézier curve is expressed in parametric form, and the derivative with respect to the parameter can be zero or nearly so, causing serious numerical problems for many stroke expansion algorithms.
+A Bézier curve is expressed in parametric form, and its derivative with respect to the parameter can be zero or nearly so, causing serious numerical problems for many stroke expansion algorithms.
 In the limit, as the Bézier curve describes a semi-cubical parabola, the curvature can become infinite.
@@
     \includegraphics[scale=0.19]{stroke_stills}
-    \caption{Renderings of stroked cubic Béziers exhibiting strong and weak correctness.
-The outlines generated by our GPU stroking compute shader are visualized in green and blue.
+    \caption{Stroked cubic Béziers exhibiting strong and weak correctness, rendered by our implementation.
+The contents of the line soup buffer are visualized as green and blue outlines.
 Segments with a winding direction that points \emph{up} are colored blue while those pointing \emph{down} are colored green.
-The curves were adapted to from the \emph{stills} dataset by \citet{Nehab2020} and the Skia test corpus.}
+The curves were adapted from the \emph{stills} dataset by \citet{Nehab2020} and the Skia test corpus.}
     \Description{Selection of stroked cubic Béziers exhibiting strong and weak correctness}
@@
 Notably, our implementation is fully data-parallel on input path segments and does not require any computationally expensive curve evaluation on the CPU.
-We implemented our shader on general-purpose GPU compute primitives supported by all modern graphics APIs, particularly Metal, Vulkan, D3D12, and WebGPU[\citenum{WebGPU}].
+We implemented our shader on general-purpose GPU compute primitives supported by all modern graphics APIs, particularly Metal, Vulkan, D3D12, and WebGPU~[\citenum{WebGPU}].
 As such, the ideas presented in this section are portable to various GPU platforms.
@@
 
-An SVG path consists of a sequence of instructions (or "verbs").
+An SVG path consists of a sequence of instructions (or ``verbs'').
 The \emph{lineto} and \emph{curveto} instructions denote an individual \emph{path segment} consisting of either a straight line or a cubic Bézier.
-A given subsequence of path segments can be bounded by a \emph{moveto} and/or \emph{closepath} instruction that marks the beginning and end of a \emph{subpath}.
+The \emph{moveto} instruction is used to begin a new \emph{subpath} and the \emph{closepath} instruction can be used to create a closed contour (by inserting a line connecting the current endpoint to the start of the subpath).
 A subpath must form a closed contour if painted as a fill.
@@
   \eIf{$style$ is $fill$}{
-      \tcp{Output source curve}
+      \tcp{Output source curve (offset = 0)}
       flatten\_cubic(segment, $0$)\;
@@
       }{
-          \tcp{Output both parallel curves}
+          \tcp{Output parallel curves}
           flatten\_cubic(segment,  style.offset)\;
@@
 For a filled primitive, a thread only outputs the flattened approximation of the path segment.
-For a stroke, a thread outputs the polylines approximating multiple curves: a) the two parallel curves on both sides of the source segment, offset by the desired stroke \emph{halfwidth}, and b) the joins that connect the path segment to the next adjacent segment, and c) evolute patches in regions of high curvature.
+For a stroke, a thread outputs the polylines approximating multiple curves: a) the two parallel curves on both sides of the source segment, offset by the desired stroke \emph{half-width}, and b) the joins that connect the path segment to the next adjacent segment, and c) evolute patches in regions of high curvature.
 We also store metadata in our input encoding that marks the position of a path segment within the containing subpath.
@@
     TOL $\gets$ Euler spiral fit tolerance\;
-    \tcp{'t0\_u' and 'dt' track the recursion stack}
+    \tcp{`t0\_u' and `dt' track the recursion stack}
     t0\_u $\gets$ 0\;
@@
             \tcp{Find location of any cusp in outline.}
-            \tcp{If $0 \leq t < 1$, draw evolute at 't'.}
+            \tcp{If $0 \leq t < 1$, draw evolute at `t'.}
             cusp0 $\gets$ es\_curvature(es, 0) * offset + chord\_len\;
@@
             shift $\gets$ countTrailingZeros(t0\_u)\;
-            t0\_u $\gets$ t0\_u >> shift;
-            dt $\gets$ dt * f32(1 << shift)\;
+            t0\_u $\gets$ t0\_u $\gg$ shift\;
+            dt $\gets$ dt * f32(1 $\ll$ shift)\;
             last\_p $\gets$ next\_p\;
@@
 If the error exceeds the desired tolerance threshold, the cubic segment is iteratively subdivided in half and a new error estimate is computed.
-Once the error satisfies the tolerance, the Euler spiral segment is directly flattened to a polyline (or, alternately, converted to a sequence of arc segments) by invoking a subroutine that is parameterized by the stroke halfwidth (see \texttt{es\_seg\_flatten\_offset} in Algoritm~\ref{alg:flatten-cubic}).
+Once the error satisfies the tolerance, the Euler spiral segment is directly flattened to a polyline (or, alternately, converted to a sequence of arc segments) by invoking a subroutine that is parameterized by the stroke half-width (see \texttt{es\_seg\_flatten\_offset} in Algoritm~\ref{alg:flatten-cubic}).
 For a regular fill, this parameter is set to $0$.
@@
 If the outline contains a cusp -- which can be determined by evaluating the curvature variation along the Euler spiral segment -- we output an evolute patch (as shown in Figure~\ref{fig:evolutes}) in order to achieve strong correctness.
-A selection of renderings produced by our algorithm that demonstrate both strong and weak correctness are shown in Figure~\ref{fig:stroke_stills}.
+A selection of renderings produced by our algorithm that demonstrate both strong and weak correctness is shown in Figure~\ref{fig:stroke_stills}.
 
@@
 Every element in the output contains the viewport coordinates of the two endpoints of a single line segment.
+Threads reserve storage buffer regions for their output by incrementing an atomic index stored in GPU device memory.
 A property of the data-parallel structure is that the line segments are unordered.
@@
 An application may want to sort the line soup, depending on the requirements of the chosen rendering algorithm.
-For instance, our own renderer employs tile-based scanline rasterization in a compute shader, with a pipeline architecture that resembles \emph{cudaraster}[\citenum{Laine2011}] and \emph{MPVG}[\citenum{Ganacim2014}].
+For instance, our own renderer employs tile-based scanline rasterization in a compute shader, with a pipeline architecture that resembles \emph{cudaraster}~[\citenum{Laine2011}] and \emph{MPVG}~[\citenum{Ganacim2014}].
 In this design, the line soup elements are spatially sorted by subsequent compute stages into a grid of 16x16 pixel tiles.
@@
     \caption{We integrated our curve flattening shader into our own renderer implemented entirely in compute shaders.
-The flattening stage (described in Section~\ref{section:gpu-impl}) converts stroked paths to filled outlines and flattens them into an unordered soup of line segments.
+The flattening stage (described in Section~\ref{section:gpu-impl}) converts stroked paths to filled outlines and flattens them to an unordered soup of line segments.
 Our renderer spatially sorts the line soup into tiles before computing pixel coverage using a highly parallel scanline rasterizer.}
@@
 
-We also implemented a variant that outputs circular arc segments instead of lines, in which the output entries contain the arc endpoints in addition to its curvature.
+We also implemented a variant that outputs circular arc segments instead of lines, in which the output entries store curvature in addition to the endpoints.
 For both primitive types, our renderer also outputs a 32-bit \emph{path ID} used by downstream pipeline stages to identify the path that a segment belongs to, in order to retrieve the correct paint (solid or gradient color) for rendering.
@@
 We evaluated the runtime performance of our compute shader on 4 GPU models and multiple form factors: Google Pixel 6 equipped with the Arm \emph{Mali-G78 MP20} mobile GPU, the Apple \emph{M1 Max} integrated laptop GPU, and two discrete desktop GPUs: the mid-2015 NVIDIA {GeForce GTX 980Ti} and the late-2022 NVIDIA {GeForce RTX 4090}.
-We authored our entire GPU pipeline in the \emph{WebGPU Shading Language} (WGSL) and used the \emph{wgpu}[\citenum{wgpu}] framework to interoperate with the native graphics APIs (we tested the M1 Max on macOS via the Metal API; the mobile and desktop GPUs were tested via the Vulkan API on Android and Ubuntu systems).
+We authored our entire GPU pipeline in the \emph{WebGPU Shading Language} (WGSL) and used the \emph{wgpu}~[\citenum{wgpu}] framework to interoperate with the native graphics APIs (we tested the M1 Max on macOS via the Metal API; the mobile and desktop GPUs were tested via the Vulkan API on Android and Ubuntu systems).
 
@@
 We authored two additional test scenes in order to stress the GPU with a higher workload.
-The first is a very large stroked path with ~500,000 individually styled dashed segments which we adapted from Skia's open-source test corpus.
+The first is a very large stroked path with \textasciitilde500,000 individually styled dashed segments which we adapted from Skia's open-source test corpus.
 We used two variants of this scene with \emph{butt} and \emph{round} cap styles.
-For the second scene we adapted the \emph{Canvas Paths} test case from MotionMark[\citenum{MotionMark}], which renders a large number of randomly generated stroked line, quadratic, and cubic Bézier paths.
+For the second scene we adapted the \emph{Canvas Paths} test case from MotionMark~[\citenum{MotionMark}], which renders a large number of randomly generated stroked line, quadratic, and cubic Bézier paths.
 We made two versions with different sizes: mmark-70k and mmark-120k with 70,000 and 120,000 path segments, respectively.
@@
     \hline
-        waves.svg & 13,308 & 406,059 & 111,965 \\
-        mmark-70k & 70,000 & 1,039,899 & 500,351 \\
-        mmark-120k & 120,000 & 1,784,896 & 858,162 \\
-        long dash (butt) & 503,304 & 1,060,476 & 1,060,476 \\
-        long dash (round) & 503,304 & 2,013,216 & 1,060,476 \\
+        waves.svg & 13,308 & 475,855 & 181,229 \\
+        mmark-70k & 70,000 & 1,577,705 & 1,162,117 \\
+        mmark-120k & 120,000 & 2,709,013 & 1,994,946 \\
+        long dash (butt) & 503,304 & 1,672,561 & 1,672,561 \\
+        long dash (round) & 503,304 & 2,625,300 & 1,672,561 \\
     \hline
@@
 The hatched bars show the timings for the version that outputs line primitives while the solid bars show that for arcs.
-Mali-G78 (the lowest-end GPU we tested) converts \emph{waves} (the largest scene in the set) to lines in \~3~ms.
+Mali-G78 (the lowest-end GPU we tested) converts \emph{waves} (the largest scene in the set) to lines in \textasciitilde3 ms.
 The compute shader executes at least an order of magnitude faster on the other GPUs.}
@@
 The entire \citet{Nehab2020} timings set adds up to less than 1 ms on all GPUs except the mobile unit.
-Our algorithm is ~14x slower on the Mali-G78 compared to the M1 Max but it is still capable of processing \emph{waves} in 3 ms.
+Our algorithm is \textasciitilde14$\times$ slower on the Mali-G78 compared to the M1 Max but it is still capable of processing \emph{waves} in 3~ms.
 We observed that the performance of the kernel scales well with increasing GPU ability even at very large workloads, as demonstrated by the order-of-magnitude difference in run time between the mobile, laptop, and high-end desktop form factors we tested.
-We also confirmed that outputting arcs instead of lines can lead to a 2x decrease in execution time (particularly in scenes with high curve content, like \emph{waves} and \emph{mmark}) as predicted by the lower overall primitive count.
+We also confirmed that outputting arcs instead of lines can lead to a 2$\times$ decrease in execution time (particularly in scenes with high curve content, like \emph{waves} and \emph{mmark}) as predicted by the lower overall primitive count.
 
@@
 The CPU encoding times (prior to the GPU prefix sums) for some of the large scenes is shown in Table~\ref{table:bump-estimate-cost}.
-Our implementation still applies dashing on the CPU.
-It is conceptually fairly straightforwad to implement on GPU; the core algorithm is prefix sum applied to segment length, and the Euler spiral representation is particularly well suited.
-However, there are complex implementatoin tradeoffs, so the dash style applied to the \emph{long dash} scene had to be processed on the CPU and dashes were sequentially encoded as individual strokes.
-Including CPU-side dashing, the time to encode \emph{long dash} on the Apple M1 Max was approximately 25 ms on average, while the average time to encode the path after dashing was 4.14 ms.
+Our implementation applies dashing on the CPU.
+It is conceptually straightforward to implement on GPU; the core algorithm is prefix sum applied to segment length, and the Euler spiral representation is particularly well suited.
+However, we believe the design of an efficient GPU pipeline architecture involves comples tradeoffs which we did not sufficiently explore for this paper.
+Therefore the dash style applied to the \emph{long dash} scene was processed on the CPU and dashes were sequentially encoded as individual strokes.
+Including CPU-side dashing, the time to encode \emph{long dash} on the Apple M1 Max was approximately 25~ms on average, while the average time to encode the path after dashing was 4.14~ms.
 
@@
 We augmented our encoding logic to compute a conservative estimate of the required output buffer size based on an upper bound on the number of line primitives per input path segment.
-We found that Wang's formula[\citenum{Goldman2003}] works well in practice, as it is relatively inexpensive to evaluate, though it results in a higher memory estimate than required (see Table~\ref{table:bump-estimate-mem}).
+We found that Wang's formula~[\citenum{Goldman2003}] works well in practice, as it is relatively inexpensive to evaluate, though it results in a higher memory estimate than required (see Table~\ref{table:bump-estimate-mem}).
 For parallel curves, Wang's formula does not guarantee an accurate upper bound on the required subdivisions, which we worked around by inflating the estimate for the source segment by a scale factor derived from the linewidth.
-We observed that enabling estimation increases our CPU pre-processing time by 2--3x on the largest scenes but the overall impact is negligible when the scene size is modest.
+We observed that enabling estimation increases our CPU pre-processing time by 2--3$\times$ on the largest scenes but overall the impact is negligible when the scene size is modest.
 
@@
 Our estimation uses Wang's formula combined with a scaling heuristic based on the linewidth.
-In the worst case, our buffer size estimator overshoots by a factor of \~4.}
+The estimate is conservative and always overshoots the actual memory requirement.}
     \begin{tabular}{|l|c|c|}
@@
     \hline
-        roads.svg  & 958.78 KB & 2.45 MB    \\
-        waves.svg  & 9.37 MB   & 10.44 MB   \\
-        mmark-70k  & 23.81 MB  & 100.79 MB  \\
-        mmark-120k & 40.86 MB  & 173.56 MB  \\
+        roads.svg  & 2.76 MB   & 8.35 MB    \\
+        waves.svg  & 10.89 MB  & 80.69 MB   \\
+        mmark-70k  & 36.10 MB  & 103.06 MB  \\
+        mmark-120k & 60.01 MB  & 177.33 MB  \\
     \hline
@@
     \includegraphics[scale=0.2]{test_renderings}
-    \caption{Renderings of some of our test scenes created using our algorithm and associated frame rates.
+    \caption{Renderings of some of our test scenes created using our algorithm.
 All images were rendered on an Apple M1 Max at 2088x1600 resolution using our GPU pipeline.
-\textbf{(a)} \emph{waves.svg} from the Nehab2020 timings data set (340fps); \textbf{(b)} The \emph{mmark-120k} scene with 120,000 input segments (72fps); \textbf{(c)} The \emph{long path} scene with ~500k input segments and round caps (40fps with CPU-side dashing - the GPU time to render the scene is 9ms); \textbf{(d)}; A simple test scene showcasing various cap and join styles (950fps); \textbf{(e)} The famous \emph{Ghostscript Tiger} containing both stroked and filled paths (810fps).}
+\textbf{(a)} \emph{waves.svg} from the Nehab2020 timings data set (3.09~ms); \textbf{(b)} The \emph{mmark-120k} scene with 120,000 input segments (19.86~ms); \textbf{(c)} The \emph{long path} scene with \textasciitilde500k input segments and round caps (25~ms including CPU-side dashing -- the GPU time to render the scene is 9~ms); \textbf{(d)}; A simple test scene showcasing various cap and join styles (1.56~ms); \textbf{(e)} The famous \emph{Ghostscript Tiger} containing both stroked and filled paths (1.69~ms).}
     \Description{Renderings of the test scenes}
@@
     \item The GPU implementation sequentializes some work that could be parallel, and is thus not as well load-balanced as it might be.
-An appealing future direction is to split the pipeline into separate stages, executed by the GPU as nodes in a work graph[\citenum{Patel2024}].
+An appealing future direction is to split the pipeline into separate stages, executed by the GPU as nodes in a work graph~[\citenum{Patel2024}].
 This structure is appealing because load balancing is done by hardware.
@@
 
-    \item Dashing stroked paths on the GPU was left out of scope for this paper. Given its parameterization by arc length, we believe the Euler spiral representation is highly suitable for dashing.
+    \item As mentioned earlier, we did not have time to sufficiently explore a pipeline architecture for GPU-based dashing.
+Given its parameterization by arc length, we believe the Euler spiral representation lends itself well to an approach based on parallel prefix sums of segment arc lengths.
 \end{itemize}

### 2024-06-14 paper.tex
Minor edits
Fixed some typos and redundancies. Clarified the renderer in the stroke
stills figure caption.

--- 2024-06-13 paper.tex
@@
 
-An SVG path consists of a sequence of instructions (or "verbs").
+An SVG path consists of a sequence of instructions (or ``verbs'').
 The \emph{lineto} and \emph{curveto} instructions denote an individual \emph{path segment} consisting of either a straight line or a cubic Bézier.
@@
     TOL $\gets$ Euler spiral fit tolerance\;
-    \tcp{'t0\_u' and 'dt' track the recursion stack}
+    \tcp{`t0\_u' and `dt' track the recursion stack}
     t0\_u $\gets$ 0u\;
@@
 If the error exceeds the desired tolerance threshold, the cubic segment is iteratively subdivided in half and a new error estimate is computed.
-Once the error satisfies the tolerance, the Euler spiral segment is directly flattened to a polyline (or, alternately, converted to a sequence of arc segments) by invoking a subroutine that is parameterized by the stroke halfwidth (see \texttt{es\_seg\_flatten\_offset} in Algoritm~\ref{alg:flatten-cubic}).
+Once the error satisfies the tolerance, the Euler spiral segment is directly flattened to a polyline by invoking a subroutine that is parameterized by the stroke half-width (see \texttt{es\_seg\_flatten\_offset} in Algoritm~\ref{alg:flatten-cubic}).
 For a regular fill, this parameter is set to $0$.

### 2024-06-14 paper.tex
More low-hanging fixes
- Add non-breaking space before citenum
- Use "ms" instead of "fps" numbers in renderings figure caption

@@
 As described in detail in \citet{Nehab2020}, very few existing implementations actually implement the strong version, so document authors have become accustomed to not depending on behavior at endpoints.
-Standards for graphics formats, such as SVG[\citenum{Svg2}], also provide an ``out,'' enabling the weaker behavior.
+Standards for graphics formats, such as SVG~[\citenum{Svg2}], also provide an ``out,'' enabling the weaker behavior.
 While our current implementation emphasizes weak correctness, we believe it can be extended to the stronger sense, by implementing evolutes and inner join contours.
@@
 Some are \emph{local,} in that they break the stroke up into smaller closed pieces, the union of which forms an approximation to the stroke.
-Included in this category is polar stroking[\citenum{Kilgard2020}], which proposes a GPU-friendly subdivision scheme based on an angle step error.
+Included in this category is polar stroking~[\citenum{Kilgard2020}], which proposes a GPU-friendly subdivision scheme based on an angle step error.
 Others are based on stroke expansion, generating an explicit outline, which is then filled to produce the stroke rendering.
@@
 
-The parallel curve of the Euler spiral (also known as ``clothoid'') was characterized by \citet{Wieleitner1907} well over a hundred years ago[\citenum{Wieleitner1907}], and has a straightforward closed-form Cesàro representation, curvature as a function of arc length.
+The parallel curve of the Euler spiral (also known as ``clothoid'') was characterized by \citet{Wieleitner1907} well over a hundred years ago~[\citenum{Wieleitner1907}], and has a straightforward closed-form Cesàro representation, curvature as a function of arc length.
 
@@
 
-In the principled, correct specification for stroking[\citenum{Nehab2020}], parallel curves are sufficient only for segments in which the curvature does not exceed the reciprocal half width.
+In the principled, correct specification for stroking~[\citenum{Nehab2020}], parallel curves are sufficient only for segments in which the curvature does not exceed the reciprocal half width.
 When it does, additional segments must be drawn, including evolutes of the original curve.
 In general, the evolute of a cubic Bézier is a very complex curve, requiring approximation techniques.
-By contrast, the evolute of an Euler spiral ($\kappa = as$) is another spiral with a simple Cesàro equation, namely $\kappa = -a^{-1}s^{-3}$, an instance of the general result that the evolute of a log-aesthetic curve is another log-aesthetic curve[\citenum{Yoshida2012}].
+By contrast, the evolute of an Euler spiral ($\kappa = as$) is another spiral with a simple Cesàro equation, namely $\kappa = -a^{-1}s^{-3}$, an instance of the general result that the evolute of a log-aesthetic curve is another log-aesthetic curve~[\citenum{Yoshida2012}].
 
@@
 We used two variants of this scene with \emph{butt} and \emph{round} cap styles.
-For the second scene we adapted the \emph{Canvas Paths} test case from MotionMark[\citenum{MotionMark}], which renders a large number of randomly generated stroked line, quadratic, and cubic Bézier paths.
+For the second scene we adapted the \emph{Canvas Paths} test case from MotionMark~[\citenum{MotionMark}], which renders a large number of randomly generated stroked line, quadratic, and cubic Bézier paths.
 We made two versions with different sizes: mmark-70k and mmark-120k with 70,000 and 120,000 path segments, respectively.
@@
 We augmented our encoding logic to compute a conservative estimate of the required output buffer size based on an upper bound on the number of line primitives per input path segment.
-We found that Wang's formula[\citenum{Goldman2003}] works well in practice, as it is relatively inexpensive to evaluate, though it results in a higher memory estimate than required (see Table~\ref{table:bump-estimate-mem}).
+We found that Wang's formula~[\citenum{Goldman2003}] works well in practice, as it is relatively inexpensive to evaluate, though it results in a higher memory estimate than required (see Table~\ref{table:bump-estimate-mem}).
 For parallel curves, Wang's formula does not guarantee an accurate upper bound on the required subdivisions, which we worked around by inflating the estimate for the source segment by a scale factor derived from the linewidth.
@@
     \includegraphics[scale=0.2]{test_renderings}
-    \caption{Renderings of some of our test scenes created using our algorithm and associated frame rates.
+    \caption{Renderings of some of our test scenes created using our algorithm.
 All images were rendered on an Apple M1 Max at 2088x1600 resolution using our GPU pipeline.
-\textbf{(a)} \emph{waves.svg} from the Nehab2020 timings data set (340fps); \textbf{(b)} The \emph{mmark-120k} scene with 120,000 input segments (72fps); \textbf{(c)} The \emph{long path} scene with ~500k input segments and round caps (40fps with CPU-side dashing - the GPU time to render the scene is 9ms); \textbf{(d)}; A simple test scene showcasing various cap and join styles (950fps); \textbf{(e)} The famous \emph{Ghostscript Tiger} containing both stroked and filled paths (810fps).}
+\textbf{(a)} \emph{waves.svg} from the Nehab2020 timings data set (2.94~ms); \textbf{(b)} The \emph{mmark-120k} scene with 120,000 input segments (13.89~ms); \textbf{(c)} The \emph{long path} scene with ~500k input segments and round caps (25~ms including CPU-side dashing - the GPU time to render the scene is 9~ms); \textbf{(d)}; A simple test scene showcasing various cap and join styles (1.05~ms); \textbf{(e)} The famous \emph{Ghostscript Tiger} containing both stroked and filled paths (1.23~ms).}
     \Description{Renderings of the test scenes}

### 2024-06-14 paper.tex
Reviewer #2 feedback (still some remaining)

@@
 Subdividing the arc in to $n$ segments, the distance error for each segment is $\frac{1}{\kappa}(1-\cos \frac{\kappa s}{2n})$.
-Solving for n, we get:
+Solving for $n$, we get:
 
@@
 
-In the principled, correct specification for stroking~[\citenum{Nehab2020}], parallel curves are sufficient only for segments in which the curvature does not exceed the reciprocal half width.
+In the principled, correct specification for stroking~[\citenum{Nehab2020}], parallel curves are sufficient only for segments in which the curvature does not exceed the reciprocal half-width.
 When it does, additional segments must be drawn, including evolutes of the original curve.
@@
 The \emph{lineto} and \emph{curveto} instructions denote an individual \emph{path segment} consisting of either a straight line or a cubic Bézier.
-A given subsequence of path segments can be bounded by a \emph{moveto} and/or \emph{closepath} instruction which mark the beginning and end of a \emph{subpath}.
+The \emph{moveto} instruction is used to begin a new \emph{subpath} and the \emph{closepath} instruction can be used to create a closed contour (by inserting a line connecting the current endpoint to the start of the subpath).
 A subpath must form a closed contour if painted as a fill.
@@
 The entire \citet{Nehab2020} timings set adds up to less than 1 ms on all GPUs except the mobile unit.
-Our algorithm is ~14x slower on the Mali-G78 compared to the M1 Max but it is still capable of processing \emph{waves} in 3 ms.
+Our algorithm is ~14$\times$ slower on the Mali-G78 compared to the M1 Max but it is still capable of processing \emph{waves} in 3~ms.
 We observed that the performance of the kernel scales well with increasing GPU ability even at very large workloads, as demonstrated by the order-of-magnitude difference in run time between the mobile, laptop, and high-end desktop form factors we tested.
-We also confirmed that outputting arcs instead of lines can lead to a 2x decrease in execution time (particularly in scenes with high curve content, like \emph{waves} and \emph{mmark}) as predicted by the lower overall primitive count.
+We also confirmed that outputting arcs instead of lines can lead to a 2$\times$ decrease in execution time (particularly in scenes with high curve content, like \emph{waves} and \emph{mmark}) as predicted by the lower overall primitive count.
 
@@
 Dashing on the GPU was out of scope, so the dash style applied to the \emph{long dash} scene had to be processed on the CPU and dashes were sequantially encoded as individual strokes.
-Including CPU-side dashing, the time to encode \emph{long dash} on the Apple M1 Max was approximately 25 ms on average, while the average time to encode the path after dashing was 4.14 ms.
+Including CPU-side dashing, the time to encode \emph{long dash} on the Apple M1 Max was approximately 25~ms on average, while the average time to encode the path after dashing was 4.14~ms.
 
@@
 For parallel curves, Wang's formula does not guarantee an accurate upper bound on the required subdivisions, which we worked around by inflating the estimate for the source segment by a scale factor derived from the linewidth.
-We observed that enabling estimation increases our CPU pre-processing time by 2-3x on the largest scenes but the impact is overall negligible when the scene size is modest.
+We observed that enabling estimation increases our CPU pre-processing time by 2-3$\times$ on the largest scenes but the impact is overall negligible when the scene size is modest.
 

### 2024-06-14 paper.tex
Update segment count and memory measurements with strong correctness

@@
 All images were rendered on an Apple M1 Max at 2088x1600 resolution using our GPU pipeline.
-\textbf{(a)} \emph{waves.svg} from the Nehab2020 timings data set (2.94~ms); \textbf{(b)} The \emph{mmark-120k} scene with 120,000 input segments (13.89~ms); \textbf{(c)} The \emph{long path} scene with ~500k input segments and round caps (25~ms including CPU-side dashing - the GPU time to render the scene is 9~ms); \textbf{(d)}; A simple test scene showcasing various cap and join styles (1.05~ms); \textbf{(e)} The famous \emph{Ghostscript Tiger} containing both stroked and filled paths (1.23~ms).}
+\textbf{(a)} \emph{waves.svg} from the Nehab2020 timings data set (3.09~ms); \textbf{(b)} The \emph{mmark-120k} scene with 120,000 input segments (19.86~ms); \textbf{(c)} The \emph{long path} scene with ~500k input segments and round caps (25~ms including CPU-side dashing - the GPU time to render the scene is 9~ms); \textbf{(d)}; A simple test scene showcasing various cap and join styles (1.56~ms); \textbf{(e)} The famous \emph{Ghostscript Tiger} containing both stroked and filled paths (1.69~ms).}
     \Description{Renderings of the test scenes}

### 2024-06-14 paper.tex
Address review feedback
I've adopted most of the suggestions and also expanded some sections. The introduction now has additional text regarding dashing.

--- 2024-06-14 paper.tex
@@
 
-Rendering of vector graphics documents requires handling both filled and stroked primitives.
-There is substantial literature on GPU rendering of filled paths, but many fewer published techniques for strokes.
-It is a more challenging problem, especially for parallel computation, because path segments cannot be processed independently of each other; the \emph{joins} between adjacent path segments depend on context.
-In addition, the path topology affects the rendered result, in particular whether subpaths are open or closed.
-In the former case, the endpoints are rendered with \emph{caps}.
-Joins and caps can have multiple styles, including different styles within the same document.
-
-Stroke expansion is the process of generating an outline that, when filled, produces the \emph{stroke} of a given input path.
-Informally, a line of the same width as the stroke is swept along the path, normal to it, and also decorated with joins and caps.
-Though stroke rendering is referenced in many 2D graphics standards, it has lacked a precise formal definition until recent work, primarily \citet{Nehab2020} and \citet{Kilgard2020}.
-
-The \citet{Nehab2020} paper gives a comprehensive survey of techniques for stroke expansion, and an algorithm which, for a number of reasons, is only suitable for implementation on CPU (though adapting it to GPU is listed as promising future work).
-It classifies techniques into \emph{local,} where each path segment generates closed geometry (which may be triangles or other primitives), and \emph{global,} where the overall result is a closed outline of the stroked path, but the partial result from each segment is in general open.
-Our technique is considered global in this scheme, yet allows independent processing of each path segment.
-
-A number of factors contribute to an algorithm being ``GPU-friendly.'' In addition to simply being able to process the input segments in parallel, such an algorithm also avoids divergent control flow (avoiding the need for explicit subdivision at special events such as inflection points and cusps) and uses robust numerical techniques not subject to particular problems when evaluated using 32-bit floating point numbers.
-
+Stroke rendering is an essential and widely part of the standard vector graphics imaging model.
+The standard representation of paths is a sequence of cubic Bézier segments.
+While there are a number of published techniques for GPU rendering of filled paths, stroke rendering is more challenging to parallelize, largely because path segments cannot be processed independently of each other; rendering of the \emph{joins} between path segments depends on both adjoining segments, and the endpoints of paths (when not closed) are rendered with \emph{caps}.
+Joins and caps can be rendered with a variety of styles within the same document.
+The style of stroke rendering also optionally includes \emph{dashing.}
+
+A typical vector graphics document may have thousands of stroked paths.
+A map or CAD drawing may have tens of thousands of paths, possibly consisting of millions of path segments.
+At this scale, the computational cost of stroke rendering on a CPU may result in loss of interactive rendering performance.
+We believe the runtime performance can be improved significantly by offloading stroke rendering to a GPU.
+This requires that the algorithm be GPU-friendly: that it exploit the available parallelism, be work efficient, and avoid unnecessarily complex and divergent control flow.
+It must also remain numerically robust when computed with 32-bit floating pointing numbers.
+
+There are a number of strategies for rendering strokes.
+Among them are \emph{local} techniques that break down the stroke into individual closed primitives, distance field techniques (or similar techniques based on point inclusion).
+In this paper, we focus on \emph{stroke expansion,} the generation of an outline that, when filled, represents the rendered stroke.
+Implementing stroke rendering using stroke expansion enables a unified approach to path rendering downstream, rendering both filled and stroked paths.
+Among other benefits, such an approach avoids a large number of draw calls when stroked and filled paths are finely interleaved, as is often the case.
+
+Correctness is another challenging aspect of stroke rendering.
 We propose that the correctness of stroke outlines be divided into \emph{weak correctness} and \emph{strong correctness.} We define strong correctness as the computation of the outline of a line swept along the segment, maintaining normal orientation, combined with stroke caps and joins.
-Weak correctness, by contrast, only requires the parallel curves of stroke segments, combined with caps and the outer contours of joins.
-The two notions are equivalent for sufficiently well-behaved input, in particular when the curvature measured at endpoints of path segments does not exceed the reciprocal of the half linewidth.
-As described in detail in \citet{Nehab2020}, very few existing implementations actually implement the strong version, so document authors have become accustomed to not depending on behavior at endpoints.
-Standards for graphics formats, such as SVG~[\citenum{Svg2}], also provide an ``out,'' enabling the weaker behavior.
-While our current implementation emphasizes weak correctness, we believe it can be extended to the stronger sense, by implementing evolutes and inner join contours.
-
-A great number of rendering techniques for strokes have been proposed.
-Some are \emph{local,} in that they break the stroke up into smaller closed pieces, the union of which forms an approximation to the stroke.
-Included in this category is polar stroking~[\citenum{Kilgard2020}], which proposes a GPU-friendly subdivision scheme based on an angle step error.
-Others are based on stroke expansion, generating an explicit outline, which is then filled to produce the stroke rendering.
-Such techniques are further divided into \emph{flattening} approaches, where the outline is approximated by a polyline, and \emph{curve-based} approaches, exemplified by the approach in \citet{Nehab2020}.
-In both cases, the outline contains approximations of parallel curves of the input curve segments, and, as proposed in \citet{Nehab2020}, their evolutes when targeting strong correctness.
-For a more detailed survey of existing stroke rendering techniques, the reader is directed to Section 4 of \citet{Nehab2020}.
-
-The subproblem of approximating parallel curves has also spawned extensive literature, none of which is entirely satisfying, especially when it comes to algorithms that can be efficiently evaluated on GPU.
+Weak correctness, by contrast, only requires the parallel curves (also known as ``offset curves'') of path segments, combined with caps and the outer contours of joins.
+Both \citet{Nehab2020} and \citet{Kilgard2020} provide useful definitions of strongly correct stroke rendering, and Nehab in particular describes how to achieve it in stroke expansion.
+Briefly, when the path curvature exceeds the reciprocal of the stroke half-width, then addition \emph{evolute} segments need to be added, in addition to the parallel curves required by weak correctness, and also the inner contours of joins are required in general.
+The technique described in this paper is intended for strongly correct stroke rendering; it produces outlines that render the same shape as \citet{Nehab2020}.
+
+Correct stroke rendering also relies on accurate approximation of the parallel curves of the input paths.
+Determining the parallel curve of a cubic Bézier is not analytically tractable, as it is a tenth order algebraic formula[\citenum{Farouki1990}].
+Thus, all practical stroke rendering techniques employ an approximation, with error tolerance as a tunable parameter, typically representing the maximum allowable Fréchet distance between the true curve and its approximation.
+A typical value is 0.25 device pixel, as that is close to the threshold of visibility.
+A correct algorithm will produce an approximate curve within the error tolerance, but an efficient algorithm will not subdivide significantly more than is necessary to meet the error bound.
+Some specifications for vector graphics, including PDF[\citenum{PDF2008}], specify the error tolerance explicitly.
+
+The subproblem of approximating parallel curves has spawned extensive literature, none of which is entirely satisfying, especially when it comes to algorithms that can be efficiently evaluated on GPU.
 An example of a reasonably good algorithm that computes flattened parallel curves is \citet{Yzerman2020}, as used in the Blend2D rendering library.
-For producing curved outlines, the Tiller and Hanson~[\citenum{Tiller1984}] algorithm is commonly implemented and cited, but its performance is quite poor when applied to cubic Béziers.
+For producing curved outlines, the Tiller and Hanson[\citenum{Tiller1984}] algorithm is commonly implemented and cited, but its performance is quite poor when applied to cubic Béziers.
 The quadratic Bézier version is adequate, and forms the basis of the algorithm in \citet{Nehab2020}.
-A variant is also used in Skia~[\citenum{Skia}]; instead of lowering the cubic Bézier to a quadratic approximation and then computing the parallel curve, Skia approximates the parallel curve directly with a quadratic Bézier, then measures the error to determine whether further subdivision is necessary.
-
-This paper presents a comprehensive solution to stroke expansion, well suited to GPU implementation.
+A variant is also used in Skia[\citenum{Skia}]; instead of lowering the cubic Bézier to a quadratic approximation and then computing the parallel curve, Skia approximates the parallel curve directly with a quadratic Bézier, then measures the error to determine whether further subdivision is necessary.
+
+Previous work on stroke rendering does not fully solve the problem of fast and correct stroke rendering.
+Most existing implementations have correctness problems, in particular failing to draw the evolutes and inner joins as required for strong correctness; \citet{Nehab2020} provides a detailed survey.
+In addition, most implementations are not suitable for running on GPU, requiring sequential execution to produce watertight outlines.
+One GPU implementation is polar stroking (\citet{Kilgard2020}), a local technique, but among other limitations, it does not bound the Fréchet distance error (rather, it bounds the angle error).
+
+Our central approach bounding the error tolerance is \emph{error metrics,} which predict the Fréchet distance error.
+The error metrics presented in this paper are both straightforward to compute, and also \emph{invertible}, which avoids excessive subdivision and also increases the amount of available parallelism to exploit.
+Previous approaches to error measurement generally follow a ``cut then measure'' approach, where the approximate curve is generated, then error measured by sampling.
+Such sampling is often time consuming, involving iteration, and even then risks underestimating the error due to inadequate sampling.
+
+The style of stroke rendering also optionally includes \emph{dashing.} Previous work, particularly \citet{Kilgard2020}, has explained how to break the input path into dash segments of the specified length on GPU using a prefix sum of segment arc lengths.
+There are implementation tradeoffs to GPU-based dash processing, including potentially a larger number of dispatch stages, and we were able to achieve acceptable performance with the dash processing on CPU.
+Thus, whether moving dashing to a GPU stage is a performance win depends on the scene and other factors in the overall rendering pipeline.
+Even so, for scenes that have a very large number of dashed strokes, we would expect further performance improvement from GPU-based dashing.
+
+Previous work, particularly \citet{Kilgard2020}, has explained how to process dashing patterns on GPU using a prefix sum of segment arc lengths.
+There are implementation tradeoffs to GPU-based dash processing, including potentially a larger number of dispatch stages, and we were able to achieve acceptable performance with the dash processing on CPU.
+Thus, whether moving dashing to a GPU stage is a performance win depends on the scene and other factors in the overall rendering pipeline.
+Even so, for scenes that have a very large number of dashed strokes, we would expect further performance improvement from GPU-based dashing.
+
+This paper presents a comprehensive solution to the core problem of stroke expansion, well suited to GPU implementation.
 It can produce both flattened polylines or an approximation consisting of circular arcs.
@@
     \includegraphics[scale=0.3]{es_flatten}
-    \caption{Flattening of an Euler spiral using points spaced by power-law}
-    \Description{An Euler spiral along with its flattened representation. The flattened curve has an error}
+    \caption{The flattened approximation to an Euler spiral using line segments is achieved by choosing subdivision points using a simple power law formula with an exponent of 2/3.}
+    \Description{An Euler spiral along with its flattened representation. The flattened curve has a consistent error, and the subdivision points are spaced according to a power law.}
     \label{fig:es_flatten}
@@
 
-The parallel curve of the Euler spiral (also known as ``clothoid'') was characterized by \citet{Wieleitner1907} well over a hundred years ago~[\citenum{Wieleitner1907}], and has a straightforward closed-form Cesàro representation, curvature as a function of arc length.
+The parallel curve of the Euler spiral (also known as ``clothoid'') was characterized by \citet{Wieleitner1907} well over a hundred years ago, and has a straightforward closed-form Cesàro representation, curvature as a function of arc length, of the form $\kappa(s) = c_0{s - s_0}^{-1/2} + c_1$.
 
@@
 Using a subpixel Fréchet distance bound guarantees that the rendered image does not deviate visibly from the exact rendering.
-Another choice would be uniform steps in tangent angle, as chosen by polar stroking~[\citenum{Kilgard2020}].
+Another choice would be uniform steps in tangent angle, as chosen by polar stroking[\citenum{Kilgard2020}].
 However, at small curvature, the stroked path can be off by several pixels, and at large curvature there may be considerably more subdivision than needed for faithful rendering.
@@
 The bottom row shows a consistent angle step, as implemented in polar stroking, but has excessive distance error at low curvature, and excessive subdivision at high curvature.
-It should be noted, to avoid the undershoot at low curvature, both the Skia~[\citenum{Skia}] and Rive~[\citenum{Rive}] renderers use a hybrid of the Wang and polar stroking error metrics.
+It should be noted, to avoid the undershoot at low curvature, both the Skia[\citenum{Skia}] and Rive[\citenum{Rive}] renderers use a hybrid of the Wang and polar stroking error metrics.
 
@@
 
-\subsection{Optimal flattening}
-
-The optimal flattening of a curve contains the minimum number of subdivisions required to satisfy a maximum error constraint, in which the error is the minimum possible with that number of subdivisions.
-
-For a convex curve, the optimal flattening is tractable to compute, though not necessarily very fast.
+%\subsection{Optimal flattening}
+%
+%The optimal flattening of a curve contains the minimum number of subdivisions required to satisfy a maximum error constraint, in which the error is the minimum possible with that number of subdivisions.
+%
+%For a convex curve, the optimal flattening is tractable to compute, though not necessarily very fast.
 When a curve is monotonic, error is also monotonic with respect to both subdivision points.
-Thus, any standard root finding technique can find a subdivision with a given error (bisection is the best known, but the ITP method \citet{Oliveira2020} is better).
+Thus, any standard root finding technique can find a subdivision with a given error (bisection is the most popular, but the ITP method[\citenum{Oliveira2020}] is better).
 A similar outer loop with a root finding technique can find the error which minimizes the global error for the entire flattening, which, due to monotonicity, is that for which the error of the final segment equals that of the previous segments.
@@
 
-In the principled, correct specification for stroking~[\citenum{Nehab2020}], parallel curves are sufficient only for segments in which the curvature does not exceed the reciprocal half width.
+In the principled, correct specification for stroking[\citenum{Nehab2020}], parallel curves are sufficient only for segments in which the curvature does not exceed the reciprocal half width.
 When it does, additional segments must be drawn, including evolutes of the original curve.
 In general, the evolute of a cubic Bézier is a very complex curve, requiring approximation techniques.
-By contrast, the evolute of an Euler spiral ($\kappa = as$) is another spiral with a simple Cesàro equation, namely $\kappa = -a^{-1}s^{-3}$, an instance of the general result that the evolute of a log-aesthetic curve is another log-aesthetic curve~[\citenum{Yoshida2012}].
+By contrast, the evolute of an Euler spiral ($\kappa = as$) is another spiral with a simple Cesàro equation, namely $\kappa = -a^{-1}s^{-3}$, an instance of the general result that the evolute of a log-aesthetic curve is another log-aesthetic curve[\citenum{Yoshida2012}].
 
@@
 
-Many techniques for stroke expansion described in the literature apply some lowering of cubic Bézier curves to a simpler curve type, more tractable for evaluating parallel curve.
+Many techniques for stroke expansion described in the literature apply some lowering of cubic Bézier curves to a simpler curve type that is more tractable for evaluating parallel curve.
 Computing parallel curves directly on cubic Bézier curve segments is not very tractable.
-In particular, the widely cited Tiller-Hanson algorithm~[\citenum{Tiller1984}] performs well for quadratic Béziers but significantly worse for cubics.
+In particular, the widely cited Tiller-Hanson algorithm[\citenum{Tiller1984}] performs well for quadratic Béziers but significantly worse for cubics.
 
@@
 Given tangent angles relative to the chord, finding the Euler spiral segment that minimizes total curvature variation is a form of geometric Hermite interpolation.
-There are a number of published solutions to this problem, involving nontrivial numerical solving techniques: gradient descent~[\citenum{Kimia2003}], bisection~[\citenum{Walton2009}], or Newton iteration~[\citenum{Connor2014}].
+There are a number of published solutions to this problem, involving nontrivial numerical solving techniques: gradient descent[\citenum{Kimia2003}], bisection[\citenum{Walton2009}], or Newton iteration[\citenum{Connor2014}].
 A more direct approach is to approximate the function as a reasonably low-order polynomial in terms of the endpoint angles.
@@
 
-The entire recursion stack for adaptive subdivision can be represented in two words: dt and scaled\_t0.
-The range is dt * scaled\_t0 ..
-dt * (scaled\_t0 + 1).
-Initial values of 1 and 0 represent the range (0, 1).
-Pushing the stack, subdividing the range in half, is represented by halving dt and doubling scaled\_t0, which leaves the start of the range invariant but its size halved.
-After accepting an approximation, the next range is determined by incrementing scaled\_t0, then repeatedly popping the stack by doubling dt and halving scaled\_t0, as long as the latter is odd.
-This can in fact be achieved without iteration by using the ``count trailing zeros'' intrinsic.
+The entire recursion stack for adaptive subdivision can be represented in two words: $dt$ and $t0\_u$ (the latter being an unsigned integer).
+The range is $dt \cdot t0\_u ..
+dt \cdot (t0\_u + 1)$.
+Initial values of $1$ and $0$ represent the range $(0, 1)$.
+Pushing the stack, subdividing the range in half, is represented by halving $dt$ and doubling $t0\_u$, which leaves the start of the range invariant but its size halved.
+After accepting an approximation, the next range is determined by incrementing $t0\_u$.
+At this point, the decision of whether to pop the stack or continue is dependent on whether $t0\_u$ is even or odd, respectively.
+If even, popping the stack is accomplished by doubling $dt$ and halving $t0\_u$, and this is repeated until $t0\_u$ becomes odd.
+The repeated stack popping can be accomplished without iteraction by using the ``count trailing zeros'' intrinsic to determine the number of stack pops $n$, then multiplying $dt$ by $2^n$, and shifting $t0\_u$ right $n$ bits.
 
@@
     \includegraphics[scale=0.19]{stroke_stills}
-    \caption{Stroked cubic Béziers exhibiting strong and weak correctness, rendered by our implementation.
-The contents of the line soup buffer are visualized as green and blue outlines.
+    \caption{Renderings of stroked cubic Béziers exhibiting strong and weak correctness.
+The outlines generated by our GPU stroking compute shader are visualized in green and blue.
 Segments with a winding direction that points \emph{up} are colored blue while those pointing \emph{down} are colored green.
-The curves were adapted from the \emph{stills} dataset by \citet{Nehab2020} and the Skia test corpus.}
+The curves were adapted to from the \emph{stills} dataset by \citet{Nehab2020} and the Skia test corpus.}
     \Description{Selection of stroked cubic Béziers exhibiting strong and weak correctness}
@@
 Notably, our implementation is fully data-parallel on input path segments and does not require any computationally expensive curve evaluation on the CPU.
-We implemented our shader on general-purpose GPU compute primitives supported by all modern graphics APIs, particularly Metal, Vulkan, D3D12, and WebGPU~[\citenum{WebGPU}].
+We implemented our shader on general-purpose GPU compute primitives supported by all modern graphics APIs, particularly Metal, Vulkan, D3D12, and WebGPU[\citenum{WebGPU}].
 As such, the ideas presented in this section are portable to various GPU platforms.
@@
 
-An SVG path consists of a sequence of instructions (or ``verbs'').
+An SVG path consists of a sequence of instructions (or "verbs").
 The \emph{lineto} and \emph{curveto} instructions denote an individual \emph{path segment} consisting of either a straight line or a cubic Bézier.
-A given subsequence of path segments can be bounded by a \emph{moveto} and/or \emph{closepath} instruction which mark the beginning and end of a \emph{subpath}.
+A given subsequence of path segments can be bounded by a \emph{moveto} and/or \emph{closepath} instruction that marks the beginning and end of a \emph{subpath}.
 A subpath must form a closed contour if painted as a fill.
@@
     TOL $\gets$ Euler spiral fit tolerance\;
-    \tcp{`t0\_u' and `dt' track the recursion stack}
-    t0\_u $\gets$ 0u\;
+    \tcp{'t0\_u' and 'dt' track the recursion stack}
+    t0\_u $\gets$ 0\;
     dt $\gets$ 1.0\;
@@
             \tcp{Find location of any cusp in outline.}
-            \tcp{If $0 \leq t < 1$, draw evolute at `t'.}
+            \tcp{If $0 \leq t < 1$, draw evolute at 't'.}
             cusp0 $\gets$ es\_curvature(es, 0) * offset + chord\_len\;
@@
 If the error exceeds the desired tolerance threshold, the cubic segment is iteratively subdivided in half and a new error estimate is computed.
-Once the error satisfies the tolerance, the Euler spiral segment is directly flattened to a polyline by invoking a subroutine that is parameterized by the stroke half-width (see \texttt{es\_seg\_flatten\_offset} in Algoritm~\ref{alg:flatten-cubic}).
+Once the error satisfies the tolerance, the Euler spiral segment is directly flattened to a polyline (or, alternately, converted to a sequence of arc segments) by invoking a subroutine that is parameterized by the stroke halfwidth (see \texttt{es\_seg\_flatten\_offset} in Algoritm~\ref{alg:flatten-cubic}).
 For a regular fill, this parameter is set to $0$.
@@
 If the outline contains a cusp -- which can be determined by evaluating the curvature variation along the Euler spiral segment -- we output an evolute patch (as shown in Figure~\ref{fig:evolutes}) in order to achieve strong correctness.
-A selection of renderings produced by our algorithm that demonstrate both strong and weak correctness is shown in Figure~\ref{fig:stroke_stills}.
+A selection of renderings produced by our algorithm that demonstrate both strong and weak correctness are shown in Figure~\ref{fig:stroke_stills}.
 
@@
 An application may want to sort the line soup, depending on the requirements of the chosen rendering algorithm.
-For instance, our own renderer employs tile-based scanline rasterization in a compute shader, with a pipeline architecture that resembles \emph{cudaraster}~[\citenum{Laine2011}] and \emph{MPVG}~[\citenum{Ganacim2014}].
+For instance, our own renderer employs tile-based scanline rasterization in a compute shader, with a pipeline architecture that resembles \emph{cudaraster}[\citenum{Laine2011}] and \emph{MPVG}[\citenum{Ganacim2014}].
 In this design, the line soup elements are spatially sorted by subsequent compute stages into a grid of 16x16 pixel tiles.
@@
     \caption{We integrated our curve flattening shader into our own renderer implemented entirely in compute shaders.
-The flattening stage (described in Section~\ref{section:gpu-impl}) converts stroked paths to filled outlines and flattens them to an unordered soup of line segments.
+The flattening stage (described in Section~\ref{section:gpu-impl}) converts stroked paths to filled outlines and flattens them into an unordered soup of line segments.
 Our renderer spatially sorts the line soup into tiles before computing pixel coverage using a highly parallel scanline rasterizer.}
@@
 We evaluated the runtime performance of our compute shader on 4 GPU models and multiple form factors: Google Pixel 6 equipped with the Arm \emph{Mali-G78 MP20} mobile GPU, the Apple \emph{M1 Max} integrated laptop GPU, and two discrete desktop GPUs: the mid-2015 NVIDIA {GeForce GTX 980Ti} and the late-2022 NVIDIA {GeForce RTX 4090}.
-We authored our entire GPU pipeline in the \emph{WebGPU Shading Language} (WGSL) and used the \emph{wgpu}~[\citenum{wgpu}] framework to interoperate with the native graphics APIs (we tested the M1 Max on macOS via the Metal API; the mobile and desktop GPUs were tested via the Vulkan API on Android and Ubuntu systems).
+We authored our entire GPU pipeline in the \emph{WebGPU Shading Language} (WGSL) and used the \emph{wgpu}[\citenum{wgpu}] framework to interoperate with the native graphics APIs (we tested the M1 Max on macOS via the Metal API; the mobile and desktop GPUs were tested via the Vulkan API on Android and Ubuntu systems).
 
@@
 
-We authored two additional test scenes in order stress the GPU with a higher workload.
+We authored two additional test scenes in order to stress the GPU with a higher workload.
 The first is a very large stroked path with ~500,000 individually styled dashed segments which we adapted from Skia's open-source test corpus.
 We used two variants of this scene with \emph{butt} and \emph{round} cap styles.
-For the second scene we adapted the \emph{Canvas Paths} test case from MotionMark~[\citenum{MotionMark}], which renders a large number of randomly generated stroked line, quadratic, and cubic Bézier paths.
+For the second scene we adapted the \emph{Canvas Paths} test case from MotionMark[\citenum{MotionMark}], which renders a large number of randomly generated stroked line, quadratic, and cubic Bézier paths.
 We made two versions with different sizes: mmark-70k and mmark-120k with 70,000 and 120,000 path segments, respectively.
@@
 The hatched bars show the timings for the version that outputs line primitives while the solid bars show that for arcs.
-Mali-G78 (the lowest end GPU we tested) converts \emph{waves} (the largest scene in the set) to lines in ~3ms.
+Mali-G78 (the lowest-end GPU we tested) converts \emph{waves} (the largest scene in the set) to lines in about ms.
 The compute shader executes at least an order of magnitude faster on the other GPUs.}
@@
 The CPU encoding times (prior to the GPU prefix sums) for some of the large scenes is shown in Table~\ref{table:bump-estimate-cost}.
-Dashing on the GPU was out of scope, so the dash style applied to the \emph{long dash} scene had to be processed on the CPU and dashes were sequantially encoded as individual strokes.
+Our implementation applies dashing on the CPU.
+It is conceptually fairly straightforwad to implement on GPU; the core algorithm is prefix sum applied to segment length, and the Euler spiral representation is particularly well suited.
+However, there are complex implementation tradeoffs, so the dash style applied to the \emph{long dash} scene had to be processed on the CPU and dashes were sequentially encoded as individual strokes.
 Including CPU-side dashing, the time to encode \emph{long dash} on the Apple M1 Max was approximately 25 ms on average, while the average time to encode the path after dashing was 4.14 ms.
@@
 We augmented our encoding logic to compute a conservative estimate of the required output buffer size based on an upper bound on the number of line primitives per input path segment.
-We found that Wang's formula~[\citenum{Goldman2003}] works well in practice, as it is relatively inexpensive to evaluate, though it results in a higher memory estimate than required (see Table~\ref{table:bump-estimate-mem}).
+We found that Wang's formula[\citenum{Goldman2003}] works well in practice, as it is relatively inexpensive to evaluate, though it results in a higher memory estimate than required (see Table~\ref{table:bump-estimate-mem}).
 For parallel curves, Wang's formula does not guarantee an accurate upper bound on the required subdivisions, which we worked around by inflating the estimate for the source segment by a scale factor derived from the linewidth.
-We observed that enabling estimation increases our CPU pre-processing time by 2-3x on the largest scenes but the impact is overall negligible when the scene size is modest.
+We observed that enabling estimation increases our CPU pre-processing time by 2--3x on the largest scenes but the overall impact is negligible when the scene size is modest.
 
@@
 Our estimation uses Wang's formula combined with a scaling heuristic based on the linewidth.
-In the worst case, our buffer size estimator overshoots by a factor ~4.}
+In the worst case, our buffer size estimator overshoots by a factor of 4.}
     \begin{tabular}{|l|c|c|}
@@
     \includegraphics[scale=0.2]{test_renderings}
-    \caption{Renderings of some of our test scenes created using our algorithm.
+    \caption{Renderings of some of our test scenes created using our algorithm and associated frame rates.
 All images were rendered on an Apple M1 Max at 2088x1600 resolution using our GPU pipeline.
-\textbf{(a)} \emph{waves.svg} from the Nehab2020 timings data set (2.94~ms); \textbf{(b)} The \emph{mmark-120k} scene with 120,000 input segments (13.89~ms); \textbf{(c)} The \emph{long path} scene with ~500k input segments and round caps (25~ms including CPU-side dashing - the GPU time to render the scene is 9~ms); \textbf{(d)}; A simple test scene showcasing various cap and join styles (1.05~ms); \textbf{(e)} The famous \emph{Ghostscript Tiger} containing both stroked and filled paths (1.23~ms).}
+\textbf{(a)} \emph{waves.svg} from the Nehab2020 timings data set (340fps); \textbf{(b)} The \emph{mmark-120k} scene with 120,000 input segments (72fps); \textbf{(c)} The \emph{long path} scene with ~500k input segments and round caps (40fps with CPU-side dashing - the GPU time to render the scene is 9ms); \textbf{(d)}; A simple test scene showcasing various cap and join styles (950fps); \textbf{(e)} The famous \emph{Ghostscript Tiger} containing both stroked and filled paths (810fps).}
     \Description{Renderings of the test scenes}
@@
     \item The GPU implementation sequentializes some work that could be parallel, and is thus not as well load-balanced as it might be.
-An appealing future direction is to split the pipeline into separate stages, executed by the GPU as nodes in a work graph~[\citenum{Patel2024}].
+An appealing future direction is to split the pipeline into separate stages, executed by the GPU as nodes in a work graph[\citenum{Patel2024}].
 This structure is appealing because load balancing is done by hardware.

### 2024-06-14 paper.tex
Fixes to code formatting and wording

--- 2024-06-14 paper.tex
@@
       }{
-          \tcp{Output both offset curves}
+          \tcp{Output parallel curves}
           flatten\_cubic(segment,  style.offset)\;
@@
             \tcp{Pop the stack}
-            t0\_u $\gets$ t0\_u + 1u\;
+            t0\_u $\gets$ t0\_u + 1\;
             shift $\gets$ countTrailingZeros(t0\_u)\;
-            t0\_u $\gets$ t0\_u >> shift;
-            dt $\gets$ dt * f32(1u << shift)\;
+            t0\_u $\gets$ t0\_u $\gg$ shift\;
+            dt $\gets$ dt * f32(1 $\ll$ shift)\;
             last\_p $\gets$ next\_p\;

### 2024-06-14 paper.tex
textasciitilde and fixes the bump memory table caption

@@
 We authored two additional test scenes in order stress the GPU with a higher workload.
-The first is a very large stroked path with ~500,000 individually styled dashed segments which we adapted from Skia's open-source test corpus.
+The first is a very large stroked path with \textasciitilde500,000 individually styled dashed segments which we adapted from Skia's open-source test corpus.
 We used two variants of this scene with \emph{butt} and \emph{round} cap styles.
@@
 The entire \citet{Nehab2020} timings set adds up to less than 1 ms on all GPUs except the mobile unit.
-Our algorithm is ~14$\times$ slower on the Mali-G78 compared to the M1 Max but it is still capable of processing \emph{waves} in 3~ms.
+Our algorithm is \textasciitilde14$\times$ slower on the Mali-G78 compared to the M1 Max but it is still capable of processing \emph{waves} in 3~ms.
 We observed that the performance of the kernel scales well with increasing GPU ability even at very large workloads, as demonstrated by the order-of-magnitude difference in run time between the mobile, laptop, and high-end desktop form factors we tested.
@@
 Our estimation uses Wang's formula combined with a scaling heuristic based on the linewidth.
-In the worst case, our buffer size estimator overshoots by a factor ~4.}
+The estimate is conservative and always overshoots the actual memory requirement.}
     \begin{tabular}{|l|c|c|}
@@
 All images were rendered on an Apple M1 Max at 2088x1600 resolution using our GPU pipeline.
-\textbf{(a)} \emph{waves.svg} from the Nehab2020 timings data set (3.09~ms); \textbf{(b)} The \emph{mmark-120k} scene with 120,000 input segments (19.86~ms); \textbf{(c)} The \emph{long path} scene with ~500k input segments and round caps (25~ms including CPU-side dashing - the GPU time to render the scene is 9~ms); \textbf{(d)}; A simple test scene showcasing various cap and join styles (1.56~ms); \textbf{(e)} The famous \emph{Ghostscript Tiger} containing both stroked and filled paths (1.69~ms).}
+\textbf{(a)} \emph{waves.svg} from the Nehab2020 timings data set (3.09~ms); \textbf{(b)} The \emph{mmark-120k} scene with 120,000 input segments (19.86~ms); \textbf{(c)} The \emph{long path} scene with \textasciitilde500k input segments and round caps (25~ms including CPU-side dashing -- the GPU time to render the scene is 9~ms); \textbf{(d)}; A simple test scene showcasing various cap and join styles (1.56~ms); \textbf{(e)} The famous \emph{Ghostscript Tiger} containing both stroked and filled paths (1.69~ms).}
     \Description{Renderings of the test scenes}

### 2024-06-14 paper.tex

--- 2024-06-14 paper.tex
@@
 Correct stroke rendering also relies on accurate approximation of the parallel curves of the input paths.
-Determining the parallel curve of a cubic Bézier is not analytically tractable, as it is a tenth order algebraic formula[\citenum{Farouki1990}].
+Determining the parallel curve of a cubic Bézier is not analytically tractable, as it is a tenth order algebraic formula~[\citenum{Farouki1990}].
 Thus, all practical stroke rendering techniques employ an approximation, with error tolerance as a tunable parameter, typically representing the maximum allowable Fréchet distance between the true curve and its approximation.
@@
 
-We authored two additional test scenes in order to stress the GPU with a higher workload.
-The first is a very large stroked path with ~500,000 individually styled dashed segments which we adapted from Skia's open-source test corpus.
+We authored two additional test scenes in order stress the GPU with a higher workload.
+The first is a very large stroked path with \textasciitilde500,000 individually styled dashed segments which we adapted from Skia's open-source test corpus.
 We used two variants of this scene with \emph{butt} and \emph{round} cap styles.
-For the second scene we adapted the \emph{Canvas Paths} test case from MotionMark[\citenum{MotionMark}], which renders a large number of randomly generated stroked line, quadratic, and cubic Bézier paths.
+For the second scene we adapted the \emph{Canvas Paths} test case from MotionMark~[\citenum{MotionMark}], which renders a large number of randomly generated stroked line, quadratic, and cubic Bézier paths.
 We made two versions with different sizes: mmark-70k and mmark-120k with 70,000 and 120,000 path segments, respectively.
@@
 The hatched bars show the timings for the version that outputs line primitives while the solid bars show that for arcs.
-Mali-G78 (the lowest-end GPU we tested) converts \emph{waves} (the largest scene in the set) to lines in about ms.
+Mali-G78 (the lowest-end GPU we tested) converts \emph{waves} (the largest scene in the set) to lines in \textasciitilde3 ms.
 The compute shader executes at least an order of magnitude faster on the other GPUs.}
@@
 Our implementation applies dashing on the CPU.
-It is conceptually fairly straightforwad to implement on GPU; the core algorithm is prefix sum applied to segment length, and the Euler spiral representation is particularly well suited.
+It is conceptually fairly straightforward to implement on GPU; the core algorithm is prefix sum applied to segment length, and the Euler spiral representation is particularly well suited.
 However, there are complex implementation tradeoffs, so the dash style applied to the \emph{long dash} scene had to be processed on the CPU and dashes were sequentially encoded as individual strokes.
-Including CPU-side dashing, the time to encode \emph{long dash} on the Apple M1 Max was approximately 25 ms on average, while the average time to encode the path after dashing was 4.14 ms.
+Including CPU-side dashing, the time to encode \emph{long dash} on the Apple M1 Max was approximately 25~ms on average, while the average time to encode the path after dashing was 4.14~ms.
 
@@
 We augmented our encoding logic to compute a conservative estimate of the required output buffer size based on an upper bound on the number of line primitives per input path segment.
-We found that Wang's formula[\citenum{Goldman2003}] works well in practice, as it is relatively inexpensive to evaluate, though it results in a higher memory estimate than required (see Table~\ref{table:bump-estimate-mem}).
+We found that Wang's formula~[\citenum{Goldman2003}] works well in practice, as it is relatively inexpensive to evaluate, though it results in a higher memory estimate than required (see Table~\ref{table:bump-estimate-mem}).
 For parallel curves, Wang's formula does not guarantee an accurate upper bound on the required subdivisions, which we worked around by inflating the estimate for the source segment by a scale factor derived from the linewidth.
-We observed that enabling estimation increases our CPU pre-processing time by 2--3x on the largest scenes but the overall impact is negligible when the scene size is modest.
+We observed that enabling estimation increases our CPU pre-processing time by 2--3$\times$ on the largest scenes but the impact is overall negligible when the scene size is modest.
 
@@
 Our estimation uses Wang's formula combined with a scaling heuristic based on the linewidth.
-In the worst case, our buffer size estimator overshoots by a factor of 4.}
+The estimate is conservative and always overshoots the actual memory requirement.}
     \begin{tabular}{|l|c|c|}

### 2024-06-14 paper.tex
Fix editing glitches
Fix some stuff that went wrong while merging.

@@
 The error metrics presented in this paper are both straightforward to compute, and also \emph{invertible}, which avoids excessive subdivision and also increases the amount of available parallelism to exploit.
-Previous approaches to error measurement generally follow a ``cut then measure'' approach, where the approximate curve is generated, then error measured by sampling.
+Previous approaches to error measurement generally follow a ``cut then measure'' approach, where the approximate curve is generated, then the error measured by sampling.
 Such sampling is often time consuming, involving iteration, and even then risks underestimating the error due to inadequate sampling.
 
-The style of stroke rendering also optionally includes \emph{dashing.} Previous work, particularly \citet{Kilgard2020}, has explained how to break the input path into dash segments of the specified length on GPU using a prefix sum of segment arc lengths.
-There are implementation tradeoffs to GPU-based dash processing, including potentially a larger number of dispatch stages, and we were able to achieve acceptable performance with the dash processing on CPU.
-Thus, whether moving dashing to a GPU stage is a performance win depends on the scene and other factors in the overall rendering pipeline.
-Even so, for scenes that have a very large number of dashed strokes, we would expect further performance improvement from GPU-based dashing.
+Regarding dashing, previous work, particularly \citet{Kilgard2020}, has explained how to break the input path into dash segments of the specified length on GPU using a prefix sum of segment arc lengths.
+There are implementation tradeoffs to GPU-based dash processing, including potentially a larger number of dispatch stages.
+Whether moving dashing to a GPU stage is a performance win depends on the scene and other factors in the overall rendering pipeline.
+While we would expect further performance improvement from GPU-based dashing, we were able to achieve acceptable performance with dash processing on CPU.
 
@@
 If even, popping the stack is accomplished by doubling $dt$ and halving $t0\_u$, and this is repeated until $t0\_u$ becomes odd.
-The repeated stack popping can be accomplished without iteraction by using the ``count trailing zeros'' intrinsic to determine the number of stack pops $n$, then multiplying $dt$ by $2^n$, and shifting $t0\_u$ right $n$ bits.
+The repeated stack popping can be accomplished without iteraction by using the ``count trailing zeros'' intrinsic to determine the number of stack pops $n$, then multiplying $dt$ by $2^n$, and shifting $t0\_u$ right by $n$ bits.
 
@@
 
-We authored two additional test scenes in order stress the GPU with a higher workload.
+We authored two additional test scenes in order to stress the GPU with a higher workload.
 The first is a very large stroked path with \textasciitilde500,000 individually styled dashed segments which we adapted from Skia's open-source test corpus.

### 2024-06-14 paper.tex
Other edits
Some of these were suggestsions from before I didn't apply, some are new comments.

@@
 Both \citet{Nehab2020} and \citet{Kilgard2020} provide useful definitions of strongly correct stroke rendering, and Nehab in particular describes how to achieve it in stroke expansion.
-Briefly, when the path curvature exceeds the reciprocal of the stroke half-width, then addition \emph{evolute} segments need to be added, in addition to the parallel curves required by weak correctness, and also the inner contours of joins are required in general.
-The technique described in this paper is intended for strongly correct stroke rendering; it produces outlines that render the same shape as \citet{Nehab2020}.
+Briefly, when the path curvature exceeds the reciprocal of the stroke half-width, \emph{evolute} segments and the \emph{inner contours of joins} are required in addition to the parallel curves.
+The technique described in this paper produces strongly correct outlines according to this definition.
 
@@
 For parallel curves, Wang's formula does not guarantee an accurate upper bound on the required subdivisions, which we worked around by inflating the estimate for the source segment by a scale factor derived from the linewidth.
-We observed that enabling estimation increases our CPU pre-processing time by 2--3$\times$ on the largest scenes but the impact is overall negligible when the scene size is modest.
+We observed that enabling estimation increases our CPU pre-processing time by 2--3$\times$ on the largest scenes but overall the impact is negligible when the scene size is modest.
 

### 2024-06-14 paper.tex
Edits around dashing
- More explicitly state that we did not fully explore the solution
  space.
- Remove repeated text an the introduction.

@@
 Joins and caps can be rendered with a variety of styles within the same document.
-The style of stroke rendering also optionally includes \emph{dashing.}
+The style of stroke rendering also optionally includes \emph{dashing}, which applies a pattern of dashes and gaps to the outline of the shape.
 
@@
 The technique described in this paper produces strongly correct outlines according to this definition.
+Examples of weakly and strongly correct rendering are shown in Figure~\ref{fig:weak_strong}.
+
+\begin{figure}
+    \includegraphics[scale=0.5]{weak_strong}
+    \caption{Example of weakly correct (top) and strongly correct (bottom) stroke rendering.
+The left example has a curvature cusp and requires evolutes to render correctly, while the right requires the inner join.}
+    \Description{Two examples of strokes.
+The top left is missing an evolute section, making it appear that a half-disc has been carved out.
+The top right is missing inner joins, looking like spokes rather than a connected path.
+The bottom two examples are rendered with the full strokes.}
+    \label{fig:weak_strong}
+\end{figure}
 
@@
 
-Regarding dashing, previous work, particularly \citet{Kilgard2020}, has explained how to break the input path into dash segments of the specified length on GPU using a prefix sum of segment arc lengths.
+Previous work, particularly \citet{Kilgard2020}, has explained how to break the input path into dash segments of the specified length on GPU using a prefix sum of segment arc lengths.
 There are implementation tradeoffs to GPU-based dash processing, including potentially a larger number of dispatch stages.
-Whether moving dashing to a GPU stage is a performance win depends on the scene and other factors in the overall rendering pipeline.
-While we would expect further performance improvement from GPU-based dashing, we were able to achieve acceptable performance with dash processing on CPU.
-
-Previous work, particularly \citet{Kilgard2020}, has explained how to process dashing patterns on GPU using a prefix sum of segment arc lengths.
-There are implementation tradeoffs to GPU-based dash processing, including potentially a larger number of dispatch stages, and we were able to achieve acceptable performance with the dash processing on CPU.
-Thus, whether moving dashing to a GPU stage is a performance win depends on the scene and other factors in the overall rendering pipeline.
-Even so, for scenes that have a very large number of dashed strokes, we would expect further performance improvement from GPU-based dashing.
+We anticipate that the potential performance benefit of implementing dashing on a GPU depends on the scene and other factors in the overall rendering pipeline.
+While we would expect further performance improvement from GPU-based dashing, we did not fully explore the solution space and we were able to achieve acceptable performance with dash processing on CPU.
 
@@
 Our implementation applies dashing on the CPU.
-It is conceptually fairly straightforward to implement on GPU; the core algorithm is prefix sum applied to segment length, and the Euler spiral representation is particularly well suited.
-However, there are complex implementation tradeoffs, so the dash style applied to the \emph{long dash} scene had to be processed on the CPU and dashes were sequentially encoded as individual strokes.
+It is conceptually straightforward to implement on GPU; the core algorithm is prefix sum applied to segment length, and the Euler spiral representation is particularly well suited.
+However, we believe the design of an efficient GPU pipeline architecture involves comples tradeoffs which we did not sufficiently explore for this paper.
+Therefore the dash style applied to the \emph{long dash} scene was processed on the CPU and dashes were sequentially encoded as individual strokes.
 Including CPU-side dashing, the time to encode \emph{long dash} on the Apple M1 Max was approximately 25~ms on average, while the average time to encode the path after dashing was 4.14~ms.

### 2024-06-14 paper.tex
Add figure illustrating simple stroke cap/join styles and dashing
Refer to Figure 1 in text

--- 2024-06-14 paper.tex
@@
 Joins and caps can be rendered with a variety of styles within the same document.
-The style of stroke rendering also optionally includes \emph{dashing}, which applies a pattern of dashes and gaps to the outline of the shape.
+The style of stroke rendering also optionally includes \emph{dashing}, which applies a pattern of dashes and gaps to the outline of the shape, as illustrated in Figure~\ref{fig:stroke_styles}.
+
+\begin{figure}
+    \vspace{15pt}
+    \includegraphics[scale=0.4]{stroke_styles}
+    \caption{Illustration of stroke styles applied to a source Bézier path (shown in black).}
+    \Description{Illustration of stroke styles applied to a source Bézier path (shown in black).}
+    \label{fig:stroke_styles}
+\end{figure}
 

### 2024-06-14 paper.tex

@@
 From this, we can derive a precise, invertible error metric.
-Subdividing the arc in to $n$ segments, the distance error for each segment is $\frac{1}{\kappa}(1-\cos \frac{\kappa s}{2n})$.
+Subdividing the arc into $n$ segments, the distance error for each segment is $\frac{1}{\kappa}(1-\cos \frac{\kappa s}{2n})$.
 Solving for $n$, we get:
@@
 In this subsection, we present a straightforward and accurate approximation, constructed piecewise from easily invertible functions.
-If higher flattening quality is desired at the expense of slower computation, this approximation can be used to determine a good initial value for numeric techniques; two iterations of Newton solving are enough to refine this guess to within 32-bit floating point accuracy.
-
-The approximation is given as follows:
+This approximation was found empirically, by interactively tuning candidate approximation functions in the Desmos graphing calculator.
+The exact integral and the approximation are shown in Figure~\ref{fig:espc}.
+Visually, it is clear that the agreement is close.
+We also built a test suite (included in our supplemental materials) to exhaustively test the subdivision count using a property testing approach, finding that the worst case discrepancy between approximate and exact results is 6\%.
+If higher flattening quality is desired at the expense of slower computation, this approximation could be used to determine a good initial value for numeric techniques; two iterations of Newton solving are enough to refine this guess to within 32-bit floating point accuracy.
+
+The approximation is as follows:
 
@@
 Negative values follow by symmetry, as the function is odd.
-
-The exact integral and the approximation given above are shown in Figure~\ref{fig:espc}.
-Visually, it is clear that the agreement is close, and in numerical testing the worst case discrepancy between approximate and exact results is approximately 6\%.
 
@@
 A good example of this pattern is \citet{Nehab2020}.
-One advantage of Euler spirals over quadratic Béziers is that they can represent inflection points just fine, so it is not necessary to compute those for additional subdivision.
+One advantage of Euler spirals over quadratic Béziers is that they can represent inflection points just fine, so it is not necessary to solve for the inflection points, or do additional subdivision.
+Avoiding these alleviates the need for conditional logic and case analysis, which makes the algorithm more amenable to GPU execution.
 

### 2024-06-14 paper.tex
Add question-mark figure

@@
 \begin{figure}
-    \includegraphics[scale=0.24]{cusp_rendering}
-    \caption{Rendering of a cubic Bézier cusp using Euler spiral segments}
-    \Description{A cubic Bézier with a cusp is approximated by Euler spiral segments with finite curvature, including one with a question mark shape.
-The resulting stroke outlines have the correct shape.}
+    \includegraphics[scale=0.5]{question}
+    \caption{Subdivision of a cubic Bézier with a cusp into continuous Euler spiral segments.}
+    \Description{A cubic Bézier with a cusp is approximated by Euler spiral segments with finite curvature, including one with a question mark shape.}
     \label{fig:cusp_rendering}

### 2024-06-15 gpu_plots.py
Point bench tool at vello:paper-revisions

@@ -17,34 +17,34 @@ save_to_file = len(sys.argv) > 1 and sys.argv[1] == 'save' or \
 
 if do_test_scenes:
 	arcs = pd.DataFrame({
-		"long dash (arcs)": [8.11, 2.90, 0.559, 0.08245],
-		"mmark-70k (arcs)": [7.50, 0.943, 1.67, 0.13434],
-		"mmark-120k (arcs)": [13.17, 1.54, 2.38, 0.19562]
+		"long dash (arcs)": [16.01, 3.24, 0.559, 0.08245],
+		"mmark-70k (arcs)": [9.93, 1.83, 1.67, 0.13434],
+		"mmark-120k (arcs)": [25.32, 2.97, 2.38, 0.19562]
 		}, index=["Mali-G78", "M1 Max", "GTX 980Ti", "RTX 4090"]
 	)
 	lines = pd.DataFrame({
-		"long dash (lines)": [13.06, 3.11, 1.35, 0.31733],
-		"mmark-70k (lines)": [12.45, 1.73, 2.95, 0.22960],
-		"mmark-120k (lines)": [22.52, 2.87, 4.34, 0.38593],
+		"long dash (lines)": [24.42, 3.38, 1.35, 0.31733],
+		"mmark-70k (lines)": [20.30, 3.42, 2.95, 0.22960],
+		"mmark-120k (lines)": [39.41, 5.63, 4.34, 0.38593],
 		}, index=["Mali-G78", "M1 Max", "GTX 980Ti", "RTX 4090"]
 	)
 else:
 	arcs = pd.DataFrame({
-		"spirograph (arcs)":[350.59, 88.01, 37.37, 13.48],
-		"lorenz (arcs)":[361.09, 65.84, 50.61, 18.48],
-		"spiral (arcs)":[324.73, 104.48, 35.97, 13.23],
-		"blender (arcs)":[384.37, 86.25, 51.05, 20.73],
-		"waves (arcs)":[1300, 135.61, 85.06, 33.75],
-		"roads (arcs)":[748.49, 102.37, 87.63, 31.39]
+		"spirograph (arcs)":[565.77, 146.03, 37.37, 13.48],
+		"lorenz (arcs)":[412.13, 82.14, 50.61, 18.48],
+		"spiral (arcs)":[529.88, 118.93, 35.97, 13.23],
+		"blender (arcs)":[599.37, 133.49, 51.05, 20.73],
[…]

### 2024-06-15 paper.tex
Update GPU timing plots with new strongly correct measurements

--- 2024-06-14 paper.tex
@@
 
-Stroke rendering is an essential and widely part of the standard vector graphics imaging model.
+Stroke rendering is an essential part of the standard vector graphics imaging model.
 The standard representation of paths is a sequence of cubic Bézier segments.
@@
 In this paper, we focus on \emph{stroke expansion,} the generation of an outline that, when filled, represents the rendered stroke.
-Implementing stroke rendering using stroke expansion enables a unified approach to path rendering both filled and stroked paths downstream.
+Implementing stroke rendering using stroke expansion enables a unified approach to rendering both filled and stroked paths downstream.
 Among other benefits, such an approach avoids a large number of draw calls when stroked and filled paths are finely interleaved, as is often the case.
@@
 
-Previous work on stroke rendering does not fully solve the problem of fast and correct stroke rendering.
+Previous work on stroke rendering does not fully solve the problem of fast and correct stroking.
 Most existing implementations have correctness problems, in particular failing to draw the evolutes and inner joins as required for strong correctness; \citet{Nehab2020} provides a detailed survey.
@@
 
-This paper presents a comprehensive solution to the core problem of stroke expansion, well suited to GPU implementation.
+This paper presents a solution to the core problem of stroke expansion, well suited to GPU implementation.
 It can produce both flattened polylines or an approximation consisting of circular arcs.
@@
 
-Many techniques for stroke expansion described in the literature apply some lowering of cubic Bézier curves to a simpler curve type that is more tractable for evaluating parallel curve.
+Many techniques for stroke expansion described in the literature apply some lowering of cubic Bézier curves to a simpler curve type that is more tractable for evaluating the parallel curve.
 Computing parallel curves directly on cubic Bézier curve segments is not very tractable.
@@
 For a filled primitive, a thread only outputs the flattened approximation of the path segment.
-For a stroke, a thread outputs the polylines approximating multiple curves: a) the two parallel curves on both sides of the source segment, offset by the desired stroke \emph{half-width}, and b) the joins that connect the path segment to the next adjacent segment, and c) evolute patches in regions of high curvature.
+For a stroke, a thread outputs the polylines approximating multiple curves: a) the two parallel curves on both sides of the source segment, offset by the desired stroke \emph{half-width}, b) the joins that connect the path segment to the next adjacent segment, and c) evolute patches in regions of high curvature.
 We also store metadata in our input encoding that marks the position of a path segment within the containing subpath.
@@
 This procedure is sufficient to generate weakly correct stroke outlines.
-If the outline contains a cusp -- which can be determined by evaluating the curvature variation along the Euler spiral segment -- we output an evolute patch (as shown in Figure~\ref{fig:evolutes}) in order to achieve strong correctness.
+If the outline contains a cusp, we output an evolute patch (as shown in Figure~\ref{fig:evolutes}) in order to achieve strong correctness.
 A selection of renderings produced by our algorithm that demonstrate both strong and weak correctness is shown in Figure~\ref{fig:stroke_stills}.
@@
 \begin{figure}
-    \includegraphics[scale=0.69]{nehab_gpu_timings}
+    \includegraphics[scale=0.69]{nehab_gpu_timings_strong}
     \caption{Mean GPU execution times of the compute shader that performs stroke expansion and flattening when run on the \citet{Nehab2020} \emph{timings} data set.
@@
 The hatched bars show the timings for the version that outputs line primitives while the solid bars show that for arcs.
-Mali-G78 (the lowest-end GPU we tested) converts \emph{waves} (the largest scene in the set) to lines in \textasciitilde3 ms.
+Mali-G78 (the lowest-end GPU we tested) converts \emph{waves} (the largest scene in the set) to lines in \textasciitilde3.5 ms.
 The compute shader executes at least an order of magnitude faster on the other GPUs.}
@@
 \begin{figure}
-    \includegraphics[scale=0.73]{test_scenes_gpu_timings}
+    \includegraphics[scale=0.73]{test_scenes_gpu_timings_strong}
     \caption{Mean GPU execution times of the compute shader that performs stroke expansion and flattening when run on the stress test scenes. The timings are shown in milliseconds.}

### 2024-06-15 gpu_plots.py

@@ -17,34 +17,34 @@ save_to_file = len(sys.argv) > 1 and sys.argv[1] == 'save' or \
 
 if do_test_scenes:
 	arcs = pd.DataFrame({
-		"long dash (arcs)": [16.01, 3.24, 0.559, 0.08245],
-		"mmark-70k (arcs)": [9.93, 1.83, 1.67, 0.13434],
-		"mmark-120k (arcs)": [25.32, 2.97, 2.38, 0.19562]
+		"long dash (arcs)": [16.01, 3.24, 0.74552, 0.11001],
+		"mmark-70k (arcs)": [9.93, 1.83, 2.60, 0.18588],
+		"mmark-120k (arcs)": [25.32, 2.97, 4.14, 0.25569]
 		}, index=["Mali-G78", "M1 Max", "GTX 980Ti", "RTX 4090"]
 	)
 	lines = pd.DataFrame({
-		"long dash (lines)": [24.42, 3.38, 1.35, 0.31733],
-		"mmark-70k (lines)": [20.30, 3.42, 2.95, 0.22960],
-		"mmark-120k (lines)": [39.41, 5.63, 4.34, 0.38593],
+		"long dash (lines)": [24.42, 3.38, 1.32, 0.34711],
+		"mmark-70k (lines)": [20.30, 3.42, 4.37, 0.30955],
+		"mmark-120k (lines)": [39.41, 5.63, 8.26, 0.48937],
 		}, index=["Mali-G78", "M1 Max", "GTX 980Ti", "RTX 4090"]
 	)
 else:
 	arcs = pd.DataFrame({
-		"spirograph (arcs)":[565.77, 146.03, 37.37, 13.48],
-		"lorenz (arcs)":[412.13, 82.14, 50.61, 18.48],
-		"spiral (arcs)":[529.88, 118.93, 35.97, 13.23],
-		"blender (arcs)":[599.37, 133.49, 51.05, 20.73],
-		"waves (arcs)":[1480, 158.11, 85.06, 33.75],
-		"roads (arcs)":[1400, 277.05, 87.63, 31.39]
+		"spirograph (arcs)":[565.77, 146.03, 77.45, 23.01],
+		"lorenz (arcs)":[412.13, 82.14, 58.17, 21.90],
+		"spiral (arcs)":[529.88, 118.93, 75.24, 24.48],
+		"blender (arcs)":[599.37, 133.49, 78.01, 24.85],
+		"waves (arcs)":[1480, 158.11, 108.83, 35.08],
[…]

### 2024-06-21 paper.tex
Revert "Re-anonymize authors"
This reverts commit 52d5d3c516ab8204b1d139b6d7b5563d4ee270f0.

@@
 \begin{figure}
-    \includegraphics[scale=0.69]{nehab_gpu_timings_strong}
+    \includegraphics[scale=0.69]{nehab_gpu_timings}
     \caption{Mean GPU execution times of the compute shader that performs stroke expansion and flattening when run on the \citet{Nehab2020} \emph{timings} data set.
@@
 The hatched bars show the timings for the version that outputs line primitives while the solid bars show that for arcs.
-Mali-G78 (the lowest-end GPU we tested) converts \emph{waves} (the largest scene in the set) to lines in \textasciitilde3.5 ms.
+Mali-G78 (the lowest-end GPU we tested) converts \emph{waves} (the largest scene in the set) to lines in \textasciitilde3 ms.
 The compute shader executes at least an order of magnitude faster on the other GPUs.}
@@
 \begin{figure}
-    \includegraphics[scale=0.73]{test_scenes_gpu_timings_strong}
+    \includegraphics[scale=0.73]{test_scenes_gpu_timings}
     \caption{Mean GPU execution times of the compute shader that performs stroke expansion and flattening when run on the stress test scenes. The timings are shown in milliseconds.}

### 2024-06-28 paper.tex

--- 2024-06-15 paper.tex
@@
 Figures~\ref{fig:nehab-gpu-timings} and \ref{fig:test-scenes-gpu-timings} show the execution times for the two data sets.
-The entire \citet{Nehab2020} timings set adds up to less than 1 ms on all GPUs except the mobile unit.
-Our algorithm is \textasciitilde14$\times$ slower on the Mali-G78 compared to the M1 Max but it is still capable of processing \emph{waves} in 3~ms.
+The entire \citet{Nehab2020} timings set adds up to less than 1.5~ms on all GPUs except the mobile unit.
+Our algorithm is \textasciitilde14$\times$ slower on the Mali-G78 compared to the M1 Max but it is still capable of processing \emph{waves} in 3.48~ms on average.
 We observed that the performance of the kernel scales well with increasing GPU ability even at very large workloads, as demonstrated by the order-of-magnitude difference in run time between the mobile, laptop, and high-end desktop form factors we tested.

### 2024-06-28 paper.tex
Rework unrolled recursion section
Explain the unrolled recursion more extensively, and move the discussion into a paragraph of the algorithm section, out of being its own subsection (and not having great flow).

The commit also contains some small fixes found on a closer read.

@@
 Both \citet{Nehab2020} and \citet{Kilgard2020} provide useful definitions of strongly correct stroke rendering, and Nehab in particular describes how to achieve it in stroke expansion.
-Briefly, when the path curvature exceeds the reciprocal of the stroke half-width, \emph{evolute} segments and the \emph{inner contours of joins} are required in addition to the parallel curves.
+Briefly, when the path curvature exceeds the reciprocal of the stroke half-width, \emph{evolute} segments are required in addition to the parallel curves, as well as \emph{inner contours of joins} when segments are short.
 The technique described in this paper produces strongly correct outlines according to this definition.
@@
 
-The parallel curve of the Euler spiral (also known as ``clothoid'') was characterized by \citet{Wieleitner1907} well over a hundred years ago, and has a straightforward closed-form Cesàro representation, curvature as a function of arc length, of the form $\kappa(s) = c_0{s - s_0}^{-1/2} + c_1$.
+The parallel curve of the Euler spiral (also known as ``clothoid'') was characterized by \citet{Wieleitner1907} well over a hundred years ago, and has a straightforward closed-form Cesàro representation, curvature as a function of arc length, of the form $\kappa(s) = c_0(s - s_0)^{-1/2} + c_1$.
 Euler spirals as an intermediate curve representation for computing parallel curves was proposed in \citet{Levien2021}, which also expands on the Wieleitner reference (including an English translation) and Cesàro equation.
@@
 
-The subdivision density for the parallel curve of an Euler spiral, normalized so that its inflection point is at $-1$ and the cusp of the parallel curve is at 1, is simply $1 - \sqrt{|1-s^2|}$.
+The subdivision density for the parallel curve of an Euler spiral, normalized so that its inflection point is at $-1$ and the cusp of the parallel curve is at 1, is simply $\sqrt{|1-s^2|}$.
 This function is plotted in Figure~\ref{fig:subdiv_density}.
@@
 The exact coefficients used are presented in Appendix \ref{appendix:gh}.
-
-\subsection{Unrolled recursion}
-
-The entire recursion stack for adaptive subdivision can be represented in two words: $dt$ and $t0\_u$ (the latter being an unsigned integer).
-The range is $dt \cdot t0\_u ..
-dt \cdot (t0\_u + 1)$.
-Initial values of $1$ and $0$ represent the range $(0, 1)$.
-Pushing the stack, subdividing the range in half, is represented by halving $dt$ and doubling $t0\_u$, which leaves the start of the range invariant but its size halved.
-After accepting an approximation, the next range is determined by incrementing $t0\_u$.
-At this point, the decision of whether to pop the stack or continue is dependent on whether $t0\_u$ is even or odd, respectively.
-If even, popping the stack is accomplished by doubling $dt$ and halving $t0\_u$, and this is repeated until $t0\_u$ becomes odd.
-The repeated stack popping can be accomplished without iteraction by using the ``count trailing zeros'' intrinsic to determine the number of stack pops $n$, then multiplying $dt$ by $2^n$, and shifting $t0\_u$ right by $n$ bits.
 
@@
 
+The adaptive subdivision in lowering cubic Béziers to Euler spiral segments is traditionally accomplished through recursion; in the case of subdivision there is a recursive call for each of the two subdivisions of the range.
+As the execution model of compute shaders does not support recursion, we \emph{unroll} recursion into an iterative loop.
+The cookbook technique for such unrolling is to store the stack explicitly as an array, which is potentially expensive (it increases pressure on registers).
+We exploit the specific nature of this recursion, encoding the entire state of the stack into two scalar values: the size of the range $dt$, and the scaled start of the range $t0\_u$, so that the range is $t0\_u \cdot dt ..
+(t0\_u + 1) \cdot dt$.
+Initial values are $1$ and $0$ respectively.
+Pushing the call stack is represented by halving $dt$ and doubling $t0\_u$, which preserves the start of the range but halves the size.
+Accepting an approximation (a leaf in the call stack) is represented by incrementing $t0\_u$.
+At this point, the decision of whether to pop the stack or continue is dependent on whether $t0\_u$ is even or odd, respectively.
+If even, popping the stack is accomplished by doubling $dt$ and halving $t0\_u$, and this is repeated until $t0\_u$ becomes odd.
+The repeated stack popping can be accomplished without iteraction by using the ``count trailing zeros'' intrinsic to determine the number of stack pops $n$, then multiplying $dt$ by $2^n$, and shifting $t0\_u$ right by $n$ bits.
+
 This procedure is sufficient to generate weakly correct stroke outlines.

### 2024-06-29 paper.tex
Fix typo

@@
 If even, popping the stack is accomplished by doubling $dt$ and halving $t0\_u$, and this is repeated until $t0\_u$ becomes odd.
-The repeated stack popping can be accomplished without iteraction by using the ``count trailing zeros'' intrinsic to determine the number of stack pops $n$, then multiplying $dt$ by $2^n$, and shifting $t0\_u$ right by $n$ bits.
+The repeated stack popping can be accomplished without iteration by using the ``count trailing zeros'' intrinsic to determine the number of stack pops $n$, then multiplying $dt$ by $2^n$, and shifting $t0\_u$ right by $n$ bits.
 

### 2024-07-02 anims/src/main.rs
Add animations
This commit just pops up a window and gets a scene animated.

@@ -1,0 +1,209 @@
+// Copyright 2024 the Vello Authors
+// SPDX-License-Identifier: Apache-2.0 OR MIT
+
+mod sequence;
+
+use anyhow::Result;
+use std::num::NonZeroUsize;
+use std::sync::Arc;
+use std::time::Instant;
+use vello::peniko::Color;
+use vello::util::{RenderContext, RenderSurface};
+use vello::{AaConfig, Renderer, RendererOptions, Scene};
+use winit::dpi::LogicalSize;
+use winit::event::*;
+use winit::event_loop::{ActiveEventLoop, ControlFlow, EventLoop};
+use winit::window::Window;
+
+use vello::wgpu;
+// Simple struct to hold the state of the renderer
+pub struct ActiveRenderState<'s> {
+    // The fields MUST be in this order, so that the surface is dropped before the window
+    surface: RenderSurface<'s>,
+    window: Arc<Window>,
+}
+
+enum RenderState<'s> {
+    Active(ActiveRenderState<'s>),
+    // Cache a window so that it can be reused when the app is resumed after being suspended
+    Suspended(Option<Arc<Window>>),
+}
+
+fn main() -> Result<()> {
+    let start_time = Instant::now();
+    // Setup a bunch of state:
+
+    // The vello RenderContext which is a global context that lasts for the lifetime of the application
+    let mut render_cx = RenderContext::new();
+
+    // An array of renderers, one per wgpu device
+    let mut renderers: Vec<Option<Renderer>> = vec![];
+
+    // State for our example where we store the winit Window and the wgpu Surface
+    let mut render_state = RenderState::Suspended(None);
+
+    // A vello Scene which is a data structure which allows one to build up a description a scene to be drawn
+    // (with paths, fills, images, text, etc) which is then passed to a renderer for rendering
+    let mut scene = Scene::new();
+
+    // Create and run a winit event loop
+    let event_loop = EventLoop::new()?;
+    #[allow(deprecated)]
+    event_loop
+        .run(move |event, event_loop| match event {
+            // Setup renderer. In winit apps it is recommended to do setup in Event::Resumed
+            // for best cross-platform compatibility
+            Event::Resumed => {
+                let RenderState::Suspended(cached_window) = &mut render_state else {
+                    return;
+                };
+
+                // Get the winit window cached in a previous Suspended event or else create a new window
+                let window = cached_window
+                    .take()
+                    .unwrap_or_else(|| create_winit_window(event_loop));
+
+                // Create a vello Surface
[…]

### 2024-07-02 anims/src/sequence.rs

@@ -1,0 +1,52 @@
+use vello::{
+    kurbo::{Affine, Line, Point, Rect, Stroke, Vec2},
+    peniko::Color,
+    Scene,
+};
+
+fn timed(t: &mut f64, duration: f64) -> bool {
+    if *t < duration {
+        true
+    } else {
+        *t -= duration;
+        false
+    }
+}
+
+pub fn render(scene: &mut Scene, mut t: f64) {
+    if timed(&mut t, 2.0) {
+        clocky(scene, t);
+    } else {
+        end_card(scene);
+    }
+}
+
+pub fn clocky(scene: &mut Scene, t: f64) {
+    let stroke = Stroke::new(6.0);
+    let p0 = Point::new(200., 200.);
+    let p1 = p0 + 180. * Vec2::from_angle(t);
+    let line = Line::new(p0, p1);
+    let line_stroke_color = Color::rgb(0.9804, 0.702, 0.5294);
+    scene.stroke(&stroke, Affine::IDENTITY, line_stroke_color, None, &line);
+    if t < 1.0 {
+        scene.stroke(
+            &stroke,
+            Affine::translate((0., 10.)),
+            line_stroke_color,
+            None,
+            &line,
+        );
+    }
+}
+
+pub fn end_card(scene: &mut Scene) {
+    let color = Color::rgb(0.1, 0.1, 0.8);
+    let rect = Rect::new(100., 100., 1000., 1000.);
+    scene.fill(
+        vello::peniko::Fill::NonZero,
+        Affine::IDENTITY,
+        &color,
+        None,
+        &rect,
+    );
+}

### 2024-07-02 anims/src/anims.rs
Checkpoint
Have basic animation of stroke in place, but obviously needs more work.

@@ -1,0 +1,65 @@
+use flatten::stroke::LoweredPath;
+use vello::{
+    kurbo::{Affine, Cap, CubicBez, Line, ParamCurve, Rect, Shape, Stroke},
+    peniko::Color,
+    Scene,
+};
+
+pub struct Anims;
+
+fn timed(t: &mut f64, duration: f64) -> bool {
+    if *t < duration {
+        true
+    } else {
+        *t -= duration;
+        false
+    }
+}
+
+const STROKE_LEN: f64 = 5.0;
+
+impl Anims {
+    pub fn new() -> Self {
+        Anims
+    }
+
+    pub fn render(&self, scene: &mut Scene, mut t: f64) {
+        if timed(&mut t, STROKE_LEN) {
+            anim_stroke(scene, t / STROKE_LEN);
+        } else {
+            end_card(scene);
+        }
+    }    
+}
+
+
+fn anim_stroke(scene: &mut Scene, t: f64) {
+    let c = CubicBez::new((110., 290.), (110., 250.), (110., 160.), (140., 160.));
+    let c = Affine::scale(3.) * c;
+    let t_adjust = (t * 1.2).min(1.0);
+    let trimmed = c.subsegment(0.0..t_adjust);
+    let flatten_style = Stroke::new(300.0).with_caps(Cap::Butt);
+    let path = trimmed.to_path(1e-9);
+    let stroked: LoweredPath<Line> = flatten::stroke::stroke_undashed(path, &flatten_style, 0.1);
+    let stroked_path = stroked.to_bez();
+    let stroke_color = Color::rgb(0.9804, 0.702, 0.5294);
+    let thin_stroke_color = Color::rgb(0.5, 0.5, 1.);
+    let fill_color = Color::rgb8(0xff, 0x93, 0x8d);
+    let stroke = Stroke::new(6.0);
+    let thin_stroke = Stroke::new(4.0);
+    scene.fill(vello::peniko::Fill::NonZero, Affine::IDENTITY, fill_color, None, &stroked_path);
+    scene.stroke(&stroke, Affine::IDENTITY, stroke_color, None, &stroked_path);
+    scene.stroke(&thin_stroke, Affine::IDENTITY, thin_stroke_color, None, &trimmed);
+}
+
+fn end_card(scene: &mut Scene) {
+    let color = Color::rgb(0.1, 0.1, 0.8);
+    let rect = Rect::new(100., 100., 1000., 1000.);
+    scene.fill(
[…]

### 2024-07-02 anims/src/main.rs

@@ -1,9 +1,10 @@
 // Copyright 2024 the Vello Authors
 // SPDX-License-Identifier: Apache-2.0 OR MIT
 
-mod sequence;
+mod anims;
 
 use anyhow::Result;
+use anims::Anims;
 use std::num::NonZeroUsize;
 use std::sync::Arc;
 use std::time::Instant;
@@ -31,6 +32,7 @@ enum RenderState<'s> {
 
 fn main() -> Result<()> {
     let start_time = Instant::now();
+    let anims = Anims::new();
     // Setup a bunch of state:
 
     // The vello RenderContext which is a global context that lasts for the lifetime of the application
@@ -130,7 +132,7 @@ fn main() -> Result<()> {
                         println!("redrawing, t = {t}!");
                         // TODO: stabilize for steady frame rates
                         // Re-add the objects to draw to the scene.
-                        sequence::render(&mut scene, t);
+                        anims.render(&mut scene, t);
 
                         // Get the RenderSurface (surface + config)
                         let surface = &render_state.surface;

### 2024-07-02 anims/src/sequence.rs

@@ -1,52 +1,0 @@
-use vello::{
-    kurbo::{Affine, Line, Point, Rect, Stroke, Vec2},
-    peniko::Color,
-    Scene,
-};
-
-fn timed(t: &mut f64, duration: f64) -> bool {
-    if *t < duration {
-        true
-    } else {
-        *t -= duration;
-        false
-    }
-}
-
-pub fn render(scene: &mut Scene, mut t: f64) {
-    if timed(&mut t, 2.0) {
-        clocky(scene, t);
-    } else {
-        end_card(scene);
-    }
-}
-
-pub fn clocky(scene: &mut Scene, t: f64) {
-    let stroke = Stroke::new(6.0);
-    let p0 = Point::new(200., 200.);
-    let p1 = p0 + 180. * Vec2::from_angle(t);
-    let line = Line::new(p0, p1);
-    let line_stroke_color = Color::rgb(0.9804, 0.702, 0.5294);
-    scene.stroke(&stroke, Affine::IDENTITY, line_stroke_color, None, &line);
-    if t < 1.0 {
-        scene.stroke(
-            &stroke,
-            Affine::translate((0., 10.)),
-            line_stroke_color,
-            None,
-            &line,
-        );
-    }
-}
-
-pub fn end_card(scene: &mut Scene) {
-    let color = Color::rgb(0.1, 0.1, 0.8);
-    let rect = Rect::new(100., 100., 1000., 1000.);
-    scene.fill(
-        vello::peniko::Fill::NonZero,
-        Affine::IDENTITY,
-        &color,
-        None,
-        &rect,
-    );
-}

### 2024-07-02 flatten/src/euler32.rs

@@ -105,7 +105,7 @@ impl CubicParams {
 }
 
 impl EulerParams {
-    pub fn from_f64(ep: &crate::EulerParams) -> Self {
+    pub fn from_f64(ep: &crate::euler::EulerParams) -> Self {
         EulerParams {
             th0: ep.th0 as f32,
             th1: ep.th1 as f32,
@@ -178,6 +178,7 @@ impl EulerParams {
 }
 
 impl EulerSeg {
+    #[allow(unused)]
     fn from_f64(es: &crate::euler::EulerSeg) -> Self {
         EulerSeg {
             p0: Point::from_kurbo(es.p0),

### 2024-07-02 flatten/src/main.rs

@@ -1,42 +1,26 @@
 // Copyright 2024 the Vello Authors
 // SPDX-License-Identifier: Apache-2.0
 
-use clap::{Arg, Parser};
-use cubic32::{Cubic, Point};
-use euler32::CubicToEulerIter;
-use flatten::{espc_integral_inv, n_subdiv_analytic};
-use flatten32::flatten_offset;
+use clap::Parser;
+use flatten::cubic32::{Cubic, Point};
+use flatten::euler32::CubicToEulerIter;
+use flatten::flatten::{espc_integral_inv, n_subdiv_analytic};
+use flatten::flatten32::flatten_offset;
 
-use crate::euler::EulerParams;
-
-mod arc_segment;
-mod cubic32;
-mod cubic_err_plot;
-mod euler;
-mod euler32;
-mod euler_arc;
-mod evolute;
-mod flatten;
-mod flatten32;
-mod perf_graph;
-#[cfg(feature = "skia-safe")]
-mod skia;
-mod stroke;
-mod svg;
-mod to_rvg;
+use flatten::euler::EulerParams;
 
 #[derive(Parser)]
 enum Args {
     Arc,
     Cubic,
-    CubicErr(cubic_err_plot::CubicErrPlot),
+    CubicErr(flatten::cubic_err_plot::CubicErrPlot),
     Evolute,
     Espc,
     EstFlattenErr,
-    PrimCountGraph(perf_graph::PrimCountArgs),
+    PrimCountGraph(flatten::perf_graph::PrimCountArgs),
     Stroke,
-    Svg(svg::SvgArgs),
-    ToRvg(to_rvg::ToRvgArgs),
+    Svg(flatten::svg::SvgArgs),
+    ToRvg(flatten::to_rvg::ToRvgArgs),
 }
 
 fn main_est_flatten_err() {
@@ -87,7 +71,7 @@ fn main_espc() {
         // Note: this isn't necessarily accurate when k1 and dist are tiny
         let analytic = n_subdiv_analytic(k0, k1, scale, dist, tol);
         let approx_f32 =
-            flatten32::n_subdiv_robust(k0 as f32, k1 as f32, scale as f32, dist as f32, tol as f32);
+            flatten::flatten32::n_subdiv_robust(k0 as f32, k1 as f32, scale as f32, dist as f32, tol as f32);
         println!("{k0} {k1:.1e} {dist}: {} {}", analytic, approx_f32);
     }
 }
[…]

### 2024-07-02 flatten/src/stroke.rs

@@ -106,8 +106,7 @@ impl<L: Lowering> LoweredPath<L> {
         self.last_pt = p;
     }
 
-    #[allow(unused)]
-    fn to_bez(&self) -> BezPath {
+    pub fn to_bez(&self) -> BezPath {
         let mut result = BezPath::new();
         for seg in &self.path {
             seg.to_bez(&mut result);

### 2024-07-03 anims/src/anims.rs
Add a couple more sequences
Show density function. Show Nimbus 'g' being flattened to lines and arcs with varying tolerance.

@@ -1,11 +1,14 @@
-use flatten::stroke::LoweredPath;
+use flatten::{stroke::LoweredPath, ArcSegment};
 use vello::{
-    kurbo::{Affine, Cap, CubicBez, Line, ParamCurve, Rect, Shape, Stroke},
+    kurbo::{Affine, BezPath, Cap, CubicBez, Line, ParamCurve, Point, Rect, Shape, Stroke},
     peniko::Color,
     Scene,
 };
 
-pub struct Anims;
+pub struct Anims {
+    dens_curve: BezPath,
+    g_path: BezPath,
+}
 
 fn timed(t: &mut f64, duration: f64) -> bool {
     if *t < duration {
@@ -18,20 +21,49 @@ fn timed(t: &mut f64, duration: f64) -> bool {
 
 const STROKE_LEN: f64 = 5.0;
 
+const G_PATH_STR: &str = "M470 295h-83c14 32 19 55 19 84c0 50 -15 85 -48 113c-31 27 -71 42 -108 42c-2 0 -21 -2 -57 -5c-27 8 -60 43 -60 63c0 16 24 25 78 27l129 6c74 3 121 45 121 107c0 38 -17 70 -55 101c-52 42 -130 68 -205 68c-96 0 -173 -44 -173 -97c0 -37 26 -70 98 -122
+c-42 -20 -53 -31 -53 -53c0 -17 6 -28 27 -50c3 -4 15 -15 36 -35l26 -24c-67 -33 -93 -71 -93 -134c0 -92 74 -163 167 -163c26 0 53 5 80 15l22 8c20 7 34 10 55 10h77v39zM147 685c-40 48 -49 63 -49 86c0 44 57 73 146 73c113 0 189 -39 189 -97c0 -36 -33 -49 -124 -49
+c-49 0 -128 -6 -162 -13zM152 345v3c0 96 41 161 103 161c46 0 74 -35 74 -91c0 -40 -11 -85 -30 -120c-16 -30 -42 -47 -73 -47c-46 0 -74 35 -74 94z";
+
 impl Anims {
     pub fn new() -> Self {
-        Anims
+        let dens_curve = density_curve();
+        let g_path = BezPath::from_svg(G_PATH_STR).unwrap();
+        Anims { dens_curve, g_path }
     }
 
     pub fn render(&self, scene: &mut Scene, mut t: f64) {
         if timed(&mut t, STROKE_LEN) {
             anim_stroke(scene, t / STROKE_LEN);
[…]

### 2024-07-03 anims/src/main.rs

@@ -3,8 +3,8 @@
 
 mod anims;
 
-use anyhow::Result;
 use anims::Anims;
+use anyhow::Result;
 use std::num::NonZeroUsize;
 use std::sync::Arc;
 use std::time::Instant;
@@ -129,7 +129,6 @@ fn main() -> Result<()> {
                         scene.reset();
 
                         let t = start_time.elapsed().as_secs_f64();
-                        println!("redrawing, t = {t}!");
                         // TODO: stabilize for steady frame rates
                         // Re-add the objects to draw to the scene.
                         anims.render(&mut scene, t);

### 2024-07-19 paper.tex
ACM formatting fixes
Minor formatting tweaks
Apply acmauthoryear cite style as requested. Small changes to the figures, making lines bolder and adding axis labels.

Also add a mention to the Vello rendering engine, which was missing due to anonymity.
Apply ACM bibliographic metadata
This applies the DOI and the rights statement as appropriate for publication in PACMCGIT.
More formatting fixes
Use acmauthoryear cite style. Add email addresses. Squash linespace of algorithm 2 to fit page margins.

With this cite style, the `\citenum` command is no longer valid, so those are all changed to `\cite

@@
-\documentclass[sigconf, nonacm]{acmart}
+\documentclass[acmsmall]{acmart}
 \usepackage[linesnumbered]{algorithm2e}
-\usepackage[format=plain, labelfont={bf}, textfont=normal]{caption}
-\usepackage{booktabs,xcolor}
+\usepackage{setspace}
+%\usepackage[format=plain, labelfont={bf}, textfont=normal]{caption}
+\usepackage{xcolor}
 \usepackage{natbib}
-\setcitestyle{authoryear}
-\SetAlFnt{\small\sffamily}
+\setcopyright{rightsretained}
+\acmJournal{PACMCGIT}
+\acmYear{2024} \acmVolume{7} \acmNumber{3} \acmArticle{} \acmMonth{7}\acmDOI{10.1145/3675390}
+\acmArticle{1}
+\citestyle{acmauthoryear}
+\SetAlFnt{\footnotesize\sffamily}
 
@@
 \author{Raph Levien}
+\email{raph@google.com}
 \author{Arman Uguray}
+\email{arman.uguray@gmail.com}
 \affiliation{
@@
 % Re-enable these if/when published, also remove the `nonacm` version in \documentclass above.
-%\copyrightyear{2024}
-%\setcopyright{acmcopyright}
-%\conferenceinfo{Foo}{Month Day-Day, Year, Location}
+\copyrightyear{2024}
+\setcopyright{acmcopyright}
+%\conferenceinfo{High Performance Graphics}{July 26-28 2024, Denver}
 %\isbn{FOO}\acmPrice{BAR}
@@
 Correct stroke rendering also relies on accurate approximation of the parallel curves of the input paths.
-Determining the parallel curve of a cubic Bézier is not analytically tractable, as it is a tenth order algebraic formula~[\citenum{Farouki1990}].
+Determining the parallel curve of a cubic Bézier is not analytically tractable, as it is a tenth order algebraic formula~\cite{Farouki1990}.
 Thus, all practical stroke rendering techniques employ an approximation, with error tolerance as a tunable parameter, typically representing the maximum allowable Fréchet distance between the true curve and its approximation.
@@
 A correct algorithm will produce an approximate curve within the error tolerance, but an efficient algorithm will not subdivide significantly more than is necessary to meet the error bound.
-Some specifications for vector graphics, including PDF~[\citenum{PDF2008}], specify the error tolerance explicitly.
+Some specifications for vector graphics, including PDF~\cite{PDF2008}, specify the error tolerance explicitly.
 
@@
 An example of a reasonably good algorithm that computes flattened parallel curves is \citet{Yzerman2020}, as used in the Blend2D rendering library.
-For producing curved outlines, the Tiller and Hanson~[\citenum{Tiller1984}] algorithm is commonly implemented and cited, but its performance is quite poor when applied to cubic Béziers.
+For producing curved outlines, the Tiller and Hanson~\cite{Tiller1984} algorithm is commonly implemented and cited, but its performance is quite poor when applied to cubic Béziers.
 The quadratic Bézier version is adequate, and forms the basis of the algorithm in \citet{Nehab2020}.
-A variant is also used in Skia~[\citenum{Skia}]; instead of lowering the cubic Bézier to a quadratic approximation and then computing the parallel curve, Skia approximates the parallel curve directly with a quadratic Bézier, then measures the error to determine whether further subdivision is necessary.
+A variant is also used in Skia~\cite{Skia}; instead of lowering the cubic Bézier to a quadratic approximation and then computing the parallel curve, Skia approximates the parallel curve directly with a quadratic Bézier, then measures the error to determine whether further subdivision is necessary.
 
@@
 \begin{figure}
-    \includegraphics[scale=0.3]{es_flatten}
+    \includegraphics[scale=0.5]{es_flatten}
     \caption{The flattened approximation to an Euler spiral using line segments is achieved by choosing subdivision points using a simple power law formula with an exponent of 2/3.}
@@
 The geometry of a stroke outline consists of joins, caps, and the two parallel curves on either side of the input path segments, offset by the half linewidth.
-The joins and caps are not particularly difficult to calculate, but parallel curves of cubic Béziers are notoriously tricky~[\citenum{Kilgard2020a}].
+The joins and caps are not particularly difficult to calculate, but parallel curves of cubic Béziers are notoriously tricky~\cite{Kilgard2020a}.
 Analytically, it is a tenth order algebraic curve, which is not particularly feasible to compute directly.
@@
 Using a subpixel Fréchet distance bound guarantees that the rendered image does not deviate visibly from the exact rendering.
-Another choice would be uniform steps in tangent angle, as chosen by polar stroking~[\citenum{Kilgard2020}].
+Another choice would be uniform steps in tangent angle, as chosen by polar stroking~\cite{Kilgard2020}.
 However, at small curvature, the stroked path can be off by several pixels, and at large curvature there may be considerably more subdivision than needed for faithful rendering.
@@
 The bottom row shows a consistent angle step, as implemented in polar stroking, but has excessive distance error at low curvature, and excessive subdivision at high curvature.
-It should be noted, to avoid the undershoot at low curvature, both the Skia~[\citenum{Skia}] and Rive~[\citenum{Rive}] renderers use a hybrid of the Wang and polar stroking error metrics.
+It should be noted, to avoid the undershoot at low curvature, both the Skia~\cite{Skia} and Rive~\cite{Rive} renderers use a hybrid of the Wang and polar stroking error metrics.
 
@@
 When a curve is monotonic, error is also monotonic with respect to both subdivision points.
-Thus, any standard root finding technique can find a subdivision with a given error (bisection is the most popular, but the ITP method[\citenum{Oliveira2020}] is better).
+Thus, any standard root finding technique can find a subdivision with a given error (bisection is the most popular, but the ITP method\cite{Oliveira2020} is better).
 A similar outer loop with a root finding technique can find the error which minimizes the global error for the entire flattening, which, due to monotonicity, is that for which the error of the final segment equals that of the previous segments.
@@
 
-In the principled, correct specification for stroking~[\citenum{Nehab2020}], parallel curves are sufficient only for segments in which the curvature does not exceed the reciprocal half-width.
+In the principled, correct specification for stroking~\cite{Nehab2020}, parallel curves are sufficient only for segments in which the curvature does not exceed the reciprocal half-width.
 When it does, additional segments must be drawn, including evolutes of the original curve.
 In general, the evolute of a cubic Bézier is a very complex curve, requiring approximation techniques.
-By contrast, the evolute of an Euler spiral ($\kappa = as$) is another spiral with a simple Cesàro equation, namely $\kappa = -a^{-1}s^{-3}$, an instance of the general result that the evolute of a log-aesthetic curve is another log-aesthetic curve~[\citenum{Yoshida2012}].
+By contrast, the evolute of an Euler spiral ($\kappa = as$) is another spiral with a simple Cesàro equation, namely $\kappa = -a^{-1}s^{-3}$, an instance of the general result that the evolute of a log-aesthetic curve is another log-aesthetic curve~\cite{Yoshida2012}.
 
@@
 Computing parallel curves directly on cubic Bézier curve segments is not very tractable.
-In particular, the widely cited Tiller-Hanson algorithm~[\citenum{Tiller1984}] performs well for quadratic Béziers but significantly worse for cubics.
+In particular, the widely cited Tiller-Hanson algorithm~\cite{Tiller1984} performs well for quadratic Béziers but significantly worse for cubics.
 
@@
 Given tangent angles relative to the chord, finding the Euler spiral segment that minimizes total curvature variation is a form of geometric Hermite interpolation.
-There are a number of published solutions to this problem, involving nontrivial numerical solving techniques: gradient descent~[\citenum{Kimia2003}], bisection~[\citenum{Walton2009}], or Newton iteration~[\citenum{Connor2014}].
+There are a number of published solutions to this problem, involving nontrivial numerical solving techniques: gradient descent~\cite{Kimia2003}, bisection~\cite{Walton2009}, or Newton iteration~\cite{Connor2014}.
 A more direct approach is to approximate the function as a reasonably low-order polynomial in terms of the endpoint angles.
@@
 \begin{figure*}
-    \includegraphics[scale=0.19]{stroke_stills}
+    \includegraphics[scale=0.15]{stroke_stills}
     \caption{Stroked cubic Béziers exhibiting strong and weak correctness, rendered by our implementation.
@@
 
-One of our artifacts is a GPU compute shader that satisfies these criteria.
+One of our artifacts is a GPU compute shader that satisfies these criteria, in the open source Vello rendering engine.
 We coupled this with a simple and efficient data layout for parallel processing.
 Notably, our implementation is fully data-parallel on input path segments and does not require any computationally expensive curve evaluation on the CPU.
-We implemented our shader on general-purpose GPU compute primitives supported by all modern graphics APIs, particularly Metal, Vulkan, D3D12, and WebGPU~[\citenum{WebGPU}].
+We implemented our shader on general-purpose GPU compute primitives supported by all modern graphics APIs, particularly Metal, Vulkan, D3D12, and WebGPU~\cite{WebGPU}.
 As such, the ideas presented in this section are portable to various GPU platforms.
@@
 \begin{algorithm}
+\setstretch{0.93}
 \SetKwProg{Fn}{fn}{}{end}
@@
 An application may want to sort the line soup, depending on the requirements of the chosen rendering algorithm.
-For instance, our own renderer employs tile-based scanline rasterization in a compute shader, with a pipeline architecture that resembles \emph{cudaraster}~[\citenum{Laine2011}] and \emph{MPVG}~[\citenum{Ganacim2014}].
+For instance, our own renderer employs tile-based scanline rasterization in a compute shader, with a pipeline architecture that resembles \emph{cudaraster}~\cite{Laine2011} and \emph{MPVG}~\cite{Ganacim2014}.
 In this design, the line soup elements are spatially sorted by subsequent compute stages into a grid of 16x16 pixel tiles.
@@
 \begin{figure*}
-    \includegraphics[scale=0.275]{pipeline}
+    \includegraphics[scale=0.21]{pipeline}
     \caption{We integrated our curve flattening shader into our own renderer implemented entirely in compute shaders.
@@
 We evaluated the runtime performance of our compute shader on 4 GPU models and multiple form factors: Google Pixel 6 equipped with the Arm \emph{Mali-G78 MP20} mobile GPU, the Apple \emph{M1 Max} integrated laptop GPU, and two discrete desktop GPUs: the mid-2015 NVIDIA {GeForce GTX 980Ti} and the late-2022 NVIDIA {GeForce RTX 4090}.
-We authored our entire GPU pipeline in the \emph{WebGPU Shading Language} (WGSL) and used the \emph{wgpu}~[\citenum{wgpu}] framework to interoperate with the native graphics APIs (we tested the M1 Max on macOS via the Metal API; the mobile and desktop GPUs were tested via the Vulkan API on Android and Ubuntu systems).
+We authored our entire GPU pipeline in the \emph{WebGPU Shading Language} (WGSL) and used the \emph{wgpu}~\cite{wgpu} framework to interoperate with the native graphics APIs (we tested the M1 Max on macOS via the Metal API; the mobile and desktop GPUs were tested via the Vulkan API on Android and Ubuntu systems).
 
@@
 We used two variants of this scene with \emph{butt} and \emph{round} cap styles.
-For the second scene we adapted the \emph{Canvas Paths} test case from MotionMark~[\citenum{MotionMark}], which renders a large number of randomly generated stroked line, quadratic, and cubic Bézier paths.
+For the second scene we adapted the \emph{Canvas Paths} test case from MotionMark~\cite{MotionMark}, which renders a large number of randomly generated stroked line, quadratic, and cubic Bézier paths.
 We made two versions with different sizes: mmark-70k and mmark-120k with 70,000 and 120,000 path segments, respectively.
@@
 We augmented our encoding logic to compute a conservative estimate of the required output buffer size based on an upper bound on the number of line primitives per input path segment.
-We found that Wang's formula~[\citenum{Goldman2003}] works well in practice, as it is relatively inexpensive to evaluate, though it results in a higher memory estimate than required (see Table~\ref{table:bump-estimate-mem}).
+We found that Wang's formula~\cite{Goldman2003} works well in practice, as it is relatively inexpensive to evaluate, though it results in a higher memory estimate than required (see Table~\ref{table:bump-estimate-mem}).
 For parallel curves, Wang's formula does not guarantee an accurate upper bound on the required subdivisions, which we worked around by inflating the estimate for the source segment by a scale factor derived from the linewidth.
@@
     \centering
-    \includegraphics[scale=0.2]{test_renderings}
+    \includegraphics[scale=0.18]{test_renderings}
     \caption{Renderings of some of our test scenes created using our algorithm.
@@
     \item The GPU implementation sequentializes some work that could be parallel, and is thus not as well load-balanced as it might be.
-An appealing future direction is to split the pipeline into separate stages, executed by the GPU as nodes in a work graph~[\citenum{Patel2024}].
+An appealing future direction is to split the pipeline into separate stages, executed by the GPU as nodes in a work graph[\cite{Patel2024}.
 This structure is appealing because load balancing is done by hardware.
@@
 \bibliographystyle{ACM-Reference-Format}
-\setcitestyle{numbers} % Switch back to numerical mode to show citation numbers in bibliography
+%\setcitestyle{numbers} % Switch back to numerical mode to show citation numbers in bibliography
 \bibliography{paper}

### 2024-07-28 paper.tex

@@
 Weak correctness, by contrast, only requires the parallel curves (also known as ``offset curves'') of path segments, combined with caps and the outer contours of joins.
-Both \citet{Nehab2020} and \citet{Kilgard2020} provide useful definitions of strongly correct stroke rendering, and Nehab in particular describes how to achieve it in stroke expansion.
+Both \cite{Nehab2020} and \cite{Kilgard2020} provide useful definitions of strongly correct stroke rendering, and Nehab in particular describes how to achieve it in stroke expansion.
 Briefly, when the path curvature exceeds the reciprocal of the stroke half-width, \emph{evolute} segments are required in addition to the parallel curves, as well as \emph{inner contours of joins} when segments are short.
@@
 The subproblem of approximating parallel curves has spawned extensive literature, none of which is entirely satisfying, especially when it comes to algorithms that can be efficiently evaluated on GPU.
-An example of a reasonably good algorithm that computes flattened parallel curves is \citet{Yzerman2020}, as used in the Blend2D rendering library.
-For producing curved outlines, the Tiller and Hanson~\cite{Tiller1984} algorithm is commonly implemented and cited, but its performance is quite poor when applied to cubic Béziers.
-The quadratic Bézier version is adequate, and forms the basis of the algorithm in \citet{Nehab2020}.
+An example of a reasonably good algorithm that computes flattened parallel curves is \cite{Yzerman2020}, as used in the Blend2D rendering library.
+For producing curved outlines, the \cite{Tiller1984} algorithm is commonly implemented and cited, but its performance is quite poor when applied to cubic Béziers.
+The quadratic Bézier version is adequate, and forms the basis of the algorithm in \cite{Nehab2020}.
 A variant is also used in Skia~\cite{Skia}; instead of lowering the cubic Bézier to a quadratic approximation and then computing the parallel curve, Skia approximates the parallel curve directly with a quadratic Bézier, then measures the error to determine whether further subdivision is necessary.
@@
 Previous work on stroke rendering does not fully solve the problem of fast and correct stroking.
-Most existing implementations have correctness problems, in particular failing to draw the evolutes and inner joins as required for strong correctness; \citet{Nehab2020} provides a detailed survey.
+Most existing implementations have correctness problems, in particular failing to draw the evolutes and inner joins as required for strong correctness; \cite{Nehab2020} provides a detailed survey.
 In addition, most implementations are not suitable for running on GPU, requiring sequential execution to produce watertight outlines.
-One GPU implementation is polar stroking (\citet{Kilgard2020}), a local technique, but among other limitations, it does not bound the Fréchet distance error (rather, it bounds the angle error).
+One GPU implementation is polar stroking (\cite{Kilgard2020}), a local technique, but among other limitations, it does not bound the Fréchet distance error (rather, it bounds the angle error).
 
@@
 
-Previous work, particularly \citet{Kilgard2020}, has explained how to break the input path into dash segments of the specified length on GPU using a prefix sum of segment arc lengths.
+Previous work, particularly \cite{Kilgard2020}, has explained how to break the input path into dash segments of the specified length on GPU using a prefix sum of segment arc lengths.
 There are implementation tradeoffs to GPU-based dash processing, including potentially a larger number of dispatch stages.
@@
 Otherwise, the curve is subdivided (usually at $t = 0.5$) and each half is recursively approximated.
-A substantial fraction of all curve approximation methods in the literature are of this form, including \citet{Nehab2020}.
+A substantial fraction of all curve approximation methods in the literature are of this form, including \cite{Nehab2020}.
 The main disadvantage is the cost of computing the error metric.
@@
 If the error metric is accurate, then approximation is near-optimal.
-One example of an invertible error metric is angle step, used in polar stroking (\citet{Kilgard2020}); the number of subdivisions is the total angle subsumed by the curve divided by the angle step size, and the parameter value for each subdivision is the result of solving for a tangent direction.
-Another widely used invertible error metric is Wang's formula (\citet{Goldman2003}, Section 5.6.3), which gives a bound on the flattening error based on the second derivative of the curve.
+One example of an invertible error metric is angle step, used in polar stroking (\cite{Kilgard2020}); the number of subdivisions is the total angle subsumed by the curve divided by the angle step size, and the parameter value for each subdivision is the result of solving for a tangent direction.
+Another widely used invertible error metric is Wang's formula (\cite{Goldman2003}, Section 5.6.3), which gives a bound on the flattening error based on the second derivative of the curve.
 This metric is conservative but works well in practice; among other applications, it is used in Skia for path flattening.
@@
 It is common to approximate cubic Béziers to some intermediate curve format more conducive to offsetting and flattening.
-A number of published solutions (\citet{Yzerman2020}, \citet{Nehab2020}) use quadratic Béziers, as it is well suited for computation of parallel curves.
+A number of published solutions (\cite{Yzerman2020}, \cite{Nehab2020}) use quadratic Béziers, as it is well suited for computation of parallel curves.
 Even so, this curve has some disadvantages.
@@
 
-The parallel curve of the Euler spiral (also known as ``clothoid'') was characterized by \citet{Wieleitner1907} well over a hundred years ago, and has a straightforward closed-form Cesàro representation, curvature as a function of arc length, of the form $\kappa(s) = c_0(s - s_0)^{-1/2} + c_1$.
-Euler spirals as an intermediate curve representation for computing parallel curves was proposed in \citet{Levien2021}, which also expands on the Wieleitner reference (including an English translation) and Cesàro equation.
+The parallel curve of the Euler spiral (also known as ``clothoid'') was characterized by \cite{Wieleitner1907} well over a hundred years ago, and has a straightforward closed-form Cesàro representation, curvature as a function of arc length, of the form $\kappa(s) = c_0(s - s_0)^{-1/2} + c_1$.
+Euler spirals as an intermediate curve representation for computing parallel curves was proposed in \cite{Levien2021}, which also expands on the Wieleitner reference (including an English translation) and Cesàro equation.
 
@@
 The problem of approximating a curve by a sequence of arc segments has extensive literature, but none of the published solutions are quite suitable for our application.
-The specific problem of approximating an Euler spiral by arcs is considered in \citet{Meek2004} using a ``cut then measure'' adaptive subdivision scheme, but their solution has poor quality; it scales as $O(1/n^2)$, while $O(1/n^3)$ is attainable.
-The result was improved ``slightly'' by \citet{Narayan2014}.
-The literature also contains optimal results, namely \citet{Maier2014} and \citet{Nuntawisuttiwong2021}, but at considerable cost; both approaches claim $O(n^2)$ time complexity.
+The specific problem of approximating an Euler spiral by arcs is considered in \cite{Meek2004} using a ``cut then measure'' adaptive subdivision scheme, but their solution has poor quality; it scales as $O(1/n^2)$, while $O(1/n^3)$ is attainable.
+The result was improved ``slightly'' by \cite{Narayan2014}.
+The literature also contains optimal results, namely \cite{Maier2014} and \cite{Nuntawisuttiwong2021}, but at considerable cost; both approaches claim $O(n^2)$ time complexity.
 The through-line for all these results is that they are solving a harder problem: adopting the constraint that the generated arc sequence is $G^1$ continuous.
@@
 Computing parallel curves directly on cubic Bézier curve segments is not very tractable.
-In particular, the widely cited Tiller-Hanson algorithm~\cite{Tiller1984} performs well for quadratic Béziers but significantly worse for cubics.
+In particular, the widely cited \cite{Tiller1984} algorithm performs well for quadratic Béziers but significantly worse for cubics.
 
@@
 Subdivisions are also indicated at special points; for example, since quadratic Béziers cannot represent inflection points, and geometric Hermite interpolation is numerically unstable if the input curve is not convex, lowering to quadratic Béziers also requires calculation of inflection points and subdividing there.
-A good example of this pattern is \citet{Nehab2020}.
+A good example of this pattern is \cite{Nehab2020}.
 One advantage of Euler spirals over quadratic Béziers is that they can represent inflection points just fine, so it is not necessary to solve for the inflection points, or do additional subdivision.
@@
 A key step in approximating one curve with another is evaluating the error of the approximation.
-A common approach (used in \citet{Nehab2020} among others) is to generate the approximate curve, then measure the distance, often by sampling at multiple points on the curve.
+A common approach (used in \cite{Nehab2020} among others) is to generate the approximate curve, then measure the distance, often by sampling at multiple points on the curve.
 All this is potentially slow, with the additional risk of underestimating the error due to undersampling.
@@
 A more direct approach is to approximate the function as a reasonably low-order polynomial in terms of the endpoint angles.
-Our approach is to use a 7th order polynomial, which is more precise than the 3rd order polynomial proposed in \citet{Reif2021}, at only modest increased cost.
+Our approach is to use a 7th order polynomial, which is more precise than the 3rd order polynomial proposed in \cite{Reif2021}, at only modest increased cost.
 The exact coefficients used are presented in Appendix \ref{appendix:gh}.
@@
 Segments with a winding direction that points \emph{up} are colored blue while those pointing \emph{down} are colored green.
-The curves were adapted from the \emph{stills} dataset by \citet{Nehab2020} and the Skia test corpus.}
+The curves were adapted from the \emph{stills} dataset by \cite{Nehab2020} and the Skia test corpus.}
     \Description{Selection of stroked cubic Béziers exhibiting strong and weak correctness}
@@
 
-As recommended in both \citet{Nehab2020} and \citet{Kilgard2020}, the rendering of a cusp with infinite curvature matches that of a near-cusp.
-The code to detect near-zero derivatives and re-evaluate is a small amount of non-divergent logic, unlike the additional ``regularization'' pass proposed in Section 3.1 of \citet{Nehab2020} to handle special cases.
+As recommended in both \cite{Nehab2020} and \cite{Kilgard2020}, the rendering of a cusp with infinite curvature matches that of a near-cusp.
+The code to detect near-zero derivatives and re-evaluate is a small amount of non-divergent logic, unlike the additional ``regularization'' pass proposed in Section 3.1 of \cite{Nehab2020} to handle special cases.
 
@@
 In particular, detecting locations in the cubic Bézier where the curvature crosses a given quantity is a medium-degree polynomial, and in general requires numerical techniques for root finding.
-In the approach of \citet{Nehab2020}, the root finder not only requires a hybrid Newton/bisection method, which requires iteration, but is also recursive in that it uses the roots of a polynomial of one degree lower as a subroutine.
+In the approach of \cite{Nehab2020}, the root finder not only requires a hybrid Newton/bisection method, which requires iteration, but is also recursive in that it uses the roots of a polynomial of one degree lower as a subroutine.
 In general, polynomials up to cubic can be considered GPU-friendly, while finding cusps in cubic Béziers requires a higher degree.
@@
 Both versions can generate stroked outlines with line and arc primitives.
-We focused our evaluation on the number of output primitives and execution time, using the \emph{timings} dataset provided by \citet{Nehab2020}, in addition to our own curve-intensive scenes to stress the GPU implementation.
+We focused our evaluation on the number of output primitives and execution time, using the \emph{timings} dataset provided by \cite{Nehab2020}, in addition to our own curve-intensive scenes to stress the GPU implementation.
 We present our findings in the remainder of this section.
@@
 The results are shown in Figure~\ref{fig:timings}.
-The time shown is the total for test files from the \emph{timings} dataset of \citet{Nehab2020}, and the tolerance was 0.25 when adjustable as a parameter.
-
-Similarly to \citet{Nehab2020}'s observations, Skia is the fastest stroke expansion algorithm we measured.
+The time shown is the total for test files from the \emph{timings} dataset of \cite{Nehab2020}, and the tolerance was 0.25 when adjustable as a parameter.
+
+Similarly to \cite{Nehab2020}'s observations, Skia is the fastest stroke expansion algorithm we measured.
 The speed is attained with some compromises, in particular an imprecise error estimation that can sometimes generate outlines exceeding the given tolerance.
@@
 This allowed us to collect measurements for the curve flattening and input processing stages in isolation.
-We ran the compute pipeline on the \emph{timings} SVG files provided by \citet{Nehab2020}.
+We ran the compute pipeline on the \emph{timings} SVG files provided by \cite{Nehab2020}.
 The largest of these SVGs is \emph{waves} (see rendering \emph{(a)} in Figure \ref{fig:renderings}) which contains 42 stroked paths and a total of 13,308 path segments.
@@
     \includegraphics[scale=0.69]{nehab_gpu_timings_strong}
-    \caption{Mean GPU execution times of the compute shader that performs stroke expansion and flattening when run on the \citet{Nehab2020} \emph{timings} data set.
+    \caption{Mean GPU execution times of the compute shader that performs stroke expansion and flattening when run on the \cite{Nehab2020} \emph{timings} data set.
 The timings are shown in milliseconds.
@@
 The compute shader executes at least an order of magnitude faster on the other GPUs.}
-    \Description{A stacked bar chart of the GPU timings for stroke expansion from \citet{Nehab2020} timings SVGs}
+    \Description{A stacked bar chart of the GPU timings for stroke expansion from \cite{Nehab2020} timings SVGs}
     \label{fig:nehab-gpu-timings}
@@
 Figures~\ref{fig:nehab-gpu-timings} and \ref{fig:test-scenes-gpu-timings} show the execution times for the two data sets.
-The entire \citet{Nehab2020} timings set adds up to less than 1.5~ms on all GPUs except the mobile unit.
+The entire \cite{Nehab2020} timings set adds up to less than 1.5~ms on all GPUs except the mobile unit.
 Our algorithm is \textasciitilde14$\times$ slower on the Mali-G78 compared to the M1 Max but it is still capable of processing \emph{waves} in 3.48~ms on average.
@@
 
-Path stroking has received attention in recent years, with \citet{Nehab2020} and \citet{Kilgard2020} both having proposed a complete theory of correct stroking in vector graphics.
+Path stroking has received attention in recent years, with \cite{Nehab2020} and \cite{Kilgard2020} both having proposed a complete theory of correct stroking in vector graphics.
 While there are a number of implementations of path filling on the GPU, the goal of also implementing stroke expansion on the GPU had not yet been realized.
-Building upon the theory proposed by \citet{Nehab2020}, we implemented an approach to stroke expansion that is GPU-friendly, and achieves high performance on commodity hardware.
+Building upon the theory proposed by \cite{Nehab2020}, we implemented an approach to stroke expansion that is GPU-friendly, and achieves high performance on commodity hardware.
 
@@
     \item The GPU implementation sequentializes some work that could be parallel, and is thus not as well load-balanced as it might be.
-An appealing future direction is to split the pipeline into separate stages, executed by the GPU as nodes in a work graph[\cite{Patel2024}.
+An appealing future direction is to split the pipeline into separate stages, executed by the GPU as nodes in a work graph\cite{Patel2024}.
 This structure is appealing because load balancing is done by hardware.
